### 2015-01-16 MScTemplateCS.tex
Template for a Master's Thesis in LaTeX, for the Department of Computer Science, Lund University.

\documentclass{cslthse-msc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{makeidx}
\usepackage{graphicx}
\usepackage[titletoc, header, page]{appendix}

\usepackage{hyperref}

%\geometry{showframe}

\author{
	Flavius Gruian \\
	{\normalsize \href{mailto:Flavius.Gruian@cs.lth.se}{\texttt{Flavius.Gruian@cs.lth.se}}}
	\and
	Camilla Lekebjer \\
    {\normalsize \href{mailto:Camilla.Lekebjer@cs.lth.se}{\texttt{Camilla.Lekebjer@cs.lth.se}}}
}

\title{Formatting a Master's Thesis}
\subtitle{A {\LaTeX} class}
%\company{The Corporation AB LTD Inc}
\supervisors{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}{Don Jeer, \href{mailto:djeer@xy.lth.se}{\texttt{djeer@xy.lth.se}}}
%\supervisor{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}
\examiner{Jane Doe, \href{mailto:jane.doe@cs.lth.se}{\texttt{jane.doe@cs.lth.se}}}

%\date{\today}
\date{January 16, 2015}

\acknowledgements{
If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.

We would like to thank Lennart Andersson for his feedback on this template.
}

\theabstract{
This document describes the Master's Thesis format for the theses carried out at 
the Department of Computer Science, Lund University.

Your abstract should capture, in English, the whole thesis with focus on the problem and solution in 150 words. It should be placed on a separate right-hand page, with an additional \textit{1cm} margin on both left and right. Avoid acronyms, footnotes, and references in the abstract if possible.

Leave a \textit{2cm} vertical space after the abstract and provide a few keywords relevant for your report. Use five to six words, of which at most two should be from the title.
}

\keywords{MSc, template, report, style, structure}

%% Only used to display font sizes
\makeatletter
\newcommand\thefontsize[1]{{#1 \f@size pt\par}}
\makeatother
%%%%%%%%%%

\begin{document}
\makefrontmatter
\chapter[Short on Formatting]{Formatting}
Avoid empty spaces between \textit{chapter}-\textit{section}, \textit{section}-\textit{sub-section}. For instance, a very brief summary of the chapter would be one way of bridging the chapter heading and the first section of that chapter.
\section{Page Size and Margins}
Use A4 paper, with the text margins given in Table \ref{tab:margins}.
\begin{table}[!hbt]
\centering
\caption{Text margins for A4.}\label{tab:margins}
\begin{tabular}{cc}
\hline
\textbf{margin} & \textbf{space} \\
\hline 
top &  3.0cm\\

bottom & 3.0cm \\

left (inside) & 2.5cm \\

right (outside) & 2.5cm \\

binding offset & 1.0cm \\ 
\hline 
\end{tabular} 
\end{table}

\section{Typeface and Font Sizes}
The fonts to use for the reports are \textbf{TeX Gyre Termes} (a \textbf{Times New Roman} clone) for serif fonts, \textsf{\textbf{TeX Gyre Heros}} (a \textsf{\textbf{Helvetica}} clone) for sans-serif fonts, and finally \texttt{\textbf{TeX Gyre Cursor}} (a \texttt{\textbf{Courier}} clone) as mono-space font. All these fonts are included with the TeXLive 2013 installation. Table \ref{tab:fonts} lists the most important text elements and the associated fonts.
\begin{table}[!hbt]
\caption{Font types, faces and sizes to be used.}\label{tab:fonts}

 \begin{tabular}{ l c c c}
\hline 
\textbf{Element} & \textbf{Face} & \textbf{Size}  & \textbf{\LaTeX size}  \\ 
\hline 
{\huge \textbf{Ch. label}} & {\huge \textbf{serif, bold}} & \thefontsize\huge & \verb+\huge+ \\ 
{\Huge \textbf{Chapter}} & {\Huge \textbf{serif, bold}} & \thefontsize\Huge & \verb+\Huge+ \\ 
{\LARGE \textsf{\textbf{Section}}} & {\Large \textsf{\textbf{sans-serif, bold}}} & \thefontsize\LARGE &  \verb+\LARGE+  \\ 
{\Large \textsf{\textbf{Subsection}}} & {\Large \textsf{\textbf{sans-serif, bold}}} & \thefontsize\Large & \verb+\Large+ \\ 
{\large \textsf{\textbf{Subsubsection}}} & {\Large \textsf{\textbf{sans-serif, bold}}} & \thefontsize\large &  \verb+\large+ \\ 
Body & serif & \thefontsize\normalsize & {\footnotesize \verb+\normalsize+} \\
%{\footnotesize Footnote} & serif  & \thefontsize\footnotesize & {\footnotesize \verb+\footnotesize+} \\
{\footnotesize \textsc{Header}} & {\footnotesize \textsc{serif, SmallCaps}} & \thefontsize\footnotesize &  \\
Footer (page numbers) & serif, regular & \thefontsize\normalsize &  \\
\hline
\textbf{Figure label} & \textbf{serif, bold} & \thefontsize\normalsize & \\
Figure caption & serif, regular & \thefontsize\normalsize & \\
\textsf{In figure} & \textsf{sans-serif} & \textit{any} & \\
\textbf{Table label} & \textbf{serif, bold} & \thefontsize\normalsize & \\
Table caption and text & serif, regular & \thefontsize\normalsize & \\
\texttt{Listings} & \texttt{mono-space} & $\le$ \thefontsize\normalsize & \\
\hline 
\end{tabular} 
\end{table}

\subsection{Headers and Footers}
Note that the page headers are aligned towards the outside of the page (right on the right-hand page, left on the left-hand page) and they contain the section title on the right and the chapter title on the left respectively, in \textsc{SmallCaps}. The footers contain only page numbers on the exterior of the page, aligned right or left depending on the page. The lines used to delimit the headers and footers from the rest of the page are $0.4 pt$ thick, and are as long as the text.

\subsection{Chapters, Sections, Paragraphs}
Chapter, section, subsection, etc. names are all left aligned, and numbered as in this document.

Chapters always start on the right-hand page, with the label and title separated from the rest of the text by a $0.4 pt$ thick line.

Paragraphs are justified (left and right), using single line spacing. Note that the first paragraph of a chapter, section, etc. is not indented, while the following are indented.

\subsection{Tables}
Table captions should be located above the table, justified, and spaced 2.0cm from left and right (important for very long captions). Tables should be numbered, but the numbering is up to you, and could be, for instance:
\begin{itemize}
\item \textbf{Table X.Y} where X is the chapter number and Y is the table number within that chapter. (This is the default in \LaTeX. More on {\LaTeX} can be found on-line, including whole books, such as \cite{goossens93}.) or
\item \textbf{Table Y} where Y is the table number within the whole report
\end{itemize}
As a recommendation, use regular paragraph text in the tables, bold headings and avoid vertical lines (see Table \ref{tab:fonts}).

\subsection{Figures}
Figure labels, numbering, and captions should be formed similarly to tables. As a recommendation, use vector graphics in figures (Figure \ref{fig:vectorg}), rather than bitmaps (Figure \ref{fig:rasterg}). Text within figures usually looks better with sans-serif fonts.
\begin{figure}[!hbt]
\centering
\includegraphics[scale=2.5]{examplepic1.pdf} 
\caption{A PDF vector graphics figure. Notice the numbering and placement of the caption. The caption text is indented 2.0cm from both left and right text margin.}\label{fig:vectorg}
\end{figure}

\begin{figure}[!hbt]
\centering
\includegraphics[scale=2.5]{examplepic3.jpg} 
\caption{A JPEG bitmap figure. Notice the bad quality of such an image when scaling it. Sometimes bitmap images are unavoidable, such as for screen dumps.}\label{fig:rasterg}
\end{figure}
For those interested in delving deeper into the design of graphical information display, please refer to books such as \cite{Tufte:1986, few2012show}.

\section{Mathematical Formulae and Equations}
You are free to use in-text equations and formulae, usually in \textit{italic serif} font. For instance: $S = \sum_i a_i$. We recommend using numbered equations when you do need to refer to the specific equations:
\begin{equation}
E = \int_0^{\delta} P(t) dt \quad \longleftrightarrow \quad E = m c^2
\end{equation}
The numbering system for equations should be similar to that used for tables and figures.

\section{References}
Your references should be gathered in a \textbf{References} section, located at the end of the document (before \textbf{Appendices}). We recommend using number style references, ordered as appearing in the document or alphabetically. Have a look at the references in this template in order to figure out the style, fonts and fields. Web references are acceptable (with restraint) as long as you specify the date you accessed the given link \cite{fontspec, CTAN}. You may of course use URLs directly in the document, using mono-space font, i.e. \url{http://cs.lth.se/}.

\section{Colours}
As a general rule, all theses are printed in black-and-white, with the exception of selected parts in selected theses that need to display colour images essential to describing the thesis outcome (\textit{computer graphics}, for instance).

A strong requirement is for using \textbf{black text on white background} in your document's main text. Otherwise we do encourage using colours in your figures, or other elements (i.e. the colour marking internal and external references) that would make the document more readable on screen. You may also emphasize table rows, columns, cells, or headers using white text on black background, or black text on light grey background.

Finally, note that the document should look good in black-and-white print. Colours are often rendered using monochrome textures in print, which makes them look different from on screen versions. This means that you should choose your colours wisely, and even opt for black-and-white textures when the distinction between colours is hard to make in print. The best way to check how your document looks, is to print out a copy yourself.

\chapter{Language}

You are strongly encouraged to write your report in English, for two reasons. First, it will improve your use of English language. Second, it will increase visibility for you, the author, as well as for the Department of Computer Science, and for your host company (if any).

However, note that your examiner (and supervisors) are not there to provide you with extensive language feedback. We recommend that you check the language used in your report in several ways:
\begin{description}
\item[Reference books] dedicated to language issues can be very useful. \cite{heffernan2000writing} 
\item[Spelling and grammar checkers] which are usually available in the commonly used text editing environments.
\item[Colleagues and friends] willing to provide feedback your writing.
\item[Studieverkstaden] is a university level workshop, that can help you with language related problems (see \href{http://www.lu.se/studera/livet-som-student/service-och-stod/studieverkstaden}{Studieverkstaden}'s web page).
\item[Websites] useful for detecting language errors or strange expressions, such as
\begin{itemize}
\item \url{http://translate.google.com}
\item \url{http://www.gingersoftware.com/grammarcheck/}
\end{itemize}
\end{description}

\section{Style Elements}
Next, we will just give some rough guidelines for good style in a report written in English. Your supervisor and examiner as well as the aforementioned \textbf{Studieverkstad} might have a different  take on these, so we recommend you follow their advice whenever in doubt. If you want a reference to a short style guide, have a look at \cite{shortstyleguide}.

\subsubsection{Widows and Orphans}

Avoid \textit{widows} and \textit{orphans}, namely words or short lines at the beginning or end of a paragraph, which are left dangling at the top or bottom of a column, separated from the rest of the paragraph.

\subsubsection{Footnotes}

We strongly recommend you avoid footnotes. To quote from \cite{OGSW}, \textit{Footnotes are frequently misused by containing information which should either be placed in the text or excluded altogether. They should be avoided as a general rule and are acceptable only in exceptional cases when incorporation of their content in the text  [is] not possible.}

\subsubsection{Active vs. Passive Voice}

Generally active voice (\textit{I ate this apple.}) is easier to understand than passive voice (\textit{This apple has been eaten (by me).}) In passive voice sentences the actor carrying out the action is often forgotten, which makes the reader wonder who actually performed the action. In a report is important to be clear about who carried out the work. Therefore we recommend to use active voice, and preferably the plural form \textit{we} instead of \textit{I} (even in single author reports).

\subsubsection{Long and Short Sentences}
A nice brief list of sentence problems and solutions is given in \cite{yalesentences}. Using choppy sentences (too short) is a common problem of many students. The opposite, using too long sentences, occurs less often, in our experience.

\subsubsection{Subject-Predicate Agreement}
A common problem of native Swedish speakers is getting the subject-predicate (verb) agreement right in sentences. Note that a verb must agree in person and number with its subject. As a rough tip, if you have subject ending in \textit{s} (plural), the predicate should not, and the other way around. Hence, \textit{only one s}. Examples follow:
\begin{description}
\item[incorrect] He have to take this road.
\item[correct] He has to take this road.
\end{description}
\begin{description}
\item[incorrect] These words forms a sentence.
\item[correct] These words form a sentence.
\end{description}
\noindent In more complex sentences, getting the agreement right is trickier. A brief guide is given in  the \textit{20 Rules of Subject Verb Agreement} \cite{subjectverb}.

\chapter{Structure}
It is a good idea to discuss the structure of the report with your supervisor rather early in your writing. Given next is a generic structure that is a recommended starting point. Note that the naming of the chapters is not compulsory, but may be a helpful guideline.
\begin{description}
\item[Introduction] should give the background of your work. Important issues to address are:
\begin{itemize}
\item short introduction to the area
\item problem definition
\item related work (literature study)
\item contributions (This part is essential to pinpoint individual work. For theses with two authors, it is especially important to identify which author has contributed with which part, both with respect to the work and the report.)
\end{itemize}
\item[Approach] should basically contain a description of your solution(s), with all the theoretical background needed.
\item[Evaluation] should describe the method used to evaluate your solution(s)/approach, including the experimental setup.
\item[Discussion] allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact. This might also be a good place to describe your positive and negative experiences related to the work you carried out.
\item[Conclusions] should summarize your findings and possible improvements or recommendations.
\item[Bibliography] is a must in a scientific report. {\LaTeX} and \texttt{bibtex} offer great support for  handling references and automatically generating bibliographies.
\item[Appendices] should contain lengthy details of the experimental setup, mathematical proofs, code download information, and shorter code snippets. Avoid longer code listings. Source code should rather be made available for download on a website or on-line repository of your choosing.

\end{description}
\makebibliography{MyMSc}

\begin{appendices}
\chapter{About This Document}
The following environments and tools were used to create this document:
\begin{itemize}
\item operating system: Mac OS X 10.10.1
\item tex distribution: MacTeX-2014, \url{http://www.tug.org/mactex/}
\item tex editor: Texmaker 4.4.1 for Mac, \url{http://www.xm1math.net/texmaker/} for its XeLaTeX flow (recommended) or pdfLaTeX flow
\item bibtex editor: BibDesk 1.6.3 for Mac, \url{http://bibdesk.sourceforge.net/}
\item fonts \texttt{cslthse-msc.cls} document class): 
\begin{description}
\item{for XeLaTeX}: TeX Gyre Termes, \textsf{TeX Gyre Heros}, \texttt{TeX Gyre Cursor} (installed from the TeXLive 2013)
\item{for pdfLaTeX}: TeX Gyre font packages: tgtermes.sty, tgheros.sty, tgcursor.sty, gtxmath.sty (available through TeXLive 2013) 
\end{description} 
\item picture editor: OmniGraffle Professional 5.4.2
\end{itemize}

\noindent A list of the essential \LaTeX packages needed to compile this document follows (all except \texttt{hyperref} are included in the document class):
\begin{itemize}
\item \texttt{fontspec}, to access local fonts, needs the XeLaTeX flow
\item \texttt{geometry}, for page layout
\item \texttt{titling}, for formatting the title page
\item \texttt{fancyhdr}, for custom headers and footers
\item \texttt{abstract}, for customizing the abstract
\item \texttt{titlesec}, for custom chapters, sections, etc.
\item \texttt{caption}, for custom tables and figure captions
\item \texttt{hyperref}, for producing PDF with hyperlinks
\item \texttt{appendix}, for appendices
\item \texttt{printlen}, for printing text sizes
\item \texttt{textcomp}, for text companion fonts (e.g. bullet)
\end{itemize}

\noindent Other useful packages:
\begin{itemize}
\item \texttt{listings}, for producing code listings with syntax colouring and line numbers
\end{itemize}

\chapter{List of Changes}
\subsubsection{Since 2014/02/18}
\begin{itemize}
\item Added the possibility to specify two supervisors. Use either of the \verb+\supervisor{}+ or \verb+\supervisors{}{}+ commands to set the names and contacts on the first page.
\end{itemize}

\subsubsection{Since 2013/09/23}
\begin{itemize}
\item Added missing colon ":" after \textit{Examiner} on the front page. 
\end{itemize}

\subsubsection{Since 2013/08/30}
\begin{itemize}
\item Changed fonts from Garamond (Times New Roman), Helvetica (Arial), Courier (Source Code Pro) to Tex Gyre fonts, namely Termes, Heros, Cursor, which are freely available with TexLive 2013 installation. These are all clones of Times New Roman, Helvetica and Courier, respectively. Garamond is problematic on some systems, being a non-freely available font.
\item Corrected the \textit{Face} column in Table \ref{tab:fonts} to correctly depict the font face.
\end{itemize}

\subsubsection{Since 2013/02/22}
\begin{itemize}
\item Number of words required in the abstract changed to 150 (from 300).
\end{itemize}

\subsubsection{Since 2013/02/15}
\begin{itemize}
\item Made a separate document class, for clarity.
\item made it work with pdfLaTeX and garamond.sty, in addition to XeLaTeX and true type fonts. It is up to the user to get the hold of the garamond.zip from \url{http://gael-varoquaux.info/computers/garamond/index.html}.
\end{itemize}
\end{appendices}

\end{document}

### 2015-09-11 MScTemplateCS.tex
Minor bug fixed: when \supervisor or \supervisors are not used, a specific error is shown now, rather than a "no line to end".
Improved Chapter 3: Structure, to better describe the different parts.

@@
 
-%\date{\today}
-\date{January 16, 2015}
+\date{\today}
+%\date{January 16, 2015}
 
@@
 It is a good idea to discuss the structure of the report with your supervisor rather early in your writing.
-Given next is a generic structure that is a recommended starting point.
-Note that the naming of the chapters is not compulsory, but may be a helpful guideline.
+Given next is a generic structure that is a starting point, but by no means the absolute standard.
+Your supervisor should provide a better structure for the specific field you are writing your thesis in.
+Note also that the naming of the chapters is not compulsory, but may be a helpful guideline.
 \begin{description}
-\item[Introduction] should give the background of your work. Important issues to address are:
+\item[Introduction] should give the background of your work. Important parts to cover:
 \begin{itemize}
-\item short introduction to the area
-\item problem definition
-\item related work (literature study)
-\item contributions (This part is essential to pinpoint individual work.
-For theses with two authors, it is especially important to identify which author has contributed with which part, both with respect to the work and the report.)
+\item Give the context of your work, have a short introduction to the area.
+\item Define the problem you are solving (or trying to solve).
+\item Specify your contributions.
+What does this particular work/report bring to the research are or to the body of knowledge?
+How is the work divided between the co-authors?
+(This part is essential to pinpoint individual work.
+For theses with two authors, it is compulsory to identify which author has contributed with which part, both with respect to the work and the report.)
+\item Describe related work (literature study).
+Besides listing other work in the area, mention how is it related or relevant to your work.
+The tradition in some research area is to place this part at the end of the report (check with your supervisor).
 \end{itemize}
-\item[Approach] should basically contain a description of your solution(s), with all the theoretical background needed.
-\item[Evaluation] should describe the method used to evaluate your solution(s)/approach, including the experimental setup.
-\item[Discussion] allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
+\item[Approach] should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:
+\begin{itemize}
+\item \textbf{Method}: describe how you go about solving the problem you defined. Also how do you show/prove that your solution actually works, and how well does it work.
+\item \textbf{Theory}: should contain the theoretical background needed to understand your work, if necessary.
+\item \textbf{Implementation}: if your work involved building an artefact/implementation, give the details here.
+Note, that this should not, as a rule, be a chronological description of your efforts, but a view of the result.
+There is a place for insights and lamentation later on in the report, in the Discussion section.
+\end{itemize}
+\item[Evaluation] is the part where you present the finds. Depending on the area this part contains a subset or all of the following: 
+\begin{itemize}
+\item \textbf{Experimental Setup} should describe the details of the method used to evaluate your solution(s)/approach.
+Sometimes this is already addressed in the \textbf{Method}, sometimes this part replaces \textbf{Method}.
+\item \textbf{Results} contains the data (as tables, graphs) obtained via experiments  (benchmarking, polls, interviews).
+\item \textbf{Discussion} allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
 This might also be a good place to describe your positive and negative experiences related to the work you carried out.
+\end{itemize} 
+Occasionally these sections are intermingled, if this allows for a better presentation of your work.
+However, try to distinguish between measurements or hard data (results) and extrapolations, interpretations, or speculations (discussion).
 \item[Conclusions] should summarize your findings and possible improvements or recommendations.
@@
 \chapter{List of Changes}
+\subsubsection{Since 2015/04/27}
+\begin{itemize}
+\item Improved the \textbf{Structure} chapter and added more detailed comments for each part.
+\end{itemize}
+
 \subsubsection{Since 2014/02/18}

### 2016-05-11 MScTemplateCS.tex
Added a popular science summary template, that is also appended to the end of your report (1 page only).
Better template for the pop-sci summary.

@@
 \usepackage{hyperref}
+\usepackage{pdfpages}
 
@@
 	{\normalsize \href{mailto:Flavius.Gruian@cs.lth.se}{\texttt{Flavius.Gruian@cs.lth.se}}}
-	\and
-	Camilla Lekebjer \\
-    {\normalsize \href{mailto:Camilla.Lekebjer@cs.lth.se}{\texttt{Camilla.Lekebjer@cs.lth.se}}}
+%	\and
+%	Camilla Lekebjer \\
+%    {\normalsize \href{mailto:Camilla.Lekebjer@cs.lth.se}{\texttt{Camilla.Lekebjer@cs.lth.se}}}
 }
@@
 We would like to thank Lennart Andersson for his feedback on this template.
+
+We would also like thank Camilla Lekebjer for her contribution on this template, as well as Magnus Hultin for his popular science summary class and example document.
 }
@@
 \item \texttt{textcomp}, for text companion fonts (e.g. bullet)
+\item \texttt{pdfpages}, to include the popular science summary page at the end
 \end{itemize}
@@
 \chapter{List of Changes}
+
+\subsubsection{Since 2016/04/29}
+\begin{itemize}
+\item A better template for the popular science summary, by Magnus Hultin.
+\end{itemize}
+
+\subsubsection{Since 2015/09/11}
+\begin{itemize}
+\item Added a template for the popular science summary, in the \verb+popsci+ directory.
+\item Added code in the report that imports the one page popular science \texttt{pdf} at the end of the document.
+\end{itemize}
+
 \subsubsection{Since 2015/04/27}
@@
 \end{itemize}
+
+\includepdf[pages={1}]{popsci/popsci.pdf}
 \end{appendices}

### 2016-05-11 popsci/popsci.tex
Better template for the pop-sci summary.

\documentclass{popsci}

\usepackage[swedish, english]{babel}
%\usepackage[utf8]{inputenc}
\usepackage{fancyhdr}
\usepackage{titling}
\usepackage{color}
\usepackage{colortbl}

\usepackage{ifxetex}

\ifxetex
  \usepackage{fontspec}
\else
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{lmodern}
\fi

\presentationsdag{2016-04-28}

\examensarbete{}
%To create a title in two rows, leave examensarbete blank and fill in examensarbeteTwoRows.
\examensarbeteTwoRows{Application Specific Instruction-set Processor Using a Parametrizable multi-SIMD}{Synthesizeable Model Supporting Design Space Exploration}

\author{Magnus Hultin}
\supervisor {Flavius Gruian (LTH)}
\examiner{Krzysztof Kuchcinski (LTH)}

\title{Parametrisk processor modell för design utforskning}

\begin{document}

\theabstract{Applikations-specifika processorer är allt mer vanligt för få ut rätt prestanda med så lite resurser som möjligt. Detta arbete har en parametrisk modell för att kunna testa hur mycket resurser som behövs för en specifik applikation.}

{\noindent För att öka prestandan i dagens processorer finns det vektorenheter och flera kärnor i processorer. Vektorenheten finns till för att kunna utföra en operation på en mängd data samtidigt och flera kärnor gör att man kan utföra fler instruktioner samtidigt. Ofta är processorerna designade för att kunna stödja en mängd olika datorprogram. Detta resulterar i att det blir kompromisser som kan påverka prestandan för vissa program och vara överflödigt för andra. I t.ex. videokameror, mobiltelefoner, medicinsk utrustning, digital kameror och annan inbyggd elektronik, kan man istället använda en processor som saknar vissa funktioner men som istället är mer energieffektiv. Man kan jämföra det med att frakta ett paket med en stor lastbil istället för att använda en mindre bil där samma paketet också skulle få plats.

I mitt examensarbete har jag skrivit en modell som kan användas för att snabbt designa en processor enligt vissa parametrar. Dessa parametrar väljs utifrån vilket eller vilka program man tänkta köra på den. Vissa program kan t.ex. lättare använda flera kärnor och vissa program kan använda korta eller längre vektorenheter för dess data.

För att kunna välja vilken typ av processor som är rätt för den specifika applikationen krävs det ofta att man snabbt kan testa olika prototyper. Att implementera dessa till hårdvara kan ofta vara tidskrävande och ifall det visar sig att implementationen inte klarar dem kraven man ställt för prestanda och energieffektivitet, måste man designa för nya parametrar och mer tid har blivit slösat. Om den här processen istället kan göras automatiskt utifrån dessa design-parametrar kan man teoretiskt spara en massa tid.
Modellen testades med olika multimedia program. Den mest beräkningsintensiva och mest upprepande delen av programmen användes. Dessa kallas för kärnor av programmen. Kärnorna som användes var ifrån MPEG och JPEG, som används för bildkomprimering och videokomprimering.

Resultatet visar att det finns en prestanda vinst jämfört med generella processorer men att detta också ökar resurserna som behövs. Detta trots att den generella processorn har nästan dubbelt så hög klockfrekvens än dem applikations-specifika processorerna. Resultatet visar också att schemaläggning av instruktionerna i programmen spelar en stor roll för att kunna utnyttja resurserna som finns tillgängliga och därmed öka prestandan. Med den schemaläggningen som utnyttjade resurserna bäst var prestandan minst 79\% bättre än den generella processorn.
}

\end{document}

### 2018-12-03 MScTemplateCS.tex
Template changed to include generating the covers.

@@
 \usepackage[titletoc, header, page]{appendix}
-
-\usepackage{hyperref}
-\usepackage{pdfpages}
+\usepackage{transparent}
 
 %\geometry{showframe}
+%better like this?
+\student{Flavius Gruian}{Flavius.Gruian@cs.lth.se}
+%\students{Flavius Gruian}{Flavius.Gruian@cs.lth.se}{Camilla Lekebjer}{Camilla.Lekebjer@cs.lth.se}
 
-\author{
-	Flavius Gruian \\
-	{\normalsize \href{mailto:Flavius.Gruian@cs.lth.se}{\texttt{Flavius.Gruian@cs.lth.se}}}
-%	\and
-%	Camilla Lekebjer \\
-%    {\normalsize \href{mailto:Camilla.Lekebjer@cs.lth.se}{\texttt{Camilla.Lekebjer@cs.lth.se}}}
-}
+\thesisnumber{LU-CS-EX: 2019-XX} % Birger Swahn will provide this number to you, once the thesis is ready for publication
+% default is Master. Uncomment the following for "kandidatarbete"/Bachelor's thesis
+%\thesistype{Bachelor}
 
 \title{Formatting a Master's Thesis}
+
 \subtitle{A {\LaTeX} class}
@@
 
-\keywords{MSc, template, report, style, structure}
+\keywords{MSc, BSc, template, report, style, structure}
 
@@
 
+Make sure you add references as close to the claim as possible~\cite{CTAN}, as shown, not at the end of a whole paragraph.
+Notice also that there is a space before the reference; best is to use \verb+~\cite{ref}+, to allow for unbreakable spaces.
+References should not be used after the period marking the end of sentence.
+Using the reference as follows (end of paragrah, after period) is {\em{strongly discouraged}}, since it says nothing about which specific claim you provide the reference for.
+\cite{fontspec}
+
 \section{Colours}
@@
 
-Finally, note that the document should look good in black-and-white print.
+Note that the document should look good in black-and-white print.
 Colours are often rendered using monochrome textures in print, which makes them look different from on screen versions.
@@
 The best way to check how your document looks, is to print out a copy yourself.
+
+The {\LaTeX}  class defines also a few {\em LTH} standard colours, which you could use in your document for various elements, to adhere to the standard university profile.
+\\These are: {\color{LTHblue}LTHblue}, {\color{LTHbronze}LTHbronze}, {\color{LTHgreen}LTHgreen}, {\color{LTHpink}LTHpink}, {\color{LTHcyan}LTHcyan}, {\color{LTHgrey}LTHgrey}.
 
@@
 \item \textbf{Results} contains the data (as tables, graphs) obtained via experiments  (benchmarking, polls, interviews).
+Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
 \item \textbf{Discussion} allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
-This might also be a good place to describe your positive and negative experiences related to the work you carried out.
+Focus here on a broader view of the results, talking about the relation between the different finds.\footnote{Bad practice is to display graphs in Results and then describe them textually one by one in here.
+No!
+Both sections should have some discussion, but one targets individual finds and the other tries to bridge between these adopting a more overarching viewpoint.} This might also be a good place to describe your positive and negative experiences related to the work you carried out.
 \end{itemize} 
@@
 \end{description}
+% Should use consistent formatting when it comes to Names ("FirstName LastName", or "F. LastName")
 \makebibliography{MyMSc}
@@
 \begin{itemize}
-\item operating system: Mac OS X 10.10.1
+\item operating system: Mac OS X 10.14
 \item tex distribution: MacTeX-2014, \url{http://www.tug.org/mactex/}
-\item tex editor: Texmaker 4.4.1 for Mac, \url{http://www.xm1math.net/texmaker/} for its XeLaTeX flow (recommended) or pdfLaTeX flow
+\item tex editor: Texmaker 5.0.2 for Mac, \url{http://www.xm1math.net/texmaker/} for its XeLaTeX flow (recommended) or pdfLaTeX flow
 \item bibtex editor: BibDesk 1.6.3 for Mac, \url{http://bibdesk.sourceforge.net/}
@@
 
+\subsubsection{Since 2018/11/21}
+\begin{itemize}
+\item Added the report cover to the template (along with the archive cover, as the 2nd page).
+This includes the front cover and back as the pop-sci article.
+Once the report is ready for print, the authors will receive a report number, of the format \texttt{LU-CS-EX 2019-NN}.
+Generate the report using this number in the template, to get a print-ready pdf.
+\item Changed both the report and pop-sci template to use \verb+\student+ or \verb+\students+ commands instead of \verb+\author+.
+\item Clarified the difference between {\em Results} and {\em Discussion} sections in a report.
+\item Added more information about the use of {\em References}.
+\item Added {\em LTH} colours in the template and description in {\em Colours}.
+\end{itemize}
+
 \subsubsection{Since 2016/04/29}
@@
 \begin{itemize}
-\item Changed fonts from Garamond (Times New Roman), Helvetica (Arial), Courier (Source Code Pro) to Tex Gyre fonts, namely Termes, Heros, Cursor, which are freely available with TexLive 2013 installation.
+\item Changed fonts from Garamond ( Times New Roman ), Helvetica ( Arial ), Courier ( Source Code Pro ) to Tex Gyre fonts, namely Termes, Heros, Cursor, which are freely available with TexLive 2013 installation.
 These are all clones of Times New Roman, Helvetica and Courier, respectively.
@@
 
+%make sure we're on even page with the pop-sci
+\checkoddpage
+\ifoddpage
+\else
+   \newpage
+   \thispagestyle{empty}
+   \mbox{ }
+\fi
 \includepdf[pages={1}]{popsci/popsci.pdf}

### 2019-04-16 MScTemplateCS.tex
Fixed the title box and added commands for one, two or three lines titles.tr
Changed some of the fonts to allow for better output in Overleaf. (It's still recommended to use XeLaTeX on a local environment)
Fixed the issue with missing names on the cover, in case of two authors.
Added the list of used packages at the end of the document, before the pop-sci. To hide it, use the nofilelist option for the package.

@@
+%\documentclass[nofilelist]{cslthse-msc}
+% to show a list of used packages at the end of the document, delete the nofilelist option
 \documentclass{cslthse-msc} 
@@
 \usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
+%\usepackage{amsfonts}
+%%\usepackage{amssymb}
 \usepackage{amsthm}
@@
 
+% used to display the used files at the end. Select nofilelist as a package option to disable this
+\listfiles % initialize
+
 %\geometry{showframe}
 %better like this?
-\student{Flavius Gruian}{Flavius.Gruian@cs.lth.se}
-%\students{Flavius Gruian}{Flavius.Gruian@cs.lth.se}{Camilla Lekebjer}{Camilla.Lekebjer@cs.lth.se}
+%\student{Flavius Gruian}{Flavius.Gruian@cs.lth.se}
+\students{Flavius Gruian}{Flavius.Gruian@cs.lth.se}{Camilla Lekebjer}{Camilla.Lekebjer@cs.lth.se}
 
@@
 % default is Master. Uncomment the following for "kandidatarbete"/Bachelor's thesis
-%\thesistype{Bachelor}
+%\thesistype{Bachelor}{Kandidatarbete}
 
-\title{Formatting a Master's Thesis}
+%\title{Formatting a Master's Thesis}
+\title{Formatting a Master's Thesis and a bunch of other things that are not really needed in here}
+
+%\onelinetitle
+%\twolinestitle
+\threelinestitle
+%\fourlinestitle
 
@@
 
+\subsubsection{Since 2019/04/01}
+\begin{itemize}
+\item Added the \texttt{nofilelist} option in to the package.
+By default the list of packages is displayed at the end of the document, before the pop-sci.
+With the option, the file list is not displayed.
+Used mainly for debug purposes.
+Ignore warnings and the two errors about the \texttt{fontenc.sty}
+\end{itemize}
+
 \subsubsection{Since 2018/11/21}
@@
 
+% display used packages information unless noflielist is used in the cslthse-msc package option
+\printfilelist
+
 %make sure we're on even page with the pop-sci

### 2020-09-18 MScTemplateCS.tex
Fixed uneven line spacing in the title on the cover.
Changed Bibliography name to References - this can be easily done in the document now.
Added Swedish title request, magic number for LU-CS-EX

@@
-%\documentclass[nofilelist]{cslthse-msc}
+\documentclass[nofilelist]{cslthse-msc}
 % to show a list of used packages at the end of the document, delete the nofilelist option
-\documentclass{cslthse-msc} 
+%\documentclass{cslthse-msc} 
 \usepackage[utf8]{inputenc}
@@
 
-\thesisnumber{LU-CS-EX: 2019-XX} % Birger Swahn will provide this number to you, once the thesis is ready for publication
+\thesisnumber{LU-CS-EX: 2023-79} % Magic Number! Do not change unless Birger Swahn asks you to do so!
 % default is Master. Uncomment the following for "kandidatarbete"/Bachelor's thesis
@@
 \title{Formatting a Master's Thesis and a bunch of other things that are not really needed in here}
+\svensktitel{Infoga den Svenska titel här!}
 
@@
 We would also like thank Camilla Lekebjer for her contribution on this template, as well as Magnus Hultin for his popular science summary class and example document.
+
+Thanks also go to the following (former) students for helping with feedback and suggestions on this template: Mikael Persson, Christoffer Lundgren, Mahmoud Nasser.
 }
@@
 \begin{document}
+\renewcommand{\bibname}{References}
+
 \makefrontmatter
@@
 % Should use consistent formatting when it comes to Names ("FirstName LastName", or "F. LastName")
+%\printbibliography
 \makebibliography{MyMSc}
@@
 \chapter{List of Changes}
+
+\subsubsection{Since 2020/09/18}
+\begin{itemize}
+\item Added the Swedish title displayed on page 3. Use \verb+\svensktitel{...}+ command to set it.
+\item Converted the default LU-CS-EX number to a magic number, \verb+2023-79+, easy to find and change in PDF with \verb+changepagestring.pl+; this is a \verb+perl+ command line utility found in \verb+CAM::PDF+ module -- but only works on some PDFs (e.g.
+generated with \verb+pdflatex+) but not others (e.g.
+\verb+xelatex+) \textbf{[CS internal use]}
+\end{itemize}
+
+\subsubsection{Since 2019/10/30}
+\begin{itemize}
+\item Bibliography is now called References. This can be easily adjusted by changing \\ \verb+\renewcommand{\bibname}{References}+ after \verb+\begin{document}+.
+\end{itemize}
 

### 2023-01-24 GoalDocument.tex
Set up goal document outline

\documentclass{article}

% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[UKenglish]{babel}

% Set page size and margins
% Replace `letterpaper' with `a4paper' for UK/EU standard size
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=false, allcolors=blue]{hyperref}

\title{Goal Document: Translating Typing Rules into JastAdd}
\author{Nicholas Boyd Isacsson}

\begin{document}
\maketitle

\section{Involved parties}
\subsubsection*{Student}
Nicholas Boyd Isacsson\\ \href{mailto:nicholas@isacsson.se}{nicholas@isacsson.se}
\\
\subsubsection*{Supervisor}
Christoph Reichenbach \\ \href{mailto:christoph.reichenbach@cs.lth.se}{christoph.reichenbach@cs.lth.se}

\subsubsection*{Examiner}
Niklas Fors \\ \href{mailto:niklas.fors@cs.lth.se}{niklas.fors@cs.lth.se}

\section{Background}

\section{Project description}

\section{Motivation}

\section{Previous work}

\section{Evaluation}

\section{Methodology}

\end{document}

### 2023-02-03 GoalDocument.tex
Reformat involved parties

@@
 \subsubsection*{Student}
-Nicholas Boyd Isacsson\\ \href{mailto:nicholas@isacsson.se}{nicholas@isacsson.se}
-\\
+  Nicholas Boyd Isacsson \\
+  \href{mailto:nicholas@isacsson.se}{nicholas@isacsson.se}
+
 \subsubsection*{Supervisor}
-Christoph Reichenbach \\ \href{mailto:christoph.reichenbach@cs.lth.se}{christoph.reichenbach@cs.lth.se}
+  Christoph Reichenbach \\
+  \href{mailto:christoph.reichenbach@cs.lth.se}{christoph.reichenbach@cs.lth.se}
 
 \subsubsection*{Examiner}
-Niklas Fors \\ \href{mailto:niklas.fors@cs.lth.se}{niklas.fors@cs.lth.se}
+  Niklas Fors \\
+  \href{mailto:niklas.fors@cs.lth.se}{niklas.fors@cs.lth.se}
 

### 2023-02-06 GoalDocument.tex
Add comments about what to cover
goaldoc: Add research questions note
goaldoc: Add first draft of written sections

@@
 \section{Background}
+In programming languages there are many types of data, and for any real world use case there will be constraints upon these types, such as what types of input a function can contain.
+Type checking is the method of verifying, either at runtime (dynamic checking) or compile time (static checking), that these constraints are held.
+
+JastAdd is a meta-compilation system, allowing one to work with and extend abstract syntax trees with great flexibility and is extensively used within the SDE group at LTH.
+However, implementing a type checker in JastAdd by hand currently requires a lot of boilerplate code, slowing down iteration.
+Typing rules offer a condensed, standardised syntax for defining simple languages, allowing fast iterations with minimal work, but these rules then typically need to be implemented by hand.
 
 \section{Project description}
+% Build JastAdd Specification from Typing Rules
+% Typing rules are an efficient way for expressing flow-insensitive program analyses.
+The goal for this thesis project is to automatically translate typing rules into JastAdd rules, automatically selecting different type inference schemes (monomorphic, polymorphic, syntax-directed or not, ...).
 
-\section{Motivation}
+The goal of this project is to translate the concisely formatted typing rules into a working type-checker in JastAdd.
+We'll start with a basic implementation able to handle simple syntax-directed rules.
+Once this is working, there are two possible directions to take the project.
+One is to support object-oriented languages, by adding support for subtyping and ad hoc polymorphism.
+The other is to support a more functional approach by implementing a Hindley-Milner style type system.
+Which of these routes to take will be decided based on the experience of implementing the basic system.
+
+Research questions include what language features we can support and what approaches we can use to translate different cases and how best to leverage JastAdd's capabilities.
+
+% deliniate scope, syntax-directed rules, need not be fully self-contained
+% eventually break out into one of non syntax-directed rules Hindley-Milner
+% or subtyping and adhoc polymorphism
+
+% Research questions:
+%   What can be translated
+%   *How to translate*, different approaches
+
+%\section{Motivation} % merge with background
+% Why compile to JastAdd?
 
 \section{Previous work}
+\begin{itemize}
+  \item
+    André Pacak, Sebastian Erdweg, and Tamás Szabó.
+2020.
+A systematic approach to deriving incremental type checkers.
+Proc.
+ACM Program.
+Lang.
+4, OOPSLA, Article 127 (November 2020), 28 pages.
+https://doi.org/10.1145/3428195 \\
+    \url{https://szabta89.github.io/publications/inca-typechecking.pdf} \\
+    A paper about implementing a compiler, capable of deriving incremental type checking from typing rules.
+    Similar goal to ours, but with a significantly different compilation target, meaning we will likely differ in architecture and what trade-offs are made.
+
+  \item
+    Mikael Pettersson. 1996. A Compiler for Natural Semantics. In Proceedings of the 6th International Conference on Compiler Construction (CC '96). Springer-Verlag, Berlin, Heidelberg, 177–191. \\
+    \url{https://link.springer.com/chapter/10.1007/3-540-61053-7\_61} \\
+
+  \item
+    Grigore Roșu, Traian Florin Șerbănută. An overview of the K semantic framework.
+The Journal of Logic and Algebraic Programming, Volume 79, Issue 6, 2010, Pages 397-434, ISSN 1567-8326, https://doi.org/10.1016/j.jlap.2010.03.012. \\
+    \url{https://www.sciencedirect.com/science/article/pii/S1567832610000160} \\
+
+\end{itemize}
 
 \section{Evaluation}
+The first step of evaluating the project is to see what language features it supports.
+Once the program is capable of compiling a basic language, we can also compare it to existing handwritten (or written ourselves) JastAdd specifications for those same languages.
+Here we could evaluate the size of the finished specification as well as the performance of the type checking.
+
+An additional point of comparison is the Datalog paper, to which we could compare supported features as well as output size and performance.
+%  Case studies, implementation possible by hand or already finished implementations
+%  Compare with handwritten implementations:
+%    - Spec size
+%    - Performance
+%  Compare with datalog paper
 
 \section{Methodology}
+% Steps:
+%   - Input language and parsing
+%   - Simply typed lambda calculus
+%   - Additional features
+%   ...
+%   - Evaluation
+%   - Finalise the thesis document (collect bullet points throughout and organise them but don't write text)
+
+The timeline for the project is spread out over 20 weeks, from now until the beginning of June.
+The following steps are planned:
+\begin{itemize}
+  \item Define an input language and handle parsing
+  \item Implement the basics required for simply typed lambda calculus
+  \item Implement as many additional language features as time allows
+  \item Evaluation
+  \item Finalise the thesis document, starting around the end of April %(collect bullet points throughout and organise them but don't write text)
+\end{itemize}
 

### 2023-02-09 GoalDocument.tex
goaldoc: Minor adjustments as per feedback

@@
 In programming languages there are many types of data, and for any real world use case there will be constraints upon these types, such as what types of input a function can contain.
-Type checking is the method of verifying, either at runtime (dynamic checking) or compile time (static checking), that these constraints are held.
+Type checking is a method of verifying, either at runtime (dynamic checking) or compile time (static checking), that these constraints are held.
 
-JastAdd is a meta-compilation system, allowing one to work with and extend abstract syntax trees with great flexibility and is extensively used within the SDE group at LTH.
+JastAdd is a meta-compilation system, allowing one to work with and extend abstract syntax trees with great flexibility. % and is extensively used within the SDE group at LTH.
 However, implementing a type checker in JastAdd by hand currently requires a lot of boilerplate code, slowing down iteration.
@@
 The goal of this project is to translate the concisely formatted typing rules into a working type-checker in JastAdd.
-We'll start with a basic implementation able to handle simple syntax-directed rules.
+We will start with a basic implementation able to handle simple syntax-directed rules.
 Once this is working, there are two possible directions to take the project.
@@
 
-\section{Previous work}
+\section{Related work}
 \begin{itemize}
@@
 The first step of evaluating the project is to see what language features it supports.
+% What are the challenges?
+What properties of typesystems make them difficult/easy to embed?
+Efficiency, spec size
 Once the program is capable of compiling a basic language, we can also compare it to existing handwritten (or written ourselves) JastAdd specifications for those same languages.
-Here we could evaluate the size of the finished specification as well as the performance of the type checking.
+Here we will evaluate the size of the finished specification as well as the performance of the type checking.
 
-An additional point of comparison is the Datalog paper, to which we could compare supported features as well as output size and performance.
+An additional point of comparison is Pacak et al.'s \emph{A systematic approach to deriving incremental type checkers}, to which we will compare supported features as well as output size and performance.
 %  Case studies, implementation possible by hand or already finished implementations

### 2023-02-09 GoalDocument.tex
goaldoc: Add approximate timeframes for each step

@@
 The timeline for the project is spread out over 20 weeks, from now until the beginning of June.
-The following steps are planned:
+The following steps are planned, with approximate timeframes:
 \begin{itemize}
-  \item Define an input language and handle parsing
-  \item Implement the basics required for simply typed lambda calculus
-  \item Implement as many additional language features as time allows
-  \item Evaluation
-  \item Finalise the thesis document, starting around the end of April %(collect bullet points throughout and organise them but don't write text)
+  \item Define an input language and handle parsing, 2 weeks
+  \item Implement the basics required for simply typed lambda calculus, 2 weeks
+  \item Implement as many additional language features as time allows, 8 weeks
+  \item Evaluation, 2 weeks
+  \item Finalise the thesis document, 6 weeks %(collect bullet points throughout and organise them but don't write text)
 \end{itemize}

### 2023-02-16 GoalDocument.tex
goaldoc: Add Saioc Hüttel paper to related work
goaldoc: Reorganise previous work

@@
 \section{Related work}
+A paper about implementing a compiler, capable of deriving incremental type checking from typing rules.
+Similar goal to ours, but with a significantly different compilation target, meaning we will likely differ in architecture and what trade-offs are made.
 \begin{itemize}
@@
     \url{https://szabta89.github.io/publications/inca-typechecking.pdf} \\
-    A paper about implementing a compiler, capable of deriving incremental type checking from typing rules.
-    Similar goal to ours, but with a significantly different compilation target, meaning we will likely differ in architecture and what trade-offs are made.
-
+\end{itemize}
+The following papers are more general compilers for natural semantics, without focusing on type checking.
+\begin{itemize}
   \item
@@
     \url{https://www.sciencedirect.com/science/article/pii/S1567832610000160} \\
+
+  \item
+    Saioc, G., Hüttel, H. (2022). A Tool for Describing and Checking Natural Semantics Definitions of Programming Languages. FROM. \\
+    \url{https://arxiv.org/abs/2209.09471}
 

### 2023-09-25 MSc.tex
Init MSc.tex
Outline MSc structure

--- 2018-12-03 MScTemplateCS.tex
@@
-\documentclass{cslthse-msc}
+\documentclass[nofilelist]{cslthse-msc}
+% to show a list of used packages at the end of the document, delete the nofilelist option
+%\documentclass{cslthse-msc}
 \usepackage[utf8]{inputenc}
@@
 
+% used to display the used files at the end. Select nofilelist as a package option to disable this
+\listfiles % initialize
+
 %\geometry{showframe}
 %better like this?
-\student{Flavius Gruian}{Flavius.Gruian@cs.lth.se}
+\student{Nicholas Boyd Isacsson}{nicholas@isacsson.se}
 %\students{Flavius Gruian}{Flavius.Gruian@cs.lth.se}{Camilla Lekebjer}{Camilla.Lekebjer@cs.lth.se}
 
-\thesisnumber{LU-CS-EX: 2019-XX} % Birger Swahn will provide this number to you, once the thesis is ready for publication
+\thesisnumber{LU-CS-EX: 2023-79} % Magic Number! Do not change unless Birger Swahn asks you to do so!
 % default is Master. Uncomment the following for "kandidatarbete"/Bachelor's thesis
-%\thesistype{Bachelor}
+%\thesistype{Bachelor}{Kandidatarbete}
 
-\title{Formatting a Master's Thesis}
+%\title{Formatting a Master's Thesis}
+\title{Typechecker Generation using Reference Attribute Grammars}
+\svensktitel{Infoga den Svenska titeln här!}
 
-\subtitle{A {\LaTeX} class}
+%\onelinetitle
+%\twolinestitle
+\threelinestitle
+%\fourlinestitle
+
+%\subtitle{A {\LaTeX} class}
 %\company{The Corporation AB LTD Inc}
-\supervisors{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}{Don Jeer, \href{mailto:djeer@xy.lth.se}{\texttt{djeer@xy.lth.se}}}
+%\supervisors{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}{Don Jeer, \href{mailto:djeer@xy.lth.se}{\texttt{djeer@xy.lth.se}}}
+\supervisor{Christoph Reichenbach, \href{mailto:christoph.reichenbach@cs.lth.se}{\texttt{christoph.reichenbach@cs.lth.se}}}
 %\supervisor{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}
-\examiner{Jane Doe, \href{mailto:jane.doe@cs.lth.se}{\texttt{jane.doe@cs.lth.se}}}
+\examiner{Niklas Fors, \href{mailto:niklas.fors@cs.lth.se}{\texttt{niklas.fors@cs.lth.se}}}
 
@@
 
-\acknowledgements{
-If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.
-
-We would like to thank Lennart Andersson for his feedback on this template.
-
-We would also like thank Camilla Lekebjer for her contribution on this template, as well as Magnus Hultin for his popular science summary class and example document.
-}
+%\acknowledgements{
+%If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.
+%
+%We would like to thank Lennart Andersson for his feedback on this template.
+%
+%We would also like thank Camilla Lekebjer for her contribution on this template, as well as Magnus Hultin for his popular science summary class and example document.
+%
+%Thanks also go to the following (former) students for helping with feedback and suggestions on this template: Mikael Persson, Christoffer Lundgren, Mahmoud Nasser.
+%}
 
@@
 \begin{document}
+\renewcommand{\bibname}{References}
+
 \makefrontmatter
-\chapter[Short on Formatting]{Formatting}
-Avoid empty spaces between \textit{chapter}-\textit{section}, \textit{section}-\textit{sub-section}.
-For instance, a very brief summary of the chapter would be one way of bridging the chapter heading and the first section of that chapter.
-\section{Page Size and Margins}
-Use A4 paper, with the text margins given in Table \ref{tab:margins}.
-\begin{table}[!hbt]
-\centering
-\caption{Text margins for A4.}\label{tab:margins}
-\begin{tabular}{cc}
-\hline
-\textbf{margin} & \textbf{space} \\
-\hline 
-top &  3.0cm\\
-
-bottom & 3.0cm \\
-
-left (inside) & 2.5cm \\
-
-right (outside) & 2.5cm \\
-
-binding offset & 1.0cm \\ 
-\hline 
-\end{tabular} 
-\end{table}
-
-\section{Typeface and Font Sizes}
-The fonts to use for the reports are \textbf{TeX Gyre Termes} (a \textbf{Times New Roman} clone) for serif fonts, \textsf{\textbf{TeX Gyre Heros}} (a \textsf{\textbf{Helvetica}} clone) for sans-serif fonts, and finally \texttt{\textbf{TeX Gyre Cursor}} (a \texttt{\textbf{Courier}} clone) as mono-space font.
-All these fonts are included with the TeXLive 2013 installation.
-Table \ref{tab:fonts} lists the most important text elements and the associated fonts.
-\begin{table}[!hbt]
-\caption{Font types, faces and sizes to be used.}\label{tab:fonts}
-
- \begin{tabular}{ l c c c}
-\hline 
-\textbf{Element} & \textbf{Face} & \textbf{Size}  & \textbf{\LaTeX size}  \\ 
-\hline 
-{\huge \textbf{Ch. label}} & {\huge \textbf{serif, bold}} & \thefontsize\huge & \verb+\huge+ \\ 
-{\Huge \textbf{Chapter}} & {\Huge \textbf{serif, bold}} & \thefontsize\Huge & \verb+\Huge+ \\ 
-{\LARGE \textsf{\textbf{Section}}} & {\Large \textsf{\textbf{sans-serif, bold}}} & \thefontsize\LARGE &  \verb+\LARGE+  \\ 
-{\Large \textsf{\textbf{Subsection}}} & {\Large \textsf{\textbf{sans-serif, bold}}} & \thefontsize\Large & \verb+\Large+ \\ 
-{\large \textsf{\textbf{Subsubsection}}} & {\Large \textsf{\textbf{sans-serif, bold}}} & \thefontsize\large &  \verb+\large+ \\ 
-Body & serif & \thefontsize\normalsize & {\footnotesize \verb+\normalsize+} \\
-%{\footnotesize Footnote} & serif  & \thefontsize\footnotesize & {\footnotesize \verb+\footnotesize+} \\
-{\footnotesize \textsc{Header}} & {\footnotesize \textsc{serif, SmallCaps}} & \thefontsize\footnotesize &  \\
-Footer (page numbers) & serif, regular & \thefontsize\normalsize &  \\
-\hline
-\textbf{Figure label} & \textbf{serif, bold} & \thefontsize\normalsize & \\
-Figure caption & serif, regular & \thefontsize\normalsize & \\
-\textsf{In figure} & \textsf{sans-serif} & \textit{any} & \\
-\textbf{Table label} & \textbf{serif, bold} & \thefontsize\normalsize & \\
-Table caption and text & serif, regular & \thefontsize\normalsize & \\
-\texttt{Listings} & \texttt{mono-space} & $\le$ \thefontsize\normalsize & \\
-\hline 
-\end{tabular} 
-\end{table}
-
-\subsection{Headers and Footers}
-Note that the page headers are aligned towards the outside of the page (right on the right-hand page, left on the left-hand page) and they contain the section title on the right and the chapter title on the left respectively, in \textsc{SmallCaps}.
-The footers contain only page numbers on the exterior of the page, aligned right or left depending on the page.
-The lines used to delimit the headers and footers from the rest of the page are $0.4 pt$ thick, and are as long as the text.
-
-\subsection{Chapters, Sections, Paragraphs}
-Chapter, section, subsection, etc. names are all left aligned, and numbered as in this document.
-
-Chapters always start on the right-hand page, with the label and title separated from the rest of the text by a $0.4 pt$ thick line.
-
-Paragraphs are justified (left and right), using single line spacing. Note that the first paragraph of a chapter, section, etc. is not indented, while the following are indented.
-
-\subsection{Tables}
-Table captions should be located above the table, justified, and spaced 2.0cm from left and right (important for very long captions).
-Tables should be numbered, but the numbering is up to you, and could be, for instance:
-\begin{itemize}
-\item \textbf{Table X.Y} where X is the chapter number and Y is the table number within that chapter.
-(This is the default in \LaTeX.
-More on {\LaTeX} can be found on-line, including whole books, such as \cite{goossens93}.) or
-\item \textbf{Table Y} where Y is the table number within the whole report
-\end{itemize}
-As a recommendation, use regular paragraph text in the tables, bold headings and avoid vertical lines (see Table \ref{tab:fonts}).
-
-\subsection{Figures}
-Figure labels, numbering, and captions should be formed similarly to tables.
-As a recommendation, use vector graphics in figures (Figure \ref{fig:vectorg}), rather than bitmaps (Figure \ref{fig:rasterg}).
-Text within figures usually looks better with sans-serif fonts.
-\begin{figure}[!hbt]
-\centering
-\includegraphics[scale=2.5]{examplepic1.pdf} 
-\caption{A PDF vector graphics figure. Notice the numbering and placement of the caption. The caption text is indented 2.0cm from both left and right text margin.}\label{fig:vectorg}
-\end{figure}
-
-\begin{figure}[!hbt]
-\centering
-\includegraphics[scale=2.5]{examplepic3.jpg} 
-\caption{A JPEG bitmap figure. Notice the bad quality of such an image when scaling it. Sometimes bitmap images are unavoidable, such as for screen dumps.}\label{fig:rasterg}
-\end{figure}
-For those interested in delving deeper into the design of graphical information display, please refer to books such as \cite{Tufte:1986, few2012show}.
-
-\section{Mathematical Formulae and Equations}
-You are free to use in-text equations and formulae, usually in \textit{italic serif} font.
-For instance: $S = \sum_i a_i$.
-We recommend using numbered equations when you do need to refer to the specific equations:
-\begin{equation}
-E = \int_0^{\delta} P(t) dt \quad \longleftrightarrow \quad E = m c^2
-\end{equation}
-The numbering system for equations should be similar to that used for tables and figures.
-
-\section{References}
-Your references should be gathered in a \textbf{References} section, located at the end of the document (before \textbf{Appendices}).
-We recommend using number style references, ordered as appearing in the document or alphabetically.
-Have a look at the references in this template in order to figure out the style, fonts and fields.
-Web references are acceptable (with restraint) as long as you specify the date you accessed the given link \cite{fontspec, CTAN}.
-You may of course use URLs directly in the document, using mono-space font, i.e.
-\url{http://cs.lth.se/}.
-
-Make sure you add references as close to the claim as possible~\cite{CTAN}, as shown, not at the end of a whole paragraph.
-Notice also that there is a space before the reference; best is to use \verb+~\cite{ref}+, to allow for unbreakable spaces.
-References should not be used after the period marking the end of sentence.
-Using the reference as follows (end of paragrah, after period) is {\em{strongly discouraged}}, since it says nothing about which specific claim you provide the reference for.
-\cite{fontspec}
-
-\section{Colours}
-As a general rule, all theses are printed in black-and-white, with the exception of selected parts in selected theses that need to display colour images essential to describing the thesis outcome (\textit{computer graphics}, for instance).
-
-A strong requirement is for using \textbf{black text on white background} in your document's main text.
-Otherwise we do encourage using colours in your figures, or other elements (i.e.
-the colour marking internal and external references) that would make the document more readable on screen.
-You may also emphasize table rows, columns, cells, or headers using white text on black background, or black text on light grey background.
-
-Note that the document should look good in black-and-white print.
-Colours are often rendered using monochrome textures in print, which makes them look different from on screen versions.
-This means that you should choose your colours wisely, and even opt for black-and-white textures when the distinction between colours is hard to make in print.
-The best way to check how your document looks, is to print out a copy yourself.
-
-The {\LaTeX}  class defines also a few {\em LTH} standard colours, which you could use in your document for various elements, to adhere to the standard university profile.
-\\These are: {\color{LTHblue}LTHblue}, {\color{LTHbronze}LTHbronze}, {\color{LTHgreen}LTHgreen}, {\color{LTHpink}LTHpink}, {\color{LTHcyan}LTHcyan}, {\color{LTHgrey}LTHgrey}.
-
-\chapter{Language}
-
-You are strongly encouraged to write your report in English, for two reasons.
-First, it will improve your use of English language.
-Second, it will increase visibility for you, the author, as well as for the Department of Computer Science, and for your host company (if any).
-
-However, note that your examiner (and supervisors) are not there to provide you with extensive language feedback. We recommend that you check the language used in your report in several ways:
-\begin{description}
-\item[Reference books] dedicated to language issues can be very useful. \cite{heffernan2000writing} 
-\item[Spelling and grammar checkers] which are usually available in the commonly used text editing environments.
-\item[Colleagues and friends] willing to provide feedback your writing.
-\item[Studieverkstaden] is a university level workshop, that can help you with language related problems (see \href{http://www.lu.se/studera/livet-som-student/service-och-stod/studieverkstaden}{Studieverkstaden}'s web page).
-\item[Websites] useful for detecting language errors or strange expressions, such as
-\begin{itemize}
-\item \url{http://translate.google.com}
-\item \url{http://www.gingersoftware.com/grammarcheck/}
-\end{itemize}
-\end{description}
-
-\section{Style Elements}
-Next, we will just give some rough guidelines for good style in a report written in English.
-Your supervisor and examiner as well as the aforementioned \textbf{Studieverkstad} might have a different  take on these, so we recommend you follow their advice whenever in doubt.
-If you want a reference to a short style guide, have a look at \cite{shortstyleguide}.
-
-\subsubsection{Widows and Orphans}
-
-Avoid \textit{widows} and \textit{orphans}, namely words or short lines at the beginning or end of a paragraph, which are left dangling at the top or bottom of a column, separated from the rest of the paragraph.
-
-\subsubsection{Footnotes}
-
-We strongly recommend you avoid footnotes.
-To quote from \cite{OGSW}, \textit{Footnotes are frequently misused by containing information which should either be placed in the text or excluded altogether.
-They should be avoided as a general rule and are acceptable only in exceptional cases when incorporation of their content in the text  [is] not possible.}
-
-\subsubsection{Active vs. Passive Voice}
-
-Generally active voice (\textit{I ate this apple.}) is easier to understand than passive voice (\textit{This apple has been eaten (by me).}) In passive voice sentences the actor carrying out the action is often forgotten, which makes the reader wonder who actually performed the action.
-In a report is important to be clear about who carried out the work.
-Therefore we recommend to use active voice, and preferably the plural form \textit{we} instead of \textit{I} (even in single author reports).
-
-\subsubsection{Long and Short Sentences}
-A nice brief list of sentence problems and solutions is given in \cite{yalesentences}.
-Using choppy sentences (too short) is a common problem of many students.
-The opposite, using too long sentences, occurs less often, in our experience.
-
-\subsubsection{Subject-Predicate Agreement}
-A common problem of native Swedish speakers is getting the subject-predicate (verb) agreement right in sentences.
-Note that a verb must agree in person and number with its subject.
-As a rough tip, if you have subject ending in \textit{s} (plural), the predicate should not, and the other way around.
-Hence, \textit{only one s}.
-Examples follow:
-\begin{description}
-\item[incorrect] He have to take this road.
-\item[correct] He has to take this road.
-\end{description}
-\begin{description}
-\item[incorrect] These words forms a sentence.
-\item[correct] These words form a sentence.
-\end{description}
-\noindent In more complex sentences, getting the agreement right is trickier. A brief guide is given in  the \textit{20 Rules of Subject Verb Agreement} \cite{subjectverb}.
-
-\chapter{Structure}
-It is a good idea to discuss the structure of the report with your supervisor rather early in your writing.
-Given next is a generic structure that is a starting point, but by no means the absolute standard.
-Your supervisor should provide a better structure for the specific field you are writing your thesis in.
-Note also that the naming of the chapters is not compulsory, but may be a helpful guideline.
-\begin{description}
-\item[Introduction] should give the background of your work. Important parts to cover:
-\begin{itemize}
-\item Give the context of your work, have a short introduction to the area.
-\item Define the problem you are solving (or trying to solve).
-\item Specify your contributions.
+\chapter{Introduction} % should give the background of your work. Important parts to cover:
+% Give the context of your work, have a short introduction to the area.
+% Define the problem you are solving (or trying to solve).
+% Specify your contributions.
 What does this particular work/report bring to the research are or to the body of knowledge?
@@
 For theses with two authors, it is compulsory to identify which author has contributed with which part, both with respect to the work and the report.)
-\item Describe related work (literature study).
+% Describe related work (literature study).
 Besides listing other work in the area, mention how is it related or relevant to your work.
 The tradition in some research area is to place this part at the end of the report (check with your supervisor).
-\end{itemize}
-\item[Approach] should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:
-\begin{itemize}
-\item \textbf{Method}: describe how you go about solving the problem you defined. Also how do you show/prove that your solution actually works, and how well does it work.
-\item \textbf{Theory}: should contain the theoretical background needed to understand your work, if necessary.
-\item \textbf{Implementation}: if your work involved building an artefact/implementation, give the details here.
+
+\chapter{Approach} % should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:
+\section{Method} % describe how you go about solving the problem you defined. Also how do you show/prove that your solution actually works, and how well does it work.
+\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
+\section{Implementation} % if your work involved building an artefact/implementation, give the details here.
 Note, that this should not, as a rule, be a chronological description of your efforts, but a view of the result.
 There is a place for insights and lamentation later on in the report, in the Discussion section.
-\end{itemize}
-\item[Evaluation] is the part where you present the finds. Depending on the area this part contains a subset or all of the following: 
-\begin{itemize}
-\item \textbf{Experimental Setup} should describe the details of the method used to evaluate your solution(s)/approach.
+\section{Evaluation} % is the part where you present the finds. Depending on the area this part contains a subset or all of the following:
+\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach.
 Sometimes this is already addressed in the \textbf{Method}, sometimes this part replaces \textbf{Method}.
-\item \textbf{Results} contains the data (as tables, graphs) obtained via experiments  (benchmarking, polls, interviews).
+\section{Results} % contains the data (as tables, graphs) obtained via experiments  (benchmarking, polls, interviews).
 Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
-\item \textbf{Discussion} allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
+\section{Discussion} % allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
 Focus here on a broader view of the results, talking about the relation between the different finds.\footnote{Bad practice is to display graphs in Results and then describe them textually one by one in here.
@@
 Both sections should have some discussion, but one targets individual finds and the other tries to bridge between these adopting a more overarching viewpoint.} This might also be a good place to describe your positive and negative experiences related to the work you carried out.
-\end{itemize} 
-Occasionally these sections are intermingled, if this allows for a better presentation of your work.
+% Occasionally these sections are intermingled, if this allows for a better presentation of your work.
 However, try to distinguish between measurements or hard data (results) and extrapolations, interpretations, or speculations (discussion).
-\item[Conclusions] should summarize your findings and possible improvements or recommendations.
-\item[Bibliography] is a must in a scientific report. {\LaTeX} and \texttt{bibtex} offer great support for  handling references and automatically generating bibliographies.
-\item[Appendices] should contain lengthy details of the experimental setup, mathematical proofs, code download information, and shorter code snippets.
+
+\chapter{Conclusions} % should summarize your findings and possible improvements or recommendations.
+
+% Should use consistent formatting when it comes to Names ("FirstName LastName", or "F. LastName")
+%\printbibliography
+\makebibliography{MyMSc} % is a must in a scientific report. {\LaTeX} and \texttt{bibtex} offer great support for  handling references and automatically generating bibliographies.
+
+\begin{appendices} %should contain lengthy details of the experimental setup, mathematical proofs, code download information, and shorter code snippets.
 Avoid longer code listings.
@@
 
-\end{description}
-% Should use consistent formatting when it comes to Names ("FirstName LastName", or "F. LastName")
-\makebibliography{MyMSc}
-
-\begin{appendices}
-\chapter{About This Document}
-The following environments and tools were used to create this document:
-\begin{itemize}
-\item operating system: Mac OS X 10.14
-\item tex distribution: MacTeX-2014, \url{http://www.tug.org/mactex/}
-\item tex editor: Texmaker 5.0.2 for Mac, \url{http://www.xm1math.net/texmaker/} for its XeLaTeX flow (recommended) or pdfLaTeX flow
-\item bibtex editor: BibDesk 1.6.3 for Mac, \url{http://bibdesk.sourceforge.net/}
-\item fonts \texttt{cslthse-msc.cls} document class): 
-\begin{description}
-\item{for XeLaTeX}: TeX Gyre Termes, \textsf{TeX Gyre Heros}, \texttt{TeX Gyre Cursor} (installed from the TeXLive 2013)
-\item{for pdfLaTeX}: TeX Gyre font packages: tgtermes.sty, tgheros.sty, tgcursor.sty, gtxmath.sty (available through TeXLive 2013) 
-\end{description} 
-\item picture editor: OmniGraffle Professional 5.4.2
-\end{itemize}
-
-\noindent A list of the essential \LaTeX packages needed to compile this document follows (all except \texttt{hyperref} are included in the document class):
-\begin{itemize}
-\item \texttt{fontspec}, to access local fonts, needs the XeLaTeX flow
-\item \texttt{geometry}, for page layout
-\item \texttt{titling}, for formatting the title page
-\item \texttt{fancyhdr}, for custom headers and footers
-\item \texttt{abstract}, for customizing the abstract
-\item \texttt{titlesec}, for custom chapters, sections, etc.
-\item \texttt{caption}, for custom tables and figure captions
-\item \texttt{hyperref}, for producing PDF with hyperlinks
-\item \texttt{appendix}, for appendices
-\item \texttt{printlen}, for printing text sizes
-\item \texttt{textcomp}, for text companion fonts (e.g. bullet)
-\item \texttt{pdfpages}, to include the popular science summary page at the end
-\end{itemize}
-
-\noindent Other useful packages:
-\begin{itemize}
-\item \texttt{listings}, for producing code listings with syntax colouring and line numbers
-\end{itemize}
-
-\chapter{List of Changes}
-
-\subsubsection{Since 2018/11/21}
-\begin{itemize}
-\item Added the report cover to the template (along with the archive cover, as the 2nd page).
-This includes the front cover and back as the pop-sci article.
-Once the report is ready for print, the authors will receive a report number, of the format \texttt{LU-CS-EX 2019-NN}.
-Generate the report using this number in the template, to get a print-ready pdf.
-\item Changed both the report and pop-sci template to use \verb+\student+ or \verb+\students+ commands instead of \verb+\author+.
-\item Clarified the difference between {\em Results} and {\em Discussion} sections in a report.
-\item Added more information about the use of {\em References}.
-\item Added {\em LTH} colours in the template and description in {\em Colours}.
-\end{itemize}
-
-\subsubsection{Since 2016/04/29}
-\begin{itemize}
-\item A better template for the popular science summary, by Magnus Hultin.
-\end{itemize}
-
-\subsubsection{Since 2015/09/11}
-\begin{itemize}
-\item Added a template for the popular science summary, in the \verb+popsci+ directory.
-\item Added code in the report that imports the one page popular science \texttt{pdf} at the end of the document.
-\end{itemize}
-
-\subsubsection{Since 2015/04/27}
-\begin{itemize}
-\item Improved the \textbf{Structure} chapter and added more detailed comments for each part.
-\end{itemize}
-
-\subsubsection{Since 2014/02/18}
-\begin{itemize}
-\item Added the possibility to specify two supervisors. Use either of the \verb+\supervisor{}+ or \verb+\supervisors{}{}+ commands to set the names and contacts on the first page.
-\end{itemize}
-
-\subsubsection{Since 2013/09/23}
-\begin{itemize}
-\item Added missing colon ":" after \textit{Examiner} on the front page. 
-\end{itemize}
-
-\subsubsection{Since 2013/08/30}
-\begin{itemize}
-\item Changed fonts from Garamond ( Times New Roman ), Helvetica ( Arial ), Courier ( Source Code Pro ) to Tex Gyre fonts, namely Termes, Heros, Cursor, which are freely available with TexLive 2013 installation.
-These are all clones of Times New Roman, Helvetica and Courier, respectively.
-Garamond is problematic on some systems, being a non-freely available font.
-\item Corrected the \textit{Face} column in Table \ref{tab:fonts} to correctly depict the font face.
-\end{itemize}
-
-\subsubsection{Since 2013/02/22}
-\begin{itemize}
-\item Number of words required in the abstract changed to 150 (from 300).
-\end{itemize}
-
-\subsubsection{Since 2013/02/15}
-\begin{itemize}
-\item Made a separate document class, for clarity.
-\item made it work with pdfLaTeX and garamond.sty, in addition to XeLaTeX and true type fonts.
-It is up to the user to get the hold of the garamond.zip from \url{http://gael-varoquaux.info/computers/garamond/index.html}.
-\end{itemize}
+% display used packages information unless noflielist is used in the cslthse-msc package option
+\printfilelist
 

### 2023-10-10 MSc.tex
Further outlning of structure

@@
 \theabstract{
-This document describes the Master's Thesis format for the theses carried out at
-the Department of Computer Science, Lund University.
-
-Your abstract should capture, in English, the whole thesis with focus on the problem and solution in 150 words.
+%Your abstract should capture, in English, the whole thesis with focus on the problem and solution in 150 words.
 It should be placed on a separate right-hand page, with an additional \textit{1cm} margin on both left and right.
 Avoid acronyms, footnotes, and references in the abstract if possible.
-
-Leave a \textit{2cm} vertical space after the abstract and provide a few keywords relevant for your report. Use five to six words, of which at most two should be from the title.
+%Leave a \textit{2cm} vertical space after the abstract and provide a few keywords relevant for your report. Use five to six words, of which at most two should be from the title.
 }
@@
 The tradition in some research area is to place this part at the end of the report (check with your supervisor).
+\section{Research questions}
+\begin{itemize}
+\end{itemize}
 
-\chapter{Approach} % should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:
+\chapter{Background}
+%\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
+\section{Type systems}
+\subsection{Data types}
+\subsection{Typechecking}
+\subsubsection{Dynamic Typechecking}
+\subsubsection{Static Typechecking}
+\section{Reference Attribute Grammars}
+\subsection{Reference Attribute Grammars}
+\subsection{JastAdd}
+
+\chapter{Approach/Design} % should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:
+
+\section{Architecture}
+
+\subsection{Typing rule definition}
+Custom DSL
+ASCII representation of BNF (is it BNF?)
+\subsubsection{Language}
+Figure of example BNF (from Pierce?) and ASCII representation
+\subsubsection{Parser}
+\subsubsection{Abstract syntax tree}
+
+\subsection{Typechecker template}
+For ease of use,=
+\subsubsection{Parser}
+For testing purposes, our outputted typechecker comes with a simple tree-based parser.
+Reflection-based parsing
+\subsubsection{Project files}
+\subsubsection{Typechecking aspect}
+
+\subsection{Output typechecker}
+\subsubsection{Tree-representation}
+\subsubsection{Typechecking algorithm}
+Simple algorithm
+Recursive
+\subsubsection{Error handling}
+
 \section{Method} % describe how you go about solving the problem you defined. Also how do you show/prove that your solution actually works, and how well does it work.
-\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
 \section{Implementation} % if your work involved building an artefact/implementation, give the details here.
@@
 There is a place for insights and lamentation later on in the report, in the Discussion section.
-\section{Evaluation} % is the part where you present the finds. Depending on the area this part contains a subset or all of the following:
+
+\chapter{Evaluation} % is the part where you present the finds. Depending on the area this part contains a subset or all of the following:
 \section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach.
@@
 Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
-\section{Discussion} % allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
+\subsection{Input/output examples, with descriptions}
+
+\chapter{Discussion} % allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
 Focus here on a broader view of the results, talking about the relation between the different finds.\footnote{Bad practice is to display graphs in Results and then describe them textually one by one in here.
@@
 However, try to distinguish between measurements or hard data (results) and extrapolations, interpretations, or speculations (discussion).
+\section{Supported features}
+\section{Future work}
+\section{Development learnings}
 

### 2023-10-16 MSc.tex
Various thesis work

@@
 \usepackage[utf8]{inputenc}
-\usepackage[english]{babel}
+\usepackage[british]{babel}
 \usepackage{amsmath}
@@
 %\title{Formatting a Master's Thesis}
-\title{Typechecker Generation using Reference Attribute Grammars}
+\title{Type Checker Generation using Reference Attribute Grammars}
 \svensktitel{Infoga den Svenska titeln här!}
@@
 % Give the context of your work, have a short introduction to the area.
+While writing software, developers often make mistakes.
+Catching these mistakes as soon as possible has become 
+Type checkers are one of the most common static checking methods, but creating a new type checker is often an arduous process, requiring large amounts of complex code.
+Furthermore, making changes to the typing rules of a language can require extensive rewrites in many different parts of the type checker's codebase.
+% This can hinder experimentation
+
+In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules.
+
 % Define the problem you are solving (or trying to solve).
@@
 For theses with two authors, it is compulsory to identify which author has contributed with which part, both with respect to the work and the report.)
+\section{Research questions}
+\begin{itemize}
+  \item What kinds of typing rules can we translate through RAG?
+  \item What approach?
+  \item What are the potential issues of translating typing rules into JastAdd-based RAGs?
+\end{itemize}
+
+\section{Related work} % Maybe move to discussion?
 % Describe related work (literature study).
@@
 The tradition in some research area is to place this part at the end of the report (check with your supervisor).
-\section{Research questions}
-\begin{itemize}
-\end{itemize}
 
@@
 \subsection{Data types}
-\subsection{Typechecking}
-\subsubsection{Dynamic Typechecking}
-\subsubsection{Static Typechecking}
+\subsection{Type checking}
+\subsubsection{Dynamic Type checking}
+\subsubsection{Static Type checking}
 \section{Reference Attribute Grammars}
@@
 \section{Architecture}
-
+Overview
 \subsection{Typing rule definition}
 Custom DSL
-ASCII representation of BNF (is it BNF?)
+ASCII representation of natural deduction %BNF (is it BNF?)
 \subsubsection{Language}
-Figure of example BNF (from Pierce?) and ASCII representation
+Figure of example natural deduction rules (from Pierce?) and ASCII representation
 \subsubsection{Parser}
@@
 
-\subsection{Typechecker template}
-For ease of use,=
+\subsection{Type checker template}
+For ease of use the compiler will output, in addition to the type checker, a complete runnable Gradle project.
+It can parse simplified tree-representations of programs, and utilise these to perform the type checking.
 \subsubsection{Parser}
-For testing purposes, our outputted typechecker comes with a simple tree-based parser.
+For testing purposes, our outputted type checker comes with a simple tree-based parser.
 Reflection-based parsing
 \subsubsection{Project files}
-\subsubsection{Typechecking aspect}
+\subsubsection{Type checking aspect}
 
-\subsection{Output typechecker}
+\subsection{Output type checker}
 \subsubsection{Tree-representation}
-\subsubsection{Typechecking algorithm}
+\subsubsection{Type checking algorithm}
+Generate code for each node and typing rule
 Simple algorithm
@@
 
-\section{Method} % describe how you go about solving the problem you defined. Also how do you show/prove that your solution actually works, and how well does it work.
+\section{Method (remove this?)} % describe how you go about solving the problem you defined. Also how do you show/prove that your solution actually works, and how well does it work.
+
+\cite{Hedin2011}
 \section{Implementation} % if your work involved building an artefact/implementation, give the details here.
@@
 \chapter{Evaluation} % is the part where you present the finds. Depending on the area this part contains a subset or all of the following:
-\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach.
+%\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach.
 Sometimes this is already addressed in the \textbf{Method}, sometimes this part replaces \textbf{Method}.
@@
 Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
-\subsection{Input/output examples, with descriptions}
+\subsection{Input/output examples, with descriptions} %TODO
+Bool language
+Type variables
+\section{Research questions}
+\subsection{Research q 1}
+\subsection{Research q 2}
+\subsection{Research q 3}
+List of challenges % TODO
+  Splitting code between multiple JastAdd components (and java) can get messy
+  Using JastAdd simultaneously as a library and a build component
 
@@
 \section{Supported features}
-\section{Future work}
 \section{Development learnings}
@@
 \chapter{Conclusions} % should summarize your findings and possible improvements or recommendations.
+\section{Summary of findings}
+\section{Future work}
+Improved error handling
+Check for completeness of rules
+Verify rules against AST
+Handle multiple rules for the same node (add(int, int): int, add(double, double): double)
+Better support for numbers, strings, etc.
 
@@
 Source code should rather be made available for download on a website or on-line repository of your choosing.
+Git repository link
 

### 2023-10-16 MSc.tex
Add architectural overview graph

@@
 \usepackage{transparent}
+\usepackage{tikz}
+\usetikzlibrary{fit,positioning,shapes.geometric}
 
@@
 Overview
+
+\noindent\resizebox{\textwidth}{!}{%
+\begin{tikzpicture}[
+  every node/.style = {draw=black, rounded corners, align=center},
+  every path/.style = {draw, -latex}
+                          ]
+  \node (typerules)                               {Object language typing\\ rules definition};
+  \node (trparsing) [right=of typerules, ellipse] {scan,\\parse,\\build};
+  \node (trparser)  [above=of trparsing]          {Typing rule parser};
+  \node (trast)     [right=of trparsing]          {Typing rules AST};
+
+  \node (typecheckgen) [below=of trast, ellipse] {
+    \textbf{Type checker generation}\\
+    - Link rules and language AST\\
+    - Fill in template
+  };
+
+  \node (olast)     [below=of typerules] {Object language AST};
+  \node (inputprog) [below=of olast]     {Input program};
+
+  \node (outputparsing)      [below=of typecheckgen, ellipse]  {Scan,\\parse,\\build};
+  \node (outputtypechecking) [right=of outputparsing, ellipse] {Perform\\typechecking};
+  \node (typeerrors)         [right=of outputtypechecking]     {List of type errors};
+
+  \node (template) [right=of typecheckgen] {
+    \textbf{Type checker template}\\
+    - Scanner, parser, builder\\
+    - Project files\\
+    - Type checking algorithm\\
+  };
+
+  \node (userinput)   [dashed, fill=none, fit=(typerules) (olast)]                  {};
+  \node               [above, draw=none] at (userinput.north)                       {User input};
+  \node (typechecker) [dashed, fill=none, fit=(outputparsing) (outputtypechecking)] {};
+  \node               [above, draw=none] at (typechecker.north)                     {Type checker};
+
+  \draw   (typerules)  to  (trparsing);
+  \draw   (trparser)  to  (trparsing);
+  \draw   (trparsing)  to  (trast);
+  \draw   (trast)  to  (typecheckgen);
+  \draw   (olast)  to  (typecheckgen);
+  \draw   (olast)  to  (outputparsing);
+  \draw   (outputparsing)  to  (outputtypechecking);
+  \draw   (outputtypechecking)  to  (typeerrors);
+  \draw   (typecheckgen)  to  (typechecker);
+  \draw   (inputprog)  to  (outputparsing);
+  \draw   (template)  to  (typecheckgen);
+
+\end{tikzpicture}
+}
+
 \subsection{Typing rule definition}

### 2023-10-25 MSc.tex
Set up swedish as secondary language
Rework overview graph

--- 2023-10-16 MSc.tex
@@
 \usepackage[utf8]{inputenc}
-\usepackage[british]{babel}
+\usepackage[main=british,swedish]{babel}
 \usepackage{amsmath}
@@
 \usepackage{transparent}
+\usepackage{tikz}
+\usepackage{enumitem}
+\usetikzlibrary{fit,positioning,shapes.geometric}
 
@@
 \title{Type Checker Generation using Reference Attribute Grammars}
+\begin{otherlanguage}{swedish}
 \svensktitel{Infoga den Svenska titeln här!}
+\end{otherlanguage}
 
@@
 Overview
+
+\begin{figure}[]
+  \resizebox{\textwidth}{!}{%
+  \begin{tikzpicture}[
+      every node/.style = {draw=black, rounded corners, align=center},
+      every path/.style = {draw, -latex}
+    ]
+    \node (typerules)                               {Typing rules\\definition};
+    \node (trparsing) [right=of typerules, ellipse] {Scan,\\parse,\\build};
+    \node (trparser)  [above=of trparsing]          {Typing rule parser};
+    \node (trast)     [right=of trparsing]          {Typing rules AST};
+
+    %\node (typecheckgen) [below=of trast, ellipse, text width=5cm] {
+    %  \textbf{Type checker generation}\\
+    %  \hrule
+    %  \begin{itemize}[noitemsep,topsep=2pt]
+    %    \item Copy template
+    %    \item Generate type checker from typing rules
+    %  \end{itemize}
+    %};
+    \node (typecheckgen)       [below=of trast, ellipse] {Generate\\type checker};
+    \node (typingrulecompiler) [right=of trast]          {Typing rule\\compiler (\ref{typingrulecompiler})};
+    \node (typechecker)        [below=of typecheckgen]   {Type checker};
+
+    \node (olast)     [below=of typerules] {AST definition};
+
+    \node (outputtypechecking) [below=of typechecker, ellipse] {Perform\\typechecking};
+    \node (inputast)           [left=of outputtypechecking]    {Input program\\AST};
+    \node (outputparsing)      [left=of inputast, ellipse]     {Scan,\\parse,\\build};
+    \node (typeerrors)         [right=of outputtypechecking]   {Type errors};
+    \node (inputprog)          [left=of outputparsing]         {Input program};
+
+    %\node (template) [below=of inputast, text width=6cm] {
+    %  \textbf{Type checker template}\\
+    %  \hrule
+    %  \begin{itemize}[noitemsep,topsep=2pt]
+    %    \item Scanner, parser, builder
+    %    \item Project files
+    %    \item Type checking algorithm (\ref{typecheckingalgorithm})
+    %  \end{itemize}
+    %};
+    \node (templateparser)        [below=of outputparsing]                               {Tree-based\\parser};
+    \node (templateprojectfiles)  at (inputast |- templateparser)                        {Project files};
+    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser)              {Type check\\algorithm};
+    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm)] {};
+    \node                         [above, draw=none] at (template.north)                 {Template};
+
+    \node (userinput)   [dashed, fill=none, fit=(typerules) (olast)]                  {};
+    \node               [above, draw=none] at (userinput.north)                       {Object language\\static semantics};
+    %\node (typechecker) [dashed, fill=none, fit=(outputparsing) (outputtypechecking)] {};
+    %\node               [above, draw=none] at (typechecker.north)                     {Type checker};
+
+    \draw (typerules)          to (trparsing);
+    \draw (trparser)           to (trparsing);
+    \draw (trparsing)          to (trast);
+    \draw (trast)              to (typecheckgen);
+    \draw (typingrulecompiler) to (typecheckgen);
+    \draw (typechecker)        to (outputtypechecking);
+    \draw (olast)              to (outputparsing);
+    \draw (outputparsing)      to (inputast);
+    \draw (inputast)           to (outputtypechecking);
+    \draw (outputtypechecking) to (typeerrors);
+    \draw (typecheckgen)       to (typechecker);
+    \draw (inputprog)          to (outputparsing);
+
+    \draw[dashed] (templateparser)                  to (outputparsing);
+    \draw         (typecheckingalgorithm)           to (outputtypechecking);
+    \draw[dashed] (templateprojectfiles.north west) to (outputparsing);
+    \draw[dashed] (templateprojectfiles.north east) to (outputtypechecking);
+
+  \end{tikzpicture}
+  }
+  \caption{An overview of the project components and flow.}
+  \label{overviewgraph}
+\end{figure}
+
 \subsection{Typing rule definition}

### 2023-11-02 MSc.tex
Further work on architecture

@@
 While writing software, developers often make mistakes.
-Catching these mistakes as soon as possible has become 
-Type checkers are one of the most common static checking methods, but creating a new type checker is often an arduous process, requiring large amounts of complex code.
+Catching these mistakes as early as possible can prevent these bugs from ever reaching production code.
+Type checkers are one of the most common static checking methods, ensuring that the correct data types are used, but creating a new type checker is often an arduous process, requiring large amounts of complex code.
 Furthermore, making changes to the typing rules of a language can require extensive rewrites in many different parts of the type checker's codebase.
@@
 \section{Research questions}
+To formulate our goals more concretely, we've set out to answer the following research questions:
+\newcommand{\rqone}{What kinds of typing rules can we translate through RAG?}
+\newcommand{\rqtwo}{How can we implement the translations and into what form?}
+\newcommand{\rqthree}{What are the potential issues of translating typing rules into JastAdd-based RAGs?}
 \begin{itemize}
-  \item What kinds of typing rules can we translate through RAG?
-  \item What approach?
-  \item What are the potential issues of translating typing rules into JastAdd-based RAGs?
+  \item \rqone
+  \item \rqtwo
+  \item \rqthree
 \end{itemize}
@@
 \subsection{Type checking}
-\subsubsection{Dynamic Type checking}
-\subsubsection{Static Type checking}
+\subsubsection{Dynamic type checking}
+\subsubsection{Static type checking}
+
 \section{Reference Attribute Grammars}
@@
 \subsection{JastAdd}
+\subsubsection{AST definitions}\label{astdef}
 
@@
 \section{Architecture}
-Overview
-
+In this section we'll be providing an overview of our architecture.
+Since it contains many different components, it may be helpful to refer to the overview graph in figure \ref{overviewgraph} while reading.
+It represents the flow of the program, with components and artefacts (represented as rectangular boxes) with arrows linking them to actions (elliptical boxes) which in turn have arrows to the new components and artefacts they produce.
+The dashed arrows from represent links from the template components that exist in the current experimental state of the project, but in a finished product would instead link to component from the object language's own compilation environment.
+%It uses rectangular boxes to represent the concrete data or artefacts of the procedure, and elliptical boxes to represent the actions performed with them.
 \begin{figure}[]
@@
     ]
-    \node (typerules)                               {Typing rules\\definition};
+    \node (typerules)                               {Typing rules\\definition (\ref{typingruledefinition})};
     \node (trparsing) [right=of typerules, ellipse] {Scan,\\parse,\\build};
@@
 
-    \node (olast)     [below=of typerules] {AST definition};
+    \node (olast)     [below=of typerules] {AST definition\\(\ref{astdef})};
 
@@
     %};
-    \node (templateparser)        [below=of outputparsing]                               {Tree-based\\parser};
-    \node (templateprojectfiles)  at (inputast |- templateparser)                        {Project files};
-    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser)              {Type check\\algorithm};
+    \node (templateparser)        [below=of outputparsing]                               {Tree-based\\parser\\(\ref{treebasedparser})};
+    \node (templateprojectfiles)  at (inputast |- templateparser)                        {Project files\\(\ref{projectfiles})};
+    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser)              {Type checking\\aspect (\ref{typecheckingaspect})};
     \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm)] {};
@@
 
-\subsection{Typing rule definition}
+\subsection{Typing rule definition}\label{typingruledefinition}
+To express typing rules, we created a simple ASCII representation of the natural deduction notation used by Pierce\cite{Pierce}.
 Custom DSL
+Include typing rules AST definition (or BNF?)
 ASCII representation of natural deduction %BNF (is it BNF?)
 \subsubsection{Language}
-Figure of example natural deduction rules (from Pierce?) and ASCII representation
+- Figure of example natural deduction rules (from Pierce?) and ASCII representation
+No infix notation
+No environments
+
+Our language can express a subset of the notation used by Pierce, with some limitations.
+Expressions aren't written in the language's native syntax, instead using a simplified syntax resembling the AST classes.
+Values and functions take their names directly from the AST-nodes, and any parameters are included within parenthesis.
+This avoids much of the difficulty in parsing different languages, which can have wildly different syntax rules.
+
+Numerical values have not been implemented, but natural numbers can be expressed using Peano numbers\cite{peano}, utilising a zero value and successor function which represents the number incremented by one.
+If a language has the syntax \verb|0 + 1| it may instead be represented as \verb|Add(Zero, Succ(Zero))|.
+
 \subsubsection{Parser}
@@
 \subsection{Type checker template}
-For ease of use the compiler will output, in addition to the type checker, a complete runnable Gradle project.
-It can parse simplified tree-representations of programs, and utilise these to perform the type checking.
-\subsubsection{Parser}
+In addition to the compiled files, there are a number of files that are added to the output file that are identical for all generated typecheckers, which we referr to as the template.
+Some of these are essential to the functions of the type checker, and others are included for convenience of testing.
+We expect that in a real world use case, our software would integrate into an object language's compiler pipeline, and utilise its own parser and project files to provide the abstract syntax tree for our type checking to tie into.
+\subsubsection{Parser}\label{treebasedparser}
 For testing purposes, our outputted type checker comes with a simple tree-based parser.
 Reflection-based parsing
-\subsubsection{Project files}
-\subsubsection{Type checking aspect}
+
+\subsubsection{Project files}\label{projectfiles}
+For ease of use the compiler will output a complete runnable Gradle project.
+It can parse simplified tree-representations of programs, and utilise these to perform the type checking.
+
+\subsubsection{Type checking aspect}\label{typecheckingaspect}
+
+\subsubsection{Executable}
+
+\subsection{Typing rule compiler}\label{typingrulecompiler}
 
@@
 \subsubsection{Tree-representation}
-\subsubsection{Type checking algorithm}
+\subsubsection{Type checking algorithm}\label{typecheckingalgorithm}
 Generate code for each node and typing rule
@@
 
-\section{Method (remove this?)} % describe how you go about solving the problem you defined. Also how do you show/prove that your solution actually works, and how well does it work.
+%\section{Method (remove this?)} % describe how you go about solving the problem you defined. Also how do you show/prove that your solution actually works, and how well does it work.
 
@@
 There is a place for insights and lamentation later on in the report, in the Discussion section.
+%TODO
+\subsection{Type checker generation}
+\begin{figure}[h]
+\begin{lstlisting}[]
+RuleSet ::= Rule*;
+
+abstract Term;
+Function : Term ::= <ID> Term*;
+Value    : Term ::= <ID>;
+
+Rule ::= <Name> Conclusion:Formula Premises:Formula*;
+
+abstract Formula;
+HasType : Formula ::= Expr:Term Ty:TyTerm;
+
+abstract TyTerm;
+TyVal   : TyTerm ::= <ID>;
+TyVar : TyTerm ::= <ID>;
+\end{lstlisting}
+  \caption{Typing rules abstract syntax tree specification}
+  \label{trastspec}
+\end{figure}
+
+The type checker generation is based on a single-responsibility principle, where each node of the abstract syntax tree is responsible for generating the code it corresponds to.
+Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of figure \ref{trastspec}, where the \verb|RuleSet| is the root node.
+The \verb|RuleSet| node consists of a list of \verb|Rule| nodes, and begins by sorting each rule by which object language node it describes.
+
+The \verb|RuleSet| node then begins by generating the outline of a JastAdd aspect \verb|TypingRules|, and the definition for the method \verb|.type()| for each object language node.
+It then calls on the \verb|Rule| node to fill in the implementation of the method.
+
+A \verb|Rule| consists of a name, a conclusion, and a possibly empty list of premises that need to be true for the conclusion to be returned.
+In the simplest case of an empty list of premises, the entire content of the rule is generated by the conclusion.
+For rules containing premises, the code becomes more involved.
+Firstly, any variables or type variables used in the rule need to be declared, followed by a check to ensure each use of a type variable represents a node with the same type.
+Then it generates an if-statement corresponding to our premises, by letting each \verb|Premise| generate a boolean expression, all of which get joined by the AND operator.
+
+Both \verb|Premise|s and \verb|Conclusion|s are represented by the same abstract class \verb|Formula|, which currently contains a single implementation \verb|HasType|, consisting of an expression and a type.
+Depending on whether it is used as a premise or conclusion, it can either represent that the expression must have the specified type or that the expression should be given the specified type.
+
+\subsection{Type checker design}
+The provided typing rules compile into \verb|type()| methods, which return a value from the Type interface.
+For the most simple typing rules, this method's implementation will consist of a single return statement.
+
+Types are declared in an AST specification, currently provided by the template, allowing simple type hierarchies.
 

### 2023-11-02 MSc.tex
Add some code samples and descriptions

@@
 \usepackage{enumitem}
+\usepackage{listings}
 \usetikzlibrary{fit,positioning,shapes.geometric}
+
+\lstdefinestyle{mystyle}{
+    basicstyle=\ttfamily,
+    breakatwhitespace=false,
+    breaklines=true,
+    captionpos=b,
+    keepspaces=true,
+    numbers=left,
+    numbersep=5pt,
+    showspaces=false,
+    showstringspaces=false,
+    showtabs=false,
+    tabsize=2
+}
+\lstset{style=mystyle}
 
@@
 Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
-\subsection{Input/output examples, with descriptions} %TODO
-Bool language
-Type variables
+%\subsection{Input/output examples, with descriptions}
+\subsection{Bools}
+
+\begin{figure}[h]
+\begin{lstlisting}[]
+Expression ::= Term;
+
+abstract Term;
+True  : Term;
+False : Term;
+Or    : Term ::= Left:Term Right:Term;
+Zero  : Term;
+\end{lstlisting}
+  \caption{AST specification for the Bools language}
+  \label{boolsast}
+\end{figure}
+\begin{figure}[h]
+\begin{lstlisting}[]
+[T-True]
+-----------
+True : Bool
+
+[T-False]
+------------
+False : Bool
+
+[T-Or]
+left : Bool,
+right : Bool
+----------------------
+Or(left, right) : Bool
+
+[T-Zero]
+----------
+Zero : Int
+\end{lstlisting}
+  \caption{Typing rules for the Bools language}
+  \label{boolstr}
+\end{figure}
+
+\begin{figure}[h]
+\begin{lstlisting}[]
+aspect TypingRules {
+
+syn Type Zero.type() {
+return new Int();}
+syn Type Or.type() {
+ASTNode left = getChild(0);
+ASTNode right = getChild(1);
+
+if(left.type().matches(new Bool()) && right.type().matches(new Bool())) {
+return new Bool();
+}
+throw new RuntimeException("Typechecking failed");
+}
+syn Type True.type() {
+return new Bool();}
+syn Type False.type() {
+return new Bool();}
+}
+\end{lstlisting}
+  \caption{Generated typing rules for the Bools language}
+  \label{boolstrgen}
+\end{figure}
+A simple example of a language is provided in figure \ref{boolsast}, consisting of three values, \verb|True|, \verb|False| and \verb|Zero|; and the boolean operator \verb|Or| with two parameters, Left and Right.
+In the AST specification, there are no limitations on what terms can be provided as parameters to the \verb|Or| operator, allowing illogical constructs such as \verb|Or(True, Zero)|.
+
+To prevent cases like these, we define typing rules via the specification in figure \ref{boolstr}.
+The typing rules for \verb|True|, \verb|False| and \verb|Zero| contain no premises and define the type as \verb|Bool| and \verb|Int| as appropriate.
+The typing rule for the \verb|Or| operator, which takes two parameters \verb|left| and \verb|right|, contains two premises, specifying that the \verb|Or| operator has the type \verb|Bool| only if both the \verb|left| and \verb|right| terms also have the type \verb|Bool|.
+
+The typing rules from figure \ref{boolstr} generate the type definitions in \ref{boolstrgen}.
+The simple rule definitions for the values compile into equally simple implementations, consisting of a single return statement of a newly constructed \verb|Type| object.
+For the \verb|Or| node, the code becomes slightly more complex.
+First two variables are defined, corresponding to the parameters in the typing rule, and assigned to the children with the index corresponding to the position among the parameters.
+This is followed by an if-statement, recusively evaluating the children's types and comparing the result the type demanded in the premise.
+If the condition is true, a \verb|Bool| type is returned.
+If not, a runtime exception is thrown to convey a type error.
+
+\subsection{If-else extension}
+
+\begin{figure}[h]
+\begin{lstlisting}[]
+IfElse : Term ::= Cond:Term Then:Term Else:Term;
+\end{lstlisting}
+  \caption{Extension for the Bools AST}
+  \label{ifelseast}
+\end{figure}
+\begin{figure}[h]
+\begin{lstlisting}[]
+[T-IfElse]
+x: Bool,
+y: a,
+z: a
+------------------
+IfElse(x, y, z): a
+\end{lstlisting}
+  \caption{Extension for the Bools typing rules with an if-else term}
+  \label{ifelsetr}
+\end{figure}
+\begin{figure}[h]
+\begin{lstlisting}[]
+syn Type IfElse.type() {
+ASTNode x = getChild(0);
+ASTNode y = getChild(1);
+ASTNode z = getChild(2);
+
+Type tyvar_a = null;
+if(tyvar_a == null) tyvar_a = y.type();
+else if (!tyvar_a.matches(y.type()))
+throw new RuntimeException("Typechecking failed: Type variable mismatch");
+if(tyvar_a == null) tyvar_a = z.type();
+else if (!tyvar_a.matches(z.type()))
+throw new RuntimeException("Typechecking failed: Type variable mismatch");
+if(x.type().matches(new Bool()) && y.type().matches(tyvar_a) && z.type().matches(tyvar_a)) {
+return tyvar_a;
+}
+throw new RuntimeException("Typechecking failed");
+}
+\end{lstlisting}
+  \caption{Generated typing rules for the if-else term}
+  \label{boolstrgen}
+\end{figure}
+
+%\begin{figure}
+%\centering
+%  \begin{minipage}[b]{.4\textwidth}
+%  \centering
+%\begin{lstlisting}[]
+%[T-IfElse]
+%x: Bool,
+%y: a,
+%z: a
+%---------
+%IfElse(x, y, z): a
+%\end{lstlisting}
+%  \captionof{figure}{A figure}
+%  \label{fig:test1}
+%\end{minipage}%
+%\begin{minipage}[b]{.6\textwidth}
+%  \centering
+%\begin{lstlisting}[]
+%syn Type IfElse.type() {
+%ASTNode x = getChild(0);
+%ASTNode y = getChild(1);
+%ASTNode z = getChild(2);
+%
+%Type tyvar_a = null;
+%if(tyvar_a == null) tyvar_a = y.type();
+%else if (!tyvar_a.matches(y.type()))
+%throw new RuntimeException("Typechecking failed: Type variable mismatch");
+%if(tyvar_a == null) tyvar_a = z.type();
+%else if (!tyvar_a.matches(z.type()))
+%throw new RuntimeException("Typechecking failed: Type variable mismatch");
+%if(x.type().matches(new Bool()) && y.type().matches(tyvar_a) && z.type().matches(tyvar_a)) {
+%return tyvar_a;
+%}
+%throw new RuntimeException("Typechecking failed");
+%}
+%\end{lstlisting}
+%  \captionof{figure}{Another figure}
+%  \label{fig:test2}
+%\end{minipage}
+%\end{figure}
+Extend bool with if-statement to show type variables
 \section{Research questions}

### 2023-11-02 MSc.tex
Begin answering research questions

@@
 \section{Research questions}
-\subsection{Research q 1}
-\subsection{Research q 2}
-\subsection{Research q 3}
-List of challenges % TODO
-  Splitting code between multiple JastAdd components (and java) can get messy
+\subsection{\rqone}
+
+\subsection{\rqtwo}
+\subsection{\rqthree}
+%Splitting code between multiple JastAdd components (and java) can get messy
+Writing idiomatic JastAdd components often requires splitting your code into many different places.
+We attempted to keep each node in the tree responsible for generating the code it corresponds to, which avoids parent nodes having to check their children's structure with \verb|isInstanceOf|, but also means even generating a simple rule can require a dozen calls between different nodes.
+There's a tradeoff to be made between the single responsibility of each component making it easy to verify its correctness, but the complicated interdependence making it hard to grasp the full picture.
+
 Using JastAdd simultaneously as a library and a build component
+
+%Having to use a hacky method to access private JastAdd methods to parse a second AST at runtime. (Beyond jastadd's scope?)
+Some of the features we intended to support, such as verifying that typing rules correspond to valid nodes within the object language AST, were complicated by having to parse the object language AST at runtime, using JastAdd as a library rather than a build tool.
+JastAdd does not expose any methods for parsing an AST tree at runtime, and we had to access undocumented, private methods through Java's reflection API to get the relevant information.
+Unfortunately, the resulting data was hard to work with and we could not allocate enough time to implement the planned features.
+It may be that this use case is beyond the intended scope of JastAdd, or it is perhaps only poorly documented.
+In either case, while our experimentation has shown it to be technically possible, the complexity is worth bearing in mind for future work.
 

### 2023-11-06 MSc.tex
Fully rewrite implementation section

@@
 Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of figure \ref{trastspec}, where the \verb|RuleSet| is the root node.
-The \verb|RuleSet| node consists of a list of \verb|Rule| nodes, and begins by sorting each rule by which object language node it describes.
 
-The \verb|RuleSet| node then begins by generating the outline of a JastAdd aspect \verb|TypingRules|, and the definition for the method \verb|.type()| for each object language node.
-It then calls on the \verb|Rule| node to fill in the implementation of the method.
+\subsubsection{Aspect generation}
+The \verb|RuleSet| node is responsible for generating the overall structure of the output file, a JastAdd aspect called \verb|TypingRules|.
+It sorts its \verb|Rule| children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation.
 
-A \verb|Rule| consists of a name, a conclusion, and a possibly empty list of premises that need to be true for the conclusion to be returned.
-In the simplest case of an empty list of premises, the entire content of the rule is generated by the conclusion.
-For rules containing premises, the code becomes more involved.
-Firstly, any variables or type variables used in the rule need to be declared, followed by a check to ensure each use of a type variable represents a node with the same type.
-Then it generates an if-statement corresponding to our premises, by letting each \verb|Premise| generate a boolean expression, all of which get joined by the AND operator.
+\subsubsection{Method implementation}
+The implementation of a typing rule can range from a single return statement to a system of declarations of variables, type variables and control flow.
 
-Both \verb|Premise|s and \verb|Conclusion|s are represented by the same abstract class \verb|Formula|, which currently contains a single implementation \verb|HasType|, consisting of an expression and a type.
-Depending on whether it is used as a premise or conclusion, it can either represent that the expression must have the specified type or that the expression should be given the specified type.
+\subsubsection{Parameters}
+When the typing rules concern a node with parameters, these can be named arbitrarily, and a variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
+These variables are then utilised in later premises or type variable declarations.
 
-\subsection{Type checker design}
-The provided typing rules compile into \verb|type()| methods, which return a value from the Type interface.
-For the most simple typing rules, this method's implementation will consist of a single return statement.
+\subsubsection{Premises/control flow}
+The rule's premises are translated into an if-statement, which checks all types using a method \verb|matches| added to all Type nodes.
+Its current implementation merely checks that the Type nodes are part of the same class, though it would be possible to override this with deeper comparisons for specific types.
 
-Types are declared in an AST specification, currently provided by the template, allowing simple type hierarchies.
+\subsubsection{Type variables}
+In the case of type variables, the generated code takes a dynamic approach to simplify the implementation.
+Each type variable that appears in the conclusion is instantiated to a null value, and the code generated by each premise first checks if the value is still null and either assigns the variable its own type or tests to make sure the existing type matches its own.
+
+\subsubsection{Error handling}
+The current error handling consists of throwing a runtime exception at the first discovered type error, ceasing the type checking immediately.
+This method was chosen for its simplicity of implementation, rather
 

### 2023-11-06 MSc.tex
Various formatting improvements

@@
 \lstdefinestyle{mystyle}{
-    basicstyle=\ttfamily,
+  basicstyle=\ttfamily\footnotesize,
   breakatwhitespace=false,
@@
 \lstset{style=mystyle}
+\lstdefinelanguage{jrag}[]{java}{
+  morekeywords={aspect, syn, eq, inh}
+}
 
@@
 \begin{figure}[h]
-\begin{lstlisting}[]
+\begin{lstlisting}[language=jrag] 
 aspect TypingRules {
@@
   syn Type Zero.type() {
-return new Int();}
+    return new Int();
+  }
+
   syn Type Or.type() {
@@
 
-if(left.type().matches(new Bool()) && right.type().matches(new Bool())) {
+    if(left.type().matches(new Bool()) &&
+       right.type().matches(new Bool())) {
       return new Bool();
@@
   }
+
   syn Type True.type() {
-return new Bool();}
+    return new Bool();
+  }
+
   syn Type False.type() {
-return new Bool();}
+    return new Bool();
+  }
 }
 \end{lstlisting}
-  \caption{Generated typing rules for the Bools language}
+  \caption{Generated typing rules for the Bools language (reformatted for readability)}
   \label{boolstrgen}
@@
 First two variables are defined, corresponding to the parameters in the typing rule, and assigned to the children with the index corresponding to the position among the parameters.
-This is followed by an if-statement, recusively evaluating the children's types and comparing the result the type demanded in the premise.
+This is followed by an if-statement, recursively evaluating the children's types and comparing the result the type demanded in the premise.
 If the condition is true, a \verb|Bool| type is returned.
@@
 \end{lstlisting}
-  \caption{Extension for the Bools AST}
+  \caption{Extension for the Bools AST in figure \ref{boolsast}}
   \label{ifelseast}
@@
 \end{lstlisting}
-  \caption{Extension for the Bools typing rules with an if-else term}
+  \caption{Extension for the Bools typing rules in figure \ref{boolstr} with an if-else term}
   \label{ifelsetr}
@@
 \begin{figure}[h]
-\begin{lstlisting}[]
+\begin{lstlisting}[language=jrag]
 syn Type IfElse.type() {
@@
   Type tyvar_a = null;
-if(tyvar_a == null) tyvar_a = y.type();
+  if(tyvar_a == null)
+    tyvar_a = y.type();
   else if (!tyvar_a.matches(y.type()))
     throw new RuntimeException("Typechecking failed: Type variable mismatch");
-if(tyvar_a == null) tyvar_a = z.type();
+  if(tyvar_a == null)
+    tyvar_a = z.type();
   else if (!tyvar_a.matches(z.type()))
     throw new RuntimeException("Typechecking failed: Type variable mismatch");
-if(x.type().matches(new Bool()) && y.type().matches(tyvar_a) && z.type().matches(tyvar_a)) {
+  if(x.type().matches(new Bool()) &&
+     y.type().matches(tyvar_a) &&
+     z.type().matches(tyvar_a)) {
     return tyvar_a;
@@
 \end{lstlisting}
-  \caption{Generated typing rules for the if-else term}
-  \label{boolstrgen}
+  \caption{Generated typing rules for the if-else term, extending figure \ref{boolstrgen} (reformatted for readability)}
+  \label{ifelsetrgen}
 \end{figure}

### 2023-11-16 MSc.tex
Remove seperator from typing rules without premise
Flesh out language description

@@
 \subsubsection{Language}
-- Figure of example natural deduction rules (from Pierce?) and ASCII representation
-No infix notation
-No environments
 
-Our language can express a subset of the notation used by Pierce, with some limitations.
-Expressions aren't written in the language's native syntax, instead using a simplified syntax resembling the AST classes.
-Values and functions take their names directly from the AST-nodes, and any parameters are included within parenthesis.
-This avoids much of the difficulty in parsing different languages, which can have wildly different syntax rules.
+\begin{figure}[h]
+\centering
+\begin{minipage}{.45\textwidth}
+  \centering
+  \begin{equation}
+    \tag{T-True}
+    \text{true} : \text{Bool}
+  \end{equation}
+  %\caption*{Pierce syntax}
+  %\label{fig:sub1}
+\end{minipage}%
+\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
+  \centering
+\begin{lstlisting}[]
+[T-True]
+True : Bool
+\end{lstlisting}
+  %\caption*{ASCII representation}
+  %\label{fig:sub2}
+\end{minipage}
+\caption{A typing rule for the value true in natural semantic syntax on the left and its equivalent ASCII representation on the right}
+\label{trexampletrue}
+\end{figure}
+\begin{figure}[h]
+\centering
+\begin{minipage}{.45\textwidth}
+  \centering
+  \begin{equation}
+    \tag{T-If}
+    \frac{
+      t_1 : \text{Bool}\quad
+      t_2 : \text{T}\quad
+      t_3 : \text{T}
+    }{
+      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \text{T}
+    }
+  \end{equation}
+  %\caption*{Pierce syntax}
+  %\label{fig:sub1}
+\end{minipage}%
+\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
+  \centering
+\begin{lstlisting}[]
+[T-If]
+t1 : Bool,
+t2 : t,
+t3 : t
+-----------------
+If(t1, t2, t3): t
+\end{lstlisting}
+  %\caption*{ASCII representation}
+  %\label{fig:sub2}
+\end{minipage}
+\caption{A typing rule describing an if-expression in natural semantic syntax and its equivalent ASCII representation}
+\label{trexampleif}
+\end{figure}
+
+%- Figure of example natural deduction rules (from Pierce?) and ASCII representation
+%No infix notation
+%No environments
+
+To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{pierce}.
+Our language can express a subset of the notation used by Pierce, with some differences and limitations.
+
+Figures \ref{trexampletrue} and \ref{trexampleif} show two rules from Pierce in both their original syntax and our ASCII representation side by side.
+The former figure describes a simple rule named T-True, which assigns the \verb|true| expression the type \verb|Bool|.
+
+The largest difference can be seen in the expression to be typed, where Pierce uses the object language's native syntactic form, whereas our representation is based on the corresponding AST nodes.
+In the first example, the difference is minor (\verb|true| and \verb|True| respectively) but the difference is more apparent in figure \ref{trexampleif}, where instead of the full syntactic form, the expression is written as an AST node with its children represented as parameters.
+%Note the difference between the \verb|true| written in Pierce, representing the native syntactic form of the expression, and the \verb|True| in the ASCII representation, representing the name of the corresponding AST node.
+%This difference is more apparent in figure \ref{trexampleif}, where instead of the full syntactic form, the expression is written as an AST node with its children represented as parameters.
+This change is necessary to allow us to type arbitrary languages without needing to understand their syntax and grammar, though it does make translating rules from one syntax to the other less straightforward.
+
+Another difference is the use of the metavariable \verb|T|, which has been switched to the lowercase \verb|t| in order to distinguish concrete types and type variables.
+This change also aligns the syntax for type terms with the regular terms, where concrete nodes (such as \verb|If|) are written uppercase and variables (\verb|t1|, \verb|t2|, \verb|t3|) in lowercase.
+Commas are also used to separate the premises, and the horizontal line separating premises from the conclusion is represented as a series of dashes.
 
@@
 [T-True]
------------
 True : Bool
@@
 [T-False]
-------------
 False : Bool
@@
 [T-Zero]
-----------
 Zero : Int

### 2023-11-16 MSc.tex
Describe template tree-based parser

@@
 \subsubsection{Parser}\label{treebasedparser}
-For testing purposes, our outputted type checker comes with a simple tree-based parser.
-Reflection-based parsing
+For testing purposes, our template comes with a generic tree-based parser.
+Programs are written as a tree of AST node names followed by potential parenthesis within which a comma separated list of child nodes are written.
+It recursively parses the tree, using Java's reflection API to find a class with a matching name, and a constructor with the matching number of parameters.
+Figure \ref{boolinputexample} shows an example input file for the Bools language, which will be discussed further in section \ref{boolsast}.
+
+\begin{figure}[h]
+\begin{lstlisting}[]
+Expression(
+  Or(
+    True,
+    Or(
+      True,
+      False
+    )
+  )
+)
+\end{lstlisting}
+  \caption{Example input file for the Bools language}
+  \label{boolinputexample}
+\end{figure}
+
+This approach does have the issue of occasionally parsing invalid programs, due to the existence of a hidden default constructor.
+This leads the parser to always accept nodes with no children, such as \lstinline{Expression(Or())}, even if the AST definition requires them.
 

### 2023-11-16 MSc.tex
Expand on method implementation

@@
 Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of figure \ref{trastspec}, where the \verb|RuleSet| is the root node.
+From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.
 
@@
 
-\subsubsection{Method implementation}
+\subsubsection{Method implementation overview}
 The implementation of a typing rule can range from a single return statement to a system of declarations of variables, type variables and control flow.
+These are explained in further detail in subsequent sections, but here follows a quick overview.
+
+Firstly, the algorithm checks if the number of premises is zero, in which case it immediately generates the conclusion, consisting of a return statement.
+If there are premises, it first generates code for the declarations of variables and type variables.
+This is followed by an if-statement to test whether the premises hold true, containing a return statement.
+Should the premises not hold, a type error is produced.
 
@@
 The rule's premises are translated into an if-statement, which checks all types using a method \verb|matches| added to all Type nodes.
-Its current implementation merely checks that the Type nodes are part of the same class, though it would be possible to override this with deeper comparisons for specific types.
+
+The current implementation of \verb|Type.matches(Type t)| checks that the Type it is called on is either of the same class or a subtype of the type \verb|t|.
+This allows the language limited sub-typing at the type checker generation step, though the pros and cons of this haven't been evaluated.
+It is also possible to override the \verb|matches| implementation for a specific Type, by adding a definition to a JastAdd aspect included in the output type checker.
 
@@
 
-\subsubsection{Error handling}
+\subsubsection{Error handling}\label{errorhandling}
 The current error handling consists of throwing a runtime exception at the first discovered type error, ceasing the type checking immediately.
-This method was chosen for its simplicity of implementation, rather
+This method was chosen solely for its simplicity of implementation, rather than its functionality.
+
+It provides a stack trace allowing you to see where in the type checking code the error occurs (such as what typing rule fails), but doesn't currently provide any information about what section of the input program causes the error.
+This tends to result in errors that are more useful for debugging the typing rule implementation, rather than providing useful information to the user of the type checker, where there is much room for improvements.
+
+%Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of figure \ref{trastspec}, where the \verb|RuleSet| is the root node.
+%The \verb|RuleSet| node consists of a list of \verb|Rule| nodes, and begins by sorting each rule by which object language node it describes.
+%
+%The \verb|RuleSet| node then begins by generating the outline of a JastAdd aspect \verb|TypingRules|, and the definition for the method \verb|.type()| for each object language node.
+%It then calls on the \verb|Rule| node to fill in the implementation of the method.
+%
+%A \verb|Rule| consists of a name, a conclusion, and a possibly empty list of premises that need to be true for the conclusion to be returned.
+%In the simplest case of an empty list of premises, the entire content of the rule is generated by the conclusion.
+%For rules containing premises, the code becomes more involved.
+%Firstly, any variables or type variables used in the rule need to be declared, followed by a check to ensure each use of a type variable represents a node with the same type.
+%Then it generates an if-statement corresponding to our premises, by letting each \verb|Premise| generate a boolean expression, all of which get joined by the AND operator.
+%
+%Both \verb|Premise|s and \verb|Conclusion|s are represented by the same abstract class \verb|Formula|, which currently contains a single implementation \verb|HasType|, consisting of an expression and a type.
+%Depending on whether it is used as a premise or conclusion, it can either represent that the expression must have the specified type or that the expression should be given the specified type.
+
+%\subsection{Type checker design}
+%The provided typing rules compile into \verb|type()| methods, which return a value from the Type interface.
+%For the most simple typing rules, this method's implementation will consist of a single return statement.
+%
+%Types are declared in an AST specification, currently provided by the template, allowing simple type hierarchies.
 

### 2023-11-16 MSc.tex
Make various minor changes

@@
 \usepackage{listings}
+\usepackage{subcaption}
 \usetikzlibrary{fit,positioning,shapes.geometric}
@@
 }
+
+\setcounter{secnumdepth}{4} % Number subsubsections
 
@@
 
-\keywords{MSc, BSc, template, report, style, structure}
+%\keywords{MSc, BSc, template, report, style, structure}
+\keywords{type checking, type theory, code generation}
 
@@
 \subsubsection{Static type checking}
+\subsection{Typing rules}
+use left hand side of these again \ref{trexampleif}
 
 \section{Reference Attribute Grammars}
+\cite{RAG}
 \subsection{Reference Attribute Grammars}
@@
     \node (trparsing) [right=of typerules, ellipse] {Scan,\\parse,\\build};
-    \node (trparser)  [above=of trparsing]          {Typing rule parser};
-    \node (trast)     [right=of trparsing]          {Typing rules AST};
+    \node (trparser)  [above=of trparsing]          {Typing rule parser\\(\ref{typingrulesparser})};
+    \node (trast)     [right=of trparsing]          {Typing rules AST\\(\ref{typingrulesast})};
 
@@
     \node (outputparsing)      [left=of inputast, ellipse]     {Scan,\\parse,\\build};
-    \node (typeerrors)         [right=of outputtypechecking]   {Type errors};
+    \node (typeerrors)         [right=of outputtypechecking]   {Type errors\\(\ref{errorhandling})};
     \node (inputprog)          [left=of outputparsing]         {Input program};
@@
 
-\subsubsection{Parser}
+\subsubsection{Parser}\label{typingrulesparser}
 \subsubsection{Abstract syntax tree}
@@
 Recursive
-\subsubsection{Error handling}
 
@@
 
-\cite{Hedin2011}
 \section{Implementation} % if your work involved building an artefact/implementation, give the details here.
@@
 There is a place for insights and lamentation later on in the report, in the Discussion section.
-%TODO
+
 \subsection{Type checker generation}
@@
 Source code should rather be made available for download on a website or on-line repository of your choosing.
-Git repository link
+\chapter{Source code}
+\url{https://github.com/nichobi/thesis-project}
 

### 2023-12-02 MSc.tex
Commit a bunch of changes

@@
 \usepackage{listings}
+\usepackage{svg}
 \usepackage{subcaption}
@@
 \setcounter{secnumdepth}{4} % Number subsubsections
+%\setcounter{tocdepth}{4} % Include subsubsections in table of contents
 
@@
 Furthermore, making changes to the typing rules of a language can require extensive rewrites in many different parts of the type checker's codebase.
-% This can hinder experimentation
-
-In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules.
 
 % Define the problem you are solving (or trying to solve).
-% Specify your contributions.
-What does this particular work/report bring to the research are or to the body of knowledge?
-How is the work divided between the co-authors?
-(This part is essential to pinpoint individual work.
-For theses with two authors, it is compulsory to identify which author has contributed with which part, both with respect to the work and the report.)
+In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using .
+
+% Specify your contributions. What does this particular work/report bring to the research are or to the body of knowledge?
 \section{Research questions}
@@
 The tradition in some research area is to place this part at the end of the report (check with your supervisor).
+Pacak et al. implemented a compiler for generating incremental type checkers using Datalog.\cite{Pacak}
+While the goal is similar to ours, they take a different approach, by converting typing rules into the declarative logic programming language.
+Their work is thereby largely focused on expressing typing rules within the limits of the Datalog language, which can only compute finite relations.
+We have a significantly different compilation target, JastAdd, which allows us to use reference attribute grammars to evaluate the typing relations, which we write in a more iterative approach.
 
@@
 %\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
+This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
 \section{Type systems}
 \subsection{Data types}
+Computer software generally needs to handle and manipulate several different types of data in one program, such as numbers or text.
+Most programming languages come with a number of built-in data types, from different kinds of numbers (integers, floating point) to strings or lists and arrays.
+
+Different parts of programs may have different constraints on the types of expressions, such as a print function requiring a string of text, or a plus function requiring some kind of number.
+
 \subsection{Type checking}
+There are two main approaches to ensuring that the aforementioned constraints are upheld, dynamic and static type-checking.
+Dynamic checking is done during runtime, with each type checked only when needed.
+Static checking is done on the code before it's run, usually as part of compilation, preventing many faulty programs from being run at all.
+
+As our project generates a static type checker, we will focus on them.
+Conservative
+
 \subsubsection{Dynamic type checking}
 \subsubsection{Static type checking}
-\subsection{Typing rules}
+\subsection{Type systems}
+A type system consists of two parts, a type syntax, describing the valid types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
+\subsubsection{Type syntax}
+A type syntax is a definition of the valid types within a language.
+\subsubsection{Typing rules}
 use left hand side of these again \ref{trexampleif}
@@
     \node (typecheckingalgorithm) at (outputtypechecking |- templateparser)              {Type checking\\aspect (\ref{typecheckingaspect})};
-    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm)] {};
+    \node (templatetypesyntax)    [right=of typecheckingalgorithm]                       {Type syntax\\AST};
+    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm) (templatetypesyntax)] {};
     \node                         [above, draw=none] at (template.north)                 {Template};
@@
     \draw[dashed] (templateprojectfiles.north east) to (outputtypechecking);
+    \draw[dashed] (templatetypesyntax)              to (outputtypechecking);
 
@@
 ASCII representation of natural deduction %BNF (is it BNF?)
-\subsubsection{Language}
+\subsubsection{Language}\label{sectionlanguage}
 
@@
 \subsubsection{Parser}\label{typingrulesparser}
-\subsubsection{Abstract syntax tree}
+The parser for typing rules is written using JFlex\cite{jflex}, a scanner generator, and Beaver\cite{beaver}, an LALR parser generator.
+JFlex takes a series of regular expressions to generate a tokenizer for the parsed language.
+The tokens are passed on to Beaver along with a parser definition in extended Backus-Naur form (EBNF), to generate a parser.
 
-\subsection{Type checker template}
-In addition to the compiled files, there are a number of files that are added to the output file that are identical for all generated typecheckers, which we referr to as the template.
-Some of these are essential to the functions of the type checker, and others are included for convenience of testing.
+\subsubsection{Abstract syntax tree}\label{typingrulesast}
+\begin{figure}[h]
+\begin{lstlisting}[]
+RuleSet ::= Rule*;
+
+Rule ::= <Name> Conclusion:Formula Premises:Formula*;
+
+abstract Formula;
+HasType : Formula ::= Expr:Term Ty:TyTerm;
+
+abstract Term;
+Function : Term ::= <ID> Term*;
+Value    : Term ::= <ID>;
+
+abstract TyTerm;
+TyVal : TyTerm ::= <ID>;
+TyVar : TyTerm ::= <ID>;
+\end{lstlisting}
+  \caption{Typing rules abstract syntax tree specification}
+  \label{trastspec}
+\end{figure}
+
+The abstract syntax tree for the typing rules language consists of a \lstinline{RuleSet} root node.
+The root node contains only a list of rules, where each rule has a name, a conclusion and a list of premises.
+The rule's name is currently unused but is intended to be utilised in error reporting, to clearly specify what rules are available for each node, or which rule is causing an error.
+
+Both conclusion and premises are represented by the abstract HasType node, consisting of a term, which may be a function or value; and a typeterm, representing either a type value or a type variable.
+For the conclusion, this node represents what type should be assigned to the term on the left, whereas for the premises, the node represents a requirement for the node on the left to have the type on the right.
+
+\subsection{Template}
+In addition to the compiled files, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the template.
+Some of these are key components of the type checker, but others are dummy components included solely for convenience of testing.
 We expect that in a real world use case, our software would integrate into an object language's compiler pipeline, and utilise its own parser and project files to provide the abstract syntax tree for our type checking to tie into.
+For testing purposes, we've written generic versions of these components, which work for any supported object language.
 \subsubsection{Parser}\label{treebasedparser}
@@
 Programs are written as a tree of AST node names followed by potential parenthesis within which a comma separated list of child nodes are written.
-It recursively parses the tree, using Java's reflection API to find a class with a matching name, and a constructor with the matching number of parameters.
+It recursively parses the tree, using Java's reflection API to find a class with a matching name, and a constructor with parameters of the same amount and types.
 Figure \ref{boolinputexample} shows an example input file for the Bools language, which will be discussed further in section \ref{boolsast}.
@@
 \end{lstlisting}
-  \caption{Example input file for the Bools language}
+  \caption{Example tree-representation input for the Bools language}
   \label{boolinputexample}
@@
 
-This approach does have the issue of occasionally parsing invalid programs, due to the existence of a hidden default constructor.
-This leads the parser to always accept nodes with no children, such as \lstinline{Expression(Or())}, even if the AST definition requires them.
+This approach does have the issue of occasionally parsing invalid programs, due to the reflection API finding an additional parameterless default constructor.
+This leads the parser to always accept nodes written with no children, such as \lstinline{Expression(Or())}, even if the AST definition requires them.
+These children will be null, causing issues further down the line, though as this parser is only intended for testing it is of little concern.
 
@@
 \subsubsection{Type checking aspect}\label{typecheckingaspect}
+The type checking aspect contains the declaration of the \lstinline{type()} attribute for all AST nodes, as well as a default implementation.
+This default implementation immediately produces a type error, to cause an error when the type of an expression not defined in the typing rules is demanded.
 
-\subsubsection{Executable}
+It also contains the implementation of the type matching method, which can be seen in figure \ref{typematches}.
+It's defined for all \lstinline{Type} nodes, and checks whether the type's class is equal to or a subclass of the parameter type.
 
-\subsection{Typing rule compiler}\label{typingrulecompiler}
+\begin{figure}[h]
+  \begin{lstlisting}[language=jrag]
+syn boolean Type.matches(Type t) {
+  return t.getClass().isAssignableFrom(getClass());
+}
+  \end{lstlisting}
+  \caption{Definition of the type matching method}
+  \label{typematches}
+\end{figure}
 
-\subsection{Output type checker}
-\subsubsection{Tree-representation}
-\subsubsection{Type checking algorithm}\label{typecheckingalgorithm}
-Generate code for each node and typing rule
-Simple algorithm
-Recursive
+%\subsubsection{Executable}
+
+%\subsection{Typing rule compiler}\label{typingrulecompiler}
+
+%\subsection{Output type checker}
+%\subsubsection{Tree-representation}
+%\subsubsection{Type checking algorithm}\label{typecheckingalgorithm}
+%Generate code for each node and typing rule
+%Simple algorithm
+%Recursive
 
@@
 
-\section{Implementation} % if your work involved building an artefact/implementation, give the details here.
+\section{Implementation}\label{typingrulecompiler} % if your work involved building an artefact/implementation, give the details here.
 Note, that this should not, as a rule, be a chronological description of your efforts, but a view of the result.
@@
 
-\subsection{Type checker generation}
-\begin{figure}[h]
-\begin{lstlisting}[]
-RuleSet ::= Rule*;
-
-abstract Term;
-Function : Term ::= <ID> Term*;
-Value    : Term ::= <ID>;
-
-Rule ::= <Name> Conclusion:Formula Premises:Formula*;
-
-abstract Formula;
-HasType : Formula ::= Expr:Term Ty:TyTerm;
-
-abstract TyTerm;
-TyVal : TyTerm ::= <ID>;
-TyVar : TyTerm ::= <ID>;
-\end{lstlisting}
-  \caption{Typing rules abstract syntax tree specification}
-  \label{trastspec}
-\end{figure}
-
+%\subsection{Type checker generation}
 The type checker generation is based on a single-responsibility principle, where each node of the abstract syntax tree is responsible for generating the code it corresponds to.
@@
 
-\subsubsection{Aspect generation}
+This section details the implementation of our type checker generator, on a feature by feature basis.
+
+%\subsubsection{Aspect generation}
+\subsection{Aspect generation}
 The \verb|RuleSet| node is responsible for generating the overall structure of the output file, a JastAdd aspect called \verb|TypingRules|.
@@
 
-\subsubsection{Method implementation overview}
+%\subsubsection{Method implementation overview}
+\subsection{Method implementation overview}
 The implementation of a typing rule can range from a single return statement to a system of declarations of variables, type variables and control flow.
@@
 
-\subsubsection{Parameters}
+%\subsubsection{Parameters}
+\subsection{Parameters}
 When the typing rules concern a node with parameters, these can be named arbitrarily, and a variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
@@
 
-\subsubsection{Premises/control flow}
+%\subsubsection{Premises/control flow}
+\subsection{Premises/control flow}
 The rule's premises are translated into an if-statement, which checks all types using a method \verb|matches| added to all Type nodes.
@@
 
-\subsubsection{Type variables}
+%\subsubsection{Type variables}
+\subsection{Type variables}
 In the case of type variables, the generated code takes a dynamic approach to simplify the implementation.
@@
 
-\subsubsection{Error handling}\label{errorhandling}
+%\subsubsection{Error handling}\label{errorhandling}
+\subsection{Error handling}\label{errorhandling}
 The current error handling consists of throwing a runtime exception at the first discovered type error, ceasing the type checking immediately.
@@
 \end{figure}
+
 \begin{figure}[h]
@@
 %\end{figure}
-Extend bool with if-statement to show type variables
+As an example of rules utilising type variables, we can extend the Bools language with an if-else term.
+As shown in the AST extension in figure \ref{ifelseast}, it consists of three terms: a conditional, a term to be evaluated if the condition is true and a condition to evaluate if false.
+A typing rule of this if-else term can be seen in figure \ref{ifelsetr}.
+The rule has three premises, the first of which declares the conditional must have the type \verb|Bool|.
+The next two declare that the two other children must be of the type \verb|a|, with the lowercase indicating a type variable.
+This type variable \verb|a| also subsequently appears in the resulting type of the if-else term.
+
+The code for this rule is shown in figure \ref{ifelsetrgen}, and begins with three parameter declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
+For each invocation of the type variable, a code section is generated, which checks if the type variable is still null, and if so assigns it to the type of the left-hand side term, or if it has been assigned checks whether the types match each other.
+Should any non-matches be found, a type error is thrown.
+
+The rest of the method is similar, with premises being checked -- including a redundant check of the type variables -- before returning the type variable, or a type error being thrown.
+
 \section{Research questions}
 \subsection{\rqone}
+Our project has shown that generating type checkers using RAGs is a viable solution for a number of different language constructs.
+%To assess our implementation, we'll go through the examples used by Pierce\cite{Pierce} to see which cases we can successfully express.
+To evaluate the completeness of our implementation, we refer to Pierce\cite{Pierce} as a reference work.
+The typing rules are introduced in chapter 8, starting with the simple examples shown in section \ref{sectionlanguage}.
+We've implemented support for simple type declarations, where a value or function always has a certain type.
+These simple rules aren't of much interest in themselves but are an important backbone to the subsequent analysis.
+We also support type declarations with premises that must hold for the rule to apply, allowing us to express relationships between types and bringing with them the possibility of type errors.
+
+Type variables extend these relationships further by providing a layer of abstraction, allowing us to write rules where the types aren't precisely defined.
+This enables rules where the return type of an expression depends on its child expressions, or where child expressions must have the same types.
+
+A major feature missing from our implementation is the concept of type environments or typing contexts, providing a mapping from variables and functions to types.
+This prevents us from generating type checkers for even simple general purpose languages.
 
 \subsection{\rqtwo}
+We have approached parsing typing rules into a JastAdd attribute \verb|type()|, which either returns a \verb|Type| value, or throws a runtime exception if the expression is invalidly typed.
+
+We've chosen to represent the right-hand side of type equations, the types themselves, as an abstract AST class.
+While our current implementation only supports simple types, this representation gives us a lot of flexibility for future extensions.
+Parametric types, such as lists where the type of the list contains information about the types of its elements, could be expressed as simply as \lstinline{MyList : Type ::= Type;}.
+The current implementation of \lstinline[language=java]{matches()} would however consider all lists to match, without comparing the child nodes.
+Perhaps the default implementation could be extended to compare all child elements, or these more advanced types would require manually overriding the method.
+
+We've implemented premises as conditionals, where we check if each premise holds true before returning the type, or falling through to a type error if the premises don't hold.
+It would also be possible to instead check each condition one by one, and throw a type error as soon as one doesn't match.
+This would perhaps make it easier to implement informative type errors, as you could report precisely which premise is failing.
+It would however make it harder to implement having multiple valid rules describing the type of an expression, as the first failure would seize the evaluation of the node.
+
+A cleaner solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
+Instead of reporting the reason any individual rule failed, the error might instead display each possible rule,
+
 \subsection{\rqthree}
@@
 JastAdd does not expose any methods for parsing an AST tree at runtime, and we had to access undocumented, private methods through Java's reflection API to get the relevant information.
-Unfortunately, the resulting data was hard to work with and we could not allocate enough time to implement the planned features.
+Unfortunately, the resulting data was hard to work with, and we could not allocate enough time to implement the planned features.
 It may be that this use case is beyond the intended scope of JastAdd, or it is perhaps only poorly documented.
 In either case, while our experimentation has shown it to be technically possible, the complexity is worth bearing in mind for future work.
+
+Not being able to perform these error checks at runtime means our translation will emit nonfunctional typing definitions if a mistake is written in the typing rules.
+These errors will then go unreported until the output project is compiled or run, at which point the user will have to attempt to trace these errors back to the typing rule definition, to find the actual cause.
 
@@
 \section{Development learnings}
+Our experiments have shown that reference attribute grammars are a flexible method for implementing type checker generators.
+Implementing the first top-to-bottom functioning product was challenging and took quite a lot of work, but extending it with additional features such as type variables turned out to be a surprisingly painless experience.
 
@@
 \section{Future work}
-Improved error handling
-Check for completeness of rules
-Verify rules against AST
-Handle multiple rules for the same node (add(int, int): int, add(double, double): double)
-Better support for numbers, strings, etc.
+\subsection{Improved error handling}
+The current exception-based error handling is rather primitive and leaves much to be desired.
+Some improvements, such as provide more useful error messages, could be possible by making minor changes to how these exceptions are generated.
+Information such as what part of the input program the type error occurred at, what the failing typing rule was and what prerequisites it has, could be added to the exception message, for a quick usability improvement.
+
+An improvement demanding more extensive changes would be collecting multiple errors, instead of seizing the type checker at the first discovered error.
+This could be implemented by adding another type class to be returned when an error occurs, representing the "Any" type.
+Any \verb|matches()| call on this type would return true, to avoid one error cascading into many, but still allowing other unrelated errors to be discovered.
+This implementation would also need new functionality to collect the error messages, perhaps by splitting the type resolution and type errors into separate attributes for each node, where the type errors propagate to a program-wide collection.
+
+\subsection{Check for completeness of rules}
+
+\subsection{Verify rules against AST}
+\subsection{Handle multiple rules for the same node (add(int, int): int, add(double, double): double)}
+\subsection{Better support for numbers, strings, etc.}
+\subsection{Improved type checking algorithm}
 

### 2023-12-07 MSc.tex
Expand on a variety of sections

@@
 %\keywords{MSc, BSc, template, report, style, structure}
-\keywords{type checking, type theory, code generation}
+\keywords{type checking, type theory, code generation, reference attribute grammars}
 
@@
 % Define the problem you are solving (or trying to solve).
-In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using .
+In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of reference attribute grammars.
+
+%TODO: Give an overview of the other sections
 
@@
 The tradition in some research area is to place this part at the end of the report (check with your supervisor).
-Pacak et al. implemented a compiler for generating incremental type checkers using Datalog.\cite{Pacak}
+Pacak et al. implemented a compiler for generating incremental type checkers using Datalog\cite{Pacak}.
 While the goal is similar to ours, they take a different approach, by converting typing rules into the declarative logic programming language.
+
 Their work is thereby largely focused on expressing typing rules within the limits of the Datalog language, which can only compute finite relations.
 We have a significantly different compilation target, JastAdd, which allows us to use reference attribute grammars to evaluate the typing relations, which we write in a more iterative approach.
+
+Additionally, their project defines a formalism for expressing typing rules within the Datalog language, ours defines a language which aims to keep as close as possible to the traditional inference rules used for typing rules, to minimise the difficulty of translation.
 
@@
 This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
-\section{Type systems}
+The first section explains abstract syntax trees, which are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
+The next section details type systems, by first introducing data types, then explaining the components required for a type system, followed by a description of static type checkers.
+
+\section{Abstract syntax trees}
+An abstract syntax tree (AST) is a tree-based representation of the syntax of a program.
+Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
+It's abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
+Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this information.
+A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if-else node, with named branches pointing towards the condition and the two values.
+
+This layer of abstraction makes it easier to write analyses for the program, as we can focus on the key elements and structure.
+
+\section{Type theory}
+In this section we'll discuss the elements of type theory required to understand this thesis.
+Firstly, we'll discuss what types are and give some examples of their use.
+Next, we'll describe type system, which form a relation between types and expressions.
+
 \subsection{Data types}
@@
 There are two main approaches to ensuring that the aforementioned constraints are upheld, dynamic and static type-checking.
-Dynamic checking is done during runtime, with each type checked only when needed.
-Static checking is done on the code before it's run, usually as part of compilation, preventing many faulty programs from being run at all.
+Dynamic checking is done performed runtime, with each type checked only when needed.
+Static checking is performed on the code before it's run, usually as part of compilation, preventing many faulty programs from being run at all.
 
-As our project generates a static type checker, we will focus on them.
+As our project generates a static type checker, we will focus our explanations on that topic.
+Static checkers analyse the structure of the source code, often in the form of an abstract syntax tree.
 Conservative
 
-\subsubsection{Dynamic type checking}
-\subsubsection{Static type checking}
 \subsection{Type systems}
-A type system consists of two parts, a type syntax, describing the valid types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
+A type system consists of two parts, a type syntax, describing the available types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
 \subsubsection{Type syntax}
 A type syntax is a definition of the valid types within a language.
+It can consist of fundamental types, such as integers or booleans.
+There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
+
 \subsubsection{Typing rules}
-use left hand side of these again \ref{trexampleif}
+The typing relation, notated \lstinline{t : T}, is a link between the expressions of a language and the types of the type syntax.
+The left-hand side is an expression in the object lanuage, typically written in its native syntax, the right-hand side a type from the aforementioned type syntax.
+
+Typing rules are a manner of inference rules, which in their simplest form, consist of no more than a typing relation.
+The typing rule \lstinline{T-True} seen in figure \ref{typingruleexampletrue}, for example, simply assigns the type of \lstinline{Bool} to the expression \lstinline{true}.
+Additionally, the typing rules in these examples have a name, given in parentheses next to the rule, though this is merely for reference.
+
+\begin{figure}[h]
+  \begin{equation}
+    \tag{T-True}
+    \text{true} : \text{Bool}
+  \end{equation}
+  \caption{Typing rule for a true expression}
+  \label{typingruleexampletrue}
+\end{figure}
+
+Typing rules may also have premises, a number of conditions which must hold for the typing rule to apply.
+The notational convention here is to put the premises above a bar, with the conclusion underneath it.
+
+\begin{figure}[h]
+  \begin{equation}
+    \tag{T-If}
+    \frac{
+      t_1 : \text{Bool}\quad
+      t_2 : \text{T}\quad
+      t_3 : \text{T}
+    }{
+      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \text{T}
+    }
+  \end{equation}
+  \caption{Typing rule for an if-expression}
+  \label{typingruleexampleif}
+\end{figure}
+
+Figure \ref{typingruleexampleif} shows an example rule, typing an if-expression.
+The bottom line, the conclusion, contains the keywords of the if-expression and three variables $t_i$ which represent any term within the language.
+Above the bar are three premises, placing constraints on the three variables, which must hold for the typing rule to apply.
+The first constrains $t_1$ to the type bool, the second two constrain $t_2$ and $t_3$ to a type variable T, which represents any single arbitrary type within the type syntax.
+While T may be any type, it must consistently be the same type within any single evaluation of the rule, i.e. the two terms must have the same type.
+This T also appears in the conclusion, as the resulting type of the if-expression, meaning the result of the entire expression will be that of the $t_2$ and $t_3$ terms.
 
 \section{Reference Attribute Grammars}
-\cite{RAG}
+This section introduces the subject of Reference Attribute Grammars (RAGs), first explaining the theory behind them, then introducing JastAdd, an implementation of RAGs within a Java meta-compiliation system\cite{jastadd}.
+
 \subsection{Reference Attribute Grammars}
+Reference Attribute Grammars (RAGs) are an extension to Attribute Grammars (AGs), a formalism used for specifying the semantics of programming languages.
+AGs provide a systematic way to define computations on the nodes of an abstract syntax tree, associating attributes with language constructs and are commonly used in the context of compiler construction, where they help define the translation of source code into executable code\cite{AGs}.
+
+RAGs extend these properties by introducing a new kind of attribute, allowing references to other nodes in the AST, and through the references accessing the attributes of other nodes.
+This provides a simplified and more flexible method for interconnectedness between nodes\cite{RAG}.
+
 \subsection{JastAdd}
-\subsubsection{AST definitions}\label{astdef}
+JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion.
+\cite{JastAdd}
+\subsubsection{AST specification}\label{astdef}
+One of the key components of a JastAdd projects are the \verb|.ast| files, which declare the structure of the abstract syntax tree.
+This specification is then used by JastAdd to generate Java classes representing the nodes of the AST.
+These AST classes can have a variety of forms, supporting both abstract classes and subclasses, and can contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.
+For a detailed overview of the syntax, see the reference manual\cite{jastaddref}.
+
+\subsubsection{RAG Modules}
+\begin{figure}[h]
+\begin{lstlisting}[language=jrag]
+aspect TypingRules {
+  syn Type Zero.type() {
+    return new Int();
+  }
+}
+\end{lstlisting}
+  \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
+  \label{boolstrgen}
+\end{figure}
+The RAG modules written in \verb|.jrag| files, declare the attributes to be included in the classes generated by the AST specification.
+The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.
 
@@
 
-    \node (olast)     [below=of typerules] {AST definition\\(\ref{astdef})};
+    \node (olast)     [below=of typerules] {AST specification\\(\ref{astdef})};
 
@@
 Custom DSL
-Include typing rules AST definition (or BNF?)
+Include typing rules AST specification (or BNF?)
 ASCII representation of natural deduction %BNF (is it BNF?)
@@
 This approach does have the issue of occasionally parsing invalid programs, due to the reflection API finding an additional parameterless default constructor.
-This leads the parser to always accept nodes written with no children, such as \lstinline{Expression(Or())}, even if the AST definition requires them.
+This leads the parser to always accept nodes written with no children, such as \lstinline{Expression(Or())}, even if the AST specification requires them.
 These children will be null, causing issues further down the line, though as this parser is only intended for testing it is of little concern.
@@
 
-We've implemented premises as conditionals, where we check if each premise holds true before returning the type, or falling through to a type error if the premises don't hold.
-It would also be possible to instead check each condition one by one, and throw a type error as soon as one doesn't match.
-This would perhaps make it easier to implement informative type errors, as you could report precisely which premise is failing.
-It would however make it harder to implement having multiple valid rules describing the type of an expression, as the first failure would seize the evaluation of the node.
+We've implemented premises as conditions of a single if-statement, where we check that each premise holds true before returning the type, or falling through to a type error if the premises don't hold.
+It might instead be worthwhile to check each condition individually, to be able to specify in the error message precisely which premises have failed and why.
+%This would perhaps make it easier to implement informative type errors, as you could report precisely which premise is failing.
+If paired with the current error handling of throwing an exception to convey type errors, it would make it harder to implement having multiple valid rules providing the type of an expression, as the first failure would seize the evaluation of the type.
+An alternative error handling would suit it better, one able to collect multiple errors without seizing evaluation, perhaps merely adding all the errors to a list and checking whether it is empty before returning a type.
+
+The variables in the typing rules are compiled into regular Java variables containing the AST node.
+Notably, the declarations of these variables type them merely as \verb|ASTNode|.
+As we currently only utilise the \lstinline{type()} attribute, which is declared for all nodes, this lost information has little impact.
+However, in the future it may become relevant to type these more specifically, to support more complex rules that require deeper analysis within the tree.
+
+Type variables similarly are implemented as Java variables, though they are initiated as null, to represent the type variable's unassigned state.
+This is followed by a check for each point of use in the typing rule, to see whether the type variable is still unassigned and if so assign it.
+This approach leaves much room for improvement and optimisation, by finding the first use of each type variable during the rule compilation, and generating a single declaration statement initialising it to the type of the corresponding expression.
 
 A cleaner solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
-Instead of reporting the reason any individual rule failed, the error might instead display each possible rule,
+Instead of reporting the reason any individual rule failed, the error might instead display all the rules it tried to apply, perhaps with a list of non-holding premises for each one.
 
@@
 However, try to distinguish between measurements or hard data (results) and extrapolations, interpretations, or speculations (discussion).
-\section{Supported features}
-\section{Development learnings}
+%\section{Supported features}
+%\section{Development learnings}
+\chapter{Conclusions} % should summarize your findings and possible improvements or recommendations.
 Our experiments have shown that reference attribute grammars are a flexible method for implementing type checker generators.
@@
 
-\chapter{Conclusions} % should summarize your findings and possible improvements or recommendations.
-\section{Summary of findings}
+However, our compiler still has a long way to go before it can be used for real world use cases and general purpose languages.
+Chief among the missing features is the concept of environments, without which we're unable to support even common language features such as variables or functions.
+
+%\section{Summary of findings}
 \section{Future work}
+\subsection{Environments}
+Implementing environments, or type contexts, would be the clear next step for this project.
+However, its implementation is not entirely clear-cut, with several aspects requiring consideration.
+
+The essence of environments is the mapping of variables or functions to types, and an appropriate representation of this would need to be found.
+A method often used by traditional type checkers is a symbol table, a separate data structure mapping language constructs to their types.
+In our case, we may instead be able to leverage our reference attribute grammars to store the information within the AST.
+By connecting each use of a variable or function to the AST node representing its declaration site, we could utilise the node to store the type information.
+This approach would however also have to consider the typing of external language constructs, such as built-in functions or those imported from external libraries, for which we wouldn't have access to the declaration site.
+
+%An additional concern is that of scoping, which different languages may have different rules for.
+
+A type checker working on these principles is certainly a possibility, though it remains to be seen how it could be generalised to the generation of typing rules for arbitrary languages.
+
 \subsection{Improved error handling}
@@
 
-\subsection{Check for completeness of rules}
+\subsection{Verify rules against AST}
+Major improvements to the user experience could be made by tying the typing rules to the object language AST before generating the type checker.
+Currently, there are no checks in place to verify the completeness of the typing rules.
+If a user accidentally leaves out the typing rule for a certain construct, they'll receive no warning until the outputted type checker attempts to check the expression's type.
 
-\subsection{Verify rules against AST}
-\subsection{Handle multiple rules for the same node (add(int, int): int, add(double, double): double)}
+Similarly, no checks are made to ensure the nodes described in the typing rules actually exist within the language's AST.
+Our project will compile invalid or misspelled typing rules without complaint, leading to errors which won't be caught until the outputted type checker is attempted to be compiled.
+
+\subsection{Handle multiple rules for the same node}
+In type systems it is generally considered valid to have multiple typing rules for the same construct, with different, non-intersecting prerequisites.
+It could for example be useful to type an addition operator for a variety of different numerical types, which may not have a common ancestor.
+
+Most of the work to support this is in place, though the first rule to fail would throw a type error exception, preventing the later rules from evaluating.
+These errors would either have to be caught, and only emitted if all rules had failed, or a new error handling system not based around exceptions could be introduced.
+Additionally, an extra layer of scoping would have to be introduced, to prevent variable naming collisions between the different rules.
+This could be solved simply by wrapping each rule within brackets, creating a separate scope for the evaluation of each rule.
+
+%add(int, int): int, add(double, double): double)
 \subsection{Better support for numbers, strings, etc.}
+
 \subsection{Improved type checking algorithm}

### 2023-12-14 MSc.tex
Various changes

@@
 
-%\acknowledgements{
+% For questions to people reviewing, remove before finalizing
+\newcommand{\reviewquestion}[1]{\textcolor{red}{\begin{itemize}\item Question: #1\end{itemize}}}
+%\newcommand{\reviewquestion}[1]{}
+
+\acknowledgements{
+Sincere thanks to Christoph Reichenbach,
+whose enthusiasm for the subject inspired me to pursue this subject
+and whose guidance in matters of both implementation and writing were key to this thesis.
+Additional thanks to Niklas Fors for providing several valuable insights, greatly simplifying certain components and Alexandru Dura for providing some guidance to JastAdd internals.
+
+I would also like to thank my partner Lo-Isabel Krantz Andrée, my family, and Moritz for their support throughout the duration of this project.
+
 %If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.
@@
 %Thanks also go to the following (former) students for helping with feedback and suggestions on this template: Mikael Persson, Christoffer Lundgren, Mahmoud Nasser.
-%}
+}
 
@@
 %TODO: Give an overview of the other sections
+The rest of this paper is outlined as follows.
+% This chapter introduces the
+Chapter 2 presents the necessary background information for the report.
+Chapter 3 describes the design and implementation of our program, followed by demonstrations of its results in chapter 4.
+Chapter 5 discusses the results in a more general sense, before the conclusion in chapter 6.
 
@@
 Additionally, their project defines a formalism for expressing typing rules within the Datalog language, ours defines a language which aims to keep as close as possible to the traditional inference rules used for typing rules, to minimise the difficulty of translation.
+
+\reviewquestion{Circular attributes?}
 
@@
 %\printbibliography
-\makebibliography{MyMSc} % is a must in a scientific report. {\LaTeX} and \texttt{bibtex} offer great support for  handling references and automatically generating bibliographies.
+\makebibliography{MSc} % is a must in a scientific report. {\LaTeX} and \texttt{bibtex} offer great support for  handling references and automatically generating bibliographies.
 

### 2023-12-14 MSc.tex
Initial comments

@@
 %\thesistype{Bachelor}{Kandidatarbete}
+
+\newcommand{\CR}[1]{\textcolor{cyan!70!black}{[\textbf{CR}:#1]}}
 
@@
 Catching these mistakes as early as possible can prevent these bugs from ever reaching production code.
+\CR{Possible papers to cite as evidence:
+  \url{https://dl.acm.org/doi/abs/10.1145/1985793.1985889}
+  \url{https://www.usenix.org/legacy/event/sec04/tech/full_papers/johnson/johnson_html/}
+  \url{https://ieeexplore.ieee.org/abstract/document/10190394}}
 Type checkers are one of the most common static checking methods, ensuring that the correct data types are used, but creating a new type checker is often an arduous process, requiring large amounts of complex code.

### 2023-12-14 MSc.tex
Added comments up to and including Section 4

@@
 
-\newcommand{\CR}[1]{\textcolor{cyan!70!black}{[\textbf{CR}:#1]}}
+\newcommand{\CR}[1]{\textcolor{green!60!black}{[\textbf{CR}:#1]}}
 
@@
   \url{https://ieeexplore.ieee.org/abstract/document/10190394}}
+\CR{These papers add static checkers on top of existing languages to find additional bugs; you can also
+  mention statically typed languages, but those generally don't require frequent changes to the type checker implementation, so they're less
+  compelling to motivate your work.}
 Type checkers are one of the most common static checking methods, ensuring that the correct data types are used, but creating a new type checker is often an arduous process, requiring large amounts of complex code.
@@
 % Define the problem you are solving (or trying to solve).
+\CR{Prior work has already looked into pluggable type checkers.
+  Why is translating them into JastAdd interesting?
+  (aside from theoretical questions, you can also use
+  pragmatic reasons, such as integration into our courses.)}
 In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of reference attribute grammars.
 
+\CR{Make clear to the reader what the challenges are:
+   How are these two formalisms similar, how do they differ (at a high level)?
+   You can use a simple typing rule as an example if it helps (very optional).
+   Hint: look at determinism in evaluation (syntax-directedness), unification,
+   but stick to intuition in this chapter; you can elaborate in Chapter 2 or later
+   if you think it helps.}
+
 %TODO: Give an overview of the other sections
+\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning
+  chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
+  We evaluate the approach against Foo, and demonstrate Bar.
+  Our case studies show Quux.
+  Overall, this initial exploration shows \ldots''}
+
+\CR{paper $\to$ thesis; also, this always comes at the very end of the introduction}
 The rest of this paper is outlined as follows.
@@
 \section{Research questions}
-To formulate our goals more concretely, we've set out to answer the following research questions:
-\newcommand{\rqone}{What kinds of typing rules can we translate through RAG?}
+To formulate our goals more concretely, we've \CR{Avoid contractions in technical writing} set out to answer the following research questions:
+\CR{Give the RQs numbers so you can refer to them later.  Common form: RQ1, RQ2 etc.; some authors use RQ2.3 if they split RQ2 into multiple subquestions (this can happen later).}
+\newcommand{\rqone}{What kinds of typing rules can we translate through \CR{``through?''} RAG?}
 \newcommand{\rqtwo}{How can we implement the translations and into what form?}
@@
 \begin{itemize}
-  \item \rqone
-  \item \rqtwo
-  \item \rqthree
+  \item \rqone \CR{``kinds'' of typing rules: right now the reader isn't aware that there even are different kinds (see my earlier comment on ``what the challenges are''-- sketching this out should suffice.)}
+  \item \rqtwo \CR{What does ``form'' mean, here?
+How broad is ``implement''-- are you talking algorithms (which I think you are) or including technical challenges (which I think you have in the next RQ)?}
+  \item \rqthree \CR{I'd go with `challenges in translating' instead of `potential issues of translating', language-wise}
 \end{itemize}
@@
 Pacak et al. implemented a compiler for generating incremental type checkers using Datalog\cite{Pacak}.
-While the goal is similar to ours, they take a different approach, by converting typing rules into the declarative logic programming language.
+While the goal is similar to ours, they take a different approach, by converting typing rules into the declarative logic programming language \CR{second half of this sentence confuses me}.
 
 Their work is thereby largely focused on expressing typing rules within the limits of the Datalog language, which can only compute finite relations.
-We have a significantly different compilation target, JastAdd, which allows us to use reference attribute grammars to evaluate the typing relations, which we write in a more iterative approach.
+We have a significantly different compilation target, JastAdd, which allows us to use reference attribute grammars to evaluate the typing relations, which we write in a more iterative \CR{unclear?} approach.
+\CR{Datalog is fully declarative, whereas JastAdd uses declarative rules that connect snippets of imperative Java code.  Is that what you meant?}
+\CR{Some more thoughts below to help make the distinction clearer (all optional):}
+\CR{Datalog has no direct notion of ASTs or other tree structures / algebraic terms, so trees must be encoded into relations.}
+\CR{Datalog has a simple, built-in notion of unification that can handle cases like testing sibling relations: $\textsf{Sibling}(x, y) :- Parent(p, x), \textsf{Parent}(p, y), x \ne y.$.
+    This is something you deal with later in JastAdd.}
 
 Additionally, their project defines a formalism for expressing typing rules within the Datalog language, ours defines a language which aims to keep as close as possible to the traditional inference rules used for typing rules, to minimise the difficulty of translation.
+\CR{I'd start with this difference-- the translation is targeting two rather different formalisms, which is all you need to justify the novelty.}
 
 \reviewquestion{Circular attributes?}
+\CR{Circular attributes: Recursion in typing rules: Datalog is built with support for ``infinite'' recursion (since it operates on finite domains,
+  it can use a special bottom-up evaluation scheme to ensure termination.)
+  If you need this in RAGs, it gets more complicated, but you'd need an example to explain this in the introduction-- since you don't directly address that in this work,
+  I wouldn't talk about it too much here (you could also skip it entirely here).  Minimum: talk about it in the discussion later.}
 
@@
 %\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
+\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinkingr ``I know why I want to read this''}
 This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
 The first section explains abstract syntax trees, which are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
-The next section details type systems, by first introducing data types, then explaining the components required for a type system, followed by a description of static type checkers.
+\CR{Consider saying just a little more about them here, for readers who have no idea what ASTs are and why they matter for
+    analysing programs.}
+The next section details type systems \CR{Too much detail in the following:}, by first introducing data types, then explaining the components required for a type system, followed by a description of static type checkers.
+\CR{You can use \texttt{\textbackslash label\{labelname\}} and \texttt{\textbackslash ref\{labelname\}} to reference (sub)secttions, that's usually clearer than ``next section''.}
 
 \section{Abstract syntax trees}
+\CR{What's a syntax?}
 An abstract syntax tree (AST) is a tree-based representation of the syntax of a program.
 Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
-It's abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
-Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this information.
+It's \CR{contraction} abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
+Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this information \CR{what is ``this information'', concretely?}.
 A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if-else node, with named branches pointing towards the condition and the two values.
+\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
+\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}
 
 This layer of abstraction makes it easier to write analyses for the program, as we can focus on the key elements and structure.
+\CR{Is there a connection between the AST and the behaviour of the program?}
+
+\CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}
 
 \section{Type theory}
-In this section we'll discuss the elements of type theory required to understand this thesis.
+In this section we'll \CR{contraction} discuss the elements of type theory required to understand this thesis.
 Firstly, we'll discuss what types are and give some examples of their use.
-Next, we'll describe type system, which form a relation between types and expressions.
+Next, we'll describe type system, which form\CR{forms} a relation between types and expressions.
 
 \subsection{Data types}
+\CR{think about the connection between types and values here.  You may even want to talk about values in a separate subsection.}
 Computer software generally needs to handle and manipulate several different types of data in one program, such as numbers or text.
@@
 
+\CR{if you discuss values, you can use the intuition of ``grouping values'' by certain properties (such as supported operations) to make your point clearer.}
 Different parts of programs may have different constraints on the types of expressions, such as a print function requiring a string of text, or a plus function requiring some kind of number.
+\CR{Very good point about types of expressions, but again, this is likely easier to understand by linking expressions and values.
+  I strongly recommend using a definition of the form ``all values are expressions, but not all expressions are values''.}
 
@@
 There are two main approaches to ensuring that the aforementioned constraints are upheld, dynamic and static type-checking.
+\CR{Why would we even want to ensure that they are upheld?}
 Dynamic checking is done performed runtime, with each type checked only when needed.
 Static checking is performed on the code before it's run, usually as part of compilation, preventing many faulty programs from being run at all.
+\CR{Why would I want static checking?  Dynamic only checks when needed, isn't that more efficient?}
 
@@
 Static checkers analyse the structure of the source code, often in the form of an abstract syntax tree.
-Conservative
+Conservative \CR{incomplete sentence?}
+
+\CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
+  For each concrete (and possibly very complex) operator on the left hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
+  This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
+  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractoins
+  over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
+}
 
@@
 A type system consists of two parts, a type syntax, describing the available types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
+\CR{Why do the types need a syntax?
+    (There are other approaches to types; for your own sanity, I recommend deferring to Pierce here and calling this ``typical approaches'')}
+
 \subsubsection{Type syntax}
 A type syntax is a definition of the valid types within a language.
+\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
 It can consist of fundamental types, such as integers or booleans.
 There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
+\CR{A bit vague; consider giving an example.}
 
@@
 The typing relation, notated \lstinline{t : T}, is a link between the expressions of a language and the types of the type syntax.
+\CR{It is \emph{the} link, even.}
 The left-hand side is an expression in the object lanuage, typically written in its native syntax, the right-hand side a type from the aforementioned type syntax.
@@
 
-Typing rules may also have premises, a number of conditions which must hold for the typing rule to apply.
+Typing rules may also have premises, a number of conditions which \CR{\emph{that} must hold (``which'' relates a set to a superset: ``carrots, which are vegetables'', while ``that'' restricts a set to a subset: ``vegetables that are carrots'')} must hold for the typing rule to apply.
 The notational convention here is to put the premises above a bar, with the conclusion underneath it.
@@
 This T also appears in the conclusion, as the resulting type of the if-expression, meaning the result of the entire expression will be that of the $t_2$ and $t_3$ terms.
+\newcommand{\ty}[1]{\textcolor{blue!20!black}{\textsf{#1}}}
+\newcommand{\tyvar}[1]{\ty{\textcolor{cyan!50!black}{#1}}}
+\CR{Consider typesetting T differently, e.g. as \tyvar{T}}
+\CR{\tyvar{T}, $t_1$, $t_2$, $t_3$ are \emph{free} variables here, logically speaking}.
+\CR{Please explain the purpose of typing rules in static type checking.  Using just your two examples you would be able to
+  show a well-typed program, and if you add one more rule (e.g. for integers), you can show a type error.  No need to build
+  the full derivation tree, but it's important to mention recursion here.  This entire bit may be worth a separate subsubsection.}
 
@@
 Reference Attribute Grammars (RAGs) are an extension to Attribute Grammars (AGs), a formalism used for specifying the semantics of programming languages.
+\CR{Generally only for the static semantics (RAGs can also do dynamic semantics, if you allow circular attributes, but that's not worth mentioning here.)
+Please also connect ``semantics'' to ``type checking'' (via ``static semantics'')}
 AGs provide a systematic way to define computations on the nodes of an abstract syntax tree, associating attributes with language constructs and are commonly used in the context of compiler construction, where they help define the translation of source code into executable code\cite{AGs}.
@@
 JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion.
+\CR{It's worth telling the reader here why you are explaining some details about JastAdd.}
+
 \cite{JastAdd}
@@
 \subsubsection{RAG Modules}
+\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are
+not a necessary feature for RAGs.  Please introduce them in a seprate section: you can very directly connect them to the idea of ``type syntax''.}
 \begin{figure}[h]
@@
 
+\CR{Missing: introduction to this chapter'}
+
 \section{Architecture}
 In this section we'll be providing an overview of our architecture.
+\CR{Stylistically, technical documents should avoid discussing the process of reading (it's not as frowned upon as speaking directly to the reader, though), so I recommend just saying that the figure ``provides an overview over...''}
 Since it contains many different components, it may be helpful to refer to the overview graph in figure \ref{overviewgraph} while reading.
+\CR{Please put the following in the caption of the figure instead.}
 It represents the flow of the program, with components and artefacts (represented as rectangular boxes) with arrows linking them to actions (elliptical boxes) which in turn have arrows to the new components and artefacts they produce.
@@
 %It uses rectangular boxes to represent the concrete data or artefacts of the procedure, and elliptical boxes to represent the actions performed with them.
+
+\CR{You still want an introduction to this section that explains what is following and why}
+
 \begin{figure}[]
@@
   }
-  \caption{An overview of the project components and flow.}
+  \caption{An overview of the project components and flow. \CR{Can you mark, e.g. via colours, the parts that the user provides,
+           and the parts that will be in the generated output project?}}
   \label{overviewgraph}
@@
 \subsubsection{Language}\label{sectionlanguage}
+\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
+  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
+  This would be before the current 3.1, then.
+  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
+  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
+  of the object language as well as of the types.}
 
@@
 To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{pierce}.
-Our language can express a subset of the notation used by Pierce, with some differences and limitations.
+Our language can express a subset of the notation used by Pierce, with some differences and limitations
+\CR{either forward reference to the discussion of the limitations, or remove the ``limitations'' bit here (you already say ``subset'', after all) and discuss those limitations after you define the language.}.
 
@@
 The largest difference can be seen in the expression to be typed, where Pierce uses the object language's native syntactic form, whereas our representation is based on the corresponding AST nodes.
+\CR{Given that it's nontrivial to separate out the different changes, I recommend first enumerating the differences abstractly and then discussing how the changes manifest in the figures, rather than starting with the figures to ``derive'' the differences.}
 In the first example, the difference is minor (\verb|true| and \verb|True| respectively) but the difference is more apparent in figure \ref{trexampleif}, where instead of the full syntactic form, the expression is written as an AST node with its children represented as parameters.
@@
 This change is necessary to allow us to type arbitrary languages without needing to understand their syntax and grammar, though it does make translating rules from one syntax to the other less straightforward.
+\CR{Optional: you can mention Alexandru's paper about how one could add metavariables into ``native'' syntax: \url{https://dl.acm.org/doi/pdf/10.1145/3485542}
+  (since this part has already been done, it's not something you needed to explore)}
+\CR{Capitalising the AST node types is consistent with their naming in JastAdd-- worth pointing out.}
 
@@
 This change also aligns the syntax for type terms with the regular terms, where concrete nodes (such as \verb|If|) are written uppercase and variables (\verb|t1|, \verb|t2|, \verb|t3|) in lowercase.
+\CR{I generally recommend avoiding the passive voice, but the use below is an example that would be OK}
 Commas are also used to separate the premises, and the horizontal line separating premises from the conclusion is represented as a series of dashes.
 
+\CR{Important: this affects numbers used at the type level only (which is an advanced feature mostly used in undecidable
+  typesystems like the Calculus of Constructions), and it wouldn't matter here since you can't use \ty{Succ} as a parametric
+  type anyway (cf. igure 3.4).
+  My problem is that right now this reads as if you can't typecheck programs with Integer literals,
+  but those would be e.g. \textsf{Integer} AST nodes, which (to my reading) you would have no trouble writing typing rules for.
+  (you never need to touch the actual lexeme that contains the character string that encodes the integer, in traditional
+  type analysis).  So you shouldn't actually have any real-life problem here.}
 Numerical values have not been implemented, but natural numbers can be expressed using Peano numbers\cite{peano}, utilising a zero value and successor function which represents the number incremented by one.
@@
 \subsubsection{Abstract syntax tree}\label{typingrulesast}
+\CR{I would give this AST/BNF a more specific name here, to distinguish it from the other two (as done nicely already in Figure 3.1)}
 \begin{figure}[h]
@@
 \end{figure}
+\CR{The syntax doesn't allow parametric types; please make that explicit here.}
 
@@
 \subsection{Template}
-In addition to the compiled files, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the template.
+In addition to the compiled files \CR{What are the ``compiled files?''}, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the template.
 Some of these are key components of the type checker, but others are dummy components included solely for convenience of testing.
@@
 For testing purposes, we've written generic versions of these components, which work for any supported object language.
+\CR{``Templates'' are usually things that are mostly fixed but have holes to be filled in.  I would call what you describe here the ``support library''}
+
 \subsubsection{Parser}\label{treebasedparser}
+\CR{Parser for what?}
+\CR{Is there a conection between this section and Section~\ref{typingrulesparser} wrt parsing terms of the object language?}
 For testing purposes, our template comes with a generic tree-based parser.
 Programs are written as a tree of AST node names followed by potential parenthesis within which a comma separated list of child nodes are written.
-It recursively parses the tree, using Java's reflection API to find a class with a matching name, and a constructor with parameters of the same amount and types.
+It \CR{what?} recursively parses the tree, using Java's reflection API to find a class with a matching name, and a constructor with parameters of the same amount and types.
 Figure \ref{boolinputexample} shows an example input file for the Bools language, which will be discussed further in section \ref{boolsast}.
@@
 These children will be null, causing issues further down the line, though as this parser is only intended for testing it is of little concern.
+\CR{Does this also apply to the analysis of typing rules?  If so, you can mention it as ``future work''
+  or describe how it could be fixed (it's not a fundamental limitation, so nothing that warrants a deeper discussion)}
 
@@
 
+\CR{Can you elaborate on if/how this would integrate with an existing language implementation?
+  I've always interpreted everything as that you start with some `objectlanguage.jar' file of an existing language
+  (given that you use reflection), and you get out
+  a fresh Gradle project that uses that `objectlanguage.jar' for parsing and AST construction,
+  but since you generate JastAdd source code, you can't really (to my understanding) add the `type()' attribute etc.\@
+  to the existing implementation, since JastAdd needs to run over the source jrag files.
+  Going by your overview figure, I now wonder if the generated file also incorporates all of the source code of the original
+  project?  Or do you reconstruct the AST specification via reflection somehow and simply mirror it in the generated
+  project (which would lead to two different AST representations, but with a straightforward mapping)?}
+
 \subsubsection{Type checking aspect}\label{typecheckingaspect}
@@
 
+\CR{The details of this section (esp.\@ the bit below) might fit better into Section~\ref{typingrulecompiler}?}
+\CR{What is ``type matching''?  I am not familiar with this term, and you haven't introduced it before.}
 It also contains the implementation of the type matching method, which can be seen in figure \ref{typematches}.
 It's defined for all \lstinline{Type} nodes, and checks whether the type's class is equal to or a subclass of the parameter type.
+\CR{So this is not an equivalence relation, but a partial order?  What does that mean wrt how it is used by the implementation of the type checker?}
 
@@
 %\subsection{Type checker generation}
-The type checker generation is based on a single-responsibility principle, where each node of the abstract syntax tree is responsible for generating the code it corresponds to.
+The type checker generation is based on a single-responsibility principle, where each node of the abstract syntax tree (of \emph{which} of the three (or four?) ASTs?) is responsible for generating the code it corresponds to \CR{Not sure what you mean}.
 Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of figure \ref{trastspec}, where the \verb|RuleSet| is the root node.
@@
 Should the premises not hold, a type error is produced.
+\CR{What if there are multiple rules that apply to the same AST node?  E.g., one rule that allows \ty{Int} addition and one that allows \ty{String} addition.}
 
@@
 \subsection{Parameters}
-When the typing rules concern a node with parameters, these can be named arbitrarily, and a variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
+When the typing rules concern a node with parameters, these can be named arbitrarily \CR{Who does the naming?
+Why is the ``arbitrarily'' bit important?}\CR{ Will this work if the type system specification has a type variable called ``this'' or ``null'' or ``class''?}, and a variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
 These variables are then utilised in later premises or type variable declarations.
@@
 The current implementation of \verb|Type.matches(Type t)| checks that the Type it is called on is either of the same class or a subtype of the type \verb|t|.
+\CR{is ``matches'' a good name for something that is not symmetric?}
 This allows the language limited sub-typing at the type checker generation step, though the pros and cons of this haven't been evaluated.
+\CR{Can you give an example of a case where this is helpful, and one example where it might be troublesome?
+Let's use the ``if'' rule as an example.
+If $\ty{Number} :> \ty{Int}$ (\ty{Number} is a supertype of \ty{Int}), what type will I infer if $t_2 = \ty{Number}$ and $t_3 = \ty{Int}$?
+How about the other way around?}
 It is also possible to override the \verb|matches| implementation for a specific Type, by adding a definition to a JastAdd aspect included in the output type checker.
+\CR{This sounds like this is a way to support ``ad-hoc type polymorphism''?}
 
@@
 \subsection{Type variables}
-In the case of type variables, the generated code takes a dynamic approach to simplify the implementation.
+In the case of type variables, the generated code takes a dynamic approach \CR{Meaning?} to simplify the implementation.
 Each type variable that appears in the conclusion is instantiated to a null value, and the code generated by each premise first checks if the value is still null and either assigns the variable its own type or tests to make sure the existing type matches its own.
+\CR{Is there a hidden order dependency? (Cf. my earlier question about ``matches'')}
+
+\CR{Overall, I am not convinced that this strategy for presenting the type checking algorithm
+  is very effective (though it might be better for someone who is reading the source code at the same time).
+  I would propose to merge 3.2.2, 3.2.3, 3.2.4, 3.2.5, and to split them into (a) the type checking algorithm
+  (while ignoring the code generation bit), and (b) the code generation aspect.  I.e., for (a), pretend that you are
+  interpreting the typing rules instead of compiling them, and please try to present an algorithm that captures
+  the check.  The algorithm can be in plain English with bullet points / numbered sub-steps, or in some pseudo-formal
+  notation (``foreach premise $p$ in rule $r$ ...''), as long as it is clear.}
 
@@
 
-It provides a stack trace allowing you to see where in the type checking code the error occurs (such as what typing rule fails), but doesn't currently provide any information about what section of the input program causes the error.
+It provides a stack trace allowing you \CR{don't address reader directly} to see where in the type checking code the error occurs (such as what typing rule fails), but doesn't currently provide any information about what section of the input program causes the error.
 This tends to result in errors that are more useful for debugging the typing rule implementation, rather than providing useful information to the user of the type checker, where there is much room for improvements.
+
+\CR{This section is needlessly apologetic-- error handling was not the focus of your work.  You can add a forward reference to 6.1.2, though.}
 

### 2023-12-14 MSc.tex
Expand contractions

@@
 \section{Research questions}
-To formulate our goals more concretely, we've \CR{Avoid contractions in technical writing} set out to answer the following research questions:
+To formulate our goals more concretely, we have \CR{Avoid contractions in technical writing} set out to answer the following research questions:
 \CR{Give the RQs numbers so you can refer to them later.  Common form: RQ1, RQ2 etc.; some authors use RQ2.3 if they split RQ2 into multiple subquestions (this can happen later).}
@@
 \section{Type theory}
-In this section we'll \CR{contraction} discuss the elements of type theory required to understand this thesis.
-Firstly, we'll discuss what types are and give some examples of their use.
-Next, we'll describe type system, which form\CR{forms} a relation between types and expressions.
+In this section we will discuss the elements of type theory required to understand this thesis.
+Firstly, we will discuss what types are and give some examples of their use.
+Next, we will describe type systems, which form a relation between types and expressions.
 
@@
 Dynamic checking is done performed runtime, with each type checked only when needed.
-Static checking is performed on the code before it's run, usually as part of compilation, preventing many faulty programs from being run at all.
+Static checking is performed on the code before it is run, usually as part of compilation, preventing many faulty programs from being run at all.
 \CR{Why would I want static checking?  Dynamic only checks when needed, isn't that more efficient?}
@@
 \section{Architecture}
-In this section we'll be providing an overview of our architecture.
+In this section we will be providing an overview of our architecture.
 \CR{Stylistically, technical documents should avoid discussing the process of reading (it's not as frowned upon as speaking directly to the reader, though), so I recommend just saying that the figure ``provides an overview over...''}
@@
 We expect that in a real world use case, our software would integrate into an object language's compiler pipeline, and utilise its own parser and project files to provide the abstract syntax tree for our type checking to tie into.
-For testing purposes, we've written generic versions of these components, which work for any supported object language.
+For testing purposes, we have written generic versions of these components, which work for any supported object language.
 \CR{``Templates'' are usually things that are mostly fixed but have holes to be filled in.  I would call what you describe here the ``support library''}
@@
 It also contains the implementation of the type matching method, which can be seen in figure \ref{typematches}.
-It's defined for all \lstinline{Type} nodes, and checks whether the type's class is equal to or a subclass of the parameter type.
+It is defined for all \lstinline{Type} nodes, and checks whether the type's class is equal to or a subclass of the parameter type.
 \CR{So this is not an equivalence relation, but a partial order?  What does that mean wrt how it is used by the implementation of the type checker?}
@@
 \CR{is ``matches'' a good name for something that is not symmetric?}
-This allows the language limited sub-typing at the type checker generation step, though the pros and cons of this haven't been evaluated.
+This allows the language limited sub-typing at the type checker generation step, though the pros and cons of this have not been evaluated.
 \CR{Can you give an example of a case where this is helpful, and one example where it might be troublesome?
@@
 The typing rules are introduced in chapter 8, starting with the simple examples shown in section \ref{sectionlanguage}.
-We've implemented support for simple type declarations, where a value or function always has a certain type.
-These simple rules aren't of much interest in themselves but are an important backbone to the subsequent analysis.
+We have implemented support for simple type declarations, where a value or function always has a certain type.
+These simple rules are not of much interest in themselves but are an important backbone to the subsequent analysis.
 We also support type declarations with premises that must hold for the rule to apply, allowing us to express relationships between types and bringing with them the possibility of type errors.
 
-Type variables extend these relationships further by providing a layer of abstraction, allowing us to write rules where the types aren't precisely defined.
+Type variables extend these relationships further by providing a layer of abstraction, allowing us to write rules where the types are not precisely defined.
 This enables rules where the return type of an expression depends on its child expressions, or where child expressions must have the same types.
@@
 
-We've chosen to represent the right-hand side of type equations, the types themselves, as an abstract AST class.
+We have chosen to represent the right-hand side of type equations, the types themselves, as an abstract AST class.
 While our current implementation only supports simple types, this representation gives us a lot of flexibility for future extensions.
@@
 
-We've implemented premises as conditions of a single if-statement, where we check that each premise holds true before returning the type, or falling through to a type error if the premises don't hold.
+We have implemented premises as conditions of a single if-statement, where we check that each premise holds true before returning the type, or falling through to a type error if the premises do not hold.
 It might instead be worthwhile to check each condition individually, to be able to specify in the error message precisely which premises have failed and why.
@@
 We attempted to keep each node in the tree responsible for generating the code it corresponds to, which avoids parent nodes having to check their children's structure with \verb|isInstanceOf|, but also means even generating a simple rule can require a dozen calls between different nodes.
-There's a tradeoff to be made between the single responsibility of each component making it easy to verify its correctness, but the complicated interdependence making it hard to grasp the full picture.
+There is a tradeoff to be made between the single responsibility of each component making it easy to verify its correctness, but the complicated interdependence making it hard to grasp the full picture.
 
@@
 However, our compiler still has a long way to go before it can be used for real world use cases and general purpose languages.
-Chief among the missing features is the concept of environments, without which we're unable to support even common language features such as variables or functions.
+Chief among the missing features is the concept of environments, without which we are unable to support even common language features such as variables or functions.
 
@@
 By connecting each use of a variable or function to the AST node representing its declaration site, we could utilise the node to store the type information.
-This approach would however also have to consider the typing of external language constructs, such as built-in functions or those imported from external libraries, for which we wouldn't have access to the declaration site.
+This approach would however also have to consider the typing of external language constructs, such as built-in functions or those imported from external libraries, for which we would not have access to the declaration site.
 
@@
 Currently, there are no checks in place to verify the completeness of the typing rules.
-If a user accidentally leaves out the typing rule for a certain construct, they'll receive no warning until the outputted type checker attempts to check the expression's type.
+If a user accidentally leaves out the typing rule for a certain construct, they will receive no warning until the outputted type checker attempts to check the expression's type.
 
 Similarly, no checks are made to ensure the nodes described in the typing rules actually exist within the language's AST.
-Our project will compile invalid or misspelled typing rules without complaint, leading to errors which won't be caught until the outputted type checker is attempted to be compiled.
+Our project will compile invalid or misspelled typing rules without complaint, leading to errors which will not be caught until the outputted type checker is attempted to be compiled.
 

### 2023-12-14 MSc.tex
More feedback

@@
 \chapter{Evaluation} % is the part where you present the finds. Depending on the area this part contains a subset or all of the following:
+\CR{The section has two parts, so the intro paragraph here can directly discuss their purpose.}
 %\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach.
@@
 Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
+\CR{You can't discuss results without describing your experiments, but to describe the experiments, you should first motivate them.
+  Since you are doing case studies, you can use Pierce to motivate your choice of languages (even if your language is subtly different somehow; I don't have the book in front of me).
+  Make clear to the reader that you picked your languages based on the existing literature, not on a whim.}
 %\subsection{Input/output examples, with descriptions}
@@
 \end{lstlisting}
-  \caption{Generated typing rules for the Bools language (reformatted for readability)}
+\caption{Generated typing rules for the Bools language (reformatted for readability)
+\CR{}}
   \label{boolstrgen}
@@
 For the \verb|Or| node, the code becomes slightly more complex.
+\CR{The next bit would probably be easier if you just show the code.}
 First two variables are defined, corresponding to the parameters in the typing rule, and assigned to the children with the index corresponding to the position among the parameters.
@@
 
+\CR{You may want to add a short statement that this captures the intended behaviour of the type checker.}
+
 \subsection{If-else extension}
+
+\CR{Separate case study or next step in the same case study?  Either way, the section title should make the connection to 4.1.1 (or lack thereof) clear.}
 
@@
 \end{lstlisting}
-  \caption{Extension for the Bools AST in figure \ref{boolsast}}
+  \caption{Extension for the Bools AST in figure \ref{boolsast} \CR{Consider giving this language a different name, for clarity}}
   \label{ifelseast}
@@
 \end{lstlisting}
-  \caption{Generated typing rules for the if-else term, extending figure \ref{boolstrgen} (reformatted for readability)}
+\caption{Generated typing rules for the if-else term, extending figure \ref{boolstrgen} (reformatted for readability)
+  \CR{The components of this function would be great to discuss in Section 3.
+Here in Section 4, you may want to forward refernece the discussion about redundancy/possible optimisations (Section 4.2.2) and very concretely use this code and explain the changes you have in mind (for someone else to do in future work).}
+}
   \label{ifelsetrgen}
@@
 As an example of rules utilising type variables, we can extend the Bools language with an if-else term.
+\CR{``Can extend'' sounds a bit arbitrary.  Can you motivate with Pierce (even if not exactly)?}
 As shown in the AST extension in figure \ref{ifelseast}, it consists of three terms: a conditional, a term to be evaluated if the condition is true and a condition to evaluate if false.
@@
 The rule has three premises, the first of which declares the conditional must have the type \verb|Bool|.
-The next two declare that the two other children must be of the type \verb|a|, with the lowercase indicating a type variable.
+The next two declare that the two other children must be of the type \verb|a|, with the lowercase \CR{grammar} indicating a type variable.
 This type variable \verb|a| also subsequently appears in the resulting type of the if-else term.
 
+\CR{``Shown in Figure~\ref{reference}'' is the current preferred form in our field}
 The code for this rule is shown in figure \ref{ifelsetrgen}, and begins with three parameter declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
@@
 \section{Research questions}
+
 \subsection{\rqone}
@@
 The typing rules are introduced in chapter 8, starting with the simple examples shown in section \ref{sectionlanguage}.
+\CR{Please be clearer on which examples from Pierce you implemented and where support stops.}
 We have implemented support for simple type declarations, where a value or function always has a certain type.
@@
 
+\CR{Please explain typing contexts and why/where they are important.  You can also reflect on whether typing contexts are necessary in the context of RAG-based analysis (cf. what we did in EDAP15).}
+
+\CR{I am missing the phrase ``syntax-directed'' here.}
+
 \subsection{\rqtwo}
-We have approached parsing typing rules into a JastAdd attribute \verb|type()|, which either returns a \verb|Type| value, or throws a runtime exception if the expression is invalidly typed.
+We have approached parsing typing rules into a JastAdd attribute \verb|type()|,
+\CR{parsing is the syntactic analysis of the input specification, you have done more than just parsing.}
+which either returns a \verb|Type| value, or throws a runtime exception if the expression is invalidly typed.
 
@@
 Perhaps the default implementation could be extended to compare all child elements, or these more advanced types would require manually overriding the method.
+\CR{Recursive term equivalence can become tricky when you are dealing with type variables (cf. the Hindley-Milner inference exercise in EDAP15).}
+\CR{Type equality could also be expressed in terms of typing rules.  This approach is quite common with subtyping.}
 
@@
 If paired with the current error handling of throwing an exception to convey type errors, it would make it harder to implement having multiple valid rules providing the type of an expression, as the first failure would seize the evaluation of the type.
+\CR{What do you mean by ``seize''?}
 An alternative error handling would suit it better, one able to collect multiple errors without seizing evaluation, perhaps merely adding all the errors to a list and checking whether it is empty before returning a type.
@@
 This is followed by a check for each point of use in the typing rule, to see whether the type variable is still unassigned and if so assign it.
-This approach leaves much room for improvement and optimisation, by finding the first use of each type variable during the rule compilation, and generating a single declaration statement initialising it to the type of the corresponding expression.
+This approach leaves much room for improvement \CR{no need to say ``much room for improvement'', just mention the specific disadvantages.} and optimisation, by finding the first use of each type variable during the rule compilation, and generating a single declaration statement initialising it to the type of the corresponding expression.
+\CR{I am missing a discussion of type variables and unification as we used them in EDAP15 for Hindley/Milner.}
 
@@
 Instead of reporting the reason any individual rule failed, the error might instead display all the rules it tried to apply, perhaps with a list of non-holding premises for each one.
+\CR{This is a good idea; you can even sort the rules by closest matches and only show the top ones if there are many.}
+
+\CR{Can you structure this subsection a bit more explicitly?  Could be subsubsections, but not necessarily needed; maybe a bullet list of
+challenges and for each challenge a bullet list of potential approaches could work?}
+
+\CR{Something about error handling would be good to discuss here}
 
@@
 %Splitting code between multiple JastAdd components (and java) can get messy
-Writing idiomatic JastAdd components often requires splitting your code into many different places.
-We attempted to keep each node in the tree responsible for generating the code it corresponds to, which avoids parent nodes having to check their children's structure with \verb|isInstanceOf|, but also means even generating a simple rule can require a dozen calls between different nodes.
+Writing idiomatic JastAdd components often requires splitting your \CR{Please don't address the reader directly}  code into many different places.
+We attempted to keep each node in the tree responsible for generating the code it corresponds to \CR{What does that mean?
+You need to provide a bit more background}, which avoids parent nodes having to check their children's structure with \verb|isInstanceOf|, but also means even generating a simple rule can require a dozen calls between different nodes.
 There is a tradeoff to be made between the single responsibility of each component making it easy to verify its correctness, but the complicated interdependence making it hard to grasp the full picture.
+\CR{Overall I found it difficult to process this paragraph.}
 
 Using JastAdd simultaneously as a library and a build component
+\CR{(incomplete)}
 
@@
 
-Not being able to perform these error checks at runtime means our translation will emit nonfunctional typing definitions if a mistake is written in the typing rules.
+Not being able to perform these error checks at runtime
+\CR{at runtime of which component?}
+means our translation will emit nonfunctional typing definitions if a mistake is written in the typing rules.
 These errors will then go unreported until the output project is compiled or run, at which point the user will have to attempt to trace these errors back to the typing rule definition, to find the actual cause.
+\CR{NB: this is also somewhat true of JastAdd's own compilation approach, so there might be double indirection errors ...}
+
+\CR{This is all about technical challenges.  Do you also see conceptual challenges?  (You can be less specific here, but
+  some words about support for parametric polymorphism and/or subtyping could be nice; at the very least, something
+  about the relevance of syntax-directedness woudl be valuable.)}
 
@@
 Our experiments have shown that reference attribute grammars are a flexible method for implementing type checker generators.
-Implementing the first top-to-bottom functioning product was challenging and took quite a lot of work, but extending it with additional features such as type variables turned out to be a surprisingly painless experience.
+Implementing the first top-to-bottom functioning product \CR{end-to-end prototype is what I would call it} was challenging and took quite a lot of work, but extending it with additional features such as type variables turned out to be a surprisingly painless experience.
 
@@
 This could be solved simply by wrapping each rule within brackets, creating a separate scope for the evaluation of each rule.
+\CR{This is about name clashes in the generated code?  You may want to consider automatic relabelling here anyway (``this'', ``new'' etc. as type variable names are likely going to cause issues) }
 
@@
 \subsection{Better support for numbers, strings, etc.}
+
+\CR{Not sure that there is anything needed here}
 

### 2024-01-04 MSc.tex
Add an initial abstract

@@
 %Leave a \textit{2cm} vertical space after the abstract and provide a few keywords relevant for your report. Use five to six words, of which at most two should be from the title.
+
+  Static type checkers are a popular method for early detection of bugs in computer programs, by ensuring the correct data types are used throughout the program.
+  One approach for implementing type checkers are Reference Attribute Grammars (RAGs), a formalism for specifying the static semantics of programming languages by declaring attributes on an abstract tree-representation of the program, and is used by the JastAdd meta-compilation system.
+
+  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from syntax-directed typing rules.
+  This aims to provide integration with existing compilers implemented with JastAdd, as well as a playground for type systems usable as an educational tool.
+
+  We find RAGs to be a flexible model for syntax-directed typing rules, and our software was able to support simple values, nodes with premises placing demands on their children, and those with rules requiring type variables.
+  It proved difficult to provide compile-time verification of the typing rules, and further work is required to extend the compiler to support type environments, a necessity for most real-world type systems.
 }

### 2024-01-07 MSc.tex
Address many of the comments left by supervisor

@@
 
+% disable spellcheck for this section
+% LTeX: SETTINGS enabled=false
 \lstdefinestyle{mystyle}{
@@
 }
+% LTeX: SETTINGS enabled=true
 
@@
 \newcommand{\CR}[1]{\textcolor{green!60!black}{[\textbf{CR}:#1]}}
+%\newcommand{\CR}[1]{}
+
+% For questions to people reviewing, remove before finalizing
+\newcommand{\reviewquestion}[1]{\textcolor{red!80!black}{[\textbf{Review question}:#1]}}
+%\newcommand{\reviewquestion}[1]{\textcolor{red}{\begin{itemize}\item Question: #1\end{itemize}}}
+%\newcommand{\reviewquestion}[1]{}
 
@@
 %\date{January 16, 2015}
-
-% For questions to people reviewing, remove before finalizing
-\newcommand{\reviewquestion}[1]{\textcolor{red}{\begin{itemize}\item Question: #1\end{itemize}}}
-%\newcommand{\reviewquestion}[1]{}
 
@@
 Catching these mistakes as early as possible can prevent these bugs from ever reaching production code.
-\CR{Possible papers to cite as evidence:
-  \url{https://dl.acm.org/doi/abs/10.1145/1985793.1985889}
-  \url{https://www.usenix.org/legacy/event/sec04/tech/full_papers/johnson/johnson_html/}
-  \url{https://ieeexplore.ieee.org/abstract/document/10190394}}
-\CR{These papers add static checkers on top of existing languages to find additional bugs; you can also
-  mention statically typed languages, but those generally don't require frequent changes to the type checker implementation, so they're less
-  compelling to motivate your work.}
-Type checkers are one of the most common static checking methods, ensuring that the correct data types are used, but creating a new type checker is often an arduous process, requiring large amounts of complex code.
+%\CR{Possible papers to cite as evidence:
+%  \url{https://dl.acm.org/doi/abs/10.1145/1985793.1985889}
+%  \url{https://www.usenix.org/legacy/event/sec04/tech/full_papers/johnson/johnson_html/}
+%  \url{https://ieeexplore.ieee.org/abstract/document/10190394}}
+%\CR{These papers add static checkers on top of existing languages to find additional bugs; you can also
+%  mention statically typed languages, but those generally don't require frequent changes to the type checker implementation, so they're less
+%  compelling to motivate your work.}
+Type checkers are one of the most common static checking methods, ensuring that the correct data types are used.
+While some languages include type checkers in their compilers, work has also been done to add static type checkers to existing languages.
+This requires writing large amounts of specialised code on a per language basis, but has shown to be effective at detecting bugs\cite{10.1145/1985793.1985889}\cite{10.5555/1251375.1251384}\cite{10190394}.
 Furthermore, making changes to the typing rules of a language can require extensive rewrites in many different parts of the type checker's codebase.
@@
 % Define the problem you are solving (or trying to solve).
-\CR{Prior work has already looked into pluggable type checkers.
-  Why is translating them into JastAdd interesting?
-  (aside from theoretical questions, you can also use
-  pragmatic reasons, such as integration into our courses.)}
-In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of reference attribute grammars.
-
-\CR{Make clear to the reader what the challenges are:
-   How are these two formalisms similar, how do they differ (at a high level)?
-   You can use a simple typing rule as an example if it helps (very optional).
-   Hint: look at determinism in evaluation (syntax-directedness), unification,
-   but stick to intuition in this chapter; you can elaborate in Chapter 2 or later
-   if you think it helps.}
-
-%TODO: Give an overview of the other sections
-\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning
-  chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
+%\CR{Prior work has already looked into pluggable type checkers.
+%  Why is translating them into JastAdd interesting?
+%  (aside from theoretical questions, you can also use
+%  pragmatic reasons, such as integration into our courses.)}
+In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of Reference Attribute Grammars (RAGs).
+JastAdd is an interesting target because it has already been used to implement compilers of several languages, to which our typing rule implementations could be integrated.
+JastAdd is also used extensively in courses at the department of computer science at LTH, where our project could serve as an educational tool, providing students an interactive way to define and test simple type systems.
+
+Typing rules form a complex, often ambiguous, and infinite relation.
+To be able to translate them into JastAdd code, we will need to define a limited subset of syntax-directed typing relations.
+We identify a number of different kinds of typing rules, used in previous work in the field\ref{Pierce}, which share a common set of features for us to implement.
+We then devise a strategy to translate these rules into snippets of iterative Java code, which can be evaluated by JastAdd.
+
+%\CR{Make clear to the reader what the challenges are:
+%      How are these two formalisms similar, how do they differ (at a high level)?
+%      You can use a simple typing rule as an example if it helps (very optional).
+%      Hint: look at determinism in evaluation (syntax-directedness), unification, but stick to intuition in this chapter; you can elaborate in Chapter 2 or later if you think it helps.}
+
+\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
   We evaluate the approach against Foo, and demonstrate Bar.
@@
 
-\CR{paper $\to$ thesis; also, this always comes at the very end of the introduction}
-The rest of this paper is outlined as follows.
+% Specify your contributions. What does this particular work/report bring to the research are or to the body of knowledge?
+\section{Research questions}
+To formulate our goals more concretely, we have set out to answer the following research questions:
+%\CR{Give the RQs numbers so you can refer to them later.  Common form: RQ1, RQ2 etc.; some authors use RQ2.3 if they split RQ2 into multiple subquestions (this can happen later).}
+\newcommand{\rqone}{RQ1: What kinds of typing rules can we translate using RAGs?}
+\newcommand{\rqtwo}{RQ2: How can we implement the translations and what algorithms should we use for the output code?}
+\newcommand{\rqthree}{RQ3: What are the challenges in translating typing rules into JastAdd-based RAGs?}
+\begin{itemize}
+  \item \rqone
+    %\CR{``kinds'' of typing rules: right now the reader isn't aware that there even are different kinds (see my earlier comment on ``what the challenges are''-- sketching this out should suffice.)}
+  \item \rqtwo
+    %\CR{What does ``form'' mean, here?  How broad is ``implement''-- are you talking algorithms (which I think you are) or including technical challenges (which I think you have in the next RQ)?}
+  \item \rqthree
+    %\CR{I'd go with `challenges in translating' instead of `potential issues of translating', language-wise}
+\end{itemize}
+
+\section{Chapter outline}
+%\CR{%paper $\to$ thesis; also,
+ % this always comes at the very end of the introduction}
+The rest of this thesis is outlined as follows.
 % This chapter introduces the
@@
 
-% Specify your contributions. What does this particular work/report bring to the research are or to the body of knowledge?
-\section{Research questions}
-To formulate our goals more concretely, we have \CR{Avoid contractions in technical writing} set out to answer the following research questions:
-\CR{Give the RQs numbers so you can refer to them later.  Common form: RQ1, RQ2 etc.; some authors use RQ2.3 if they split RQ2 into multiple subquestions (this can happen later).}
-\newcommand{\rqone}{What kinds of typing rules can we translate through \CR{``through?''} RAG?}
-\newcommand{\rqtwo}{How can we implement the translations and into what form?}
-\newcommand{\rqthree}{What are the potential issues of translating typing rules into JastAdd-based RAGs?}
-\begin{itemize}
-  \item \rqone \CR{``kinds'' of typing rules: right now the reader isn't aware that there even are different kinds (see my earlier comment on ``what the challenges are''-- sketching this out should suffice.)}
-  \item \rqtwo \CR{What does ``form'' mean, here?
-How broad is ``implement''-- are you talking algorithms (which I think you are) or including technical challenges (which I think you have in the next RQ)?}
-  \item \rqthree \CR{I'd go with `challenges in translating' instead of `potential issues of translating', language-wise}
-\end{itemize}
-
-\section{Related work} % Maybe move to discussion?
-% Describe related work (literature study).
-Besides listing other work in the area, mention how is it related or relevant to your work.
-The tradition in some research area is to place this part at the end of the report (check with your supervisor).
-Pacak et al. implemented a compiler for generating incremental type checkers using Datalog\cite{Pacak}.
-While the goal is similar to ours, they take a different approach, by converting typing rules into the declarative logic programming language \CR{second half of this sentence confuses me}.
-
-Their work is thereby largely focused on expressing typing rules within the limits of the Datalog language, which can only compute finite relations.
-We have a significantly different compilation target, JastAdd, which allows us to use reference attribute grammars to evaluate the typing relations, which we write in a more iterative \CR{unclear?} approach.
-\CR{Datalog is fully declarative, whereas JastAdd uses declarative rules that connect snippets of imperative Java code.  Is that what you meant?}
-\CR{Some more thoughts below to help make the distinction clearer (all optional):}
-\CR{Datalog has no direct notion of ASTs or other tree structures / algebraic terms, so trees must be encoded into relations.}
-\CR{Datalog has a simple, built-in notion of unification that can handle cases like testing sibling relations: $\textsf{Sibling}(x, y) :- Parent(p, x), \textsf{Parent}(p, y), x \ne y.$.
-    This is something you deal with later in JastAdd.}
-
-Additionally, their project defines a formalism for expressing typing rules within the Datalog language, ours defines a language which aims to keep as close as possible to the traditional inference rules used for typing rules, to minimise the difficulty of translation.
-\CR{I'd start with this difference-- the translation is targeting two rather different formalisms, which is all you need to justify the novelty.}
-
-\reviewquestion{Circular attributes?}
-\CR{Circular attributes: Recursion in typing rules: Datalog is built with support for ``infinite'' recursion (since it operates on finite domains,
-  it can use a special bottom-up evaluation scheme to ensure termination.)
-  If you need this in RAGs, it gets more complicated, but you'd need an example to explain this in the introduction-- since you don't directly address that in this work,
-  I wouldn't talk about it too much here (you could also skip it entirely here).  Minimum: talk about it in the discussion later.}
-
 \chapter{Background}
 %\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
-\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinkingr ``I know why I want to read this''}
+\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinking ``I know why I want to read this''}
 This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
-The first section explains abstract syntax trees, which are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
-\CR{Consider saying just a little more about them here, for readers who have no idea what ASTs are and why they matter for
-    analysing programs.}
-The next section details type systems \CR{Too much detail in the following:}, by first introducing data types, then explaining the components required for a type system, followed by a description of static type checkers.
-\CR{You can use \texttt{\textbackslash label\{labelname\}} and \texttt{\textbackslash ref\{labelname\}} to reference (sub)secttions, that's usually clearer than ``next section''.}
-
-\section{Abstract syntax trees}
+Section \ref{abstractsyntaxtrees} explains abstract syntax trees, a simplified tree-based representation of a program, suitable for performing static analysis on.
+These are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
+%\CR{Consider saying just a little more about them here, for readers who have no idea what ASTs are and why they matter for
+%    analysing programs.}
+Section \ref{typetheory} details type systems,
+%\CR{Too much detail in the following:}
+by explaining each of the component pieces, followed by a description of their use within static type checking.
+%\CR{You can use \texttt{\textbackslash label\{labelname\}} and \texttt{\textbackslash ref\{labelname\}} to reference (sub)secttions, that's usually clearer than ``next section''.}
+
+\section{Abstract syntax trees}\label{abstractsyntaxtrees}
 \CR{What's a syntax?}
@@
 Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
-It's \CR{contraction} abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
-Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this information \CR{what is ``this information'', concretely?}.
+It is abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
+Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this semantic information.
+
+This makes it easier to write analysis based on the behaviour of a program, without concerning oneself with its exact formulation.
+Several different formulations of a program with identical behaviour may be abstracted into identical ASTs.
+The hierarchical structure of the tree lends itself especially well for recursive analyses, and eases the passing of information between connected structures by connecting them as parent and children nodes.
+%\CR{what is ``this information'', concretely?}
 A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if-else node, with named branches pointing towards the condition and the two values.
-\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
-\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}
-
-This layer of abstraction makes it easier to write analyses for the program, as we can focus on the key elements and structure.
-\CR{Is there a connection between the AST and the behaviour of the program?}
+%\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
+%\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}
 
 \CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}
-
-\section{Type theory}
+\begin{figure}[h]
+\begin{lstlisting}[]
+Expr   -> Expr "+" Term | Term
+Term   -> Term "*" Factor | Factor
+Factor -> Int | "(" Expr ")"
+\end{lstlisting}
+  \caption{\reviewquestion{One like this? I'm unsure how to tie it in effectively.}}
+  \label{typesyntaxspec}
+\end{figure}
+
+\section{Type theory}\label{typetheory}
 In this section we will discuss the elements of type theory required to understand this thesis.
@@
 
+\section{Values and expressions}
+The main purpose of most computer software is to handle and manipulate data.
+Each of these individual pieces of data is a value, which could be anything from the number 4 or the text `Hello!'.
+
+Expressions are a broader term, indicating any syntactic structure which evaluates to a value.
+This includes but is not limited to values, which evaluate to themselves, but also constructs of values, operators or function calls, etc.
+
 \subsection{Data types}
-\CR{think about the connection between types and values here.  You may even want to talk about values in a separate subsection.}
-Computer software generally needs to handle and manipulate several different types of data in one program, such as numbers or text.
+%\CR{think about the connection between types and values here.  You may even want to talk about values in a separate subsection.}
+Data types can be considered a categorisation of values which share common properties, such as supporting numerical operations or accessing a certain element in a list of characters.
+%several different types of data in one program, such as numbers or text.
 Most programming languages come with a number of built-in data types, from different kinds of numbers (integers, floating point) to strings or lists and arrays.
 
-\CR{if you discuss values, you can use the intuition of ``grouping values'' by certain properties (such as supported operations) to make your point clearer.}
+%\CR{if you discuss values, you can use the intuition of ``grouping values'' by certain properties (such as supported operations) to make your point clearer.}
 Different parts of programs may have different constraints on the types of expressions, such as a print function requiring a string of text, or a plus function requiring some kind of number.
-\CR{Very good point about types of expressions, but again, this is likely easier to understand by linking expressions and values.
-  I strongly recommend using a definition of the form ``all values are expressions, but not all expressions are values''.}
+%\CR{Very good point about types of expressions, but again, this is likely easier to understand by linking expressions and values.
+%  I strongly recommend using a definition of the form ``all values are expressions, but not all expressions are values''.}
 
 \subsection{Type checking}
-There are two main approaches to ensuring that the aforementioned constraints are upheld, dynamic and static type-checking.
-\CR{Why would we even want to ensure that they are upheld?}
+Verifying that these constraints are upheld can assist us in catching programming mistakes, preventing the wrong data types from being used in the wrong places, which may have lead to erroneous or unpredictable behaviour.
+There are two main approaches to ensuring the constraints are upheld, dynamic and static type-checking.
+As our project generates a static type checker, we will focus our explanations on that topic, but provide a quick overview of their differences here.
+%\CR{Why would we even want to ensure that they are upheld?}
+
 Dynamic checking is done performed runtime, with each type checked only when needed.
+While this means no checks are performed unnecessarily, it makes it harder to verify the constraints over the whole program, as it won't be tested until each specific part is executed.
+
 Static checking is performed on the code before it is run, usually as part of compilation, preventing many faulty programs from being run at all.
-\CR{Why would I want static checking?  Dynamic only checks when needed, isn't that more efficient?}
-
-As our project generates a static type checker, we will focus our explanations on that topic.
+However, since it lacks knowledge of what happens at runtime, it's required to take a more conservative approach, flagging potential errors even in code which will never executed.
 Static checkers analyse the structure of the source code, often in the form of an abstract syntax tree.
-Conservative \CR{incomplete sentence?}
+
+%\CR{Why would I want static checking?  Dynamic only checks when needed, isn't that more efficient?}
 
 \CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
-  For each concrete (and possibly very complex) operator on the left hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
+  For each concrete (and possibly very complex) operator on the left-hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
   This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
-  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractoins
-  over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
+  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractions over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
 }
@@
 \subsection{Type systems}
-A type system consists of two parts, a type syntax, describing the available types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
-\CR{Why do the types need a syntax?
-    (There are other approaches to types; for your own sanity, I recommend deferring to Pierce here and calling this ``typical approaches'')}
+A type system typically consists of two parts, a type syntax, describing the available types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
+%\CR{Why do the types need a syntax?
+%    (There are other approaches to types; for your own sanity, I recommend deferring to Pierce here and calling this ``typical approaches'')}
 
 \subsubsection{Type syntax}
-A type syntax is a definition of the valid types within a language.
+A type syntax, or set of types, is a definition of the valid types within a language.
 \CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
+\reviewquestion{I believe Pierce simply refers to it as the "set of types".
+I seem to have got the term type syntax from EDAP15 \url{https://fileadmin.cs.lth.se/cs/Education/EDAP15/2022/web/skills/type-system.html}.
+In my opinion it works better to have a clear name for it, so I'm keeping it for now, unless you have further thoughts.}
 It can consist of fundamental types, such as integers or booleans.
 There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
-\CR{A bit vague; consider giving an example.}
+This enables differentiating lists that store integers from those that store strings, and ensuring that any item added to a list must be of the same type as the type parameter.
+%\CR{A bit vague; consider giving an example.}
+\begin{figure}[h]
+\begin{lstlisting}[]
+abstract Type;
+
+Bool : Type;
+Int  : Type;
+\end{lstlisting}
+  \caption{Type syntax consisting of two types, for boolean and integer values, written as a JastAdd AST specification (see Section \ref{astdef})}
+  \label{typesyntaxspec}
+\end{figure}
 
 \subsubsection{Typing rules}
-The typing relation, notated \lstinline{t : T}, is a link between the expressions of a language and the types of the type syntax.
-\CR{It is \emph{the} link, even.}
-The left-hand side is an expression in the object lanuage, typically written in its native syntax, the right-hand side a type from the aforementioned type syntax.
+The typing relation, notated \lstinline{t : T}, is the link between the expressions of a language and the types of the type syntax.
+%\CR{It is \emph{the} link, even.}
+The left-hand side is an expression in the object language, typically written in its native syntax, the right-hand side a type from the aforementioned type syntax.
 
 Typing rules are a manner of inference rules, which in their simplest form, consist of no more than a typing relation.
-The typing rule \lstinline{T-True} seen in figure \ref{typingruleexampletrue}, for example, simply assigns the type of \lstinline{Bool} to the expression \lstinline{true}.
+The typing rule \lstinline{T-True} seen in Figure \ref{typingruleexampletrue}, for example, simply assigns the type of \lstinline{Bool} to the expression \lstinline{true}.
 Additionally, the typing rules in these examples have a name, given in parentheses next to the rule, though this is merely for reference.
@@
   \end{equation}
-  \caption{Typing rule for a true expression}
+  \caption{Typing rule for a true value}
   \label{typingruleexampletrue}
@@
 
-Typing rules may also have premises, a number of conditions which \CR{\emph{that} must hold (``which'' relates a set to a superset: ``carrots, which are vegetables'', while ``that'' restricts a set to a subset: ``vegetables that are carrots'')} must hold for the typing rule to apply.
+Typing rules may also have premises, a number of conditions that must hold for the typing rule to apply.
+%\CR{\emph{that} must hold (``which'' relates a set to a superset: ``carrots, which are vegetables'', while ``that'' restricts a set to a subset: ``vegetables that are carrots'')}
 The notational convention here is to put the premises above a bar, with the conclusion underneath it.
@@
       t_1 : \text{Bool}\quad
-      t_2 : \text{T}\quad
-      t_3 : \text{T}
+      t_2 : \textsf{T}\quad
+      t_3 : \textsf{T}
     }{
-      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \text{T}
+      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \textsf{T}
     }
@@
 
+\newcommand{\ty}[1]{\textcolor{blue!20!black}{\textsf{#1}}}
+\newcommand{\tyvar}[1]{\textsf{#1}}
 Figure \ref{typingruleexampleif} shows an example rule, typing an if-expression.
-The bottom line, the conclusion, contains the keywords of the if-expression and three variables $t_i$ which represent any term within the language.
+The bottom line, the conclusion, contains the keywords of the if-expression and three free variables $t_i$ which represent any term within the language.
 Above the bar are three premises, placing constraints on the three variables, which must hold for the typing rule to apply.
-The first constrains $t_1$ to the type bool, the second two constrain $t_2$ and $t_3$ to a type variable T, which represents any single arbitrary type within the type syntax.
-While T may be any type, it must consistently be the same type within any single evaluation of the rule, i.e. the two terms must have the same type.
-This T also appears in the conclusion, as the resulting type of the if-expression, meaning the result of the entire expression will be that of the $t_2$ and $t_3$ terms.
-\newcommand{\ty}[1]{\textcolor{blue!20!black}{\textsf{#1}}}
-\newcommand{\tyvar}[1]{\ty{\textcolor{cyan!50!black}{#1}}}
-\CR{Consider typesetting T differently, e.g. as \tyvar{T}}
-\CR{\tyvar{T}, $t_1$, $t_2$, $t_3$ are \emph{free} variables here, logically speaking}.
-\CR{Please explain the purpose of typing rules in static type checking.  Using just your two examples you would be able to
-  show a well-typed program, and if you add one more rule (e.g. for integers), you can show a type error.  No need to build
-  the full derivation tree, but it's important to mention recursion here.  This entire bit may be worth a separate subsubsection.}
+The first constrains $t_1$ to the type bool, the second two constrain $t_2$ and $t_3$ to a type variable \tyvar{T}, which represents any single arbitrary type within the type syntax.
+While \tyvar{T} may be any type, it must consistently be the same type within any single evaluation of the rule, i.e. the two terms must have the same type.
+This \tyvar{T} also appears in the conclusion, as the resulting type of the if-expression, meaning the result of the entire expression will be that of the $t_2$ and $t_3$ terms.
+%\CR{Consider typesetting T differently, e.g. as \tyvar{T}}
+%\CR{\tyvar{T}, $t_1$, $t_2$, $t_3$ are \emph{free} variables here, logically speaking}.
+
+\subsubsection{Evaluating typing rules}
+While performing static type checking, the typing rules form the basis of our analysis.
+By analysing the program, we attempt to match the syntactic form with the typing rules, to find the type of the expression.
+If no matching typing rules can be found, a type error is produced.
+
+The simplest program possible with the typing rules T-True and T-If, consists of the single value \lstinline{true}.
+It can immediately be matched to the rule T-True, giving the type \lstinline{Bool}.
+
+A slightly more complex case can be expressed as \lstinline{if true then true else true}.
+The only rule that matches the structure, is T-If, with \lstinline{true} values taking the place of each of the $t_i$ variables.
+To evaluate this rule, we are thus required to recursively evaluate the types each of our $t_i$ variables, to ensure they match the premises.
+In this case, the type of $t_1$ evaluates to \lstinline{Bool}, matching the first premise of T-If.
+The types of both $t_2$ and $t_3$, represented as the type variable \tyvar{T}, also evaluate to \lstinline{Bool}.
+Since both the uses of this type variable within the premises are consistent, all three of the rule's premises have been fulfilled, and the entire expression can be typed as \tyvar{T}, which has been found to correspond to \lstinline{Bool}.
+
+\begin{figure}[h]
+  \begin{equation}
+    \tag{T-Zero}
+    \text{zero} : \text{Int}
+  \end{equation}
+  \caption{Typing rule for a zero value}
+  \label{typingruleexamplezero}
+\end{figure}
+
+Adding a second value of a different type allows us to express incorrectly typed programs, for example the program \lstinline{if zero then zero else true}.
+Recursively evaluating this according to T-If, we find that the premise concerning $t_1$ no longer upholds, as the type of it now evaluates to \lstinline{Int}.
+Additionally, the type variable \tyvar{T} typed by both $t_2$ and $t_3$ now has contradictory use, being assigned both as the type \lstinline{Int} and \lstinline{Bool}.
+As this program cannot be found to satisfy any set of typing rules in our type system, it is found to be invalid.
+
+%\CR{Please explain the purpose of typing rules in static type checking.
+Using just your two examples you would be able to show a well-typed program, and if you add one more rule (e.g.
+for integers), you can show a type error.
+No need to build the full derivation tree, but it's important to mention recursion here.
+This entire bit may be worth a separate subsubsection.}
 
 \section{Reference Attribute Grammars}
-This section introduces the subject of Reference Attribute Grammars (RAGs), first explaining the theory behind them, then introducing JastAdd, an implementation of RAGs within a Java meta-compiliation system\cite{jastadd}.
+This section introduces the subject of Reference Attribute Grammars (RAGs), first explaining the theory behind them, then introducing JastAdd, an implementation of RAGs within a Java meta-compilation system\cite{JastAdd}.
 
 \subsection{Reference Attribute Grammars}
-Reference Attribute Grammars (RAGs) are an extension to Attribute Grammars (AGs), a formalism used for specifying the semantics of programming languages.
-\CR{Generally only for the static semantics (RAGs can also do dynamic semantics, if you allow circular attributes, but that's not worth mentioning here.)
-Please also connect ``semantics'' to ``type checking'' (via ``static semantics'')}
+Reference Attribute Grammars are an extension to Attribute Grammars (AGs), a formalism used for specifying the static semantics of programming languages.
+%\CR{Generally only for the static semantics (RAGs can also do dynamic semantics, if you allow circular attributes, but that's not worth mentioning here.)
+%Please also connect ``semantics'' to ``type checking'' (via ``static semantics'')}
 AGs provide a systematic way to define computations on the nodes of an abstract syntax tree, associating attributes with language constructs and are commonly used in the context of compiler construction, where they help define the translation of source code into executable code\cite{AGs}.
@@
 
+\subsection{Non-Terminal Attributes}
+\cite{higherorderattributegrammars}
+\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are
+not a necessary feature for RAGs.  Please introduce them in a separate section: you can very directly connect them to the idea of ``type syntax''.}
+
 \subsection{JastAdd}
-JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion.
-\CR{It's worth telling the reader here why you are explaining some details about JastAdd.}
-
-\cite{JastAdd}
+JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion.\cite{JastAdd}
+Our project uses it both for the implementation of our compiler and as a compilation target, as it outputs JastAdd code.
+%\CR{It's worth telling the reader here why you are explaining some details about JastAdd.}
+
 \subsubsection{AST specification}\label{astdef}
-One of the key components of a JastAdd projects are the \verb|.ast| files, which declare the structure of the abstract syntax tree.
+One of the key components of a JastAdd projects are the \lstinline{.ast} files, which declare the structure of the abstract syntax tree.
 This specification is then used by JastAdd to generate Java classes representing the nodes of the AST.
@@
 \subsubsection{RAG Modules}
-\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are
-not a necessary feature for RAGs.  Please introduce them in a seprate section: you can very directly connect them to the idea of ``type syntax''.}
 \begin{figure}[h]
@@
 \end{figure}
-The RAG modules written in \verb|.jrag| files, declare the attributes to be included in the classes generated by the AST specification.
+The RAG modules written in \lstinline{.jrag} files, declare the attributes to be included in the classes generated by the AST specification.
 The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.
@@
 
-\CR{Missing: introduction to this chapter'}
-
-\section{Architecture}
-In this section we will be providing an overview of our architecture.
-\CR{Stylistically, technical documents should avoid discussing the process of reading (it's not as frowned upon as speaking directly to the reader, though), so I recommend just saying that the figure ``provides an overview over...''}
-Since it contains many different components, it may be helpful to refer to the overview graph in figure \ref{overviewgraph} while reading.
-\CR{Please put the following in the caption of the figure instead.}
-It represents the flow of the program, with components and artefacts (represented as rectangular boxes) with arrows linking them to actions (elliptical boxes) which in turn have arrows to the new components and artefacts they produce.
-The dashed arrows from represent links from the template components that exist in the current experimental state of the project, but in a finished product would instead link to component from the object language's own compilation environment.
-%It uses rectangular boxes to represent the concrete data or artefacts of the procedure, and elliptical boxes to represent the actions performed with them.
-
-\CR{You still want an introduction to this section that explains what is following and why}
-
+%\CR{Missing: introduction to this chapter'}
+We begin this chapter with a section presenting our language for writing typing rules, an adaptation of the natural semantics used to notate typing rules in the academic context.
+In Section \ref{sectionarchitecture} we proceed with presenting the overall architecture of our software, including an overview graph in Figure \ref{overviewgraph}, which may be useful to bear in mind while reading the rest of the report.
+Section \ref{sectionimplementation} goes into the details of our implementation, how we generate code for different parts of the typing rules.
+
+\section{Language design}\label{sectionlanguage}
 \begin{figure}[]
-  \resizebox{\textwidth}{!}{%
-  \begin{tikzpicture}[
-      every node/.style = {draw=black, rounded corners, align=center},
-      every path/.style = {draw, -latex}
-    ]
-    \node (typerules)                               {Typing rules\\definition (\ref{typingruledefinition})};
-    \node (trparsing) [right=of typerules, ellipse] {Scan,\\parse,\\build};
-    \node (trparser)  [above=of trparsing]          {Typing rule parser\\(\ref{typingrulesparser})};
-    \node (trast)     [right=of trparsing]          {Typing rules AST\\(\ref{typingrulesast})};
-
-    %\node (typecheckgen) [below=of trast, ellipse, text width=5cm] {
-    %  \textbf{Type checker generation}\\
-    %  \hrule
-    %  \begin{itemize}[noitemsep,topsep=2pt]
-    %    \item Copy template
-    %    \item Generate type checker from typing rules
-    %  \end{itemize}
-    %};
-    \node (typecheckgen)       [below=of trast, ellipse] {Generate\\type checker};
-    \node (typingrulecompiler) [right=of trast]          {Typing rule\\compiler (\ref{typingrulecompiler})};
-    \node (typechecker)        [below=of typecheckgen]   {Type checker};
-
-    \node (olast)     [below=of typerules] {AST specification\\(\ref{astdef})};
-
-    \node (outputtypechecking) [below=of typechecker, ellipse] {Perform\\typechecking};
-    \node (inputast)           [left=of outputtypechecking]    {Input program\\AST};
-    \node (outputparsing)      [left=of inputast, ellipse]     {Scan,\\parse,\\build};
-    \node (typeerrors)         [right=of outputtypechecking]   {Type errors\\(\ref{errorhandling})};
-    \node (inputprog)          [left=of outputparsing]         {Input program};
-
-    %\node (template) [below=of inputast, text width=6cm] {
-    %  \textbf{Type checker template}\\
-    %  \hrule
-    %  \begin{itemize}[noitemsep,topsep=2pt]
-    %    \item Scanner, parser, builder
-    %    \item Project files
-    %    \item Type checking algorithm (\ref{typecheckingalgorithm})
-    %  \end{itemize}
-    %};
-    \node (templateparser)        [below=of outputparsing]                               {Tree-based\\parser\\(\ref{treebasedparser})};
-    \node (templateprojectfiles)  at (inputast |- templateparser)                        {Project files\\(\ref{projectfiles})};
-    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser)              {Type checking\\aspect (\ref{typecheckingaspect})};
-    \node (templatetypesyntax)    [right=of typecheckingalgorithm]                       {Type syntax\\AST};
-    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm) (templatetypesyntax)] {};
-    \node                         [above, draw=none] at (template.north)                 {Template};
-
-    \node (userinput)   [dashed, fill=none, fit=(typerules) (olast)]                  {};
-    \node               [above, draw=none] at (userinput.north)                       {Object language\\static semantics};
-    %\node (typechecker) [dashed, fill=none, fit=(outputparsing) (outputtypechecking)] {};
-    %\node               [above, draw=none] at (typechecker.north)                     {Type checker};
-
-    \draw (typerules)          to (trparsing);
-    \draw (trparser)           to (trparsing);
-    \draw (trparsing)          to (trast);
-    \draw (trast)              to (typecheckgen);
-    \draw (typingrulecompiler) to (typecheckgen);
-    \draw (typechecker)        to (outputtypechecking);
-    \draw (olast)              to (outputparsing);
-    \draw (outputparsing)      to (inputast);
-    \draw (inputast)           to (outputtypechecking);
-    \draw (outputtypechecking) to (typeerrors);
-    \draw (typecheckgen)       to (typechecker);
-    \draw (inputprog)          to (outputparsing);
-
-    \draw[dashed] (templateparser)                  to (outputparsing);
-    \draw         (typecheckingalgorithm)           to (outputtypechecking);
-    \draw[dashed] (templateprojectfiles.north west) to (outputparsing);
-    \draw[dashed] (templateprojectfiles.north east) to (outputtypechecking);
-    \draw[dashed] (templatetypesyntax)              to (outputtypechecking);
-
-  \end{tikzpicture}
-  }
-  \caption{An overview of the project components and flow. \CR{Can you mark, e.g. via colours, the parts that the user provides,
-           and the parts that will be in the generated output project?}}
-  \label{overviewgraph}
-\end{figure}
-
-\subsection{Typing rule definition}\label{typingruledefinition}
-To express typing rules, we created a simple ASCII representation of the natural deduction notation used by Pierce\cite{Pierce}.
-Custom DSL
-Include typing rules AST specification (or BNF?)
-ASCII representation of natural deduction %BNF (is it BNF?)
-\subsubsection{Language}\label{sectionlanguage}
+\begin{lstlisting}[language=java]
+ruleset = rule_list
+
+rule_list =
+    rule
+  | rule_list rule
+
+rule =
+    '[' upperid ']' premises_list '---+' formula
+  | '[' upperid ']' formula
+
+premises_list =
+    formula
+  | premises_list ',' formula
+
+formula = term ':' tyterm
+
+term =
+    upperid '(' term_list ')'
+  | upperid '(' ')'
+  | upperid
+  | lowerid
+
+tyterm =
+    upperid                                              // TyVal
+  | lowerid                                              // TyVar
+
+term_list =
+    term
+  | term_list ',' term
+\end{lstlisting}
+  \caption{Syntax definition for the typing rule language}
+  \label{trastspec}
+\end{figure}
 \CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
@@
 
-To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{pierce}.
-Our language can express a subset of the notation used by Pierce, with some differences and limitations
-\CR{either forward reference to the discussion of the limitations, or remove the ``limitations'' bit here (you already say ``subset'', after all) and discuss those limitations after you define the language.}.
+To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{Pierce}.
+Our language can express a subset of the notation used by Pierce, with some differences.
+
+The largest difference lies in the notation used for the expression to be typed, where Pierce uses the object language's native syntactic form, whereas our representation is based on the corresponding AST nodes.
+Node names are consistent with those used by JastAdd, followed by parentheses containing variable names to denote their child nodes in the order of their definition.
+This change is necessary to allow us to type arbitrary languages without needing to understand their syntax and grammar, though it does make translating rules from one syntax to the other less straightforward.
+
+Another difference is the use of the type variables, which have been made lowercase.
+This disambiguates them from the concrete types, which as AST nodes themselves are written with uppercase letters, and also aligns the syntax for type terms with that of regular terms, where concrete nodes are written uppercase and variables in lowercase.
+\lstinline{T}, which has been switched to the lowercase \lstinline{t} in order to distinguish concrete types and type variables.
+This disambiguates them from the concrete types, which as AST nodes themselves are written with uppercase letters, and also aligns the syntax for type terms with that of regular terms, where concrete nodes (such as \lstinline{If}) are written uppercase and variables (\lstinline{t1}, \lstinline{t2}, \lstinline{t3}) in lowercase.
+%\CR{I generally recommend avoiding the passive voice, but the use below is an example that would be OK}
+Commas are also used to separate the premises, and the horizontal line separating premises from the conclusion is represented as a series of dashes.
+
+%\CR{Optional: you can mention Alexandru's paper about how one could add metavariables into ``native'' syntax: \url{https://dl.acm.org/doi/pdf/10.1145/3485542}
+%  (since this part has already been done, it's not something you needed to explore)}
+%\CR{Capitalising the AST node types is consistent with their naming in JastAdd-- worth pointing out.}
+
+%\CR{either forward reference to the discussion of the limitations, or remove the ``limitations'' bit here (you already say ``subset'', after all) and discuss those limitations after you define the language.}
 
 Figures \ref{trexampletrue} and \ref{trexampleif} show two rules from Pierce in both their original syntax and our ASCII representation side by side.
-The former figure describes a simple rule named T-True, which assigns the \verb|true| expression the type \verb|Bool|.
-
-The largest difference can be seen in the expression to be typed, where Pierce uses the object language's native syntactic form, whereas our representation is based on the corresponding AST nodes.
-\CR{Given that it's nontrivial to separate out the different changes, I recommend first enumerating the differences abstractly and then discussing how the changes manifest in the figures, rather than starting with the figures to ``derive'' the differences.}
-In the first example, the difference is minor (\verb|true| and \verb|True| respectively) but the difference is more apparent in figure \ref{trexampleif}, where instead of the full syntactic form, the expression is written as an AST node with its children represented as parameters.
-%Note the difference between the \verb|true| written in Pierce, representing the native syntactic form of the expression, and the \verb|True| in the ASCII representation, representing the name of the corresponding AST node.
-%This difference is more apparent in figure \ref{trexampleif}, where instead of the full syntactic form, the expression is written as an AST node with its children represented as parameters.
-This change is necessary to allow us to type arbitrary languages without needing to understand their syntax and grammar, though it does make translating rules from one syntax to the other less straightforward.
-\CR{Optional: you can mention Alexandru's paper about how one could add metavariables into ``native'' syntax: \url{https://dl.acm.org/doi/pdf/10.1145/3485542}
-  (since this part has already been done, it's not something you needed to explore)}
-\CR{Capitalising the AST node types is consistent with their naming in JastAdd-- worth pointing out.}
-
-Another difference is the use of the metavariable \verb|T|, which has been switched to the lowercase \verb|t| in order to distinguish concrete types and type variables.
-This change also aligns the syntax for type terms with the regular terms, where concrete nodes (such as \verb|If|) are written uppercase and variables (\verb|t1|, \verb|t2|, \verb|t3|) in lowercase.
-\CR{I generally recommend avoiding the passive voice, but the use below is an example that would be OK}
-Commas are also used to separate the premises, and the horizontal line separating premises from the conclusion is represented as a series of dashes.
-
-\CR{Important: this affects numbers used at the type level only (which is an advanced feature mostly used in undecidable
-  typesystems like the Calculus of Constructions), and it wouldn't matter here since you can't use \ty{Succ} as a parametric
-  type anyway (cf. igure 3.4).
-  My problem is that right now this reads as if you can't typecheck programs with Integer literals,
-  but those would be e.g. \textsf{Integer} AST nodes, which (to my reading) you would have no trouble writing typing rules for.
-  (you never need to touch the actual lexeme that contains the character string that encodes the integer, in traditional
-  type analysis).  So you shouldn't actually have any real-life problem here.}
-Numerical values have not been implemented, but natural numbers can be expressed using Peano numbers\cite{peano}, utilising a zero value and successor function which represents the number incremented by one.
-If a language has the syntax \verb|0 + 1| it may instead be represented as \verb|Add(Zero, Succ(Zero))|.
-
-\subsubsection{Parser}\label{typingrulesparser}
+The former figure describes a simple rule named T-True, which assigns the \lstinline{true} expression the type \lstinline{Bool}.
+
+%\CR{Given that it's nontrivial to separate out the different changes, I recommend first enumerating the differences abstractly and then discussing how the changes manifest in the figures, rather than starting with the figures to ``derive'' the differences.}
+Looking at the notation of the typed expressions in Figure \ref{trexampletrue}, the difference is minor, the syntactic term \lstinline{true} has been replaced by the name of the AST node \lstinline{True}.
+The notational change is more apparent in Figure \ref{trexampleif}, where instead of the full native syntactic form, the expression is written as an AST node with its children represented as parameters.
+%Note the difference between the \lstinline{true} written in Pierce, representing the native syntactic form of the expression, and the \lstinline{True} in the ASCII representation, representing the name of the corresponding AST node.
+%This difference is more apparent in Figure \ref{trexampleif}, where instead of the full syntactic form, the expression is written as an AST node with its children represented as parameters.
+
+%\CR{Important: this affects numbers used at the type level only (which is an advanced feature mostly used in undecidable
+%  typesystems like the Calculus of Constructions), and it wouldn't matter here since you can't use \ty{Succ} as a parametric
+%  type anyway (cf. igure 3.4).
+%  My problem is that right now this reads as if you can't typecheck programs with Integer literals,
+%  but those would be e.g. \textsf{Integer} AST nodes, which (to my reading) you would have no trouble writing typing rules for.
+%  (you never need to touch the actual lexeme that contains the character string that encodes the integer, in traditional
+%  type analysis).  So you shouldn't actually have any real-life problem here.}
+%Numerical values have not been implemented, but natural numbers can be expressed using Peano numbers\cite{peano}, utilising a zero value and successor function which represents the number incremented by one.
+%If a language has the syntax \lstinline{0 + 1} it may instead be represented as \lstinline{Add(Zero, Succ(Zero))}.
+
+\section{Architecture}\label{sectionarchitecture}
+In this section we will be presenting our architecture.
+%Since it contains many different components, it may be helpful to refer to the overview graph in Figure \ref{overviewgraph} while reading.
+Figure \ref{overviewgraph} provides an overview of all the components and the connections between them.
+The rest of the section goes into further detail about individual components.
+
+%\CR{Stylistically, technical documents should avoid discussing the process of reading (it's not as frowned upon as speaking directly to the reader, though), so I recommend just saying that the figure ``provides an overview over...''}
+%\CR{Please put the following in the caption of th/e figure instead.}
+%It uses rectangular boxes to represent the concrete data or artefacts of the procedure, and elliptical boxes to represent the actions performed with them.
+
+%\CR{You still want an introduction to this section that explains what is following and why}
+
+\begin{figure}[]
+  \resizebox{\textwidth}{!}{%
+  \begin{tikzpicture}[
+      every node/.style = {draw=black, align=center},
+      every path/.style = {draw, -latex},
+      artifact/.style={draw=black},
+      process/.style={ellipse}
+      %greennode/.style={shape=circle, draw=green, line width=2},
+      %rednode/.style={shape=circle, draw=red, line width=2}
+    ]
+    \node (typerules)                               {Typing rules\\definition (\ref{typingruledefinition})};
+    \node (trparsing) [right=of typerules, process] {Scan,\\parse,\\build};
+    \node (trparser)  [above=of trparsing]          {Typing rule parser\\(\ref{typingrulesparser})};
+    \node (trast)     [right=of trparsing]          {Typing rules AST\\(\ref{typingrulesast})};
+
+    %\node (typecheckgen) [below=of trast, process, text width=5cm] {
+    %  \textbf{Type checker generation}\\
+    %  \hrule
+    %  \begin{itemize}[noitemsep,topsep=2pt]
+    %    \item Copy template
+    %    \item Generate type checker from typing rules
+    %  \end{itemize}
+    %};
+    \node (typecheckgen)       [below=of trast, process] {Generate\\type checker};
+    \node (typingrulecompiler) [right=of trast]          {Typing rule\\compiler (\ref{sectionimplementation})};
+    \node (typechecker)        [below=of typecheckgen]   {Type checker \\(\ref{typecheckingalgorithm})};
+
+    \node (olast)     [below=of typerules] {AST specification\\(\ref{astdef})};
+
+    \node (outputtypechecking) [below=of typechecker, process] {Perform\\typechecking};
+    \node (inputast)           [left=of outputtypechecking]    {Input program\\AST};
+    \node (outputparsing)      [left=of inputast, process]     {Scan,\\parse,\\build};
+    \node (typeerrors)         [right=of outputtypechecking]   {Type errors\\(\ref{errorhandling})};
+    \node (inputprog)          [left=of outputparsing]         {Input program};
+
+    %\node (template) [below=of inputast, text width=6cm] {
+    %  \textbf{Type checker template}\\
+    %  \hrule
+    %  \begin{itemize}[noitemsep,topsep=2pt]
+    %    \item Scanner, parser, builder
+    %    \item Project files
+    %    \item Type checking algorithm (\ref{typecheckingalgorithm})
+    %  \end{itemize}
+    %};
+    \node (templateparser)        [below=of outputparsing]                               {Tree-based\\parser\\(\ref{treebasedparser})};
+    \node (templateprojectfiles)  at (inputast |- templateparser)                        {Project files\\(\ref{projectfiles})};
+    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser)              {Type checking\\aspect (\ref{typecheckingaspect})};
+    \node (templatetypesyntax)    [right=of typecheckingalgorithm]                       {Type syntax\\AST};
+    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm) (templatetypesyntax)] {};
+    \node                         [above, draw=none] at (template.north)                 {Support library};
+
+    \node (userinput)   [dashed, fill=none, fit=(typerules) (olast)]                  {};
+    \node               [above, draw=none] at (userinput.north)                       {Object language\\static semantics};
+    %\node (typechecker) [dashed, fill=none, fit=(outputparsing) (outputtypechecking)] {};
+    %\node               [above, draw=none] at (typechecker.north)                     {Type checker};
+
+    \draw (typerules)          to (trparsing);
+    \draw (trparser)           to (trparsing);
+    \draw (trparsing)          to (trast);
+    \draw (trast)              to (typecheckgen);
+    \draw (typingrulecompiler) to (typecheckgen);
+    \draw (typechecker)        to (outputtypechecking);
+    \draw (olast)              to (outputparsing);
+    \draw (outputparsing)      to (inputast);
+    \draw (inputast)           to (outputtypechecking);
+    \draw (outputtypechecking) to (typeerrors);
+    \draw (typecheckgen)       to (typechecker);
+    \draw (inputprog)          to (outputparsing);
+
+    \draw[dashed] (templateparser)                  to (outputparsing);
+    \draw         (typecheckingalgorithm)           to (outputtypechecking);
+    \draw[dashed] (templateprojectfiles.north west) to (outputparsing);
+    \draw[dashed] (templateprojectfiles.north east) to (outputtypechecking);
+    \draw[dashed] (templatetypesyntax)              to (outputtypechecking);
+
+    \matrix [below left, nodes={minimum size = 15pt, align=left}, align=left] at (current bounding box.north east) {
+      \node [artifact, label=right:Artifact] {}; \\
+      \node [process, label=right:Process] {}; \\
+      \node [draw=none, label=right:Flow] (leg1){}; \\
+      \node [draw=none, label=right:Flow (Temporary)] (leg2){}; \\
+    };
+    \draw [label=right:Line] (leg1.west) to (leg1.east);\\
+    \draw [dashed, label=right:Dashed] (leg2.west) to (leg2.east);\\
+
+  \end{tikzpicture}
+  }
+  \caption{An overview of the project components and flow.
+It represents the flow of the program, with components and artefacts (represented as rectangular boxes) with arrows linking them to actions (elliptical boxes) which in turn have arrows to the new components and artefacts they produce.
+The dashed arrows from represent links from the support library components that exist in the current experimental state of the project, but in a finished product would instead link to component from the object language's own compilation environment.
+  \CR{Can you mark, e.g. via colours, the parts that the user provides,
+           and the parts that will be in the generated output project?}}
+  \label{overviewgraph}
+\end{figure}
+
+\subsection{Typing rule definition}\label{typingruledefinition}
+To express typing rules, we created a simple ASCII representation of the natural deduction notation used by Pierce\cite{Pierce}.
+Custom DSL
+Include typing rules AST specification (or BNF?)
+ASCII representation of natural deduction %BNF (is it BNF?)
+
+\subsubsection{Typing rules parser}\label{typingrulesparser}
 The parser for typing rules is written using JFlex\cite{jflex}, a scanner generator, and Beaver\cite{beaver}, an LALR parser generator.
@@
 
-\subsubsection{Abstract syntax tree}\label{typingrulesast}
-\CR{I would give this AST/BNF a more specific name here, to distinguish it from the other two (as done nicely already in Figure 3.1)}
+\subsubsection{Typing rules AST}\label{typingrulesast}
+%\CR{I would give this AST/BNF a more specific name here, to distinguish it from the other two (as done nicely already in Figure 3.1)}
 \begin{figure}[h]
@@
 \end{figure}
-\CR{The syntax doesn't allow parametric types; please make that explicit here.}
+%\CR{The syntax doesn't allow parametric types; please make that explicit here.}
 
@@
 
-\subsection{Template}
-In addition to the compiled files \CR{What are the ``compiled files?''}, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the template.
+Currently, our compiler does not support type parameters, so they are not included in the AST.
+To support this in future, an additional \lstinline{TyTerm} could be added which includes a list of \lstinline{TyTerm}s along with its ID.
+
+\subsection{Support library}
+In addition to the generated type-checking files, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the support library.
 Some of these are key components of the type checker, but others are dummy components included solely for convenience of testing.
@@
 For testing purposes, we have written generic versions of these components, which work for any supported object language.
-\CR{``Templates'' are usually things that are mostly fixed but have holes to be filled in.  I would call what you describe here the ``support library''}
-
-\subsubsection{Parser}\label{treebasedparser}
-\CR{Parser for what?}
-\CR{Is there a conection between this section and Section~\ref{typingrulesparser} wrt parsing terms of the object language?}
-For testing purposes, our template comes with a generic tree-based parser.
+%\CR{``Templates'' are usually things that are mostly fixed but have holes to be filled in.  I would call what you describe here the ``support library''}
+
+\subsubsection{Input program parser}\label{treebasedparser}
+%\CR{Parser for what?}
+%\CR{Is there a conection between this section and Section~\ref{typingrulesparser} wrt parsing terms of the object language?}
+For testing purposes, our support library comes with a generic tree-based parser.
 Programs are written as a tree of AST node names followed by potential parenthesis within which a comma separated list of child nodes are written.
-It \CR{what?} recursively parses the tree, using Java's reflection API to find a class with a matching name, and a constructor with parameters of the same amount and types.
-Figure \ref{boolinputexample} shows an example input file for the Bools language, which will be discussed further in section \ref{boolsast}.
+The parser parses the tree recursively, using Java's reflection API to find a class with a matching name, and a constructor with parameters of the same amount and types.
+Figure \ref{boolinputexample} shows an example input file for the Bools language, which will be discussed further in Section \ref{boolsast}.
 
@@
 These children will be null, causing issues further down the line, though as this parser is only intended for testing it is of little concern.
-\CR{Does this also apply to the analysis of typing rules?  If so, you can mention it as ``future work''
-  or describe how it could be fixed (it's not a fundamental limitation, so nothing that warrants a deeper discussion)}
+%\CR{Does this also apply to the analysis of typing rules?  If so, you can mention it as ``future work''
+%  or describe how it could be fixed (it's not a fundamental limitation, so nothing that warrants a deeper discussion)}
 
@@
 It can parse simplified tree-representations of programs, and utilise these to perform the type checking.
-
-\CR{Can you elaborate on if/how this would integrate with an existing language implementation?
-  I've always interpreted everything as that you start with some `objectlanguage.jar' file of an existing language
-  (given that you use reflection), and you get out
-  a fresh Gradle project that uses that `objectlanguage.jar' for parsing and AST construction,
-  but since you generate JastAdd source code, you can't really (to my understanding) add the `type()' attribute etc.\@
+In a future real world use case, it is intended for our generated type-checking aspects to integrate into a pre-existing project, which already has all the necessary project files to parse an AST tree, to which our type aspects could integrate.
+
+%\CR{Can you elaborate on if/how this would integrate with an existing language implementation?
+%  I've always interpreted everything as that you start with some `objectlanguage.jar' file of an existing language (given that you use reflection), and you get out a fresh Gradle project that uses that `objectlanguage.jar' for parsing and AST construction, but since you generate JastAdd source code, you can't really (to my understanding) add the `type()' attribute etc.
 to the existing implementation, since JastAdd needs to run over the source jrag files.
-  Going by your overview figure, I now wonder if the generated file also incorporates all of the source code of the original
-  project?  Or do you reconstruct the AST specification via reflection somehow and simply mirror it in the generated
-  project (which would lead to two different AST representations, but with a straightforward mapping)?}
-
-\subsubsection{Type checking aspect}\label{typecheckingaspect}
-The type checking aspect contains the declaration of the \lstinline{type()} attribute for all AST nodes, as well as a default implementation.
-This default implementation immediately produces a type error, to cause an error when the type of an expression not defined in the typing rules is demanded.
-
-\CR{The details of this section (esp.\@ the bit below) might fit better into Section~\ref{typingrulecompiler}?}
-\CR{What is ``type matching''?  I am not familiar with this term, and you haven't introduced it before.}
-It also contains the implementation of the type matching method, which can be seen in figure \ref{typematches}.
-It is defined for all \lstinline{Type} nodes, and checks whether the type's class is equal to or a subclass of the parameter type.
-\CR{So this is not an equivalence relation, but a partial order?  What does that mean wrt how it is used by the implementation of the type checker?}
-
-\begin{figure}[h]
-  \begin{lstlisting}[language=jrag]
-syn boolean Type.matches(Type t) {
-  return t.getClass().isAssignableFrom(getClass());
-}
-  \end{lstlisting}
-  \caption{Definition of the type matching method}
-  \label{typematches}
-\end{figure}
+%  Going by your overview figure, I now wonder if the generated file also incorporates all of the source code of the original project?
+Or do you reconstruct the AST specification via reflection somehow and simply mirror it in the generated project (which would lead to two different AST representations, but with a straightforward mapping)?}
 
@@
 
-%\subsection{Typing rule compiler}\label{typingrulecompiler}
+%\subsection{Typing rule compiler}\label{sectionimplementation}
 
@@
 
-\section{Implementation}\label{typingrulecompiler} % if your work involved building an artefact/implementation, give the details here.
+\section{Implementation}\label{sectionimplementation} % if your work involved building an artefact/implementation, give the details here.
 Note, that this should not, as a rule, be a chronological description of your efforts, but a view of the result.
@@
 The type checker generation is based on a single-responsibility principle, where each node of the abstract syntax tree (of \emph{which} of the three (or four?) ASTs?) is responsible for generating the code it corresponds to \CR{Not sure what you mean}.
-Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of figure \ref{trastspec}, where the \verb|RuleSet| is the root node.
+Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
 From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.
@@
 \subsection{Aspect generation}
-The \verb|RuleSet| node is responsible for generating the overall structure of the output file, a JastAdd aspect called \verb|TypingRules|.
-It sorts its \verb|Rule| children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation.
+The \lstinline{RuleSet} node is responsible for generating the overall structure of the output file, a JastAdd aspect called \lstinline{TypingRules}.
+It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation.
 
@@
 \subsection{Premises/control flow}
-The rule's premises are translated into an if-statement, which checks all types using a method \verb|matches| added to all Type nodes.
-
-The current implementation of \verb|Type.matches(Type t)| checks that the Type it is called on is either of the same class or a subtype of the type \verb|t|.
+The rule's premises are translated into an if-statement, which checks all types using a method \lstinline{matches} added to all Type nodes.
+
+The current implementation of \lstinline{Type.matches(Type t)} checks that the Type it is called on is either of the same class or a subtype of the type \lstinline{t}.
 \CR{is ``matches'' a good name for something that is not symmetric?}
@@
 How about the other way around?}
-It is also possible to override the \verb|matches| implementation for a specific Type, by adding a definition to a JastAdd aspect included in the output type checker.
+It is also possible to override the \lstinline{matches} implementation for a specific Type, by adding a definition to a JastAdd aspect included in the output type checker.
 \CR{This sounds like this is a way to support ``ad-hoc type polymorphism''?}
@@
 
-\CR{Overall, I am not convinced that this strategy for presenting the type checking algorithm
-  is very effective (though it might be better for someone who is reading the source code at the same time).
-  I would propose to merge 3.2.2, 3.2.3, 3.2.4, 3.2.5, and to split them into (a) the type checking algorithm
-  (while ignoring the code generation bit), and (b) the code generation aspect.  I.e., for (a), pretend that you are
-  interpreting the typing rules instead of compiling them, and please try to present an algorithm that captures
-  the check.  The algorithm can be in plain English with bullet points / numbered sub-steps, or in some pseudo-formal
-  notation (``foreach premise $p$ in rule $r$ ...''), as long as it is clear.}
+\CR{Overall, I am not convinced that this strategy for presenting the type checking algorithm is very effective (though it might be better for someone who is reading the source code at the same time).
+  I would propose to merge 3.2.2, 3.2.3, 3.2.4, 3.2.5, and to split them into (a) the type checking algorithm (while ignoring the code generation bit), and (b) the code generation aspect.
+  I.e., for (a), pretend that you are interpreting the typing rules instead of compiling them, and please try to present an algorithm that captures
+  the check.
+  The algorithm can be in plain English with bullet points / numbered sub-steps, or in some pseudo-formal notation (``foreach premise $p$ in rule $r$ ...''), as long as it is clear.}
+
+\subsection{Type-checking algorithm}\label{typecheckingalgorithm}
+
+For each rule:
+\begin{enumerate}
+  \item For each parameter in the conclusion:
+    \begin{enumerate}
+      \item Declare a variable by the same name, initialised by fetching the child of the same index.
+    \end{enumerate}
+  \item For each type variable used in the rule:
+    \begin{enumerate}
+      \item Initialise it to null
+      \item Iterate through each premise it appears on the right-hand side of:
+        \begin{enumerate}
+          \item If the type variable is set to null, evaluate the type of the left-hand side and assign it to the type variable.
+          \item If the type variable is not set to null, evaluate the type of our left-hand side and check if it is the same type or a subtype to the type variable. If it is not, throw a type error.
+        \end{enumerate}
+    \end{enumerate}
+  \item Check whether each of the premises is true:
+    \begin{enumerate}
+      \item Evaluate the type of the left-hand side of the premise and check if it is the same type or a subtype to the right-hand side.
+    \end{enumerate}
+    If all true: Return the right-hand side of the premise
+  \item Throw a type error if we reach this point
+\end{enumerate}
+
+\begin{figure}[h]
+%foreach parameter in rule.conclusion.parameters:
+%  declare var "parameer.name" = getChild(parameter.index)
+\begin{lstlisting}[]
+foreach typevar in rule.typevars
+  declare typevar = null
+  foreach premise in rule.premises where premise.rhs == typevar:
+    if typevar == null:
+      typevar = premise.lhs.type()
+    else:
+      if typevar != premise.lhs.type():
+        throw typeerror("Type variable mismatch")
+
+if(premise1.lhs.type() == premise1.rhs &&
+   premise2.lhs.type() == premise2.rhs &&
+   ...
+  ):
+    return conclusion.rhs
+throw typeerror("Typechecking failed")
+\end{lstlisting}
+  \caption{Pseudocode for the implementation of the type() attribute}
+  \label{boolsast}
+\end{figure}
+\subsection{Code generation}
+
+\subsection{Type checking aspect}\label{typecheckingaspect}
+The type checking aspect contains the declaration of the \lstinline{type()} attribute for all AST nodes, as well as a default implementation.
+This default implementation immediately produces a type error, to cause an error when the type of an expression not defined in the typing rules is demanded.
+
+%\CR{The details of this section (esp.\@ the bit below) might fit better into Section~\ref{sectionimplementation}?}
+%\CR{What is ``type matching''?  I am not familiar with this term, and you haven't introduced it before.}
+It also contains the implementation of the type matching method, which checks whether a type is equal to or a subtype of another type, which can be seen in Figure \ref{typematches}.
+It is defined for all \lstinline{Type} nodes, and checks whether the type's class is equal to or a subclass of the parameter type.
+\CR{So this is not an equivalence relation, but a partial order?
+What does that mean wrt how it is used by the implementation of the type checker?} \reviewquestion{I think it would be easier to change this poorly thought out line of code than to fully evaluate and address the implications in the report :) Is that okay?}
+
+\begin{figure}[h]
+\begin{lstlisting}[language=jrag]
+syn boolean Type.matches(Type t) {
+  return t.getClass().isAssignableFrom(getClass());
+}
+\end{lstlisting}
+  \caption{Definition of the type matching method}
+  \label{typematches}
+\end{figure}
 
@@
 This method was chosen solely for its simplicity of implementation, rather than its functionality.
-
-It provides a stack trace allowing you \CR{don't address reader directly} to see where in the type checking code the error occurs (such as what typing rule fails), but doesn't currently provide any information about what section of the input program causes the error.
-This tends to result in errors that are more useful for debugging the typing rule implementation, rather than providing useful information to the user of the type checker, where there is much room for improvements.
-
-\CR{This section is needlessly apologetic-- error handling was not the focus of your work.  You can add a forward reference to 6.1.2, though.}
-
-%Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of figure \ref{trastspec}, where the \verb|RuleSet| is the root node.
-%The \verb|RuleSet| node consists of a list of \verb|Rule| nodes, and begins by sorting each rule by which object language node it describes.
+Possible improvements to it are discussed in Section \ref{improvederrorhandling}.
+
+It provides a stack trace with information on see where in the type checking code the error occurs (such as what typing rule fails), but doesn't currently provide any information about what section of the input program causes the error.
+%This tends to result in errors that are more useful for debugging the typing rule implementation, rather than providing useful information to the user of the type checker, where there is much room for improvements.
+
+%\CR{This section is needlessly apologetic-- error handling was not the focus of your work.  You can add a forward reference to 6.1.2, though.}
+
+%Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
+%The \lstinline{RuleSet} node consists of a list of \lstinline{Rule} nodes, and begins by sorting each rule by which object language node it describes.
 %
-%The \verb|RuleSet| node then begins by generating the outline of a JastAdd aspect \verb|TypingRules|, and the definition for the method \verb|.type()| for each object language node.
-%It then calls on the \verb|Rule| node to fill in the implementation of the method.
+%The \lstinline{RuleSet} node then begins by generating the outline of a JastAdd aspect \lstinline{TypingRules}, and the definition for the method \lstinline{.type()} for each object language node.
+%It then calls on the \lstinline{Rule} node to fill in the implementation of the method.
 %
-%A \verb|Rule| consists of a name, a conclusion, and a possibly empty list of premises that need to be true for the conclusion to be returned.
+%A \lstinline{Rule} consists of a name, a conclusion, and a possibly empty list of premises that need to be true for the conclusion to be returned.
 %In the simplest case of an empty list of premises, the entire content of the rule is generated by the conclusion.
@@
 %Firstly, any variables or type variables used in the rule need to be declared, followed by a check to ensure each use of a type variable represents a node with the same type.
-%Then it generates an if-statement corresponding to our premises, by letting each \verb|Premise| generate a boolean expression, all of which get joined by the AND operator.
+%Then it generates an if-statement corresponding to our premises, by letting each \lstinline{Premise} generate a boolean expression, all of which get joined by the AND operator.
 %
-%Both \verb|Premise|s and \verb|Conclusion|s are represented by the same abstract class \verb|Formula|, which currently contains a single implementation \verb|HasType|, consisting of an expression and a type.
+%Both \lstinline{Premise}s and \lstinline{Conclusion}s are represented by the same abstract class \lstinline{Formula}, which currently contains a single implementation \lstinline{HasType}, consisting of an expression and a type.
 %Depending on whether it is used as a premise or conclusion, it can either represent that the expression must have the specified type or that the expression should be given the specified type.
@@
 %\subsection{Type checker design}
-%The provided typing rules compile into \verb|type()| methods, which return a value from the Type interface.
+%The provided typing rules compile into \lstinline{type()} methods, which return a value from the Type interface.
 %For the most simple typing rules, this method's implementation will consist of a single return statement.
@@
 \chapter{Evaluation} % is the part where you present the finds. Depending on the area this part contains a subset or all of the following:
-\CR{The section has two parts, so the intro paragraph here can directly discuss their purpose.}
+In the first section of this chapter, we present typing rule definitions for a few simple type systems, alongside the resulting code produced as an output when they are input to our compiler.
+
+In Section \ref{sectionresearchquestions}, we evaluate the overall result of our study by answering our research questions.
+%\CR{The section has two parts, so the intro paragraph here can directly discuss their purpose.}
 %\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach.
@@
 
-\begin{figure}[h]
+\begin{figure}[]
 \begin{lstlisting}[]
@@
 \end{figure}
-\begin{figure}[h]
+\begin{figure}[]
 \begin{lstlisting}[]
@@
 
-\begin{figure}[h]
+\begin{figure}[]
 \begin{lstlisting}[language=jrag]
@@
 \end{lstlisting}
-\caption{Generated typing rules for the Bools language (reformatted for readability)
-\CR{}}
+\caption{Generated typing rules for the Bools language (reformatted for readability)}
   \label{boolstrgen}
 \end{figure}
-A simple example of a language is provided in figure \ref{boolsast}, consisting of three values, \verb|True|, \verb|False| and \verb|Zero|; and the boolean operator \verb|Or| with two parameters, Left and Right.
-In the AST specification, there are no limitations on what terms can be provided as parameters to the \verb|Or| operator, allowing illogical constructs such as \verb|Or(True, Zero)|.
-
-To prevent cases like these, we define typing rules via the specification in figure \ref{boolstr}.
-The typing rules for \verb|True|, \verb|False| and \verb|Zero| contain no premises and define the type as \verb|Bool| and \verb|Int| as appropriate.
-The typing rule for the \verb|Or| operator, which takes two parameters \verb|left| and \verb|right|, contains two premises, specifying that the \verb|Or| operator has the type \verb|Bool| only if both the \verb|left| and \verb|right| terms also have the type \verb|Bool|.
-
-The typing rules from figure \ref{boolstr} generate the type definitions in \ref{boolstrgen}.
-The simple rule definitions for the values compile into equally simple implementations, consisting of a single return statement of a newly constructed \verb|Type| object.
-For the \verb|Or| node, the code becomes slightly more complex.
+A simple example of a language is provided in Figure \ref{boolsast}, consisting of three values, \lstinline{True}, \lstinline{False} and \lstinline{Zero}; and the boolean operator \lstinline{Or} with two parameters, Left and Right.
+In the AST specification, there are no limitations on what terms can be provided as parameters to the \lstinline{Or} operator, allowing illogical constructs such as \lstinline{Or(True, Zero)}.
+
+To prevent cases like these, we define typing rules via the specification in Figure \ref{boolstr}.
+The typing rules for \lstinline{True}, \lstinline{False} and \lstinline{Zero} contain no premises and define the type as \lstinline{Bool} and \lstinline{Int} as appropriate.
+The typing rule for the \lstinline{Or} operator, which takes two parameters \lstinline{left} and \lstinline{right}, contains two premises, specifying that the \lstinline{Or} operator has the type \lstinline{Bool} only if both the \lstinline{left} and \lstinline{right} terms also have the type \lstinline{Bool}.
+
+The typing rules from Figure \ref{boolstr} generate the type definitions in Figure \ref{boolstrgen}.
+The simple rule definitions for the values compile into equally simple implementations, consisting of a single return statement of a newly constructed \lstinline{Type} object.
+For the \lstinline{Or} node, the code becomes slightly more complex.
 \CR{The next bit would probably be easier if you just show the code.}
@@
 This is followed by an if-statement, recursively evaluating the children's types and comparing the result the type demanded in the premise.
-If the condition is true, a \verb|Bool| type is returned.
+If the condition is true, a \lstinline{Bool} type is returned.
 If not, a runtime exception is thrown to convey a type error.
@@
 \end{lstlisting}
-  \caption{Extension for the Bools AST in figure \ref{boolsast} \CR{Consider giving this language a different name, for clarity}}
+  \caption{Extension for the Bools AST in Figure \ref{boolsast} \CR{Consider giving this language a different name, for clarity}}
   \label{ifelseast}
@@
 \end{lstlisting}
-  \caption{Extension for the Bools typing rules in figure \ref{boolstr} with an if-else term}
+  \caption{Extension for the Bools typing rules in Figure \ref{boolstr} with an if-else term}
   \label{ifelsetr}
 \end{figure}
-\begin{figure}[h]
+\begin{figure}[]
 \begin{lstlisting}[language=jrag]
@@
 \end{lstlisting}
-\caption{Generated typing rules for the if-else term, extending figure \ref{boolstrgen} (reformatted for readability)
+\caption{Generated typing rules for the if-else term, extending Figure \ref{boolstrgen} (reformatted for readability)
   \CR{The components of this function would be great to discuss in Section 3.
-Here in Section 4, you may want to forward refernece the discussion about redundancy/possible optimisations (Section 4.2.2) and very concretely use this code and explain the changes you have in mind (for someone else to do in future work).}
+Here in Section 4, you may want to forward reference the discussion about redundancy/possible optimisations (Section 4.2.2) and very concretely use this code and explain the changes you have in mind (for someone else to do in future work).}
 }
@@
 \CR{``Can extend'' sounds a bit arbitrary.  Can you motivate with Pierce (even if not exactly)?}
-As shown in the AST extension in figure \ref{ifelseast}, it consists of three terms: a conditional, a term to be evaluated if the condition is true and a condition to evaluate if false.
-A typing rule of this if-else term can be seen in figure \ref{ifelsetr}.
-The rule has three premises, the first of which declares the conditional must have the type \verb|Bool|.
-The next two declare that the two other children must be of the type \verb|a|, with the lowercase \CR{grammar} indicating a type variable.
-This type variable \verb|a| also subsequently appears in the resulting type of the if-else term.
-
-\CR{``Shown in Figure~\ref{reference}'' is the current preferred form in our field}
-The code for this rule is shown in figure \ref{ifelsetrgen}, and begins with three parameter declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
+As shown in the AST extension in Figure \ref{ifelseast}, it consists of three terms: a conditional, a term to be evaluated if the condition is true and a condition to evaluate if false.
+A typing rule of this if-else term can be seen in Figure \ref{ifelsetr}.
+The rule has three premises, the first of which declares the conditional must have the type \lstinline{Bool}.
+The next two declare that the two other children must be of the type \lstinline{a}, with the lowercase \CR{grammar} indicating a type variable.
+This type variable \lstinline{a} also subsequently appears in the resulting type of the if-else term.
+
+%\CR{``Shown in Figure~\ref{reference}'' is the current preferred form in our field}
+The code for this rule is shown in Figure \ref{ifelsetrgen}, and begins with three parameter declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
 For each invocation of the type variable, a code section is generated, which checks if the type variable is still null, and if so assigns it to the type of the left-hand side term, or if it has been assigned checks whether the types match each other.
@@
 
-\section{Research questions}
-
-\subsection{\rqone}
+\section{Research questions}\label{sectionresearchquestions}
+
+\subsection*{\rqone}
 Our project has shown that generating type checkers using RAGs is a viable solution for a number of different language constructs.
@@
 To evaluate the completeness of our implementation, we refer to Pierce\cite{Pierce} as a reference work.
-The typing rules are introduced in chapter 8, starting with the simple examples shown in section \ref{sectionlanguage}.
+The typing rules are introduced in chapter 8, starting with the simple examples shown in Section \ref{sectionlanguage}.
 \CR{Please be clearer on which examples from Pierce you implemented and where support stops.}
@@
 
-\subsection{\rqtwo}
-We have approached parsing typing rules into a JastAdd attribute \verb|type()|,
-\CR{parsing is the syntactic analysis of the input specification, you have done more than just parsing.}
-which either returns a \verb|Type| value, or throws a runtime exception if the expression is invalidly typed.
+\subsection*{\rqtwo}
+We have approached translating typing rules into a JastAdd attribute \lstinline{type()},
+%\CR{parsing is the syntactic analysis of the input specification, you have done more than just parsing.}
+which either returns a \lstinline{Type} value, or throws a runtime exception if the expression is invalidly typed.
 
@@
 %This would perhaps make it easier to implement informative type errors, as you could report precisely which premise is failing.
-If paired with the current error handling of throwing an exception to convey type errors, it would make it harder to implement having multiple valid rules providing the type of an expression, as the first failure would seize the evaluation of the type.
-\CR{What do you mean by ``seize''?}
+If paired with the current error handling of throwing an exception to convey type errors, it would make it harder to implement having multiple valid rules providing the type of an expression, as the first failure would throw an exception, stopping the evaluation of any further typing rules which may have been successful.
+%\CR{What do you mean by ``seize''?}
 An alternative error handling would suit it better, one able to collect multiple errors without seizing evaluation, perhaps merely adding all the errors to a list and checking whether it is empty before returning a type.
@@
 The variables in the typing rules are compiled into regular Java variables containing the AST node.
-Notably, the declarations of these variables type them merely as \verb|ASTNode|.
+Notably, the declarations of these variables type them merely as \lstinline{ASTNode}.
 As we currently only utilise the \lstinline{type()} attribute, which is declared for all nodes, this lost information has little impact.
@@
 This is followed by a check for each point of use in the typing rule, to see whether the type variable is still unassigned and if so assign it.
-This approach leaves much room for improvement \CR{no need to say ``much room for improvement'', just mention the specific disadvantages.} and optimisation, by finding the first use of each type variable during the rule compilation, and generating a single declaration statement initialising it to the type of the corresponding expression.
+This approach could be improved by finding the first use of each type variable during the rule compilation, and generating a single declaration statement initialising it to the type of the corresponding expression.
+%\CR{no need to say ``much room for improvement'', just mention the specific disadvantages.}
 \CR{I am missing a discussion of type variables and unification as we used them in EDAP15 for Hindley/Milner.}
@@
 A cleaner solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
-Instead of reporting the reason any individual rule failed, the error might instead display all the rules it tried to apply, perhaps with a list of non-holding premises for each one.
-\CR{This is a good idea; you can even sort the rules by closest matches and only show the top ones if there are many.}
+Instead of reporting the reason any individual rule failed, the error might instead display either all or only the best matching of the rules it tried to apply, perhaps with a list of non-holding premises for each one.
+%\CR{This is a good idea; you can even sort the rules by closest matches and only show the top ones if there are many.}
 
@@
 
-\subsection{\rqthree}
+\subsection*{\rqthree}
 %Splitting code between multiple JastAdd components (and java) can get messy
-Writing idiomatic JastAdd components often requires splitting your \CR{Please don't address the reader directly}  code into many different places.
+Writing idiomatic JastAdd components often requires splitting code into many different places.
 We attempted to keep each node in the tree responsible for generating the code it corresponds to \CR{What does that mean?
-You need to provide a bit more background}, which avoids parent nodes having to check their children's structure with \verb|isInstanceOf|, but also means even generating a simple rule can require a dozen calls between different nodes.
+You need to provide a bit more background}, which avoids parent nodes having to check their children's structure with \lstinline{isInstanceOf}, but also means even generating a simple rule can require a dozen calls between different nodes.
 There is a tradeoff to be made between the single responsibility of each component making it easy to verify its correctness, but the complicated interdependence making it hard to grasp the full picture.
@@
 
-Not being able to perform these error checks at runtime
-\CR{at runtime of which component?}
-means our translation will emit nonfunctional typing definitions if a mistake is written in the typing rules.
+Not being able to perform these error checks during the generation of our type checker means our translation will emit nonfunctional typing definitions if a mistake is written in the typing rules.
 These errors will then go unreported until the output project is compiled or run, at which point the user will have to attempt to trace these errors back to the typing rule definition, to find the actual cause.
-\CR{NB: this is also somewhat true of JastAdd's own compilation approach, so there might be double indirection errors ...}
+As JastAdd itself also adds a layer of indirection, by taking the attributes defined in RAG modules and compiling them into generated Java class files, this can be particularly tricky.
+%\CR{NB: this is also somewhat true of JastAdd's own compilation approach, so there might be double indirection errors ...}
 
@@
   about the relevance of syntax-directedness woudl be valuable.)}
+  % Focus on syntax directedness. Exceptions: if t has type t, it also has type t; nested rules if(xcondition): X, if(ycondition): Y
 
@@
 %\section{Development learnings}
+
+\chapter{Related work} % Maybe move to discussion?
+% Describe related work (literature study).
+Besides listing other work in the area, mention how is it related or relevant to your work.
+The tradition in some research area is to place this part at the end of the report (check with your supervisor).
+Pacak et al. implemented a compiler for generating incremental type checkers using declarative logic programming language Datalog\cite{Pacak}.
+While the goal is similar to ours, their project defines a formalism for expressing typing rules within the Datalog language, whereas ours defines a language which aims to keep as close as possible to the traditional inference rules used for typing rules, to minimise the difficulty of translation.
+%\CR{I'd start with this difference-- the translation is targeting two rather different formalisms, which is all you need to justify the novelty.}
+
+%they take a different approach, by converting typing rules into the  \CR{second half of this sentence confuses me}.
+
+Their work is ends up largely focused on expressing typing rules within the limits of the Datalog language, which can only compute finite relations.
+We have a significantly different compilation target, JastAdd, which allows us to use reference attribute grammars to evaluate the typing relations, utilising a built-in AST structure.
+Whereas Datalog is a fully declarative language, with JastAdd we express typing relations as declared attributes on the AST, with their implementations written in imperative Java code.
+%\CR{Datalog is fully declarative, whereas JastAdd uses declarative rules that connect snippets of imperative Java code.  Is that what you meant?}
+%\CR{Some more thoughts below to help make the distinction clearer (all optional):}
+%\CR{Datalog has no direct notion of ASTs or other tree structures / algebraic terms, so trees must be encoded into relations.}
+%\CR{Datalog has a simple, built-in notion of unification that can handle cases like testing sibling relations: $\textsf{Sibling}(x, y) :- Parent(p, x), \textsf{Parent}(p, y), x \ne y.$.
+%    This is something you deal with later in JastAdd.}
+% TODO: Related work section after chapter 5, one or two pages. Various papers from CR comments and goal document.
+% TODO: Discuss Pacak in introduction
+
+\reviewquestion{Circular attributes?}
+\CR{Circular attributes: Recursion in typing rules: Datalog is built with support for ``infinite'' recursion (since it operates on finite domains,
+  it can use a special bottom-up evaluation scheme to ensure termination.)
+  If you need this in RAGs, it gets more complicated, but you'd need an example to explain this in the introduction-- since you don't directly address that in this work,
+  I wouldn't talk about it too much here (you could also skip it entirely here).  Minimum: talk about it in the discussion later.}
+
 \chapter{Conclusions} % should summarize your findings and possible improvements or recommendations.
 Our experiments have shown that reference attribute grammars are a flexible method for implementing type checker generators.
-Implementing the first top-to-bottom functioning product \CR{end-to-end prototype is what I would call it} was challenging and took quite a lot of work, but extending it with additional features such as type variables turned out to be a surprisingly painless experience.
+Implementing the first end-to-end prototype was challenging and took quite a lot of work, but extending it with additional features such as type variables turned out to be a surprisingly painless experience.
 
@@
 
-\subsection{Improved error handling}
+\subsection{Improved error handling}\label{improvederrorhandling}
 The current exception-based error handling is rather primitive and leaves much to be desired.
@@
 An improvement demanding more extensive changes would be collecting multiple errors, instead of seizing the type checker at the first discovered error.
-This could be implemented by adding another type class to be returned when an error occurs, representing the "Any" type.
-Any \verb|matches()| call on this type would return true, to avoid one error cascading into many, but still allowing other unrelated errors to be discovered.
+This could be implemented by adding another type class to be returned when an error occurs, representing the `Any' type.
+Any \lstinline{matches()} call on this type would return true, to avoid one error cascading into many, but still allowing other unrelated errors to be discovered.
 This implementation would also need new functionality to collect the error messages, perhaps by splitting the type resolution and type errors into separate attributes for each node, where the type errors propagate to a program-wide collection.
+
+It may be possible to create this by wrapping around the current implementation with two additional attributes.
+One would be a total function, that returns the type evaluated by the current \lstinline{type()} method if it is successful or returns the Any type in the case of an exception.
+The second attribute would contribute an error to the error collection only in the case of an exception.
+The only other changes required would be to use the total wrapped method within the generated attributes.
 
@@
 Additionally, an extra layer of scoping would have to be introduced, to prevent variable naming collisions between the different rules.
-This could be solved simply by wrapping each rule within brackets, creating a separate scope for the evaluation of each rule.
-\CR{This is about name clashes in the generated code?  You may want to consider automatic relabelling here anyway (``this'', ``new'' etc. as type variable names are likely going to cause issues) }
+This could be solved simply by wrapping each rule within brackets, creating a separate scope for the evaluation of each rule, or relabelling the variables to prevent collisions.
+%\CR{This is about name clashes in the generated code?  You may want to consider automatic relabelling here anyway (``this'', ``new'' etc. as type variable names are likely going to cause issues) }
 
 %add(int, int): int, add(double, double): double)
-\subsection{Better support for numbers, strings, etc.}
-
-\CR{Not sure that there is anything needed here}
+%\subsection{Better support for numbers, strings, etc.}
+
+%\CR{Not sure that there is anything needed here}
 

### 2024-01-14 MSc.tex
Further iteration

@@
   Static type checkers are a popular method for early detection of bugs in computer programs, by ensuring the correct data types are used throughout the program.
+  % RAGs may not be the most convenient format for typing rules
   One approach for implementing type checkers are Reference Attribute Grammars (RAGs), a formalism for specifying the static semantics of programming languages by declaring attributes on an abstract tree-representation of the program, and is used by the JastAdd meta-compilation system.
 
-  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from syntax-directed typing rules.
+  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from a subset of possible typing rules.
   This aims to provide integration with existing compilers implemented with JastAdd, as well as a playground for type systems usable as an educational tool.
 
-  We find RAGs to be a flexible model for syntax-directed typing rules, and our software was able to support simple values, nodes with premises placing demands on their children, and those with rules requiring type variables.
+  % 'Syntax-directed' needs explaining if used, "a subset known as syntax-directed"
+  We find RAGs to be a flexible model for this subset of typing rules,
+  % This is too specific, works with small languages from research literature
+  and our software was able to support simple values, nodes with premises placing demands on their children, and those with rules requiring type variables.
   It proved difficult to provide compile-time verification of the typing rules, and further work is required to extend the compiler to support type environments, a necessity for most real-world type systems.
@@
 In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of Reference Attribute Grammars (RAGs).
-JastAdd is an interesting target because it has already been used to implement compilers of several languages, to which our typing rule implementations could be integrated.
-JastAdd is also used extensively in courses at the department of computer science at LTH, where our project could serve as an educational tool, providing students an interactive way to define and test simple type systems.
+
+Pacak et al. have previously implemented a compiler for generating incremental type checkers using declarative logic programming language Datalog\cite{Pacak}.
+It focused on defining a way of expressing typing rules as a series of relations solvable by the Datalog compiler.
+Our use of RAGs instead allows us to build our analysis directly onto a representation of the program's structure.
+JastAdd is also an interesting target because it has already been used to implement compilers of several languages, to which our typing rule implementations could potentially be integrated.
+It is also used extensively in courses within the Department of Computer Science at LTH, where our project could serve as an educational tool, providing students an interactive way to define and test simple type systems.
 
@@
 To be able to translate them into JastAdd code, we will need to define a limited subset of syntax-directed typing relations.
-We identify a number of different kinds of typing rules, used in previous work in the field\ref{Pierce}, which share a common set of features for us to implement.
+We identify a number of different kinds of typing rules, used in previous work in the field\cite{Pierce}, which share a common set of features for us to implement.
 We then devise a strategy to translate these rules into snippets of iterative Java code, which can be evaluated by JastAdd.
@@
 
-\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
-  We evaluate the approach against Foo, and demonstrate Bar.
-  Our case studies show Quux.
-  Overall, this initial exploration shows \ldots''}
+We define a custom parsable language for typing rules adapted from natural semantics, and introduce a strategy to compile it into RAG attributes.
+Our case studies show that simple type systems can be systematically compiled into attributes on the AST, and the tree syntax provides a natural approach for recursive evaluation.
+Overall, this initial exploration shows that RAGs are a flexible approach for implementing type checkers for syntax-directed typing rules, though there are difficulties when implementing rule verification at compiler runtime.
+%\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
+%  We evaluate the approach against Foo, and demonstrate Bar.
+%  Our case studies show Quux.
+%  Overall, this initial exploration shows \ldots''}
 
@@
 %\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
-\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinking ``I know why I want to read this''}
+%\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinking ``I know why I want to read this''}
 This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
@@
 %    analysing programs.}
-Section \ref{typetheory} details type systems,
-%\CR{Too much detail in the following:}
-by explaining each of the component pieces, followed by a description of their use within static type checking.
-%\CR{You can use \texttt{\textbackslash label\{labelname\}} and \texttt{\textbackslash ref\{labelname\}} to reference (sub)secttions, that's usually clearer than ``next section''.}
+Section \ref{typetheory} details type systems, by explaining each of the component pieces, followed by a description of their use within static type checking.
 
@@
 %\CR{what is ``this information'', concretely?}
-A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if-else node, with named branches pointing towards the condition and the two values.
+A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if node, with named branches pointing towards the condition and the two values.
 %\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
@@
 
-\CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}
-\begin{figure}[h]
-\begin{lstlisting}[]
-Expr   -> Expr "+" Term | Term
-Term   -> Term "*" Factor | Factor
-Factor -> Int | "(" Expr ")"
-\end{lstlisting}
-  \caption{\reviewquestion{One like this? I'm unsure how to tie it in effectively.}}
-  \label{typesyntaxspec}
-\end{figure}
+%\CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}
+%\begin{figure}[h]
+%\begin{lstlisting}[]
+%Expr   -> Expr "+" Term | Term
+%Term   -> Term "*" Factor | Factor
+%Factor -> Int | "(" Expr ")"
+%\end{lstlisting}
+%  \caption{\reviewquestion{One like this? I'm unsure how to tie it in effectively.}}
+%  \label{typesyntaxspec}
+%\end{figure}
 
@@
 Data types can be considered a categorisation of values which share common properties, such as supporting numerical operations or accessing a certain element in a list of characters.
+While expressions may be arbitrarily complex, nested structures, by assigning it a type, we can perform useful analysis on it, without having to evaluate its value.
+%\CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
+%  For each concrete (and possibly very complex) operator on the left-hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
+%  This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
+%  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractions over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
+%}
+
 %several different types of data in one program, such as numbers or text.
@@
 
-\CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
-  For each concrete (and possibly very complex) operator on the left-hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
-  This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
-  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractions over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
-}
-
 \subsection{Type systems}
@@
 
-\subsubsection{Type syntax}
-A type syntax, or set of types, is a definition of the valid types within a language.
-\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
-\reviewquestion{I believe Pierce simply refers to it as the "set of types".
+\subsubsection{Type syntax}\label{typesyntax}
+We're using the term `type syntax' to mean a definition of the valid types within a language, often referred to simply as the set of types.
+%\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
+%\reviewquestion{I believe Pierce simply refers to it as the "set of types".
 I seem to have got the term type syntax from EDAP15 \url{https://fileadmin.cs.lth.se/cs/Education/EDAP15/2022/web/skills/type-system.html}.
@@
 
-\subsection{Non-Terminal Attributes}
-\cite{higherorderattributegrammars}
-\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are
-not a necessary feature for RAGs.  Please introduce them in a separate section: you can very directly connect them to the idea of ``type syntax''.}
+%\subsection{Non-Terminal Attributes}
+%Our project also uses not\cite{higherorderattributegrammars}
+%\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are not a necessary feature for RAGs.
+%Please introduce them in a separate section: you can very directly connect them to the idea of ``type syntax''.}
 
@@
 \begin{figure}[]
-\begin{lstlisting}[language=java]
-ruleset = rule_list
+\begin{lstlisting}[]
+ruleset = rule_list;
 
@@
     rule
-  | rule_list rule
+  | rule_list rule;
 
 rule =
-    '[' upperid ']' premises_list '---+' formula
-  | '[' upperid ']' formula
+    '[' uppercase_id ']' premises_list '---+' formula
+  | '[' uppercase_id ']' formula;
 
@@
     formula
-  | premises_list ',' formula
-
-formula = term ':' tyterm
+  | premises_list ',' formula;
+
+formula = term ':' tyterm;
 
 term =
-    upperid '(' term_list ')'
-  | upperid '(' ')'
-  | upperid
-  | lowerid
+    uppercase_id '(' term_list ')'
+  | uppercase_id '(' ')'
+  | uppercase_id
+  | lowercase_id;
 
 tyterm =
-    upperid                                              // TyVal
-  | lowerid                                              // TyVar
+    uppercase_id
+  | lowercase_id;
 
@@
     term
-  | term_list ',' term
+  | term_list ',' term;
+
+uppercase_id = "[A-Z][A-Za-z0-9-]";
+uppercase_id = "[a-z][A-Za-z0-9-]";
 \end{lstlisting}
-  \caption{Syntax definition for the typing rule language}
+  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to BNF.}
   \label{trastspec}
 \end{figure}
-\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
-  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
-  This would be before the current 3.1, then.
-  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
-  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
-  of the object language as well as of the types.}
+%\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
+%  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
+%  This would be before the current 3.1, then.
+%  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
+%  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
+%  of the object language as well as of the types.}
 
@@
   \begin{tikzpicture}[
-      every node/.style = {draw=black, align=center},
+      every node/.style = {align=center, line width=0.8pt},
       every path/.style = {draw, -latex},
       artifact/.style={draw=black},
-      process/.style={ellipse}
+      userinput/.style={draw=red},
+      generated/.style={draw=blue},
+      process/.style={ellipse, draw=black}
       %greennode/.style={shape=circle, draw=green, line width=2},
@@
     ]
-    \node (typerules)                               {Typing rules\\definition (\ref{typingruledefinition})};
+    \node (typerules) [userinput]                    {Typing rules\\definition (\ref{typingruledefinition})};
     \node (trparsing) [right=of typerules, process] {Scan,\\parse,\\build};
-    \node (trparser)  [above=of trparsing]          {Typing rule parser\\(\ref{typingrulesparser})};
-    \node (trast)     [right=of trparsing]          {Typing rules AST\\(\ref{typingrulesast})};
+    \node (trparser)  [above=of trparsing, artifact]          {Typing rule parser\\(\ref{typingrulesparser})};
+    \node (trast)     [right=of trparsing, generated]          {Typing rules AST\\(\ref{typingrulesast})};
 
@@
     \node (typecheckgen)       [below=of trast, process] {Generate\\type checker};
-    \node (typingrulecompiler) [right=of trast]          {Typing rule\\compiler (\ref{sectionimplementation})};
-    \node (typechecker)        [below=of typecheckgen]   {Type checker \\(\ref{typecheckingalgorithm})};
-
-    \node (olast)     [below=of typerules] {AST specification\\(\ref{astdef})};
+    \node (typingrulecompiler) [right=of trast, artifact]          {Typing rule\\compiler (\ref{sectionimplementation})};
+    \node (typechecker)        [below=of typecheckgen, generated]   {Type checker \\(\ref{typecheckingalgorithm})};
+
+    \node (olast)     [below=of typerules, userinput] {AST specification\\(\ref{astdef})};
 
     \node (outputtypechecking) [below=of typechecker, process] {Perform\\typechecking};
-    \node (inputast)           [left=of outputtypechecking]    {Input program\\AST};
+    \node (inputast)           [left=of outputtypechecking, generated]    {Input program\\AST};
     \node (outputparsing)      [left=of inputast, process]     {Scan,\\parse,\\build};
-    \node (typeerrors)         [right=of outputtypechecking]   {Type errors\\(\ref{errorhandling})};
-    \node (inputprog)          [left=of outputparsing]         {Input program};
+    \node (typeerrors)         [right=of outputtypechecking, generated]   {Type errors\\(\ref{errorhandling})};
+    \node (inputprog)          [left=of outputparsing, artifact]         {Input program};
 
@@
     %};
-    \node (templateparser)        [below=of outputparsing]                               {Tree-based\\parser\\(\ref{treebasedparser})};
-    \node (templateprojectfiles)  at (inputast |- templateparser)                        {Project files\\(\ref{projectfiles})};
-    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser)              {Type checking\\aspect (\ref{typecheckingaspect})};
-    \node (templatetypesyntax)    [right=of typecheckingalgorithm]                       {Type syntax\\AST};
-    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm) (templatetypesyntax)] {};
+    \node (templateparser)        [below=of outputparsing, artifact]                               {Tree-based\\parser\\(\ref{treebasedparser})};
+    \node (templateprojectfiles)  at (inputast |- templateparser)           [artifact]           {Project files\\(\ref{projectfiles})};
+    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser) [artifact]           {Type checking\\aspect (\ref{typecheckingaspect})};
+    \node (templatetypesyntax)    [right=of typecheckingalgorithm, artifact]                       {Type syntax\\AST (\ref{typesyntax})};
+    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm) (templatetypesyntax), artifact] {};
     \node                         [above, draw=none] at (template.north)                 {Support library};
 
-    \node (userinput)   [dashed, fill=none, fit=(typerules) (olast)]                  {};
+    \node (userinput)   [dashed, fill=none, fit=(typerules) (olast), artifact]                  {};
     \node               [above, draw=none] at (userinput.north)                       {Object language\\static semantics};
@@
 
-    \matrix [below left, nodes={minimum size = 15pt, align=left}, align=left] at (current bounding box.north east) {
+    \matrix [left, nodes={minimum size = 15pt, align=left}, align=left, draw=black] at (current bounding box.north east) {
       \node [artifact, label=right:Artifact] {}; \\
+      \node [userinput, label=right:User provided] {}; \\
+      \node [generated, label=right:Generated] {}; \\
       \node [process, label=right:Process] {}; \\
@@
 The dashed arrows from represent links from the support library components that exist in the current experimental state of the project, but in a finished product would instead link to component from the object language's own compilation environment.
-  \CR{Can you mark, e.g. via colours, the parts that the user provides,
-           and the parts that will be in the generated output project?}}
+  %\CR{Can you mark, e.g. via colours, the parts that the user provides, and the parts that will be in the generated output project?}
+  }
   \label{overviewgraph}
@@
 %\subsection{Type checker generation}
-The type checker generation is based on a single-responsibility principle, where each node of the abstract syntax tree (of \emph{which} of the three (or four?) ASTs?) is responsible for generating the code it corresponds to \CR{Not sure what you mean}.
-Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
-From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.
-
-This section details the implementation of our type checker generator, on a feature by feature basis.
-
-%\subsubsection{Aspect generation}
-\subsection{Aspect generation}
-The \lstinline{RuleSet} node is responsible for generating the overall structure of the output file, a JastAdd aspect called \lstinline{TypingRules}.
-It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation.
-
-%\subsubsection{Method implementation overview}
-\subsection{Method implementation overview}
-The implementation of a typing rule can range from a single return statement to a system of declarations of variables, type variables and control flow.
-These are explained in further detail in subsequent sections, but here follows a quick overview.
-
-Firstly, the algorithm checks if the number of premises is zero, in which case it immediately generates the conclusion, consisting of a return statement.
-If there are premises, it first generates code for the declarations of variables and type variables.
-This is followed by an if-statement to test whether the premises hold true, containing a return statement.
-Should the premises not hold, a type error is produced.
-\CR{What if there are multiple rules that apply to the same AST node?  E.g., one rule that allows \ty{Int} addition and one that allows \ty{String} addition.}
-
-%\subsubsection{Parameters}
-\subsection{Parameters}
-When the typing rules concern a node with parameters, these can be named arbitrarily \CR{Who does the naming?
-Why is the ``arbitrarily'' bit important?}\CR{ Will this work if the type system specification has a type variable called ``this'' or ``null'' or ``class''?}, and a variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
-These variables are then utilised in later premises or type variable declarations.
-
-%\subsubsection{Premises/control flow}
-\subsection{Premises/control flow}
-The rule's premises are translated into an if-statement, which checks all types using a method \lstinline{matches} added to all Type nodes.
-
-The current implementation of \lstinline{Type.matches(Type t)} checks that the Type it is called on is either of the same class or a subtype of the type \lstinline{t}.
-\CR{is ``matches'' a good name for something that is not symmetric?}
-This allows the language limited sub-typing at the type checker generation step, though the pros and cons of this have not been evaluated.
-\CR{Can you give an example of a case where this is helpful, and one example where it might be troublesome?
-Let's use the ``if'' rule as an example.
-If $\ty{Number} :> \ty{Int}$ (\ty{Number} is a supertype of \ty{Int}), what type will I infer if $t_2 = \ty{Number}$ and $t_3 = \ty{Int}$?
-How about the other way around?}
-It is also possible to override the \lstinline{matches} implementation for a specific Type, by adding a definition to a JastAdd aspect included in the output type checker.
-\CR{This sounds like this is a way to support ``ad-hoc type polymorphism''?}
-
-%\subsubsection{Type variables}
-\subsection{Type variables}
-In the case of type variables, the generated code takes a dynamic approach \CR{Meaning?} to simplify the implementation.
-Each type variable that appears in the conclusion is instantiated to a null value, and the code generated by each premise first checks if the value is still null and either assigns the variable its own type or tests to make sure the existing type matches its own.
-\CR{Is there a hidden order dependency? (Cf. my earlier question about ``matches'')}
-
-\CR{Overall, I am not convinced that this strategy for presenting the type checking algorithm is very effective (though it might be better for someone who is reading the source code at the same time).
-  I would propose to merge 3.2.2, 3.2.3, 3.2.4, 3.2.5, and to split them into (a) the type checking algorithm (while ignoring the code generation bit), and (b) the code generation aspect.
-  I.e., for (a), pretend that you are interpreting the typing rules instead of compiling them, and please try to present an algorithm that captures
-  the check.
-  The algorithm can be in plain English with bullet points / numbered sub-steps, or in some pseudo-formal notation (``foreach premise $p$ in rule $r$ ...''), as long as it is clear.}
+%The type checker generation is based on a single-responsibility principle, where each node of the typing rule AST is responsible for generating the code it corresponds to \CR{Not sure what you mean}.
+In this section we present our approach to translating typing rules.
+We start by presenting pseudocode for our typing rule evaluation.
+%Section \ref{typecheckingalgorithm} presenting a p the algorithm we use to evaluate typing rules in pseudocode.
+This is followed by Section \ref{codegeneration}, which explains how we generate  code to implement this algorithm.
 
@@
 
-For each rule:
-\begin{enumerate}
-  \item For each parameter in the conclusion:
-    \begin{enumerate}
-      \item Declare a variable by the same name, initialised by fetching the child of the same index.
-    \end{enumerate}
-  \item For each type variable used in the rule:
-    \begin{enumerate}
-      \item Initialise it to null
-      \item Iterate through each premise it appears on the right-hand side of:
-        \begin{enumerate}
-          \item If the type variable is set to null, evaluate the type of the left-hand side and assign it to the type variable.
-          \item If the type variable is not set to null, evaluate the type of our left-hand side and check if it is the same type or a subtype to the type variable. If it is not, throw a type error.
-        \end{enumerate}
-    \end{enumerate}
-  \item Check whether each of the premises is true:
-    \begin{enumerate}
-      \item Evaluate the type of the left-hand side of the premise and check if it is the same type or a subtype to the right-hand side.
-    \end{enumerate}
-    If all true: Return the right-hand side of the premise
-  \item Throw a type error if we reach this point
-\end{enumerate}
-
-\begin{figure}[h]
+%For each rule:
+%\begin{enumerate}
+%  \item For each parameter in the conclusion:
+%    \begin{enumerate}
+%      \item Declare a variable by the same name, initialised by fetching the child of the same index.
+%    \end{enumerate}
+%  \item For each type variable used in the rule:
+%    \begin{enumerate}
+%      \item Initialise it to null
+%      \item Iterate through each premise it appears on the right-hand side of:
+%        \begin{enumerate}
+%          \item If the type variable is set to null, evaluate the type of the left-hand side and assign it to the type variable.
+%          \item If the type variable is not set to null, evaluate the type of our left-hand side and check if it is the same type or a subtype to the type variable. If it is not, throw a type error.
+%        \end{enumerate}
+%    \end{enumerate}
+%  \item Check whether each of the premises is true:
+%    \begin{enumerate}
+%      \item Evaluate the type of the left-hand side of the premise and check if it is the same type or a subtype to the right-hand side.
+%    \end{enumerate}
+%    If all true: Return the right-hand side of the premise
+%  \item Throw a type error if we reach this point
+%\end{enumerate}
+
+\begin{figure}[]
 %foreach parameter in rule.conclusion.parameters:
@@
 \begin{lstlisting}[]
-foreach typevar in rule.typevars
-  declare typevar = null
+foreach typevar in rule.typevars:
+  typevar.type = null;
   foreach premise in rule.premises where premise.rhs == typevar:
-    if typevar == null:
-      typevar = premise.lhs.type()
+    if typevar.type == null:
+      typevar.type = premise.lhs.type()
+    else if typevar.type != premise.lhs.type():
+      throw typeerror("Type variable mismatch")
+
+// Check if all evaluated left-hand side types match their right-hand side
+if allTrue(rule.premises.map(p => p.lhs.type() == p.rhs)):
+  return rule.conclusion.rhs
 else:
-      if typevar != premise.lhs.type():
-        throw typeerror("Type variable mismatch")
-
-if(premise1.lhs.type() == premise1.rhs &&
-   premise2.lhs.type() == premise2.rhs &&
-   ...
-  ):
-    return conclusion.rhs
   throw typeerror("Typechecking failed")
@@
   \caption{Pseudocode for the implementation of the type() attribute}
-  \label{boolsast}
-\end{figure}
-\subsection{Code generation}
+  \label{psuedotype}
+\end{figure}
+
+Here we aim to give an overview of the algorithm used to evaluate the typing rules.
+Figure \ref{psuedotype} contains a pseudocode algorithm for the evaluation of a typing rule.
+It is not representative of the actual generated code, but captures its behaviour in the manner of an interpreted language.
+
+\subsection{Code generation}\label{codegeneration}
+
+In this section we will present the code we generate to implement the previously discussed algorithm.
+Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
+From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.
+%TODO: Rewrite this section to be less bad
+
+This section details the implementation of our type checker generator, on a feature by feature basis.
+
+\subsubsection{Aspect generation}
+The \lstinline{RuleSet} node is responsible for generating the overall structure of the output file, a JastAdd aspect called \lstinline{TypingRules}.
+It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation (though this feature is not yet fully working).
+
+\subsubsection{Method implementation overview}
+The implementation of a typing rule can range from a single return statement to a system of declarations of variables, type variables and control flow.
+These are explained in further detail in subsequent sections, but here follows a quick overview.
+
+Firstly, the algorithm checks if the number of premises is zero, in which case it immediately generates the conclusion, consisting of a return statement.
+If there are premises, it first generates code for the declarations of variables and both declarations and unification for type variables.
+This is followed by an if-statement to test whether the premises hold true, containing a return statement.
+Should the premises not hold, a type error is produced.
+%\CR{What if there are multiple rules that apply to the same AST node?  E.g., one rule that allows \ty{Int} addition and one that allows \ty{String} addition.}
+
+\subsubsection{Parameters}
+When the typing rules concern a node with parameters, these can be named arbitrarily in the typing rule.
+%\CR{Who does the naming?  Why is the ``arbitrarily'' bit important?}
+%\CR{ Will this work if the type system specification has a type variable called ``this'' or ``null'' or ``class''?}.
+A variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
+This naming format means any errors reported will use the same variable names as the user has defined in the typing rules, in order to minimise the disconnect when analysing errors from the generated code.
+However, it also means it is not possible to name a parameter any of the reserved keywords in Java.
+This could be alleviated by adding a prefix to the variable names, at the cost of a slight indirection for the user.
+These variables are then utilised in later premises or type variable declarations.
+
+\subsubsection{Type variables}
+In the case of type variables, much of the logic is handled at runtime, to simplify the code generation.
+Each type variable that appears in the conclusion is instantiated to a null value, and the code generated by each premise first checks if the value is still null and either assigns the variable its own type or tests to make sure the existing type matches its own.
+
+The generated code could be simplified in future work by instantiating the type variable to the type of the left-hand side of the first premise it appears in, and then only checking if each of the subsequent uses match it.
+
+\subsubsection{Premises/control flow}
+For rules containing premises, the return statement is protected by an if-statement.
+At this point we've declared variable for each of the child nodes and type variables used, so the premises can very directly be translated into boolean expressions.
+A premise \lstinline{x : Bool} translates into the expression \lstinline{x.type().matches(new Bool());}.
+This includes a call to the \lstinline{type()} of each left-hand side of the premise, recursively evaluating the necessary types.
+Since our typing rules can only reference their direct children, this recursion will always progress down the program's AST, ensuring termination.
+Each of the boolean expressions then joined by the \lstinline{&&} operator into a single if-statement, protecting the return of the right-hand side of the conclusion.
+For the case where any of the boolean expressions is false, the if-statement is followed by a thrown exception representing a type error.
+
+%\CR{Overall, I am not convinced that this strategy for presenting the type checking algorithm is very effective (though it might be better for someone who is reading the source code at the same time).
+%  I would propose to merge 3.2.2, 3.2.3, 3.2.4, 3.2.5, and to split them into (a) the type checking algorithm (while ignoring the code generation bit), and (b) the code generation aspect.
+%  I.e., for (a), pretend that you are interpreting the typing rules instead of compiling them, and please try to present an algorithm that captures
+%  the check.
+%  The algorithm can be in plain English with bullet points / numbered sub-steps, or in some pseudo-formal notation (``foreach premise $p$ in rule $r$ ...''), as long as it is clear.}
 
@@
 %\CR{What is ``type matching''?  I am not familiar with this term, and you haven't introduced it before.}
-It also contains the implementation of the type matching method, which checks whether a type is equal to or a subtype of another type, which can be seen in Figure \ref{typematches}.
-It is defined for all \lstinline{Type} nodes, and checks whether the type's class is equal to or a subclass of the parameter type.
-\CR{So this is not an equivalence relation, but a partial order?
+It also contains the implementation of the type matching attribute, as defined in Figure \ref{typematches}.
+%It is defined for all \lstinline{Type} nodes, and checks whether the type's class is equal to or a subclass of the parameter type.
+%\CR{So this is not an equivalence relation, but a partial order?
 What does that mean wrt how it is used by the implementation of the type checker?} \reviewquestion{I think it would be easier to change this poorly thought out line of code than to fully evaluate and address the implications in the report :) Is that okay?}
+The current implementation of \lstinline{Type.matches(Type t)} checks that the two types are the exact same, though we also considered an implementation that checks if the Type it is called on is either of the same class or a subtype of the type \lstinline{t}.
+%\CR{is ``matches'' a good name for something that is not symmetric?}
+This would allow the language limited sub-typing at the type checker generation step, but at the risk of causing hard to predict behaviour and hidden order dependencies with type variables.
+%\CR{Can you give an example of a case where this is helpful, and one example where it might be troublesome?
+%Let's use the ``if'' rule as an example.
+If $\ty{Number} :> \ty{Int}$ (\ty{Number} is a supertype of \ty{Int}), what type will I infer if $t_2 = \ty{Number}$ and $t_3 = \ty{Int}$?
+How about the other way around?}
+
+It is also possible to override the \lstinline{matches} implementation for a specific Type, by adding a definition to a JastAdd aspect included in the output type checker.
+This could potentially be used to compare type parameters in future versions, or support for ad-hoc type polymorphism.
+%\CR{This sounds like this is a way to support ``ad-hoc type polymorphism''?}
 
@@
 syn boolean Type.matches(Type t) {
-  return t.getClass().isAssignableFrom(getClass());
+  return getClass().equals(t.getClass());
 }
@@
 Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
-\CR{You can't discuss results without describing your experiments, but to describe the experiments, you should first motivate them.
-  Since you are doing case studies, you can use Pierce to motivate your choice of languages (even if your language is subtly different somehow; I don't have the book in front of me).
-  Make clear to the reader that you picked your languages based on the existing literature, not on a whim.}
+We have chosen to evaluate our program using two example languages, inspired by the languages Booleans and Numbers introduced in Types and Programming Languages\ref{Pierce}.
+We have chosen to simply the languages somewhat, to reduce the amount of equivalent cases for our program to parse, making it easier to evaluate each feature individually.
+Thus, we have defined two languages Bools and BoolsIf, where the former is a simple language consisting of boolean values, an \lstinline{OR} operator, and a zero-value.
+This is a minimal case to produce a meaningful type checker, as it contains values of two different types and a rule with premises limiting the types.
+
+The second language BoolsIf is an extension to the Bools language.
+It adds an if-expression, which typing rules utilises type variables to ensure both if and else terms have the same type, and then uses that type as its own type.
+%\CR{You can't discuss results without describing your experiments, but to describe the experiments, you should first motivate them.
+%  Since you are doing case studies, you can use Pierce to motivate your choice of languages (even if your language is subtly different somehow; I don't have the book in front of me).
+%  Make clear to the reader that you picked your languages based on the existing literature, not on a whim.}
 %\subsection{Input/output examples, with descriptions}
@@
 For the \lstinline{Or} node, the code becomes slightly more complex.
-\CR{The next bit would probably be easier if you just show the code.}
-First two variables are defined, corresponding to the parameters in the typing rule, and assigned to the children with the index corresponding to the position among the parameters.
-This is followed by an if-statement, recursively evaluating the children's types and comparing the result the type demanded in the premise.
-If the condition is true, a \lstinline{Bool} type is returned.
-If not, a runtime exception is thrown to convey a type error.
-
-\CR{You may want to add a short statement that this captures the intended behaviour of the type checker.}
-
-\subsection{If-else extension}
-
-\CR{Separate case study or next step in the same case study?  Either way, the section title should make the connection to 4.1.1 (or lack thereof) clear.}
+%\CR{The next bit would probably be easier if you just show the code.}
+It declares variables corresponding to the child nodes given in the parameters, followed by an if-statement to check if the premises are upheld by recursively evaluating the types of the child nodes.
+Depending on the result of this evaluation, either a \lstinline{Bool} type is returned, or a type error is thrown.
+%First two variables are defined, corresponding to the parameters in the typing rule, and assigned to the children with the index corresponding to the position among the parameters.
+%This is followed by an if-statement, recursively evaluating the children's types and comparing the result the type demanded in the premise.
+%If the condition is true, a \lstinline{Bool} type is returned.
+%If not, a runtime exception is thrown to convey a type error.
+
+%\CR{You may want to add a short statement that this captures the intended behaviour of the type checker.}
+This type checker accurately captures the intended behaviour of the typing rules.
+
+\subsection{BoolsIf}
+
+%\CR{Separate case study or next step in the same case study?  Either way, the section title should make the connection to 4.1.1 (or lack thereof) clear.}
 
@@
 \begin{lstlisting}[]
-IfElse : Term ::= Cond:Term Then:Term Else:Term;
+If : Term ::= Cond:Term Then:Term Else:Term;
 \end{lstlisting}
-  \caption{Extension for the Bools AST in Figure \ref{boolsast} \CR{Consider giving this language a different name, for clarity}}
-  \label{ifelseast}
+  \caption{The additional term introduced in the BoolsIf language, extending the Bools AST in Figure \ref{boolsast}}
+  %\CR{Consider giving this language a different name, for clarity}
+  \label{ifast}
 \end{figure}
@@
 \begin{lstlisting}[]
-[T-IfElse]
+[T-If]
 x: Bool,
@@
 ------------------
-IfElse(x, y, z): a
+If(x, y, z): a
 \end{lstlisting}
-  \caption{Extension for the Bools typing rules in Figure \ref{boolstr} with an if-else term}
-  \label{ifelsetr}
+  \caption{Additional typing rules for the BoolsIf language, an extension for the Bools typing rules in Figure \ref{boolstr} with an if term}
+  \label{iftr}
 \end{figure}
@@
 \begin{lstlisting}[language=jrag]
-syn Type IfElse.type() {
+syn Type If.type() {
   ASTNode x = getChild(0);
@@
 \end{lstlisting}
-\caption{Generated typing rules for the if-else term, extending Figure \ref{boolstrgen} (reformatted for readability)
-  \CR{The components of this function would be great to discuss in Section 3.
+\caption{Generated typing rule for the if term added in the BoolsIf extension. The rest of the typing rules are identical to those in Figure \ref{boolstrgen}. (Reformatted for readability)
+  %\CR{The components of this function would be great to discuss in Section 3.
 Here in Section 4, you may want to forward reference the discussion about redundancy/possible optimisations (Section 4.2.2) and very concretely use this code and explain the changes you have in mind (for someone else to do in future work).}
 }
-  \label{ifelsetrgen}
+  \label{iftrgen}
 \end{figure}
@@
 %\begin{lstlisting}[]
-%[T-IfElse]
+%[T-If]
 %x: Bool,
@@
 %---------
-%IfElse(x, y, z): a
+%If(x, y, z): a
 %\end{lstlisting}
@@
 %\begin{lstlisting}[]
-%syn Type IfElse.type() {
+%syn Type If.type() {
 %ASTNode x = getChild(0);
@@
 %\end{figure}
-As an example of rules utilising type variables, we can extend the Bools language with an if-else term.
-\CR{``Can extend'' sounds a bit arbitrary.  Can you motivate with Pierce (even if not exactly)?}
-As shown in the AST extension in Figure \ref{ifelseast}, it consists of three terms: a conditional, a term to be evaluated if the condition is true and a condition to evaluate if false.
-A typing rule of this if-else term can be seen in Figure \ref{ifelsetr}.
+As an example of rules utilising type variables, we have extended the Bools language with an if term, as used in Pierce's Booleans language\ref{Pierce}.
+This if term captures each of the utilities of type variables in typing rules, using it both to ensure that multiple terms are of the same type and then using that same type as the result of the rule evaluation.
+%\CR{``Can extend'' sounds a bit arbitrary.  Can you motivate with Pierce (even if not exactly)?}
+As shown in the AST extension in Figure \ref{ifast}, it consists of three terms: a conditional, a term to be evaluated if the condition is true and a condition to evaluate if false.
+A typing rule of this if term can be seen in Figure \ref{iftr}.
 The rule has three premises, the first of which declares the conditional must have the type \lstinline{Bool}.
-The next two declare that the two other children must be of the type \lstinline{a}, with the lowercase \CR{grammar} indicating a type variable.
-This type variable \lstinline{a} also subsequently appears in the resulting type of the if-else term.
+The next two declare that the two other children must be of the type \lstinline{a}, with the lowercase identifier indicating a type variable.
+This type variable \lstinline{a} also subsequently appears in the resulting type of the if term.
 
 %\CR{``Shown in Figure~\ref{reference}'' is the current preferred form in our field}
-The code for this rule is shown in Figure \ref{ifelsetrgen}, and begins with three parameter declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
+The code for this rule is shown in Figure \ref{iftrgen}, and begins with three parameter declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
 For each invocation of the type variable, a code section is generated, which checks if the type variable is still null, and if so assigns it to the type of the left-hand side term, or if it has been assigned checks whether the types match each other.
@@
 
-\subsection*{\rqone}
+\subsection{\rqone}
 Our project has shown that generating type checkers using RAGs is a viable solution for a number of different language constructs.
@@
 To evaluate the completeness of our implementation, we refer to Pierce\cite{Pierce} as a reference work.
-The typing rules are introduced in chapter 8, starting with the simple examples shown in Section \ref{sectionlanguage}.
-\CR{Please be clearer on which examples from Pierce you implemented and where support stops.}
-We have implemented support for simple type declarations, where a value or function always has a certain type.
+Typing rules are first introduced in chapter 8, with two simple languages in Figure 8-1 and 8-2.
+The first, Booleans (generally referred to as B) consists of only the syntactic structures \lstinline{true}, \lstinline{false} and \lstinline{if t then t else t}.
+Extracts of the typing rules for the B language were used Section \ref{sectionlanguage}.
+The second language called Numbers (NB) is an extended version of B, adding arithmetic expressions.
+This introduces the terms \lstinline{0}, \lstinline{succ t}, \lstinline{pred t} and \lstinline{iszero t}.
+
+Both the B and NB languages are fully supported by our project, and our Bools and BoolsIf languages were devised to utilise all the same language features but in a more condensed form.
+In chapter 9, Pierce introduces the Simply Typed Lambda Calculus, which typing rules our compiler does not yet support, as they require environments, a feature we'll discuss in Section \ref{futureworkenvironments}.
+
+%\CR{Please be clearer on which examples from Pierce you implemented and where support stops.}
+We have implemented support for simple syntax-directed type declarations, where a value or function always has a certain type.
 These simple rules are not of much interest in themselves but are an important backbone to the subsequent analysis.
@@
 
-A major feature missing from our implementation is the concept of type environments or typing contexts, providing a mapping from variables and functions to types.
-This prevents us from generating type checkers for even simple general purpose languages.
-
-\CR{Please explain typing contexts and why/where they are important.  You can also reflect on whether typing contexts are necessary in the context of RAG-based analysis (cf. what we did in EDAP15).}
-
-\CR{I am missing the phrase ``syntax-directed'' here.}
-
-\subsection*{\rqtwo}
-We have approached translating typing rules into a JastAdd attribute \lstinline{type()},
-%\CR{parsing is the syntactic analysis of the input specification, you have done more than just parsing.}
-which either returns a \lstinline{Type} value, or throws a runtime exception if the expression is invalidly typed.
-
+%A major feature missing from our implementation is the concept of type environments or typing contexts, providing a mapping from variables and functions to types.
+%This prevents us from generating type checkers for even simple general purpose languages.
+%
+%\CR{Please explain typing contexts and why/where they are important.  You can also reflect on whether typing contexts are necessary in the context of RAG-based analysis (cf. what we did in EDAP15).}
+%
+%\CR{I am missing the phrase ``syntax-directed'' here.}
+
+\subsection{\rqtwo}
+The basis of our approach has been to translate typing rules into \lstinline{type()} attributes added to the program's AST.
+We opted for an approach where a successfully evaluated typing rule returns a \lstinline{Type} value, defined in a separate type syntax file, while a failed rule throws a runtime exception representing a type error.
+
+In this section we will be analysing different aspects of this approach, and compare them to potential alternative approaches.
+%We have approached translating typing rules into a JastAdd attribute \lstinline{type()}, which either returns a \lstinline{Type} value, or throws a runtime exception if the expression is invalidly typed.
+
+\subsubsection{Type syntax}
 We have chosen to represent the right-hand side of type equations, the types themselves, as an abstract AST class.
@@
 Parametric types, such as lists where the type of the list contains information about the types of its elements, could be expressed as simply as \lstinline{MyList : Type ::= Type;}.
+
 The current implementation of \lstinline[language=java]{matches()} would however consider all lists to match, without comparing the child nodes.
-Perhaps the default implementation could be extended to compare all child elements, or these more advanced types would require manually overriding the method.
-\CR{Recursive term equivalence can become tricky when you are dealing with type variables (cf. the Hindley-Milner inference exercise in EDAP15).}
-\CR{Type equality could also be expressed in terms of typing rules.  This approach is quite common with subtyping.}
-
+Perhaps the default implementation could be extended to compare all child elements, at the risk of potentially tricky evaluation, or these more advanced types would require manually overriding the method.
+Yet another approach may be to design a way to define type equality within the typing rules, allowing users to customise the behaviour without having to restort to manually writing JastAdd code.
+
+%\CR{Recursive term equivalence can become tricky when you are dealing with type variables (cf. the Hindley-Milner inference exercise in EDAP15).}
+%\CR{Type equality could also be expressed in terms of typing rules.  This approach is quite common with subtyping.}
+
+\subsubsection{Error handling}
 We have implemented premises as conditions of a single if-statement, where we check that each premise holds true before returning the type, or falling through to a type error if the premises do not hold.
@@
 
+A cleaner solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
+Instead of reporting the reason any individual rule failed, the error might instead display either all or only the best matching of the rules it tried to apply, perhaps with a list of non-holding premises for each one.
+%\CR{This is a good idea; you can even sort the rules by closest matches and only show the top ones if there are many.}
+
+\subsubsection{Variables}
 The variables in the typing rules are compiled into regular Java variables containing the AST node.
@@
 
-Type variables similarly are implemented as Java variables, though they are initiated as null, to represent the type variable's unassigned state.
-This is followed by a check for each point of use in the typing rule, to see whether the type variable is still unassigned and if so assign it.
+\subsubsection{Type variables}
+%\CR{I am missing a discussion of type variables and unification as we used them in EDAP15 for Hindley/Milner.}
+Type variables are implemented as Java variables and as we currently only support type variables scoped within a single typing rule and lack support for type parameters, all uses of it will be known directly at the evaluation of the typing rule.
+Thus, our unification algorithm can be very simple.
+All our type variables are initiated as null, to represent the type variable's unassigned state.
+This is followed by a check for each point of use in the typing rule, to see whether the type variable is still unassigned and if so assign it to the type of the left-hand side.
+
 This approach could be improved by finding the first use of each type variable during the rule compilation, and generating a single declaration statement initialising it to the type of the corresponding expression.
 %\CR{no need to say ``much room for improvement'', just mention the specific disadvantages.}
-\CR{I am missing a discussion of type variables and unification as we used them in EDAP15 for Hindley/Milner.}
-
-A cleaner solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
-Instead of reporting the reason any individual rule failed, the error might instead display either all or only the best matching of the rules it tried to apply, perhaps with a list of non-holding premises for each one.
-%\CR{This is a good idea; you can even sort the rules by closest matches and only show the top ones if there are many.}
-
-\CR{Can you structure this subsection a bit more explicitly?  Could be subsubsections, but not necessarily needed; maybe a bullet list of
-challenges and for each challenge a bullet list of potential approaches could work?}
-
-\CR{Something about error handling would be good to discuss here}
-
-\subsection*{\rqthree}
+
+%\reviewquestion{Circular attributes?}
+%\CR{Circular attributes: Recursion in typing rules: Datalog is built with support for ``infinite'' recursion (since it operates on finite domains,
+%  it can use a special bottom-up evaluation scheme to ensure termination.)
+%  If you need this in RAGs, it gets more complicated, but you'd need an example to explain this in the introduction-- since you don't directly address that in this work,
+%  I wouldn't talk about it too much here (you could also skip it entirely here).  Minimum: talk about it in the discussion later.}
+%  %TODO: Circular attributes not necessary in our case, but may be required for other types of type analysis
+
+%\CR{Can you structure this subsection a bit more explicitly?
+Could be subsubsections, but not necessarily needed; maybe a bullet list of challenges and for each challenge a bullet list of potential approaches could work?}
+
+\subsubsection{Error handling}
+We chose a very simple representation for type errors, based around runtime exceptions, which are thrown either when type variable unification fails or the premises of a typing rule fails to match.
+The exceptions include a short message stating which of the two causes has lead to the error.
+
+This approach gave us the minimum information to distinguish between successful and failing type checking, with a very simple implementation.
+It does however complicate the case where a single node has multiple applicable typing rules, depending on their premises.
+In Section \ref{improvederrorhandling} we discuss alternative approaches.
+
+%\CR{Something about error handling would be good to discuss here}
+
+\subsection{\rqthree}
+
+\subsubsection{Technical challenges}
 %Splitting code between multiple JastAdd components (and java) can get messy
-Writing idiomatic JastAdd components often requires splitting code into many different places.
-We attempted to keep each node in the tree responsible for generating the code it corresponds to \CR{What does that mean?
+Writing idiomatic JastAdd components often requires splitting a single algorithm into many different pieces of code.
+Our approach to generating type attributes was based around recursively traversing the typing rule AST to fill in each bit of information, rather than having one parent node trying to form an understanding of the entire tree.
+The plus side of this recursive approach is that each component doesn't need an intricate understanding of the nodes underneath it.
+A \lstinline{Rule} node doesn't need to know whether the right-hand side of the conclusion is a concrete type or a type variable.
+The \lstinline{HasType} node representing the conclusion generates the structure of the return statement, calling on the right-hand side node to fill in the exact type.
+
+However, this code could at times become complex, as certain lines required bits of information from several different parts of the AST, several steps apart.
+Propagating this information required adding intermediate attributes to nodes throughout the AST, leading to a complex set of interdependencies.
+If we want to change a specific token of the generated code, tracing down the exact attribute it is generated in can be tricky.
+%There is a tradeoff to be made between the single responsibility of each component making it easy to verify its correctness, but the complicated interdependence making it hard to grasp the full picture.
+
+%We attempted to keep each node in the tree responsible for generating the code it corresponds to \CR{What does that mean?
 You need to provide a bit more background}, which avoids parent nodes having to check their children's structure with \lstinline{isInstanceOf}, but also means even generating a simple rule can require a dozen calls between different nodes.
-There is a tradeoff to be made between the single responsibility of each component making it easy to verify its correctness, but the complicated interdependence making it hard to grasp the full picture.
-\CR{Overall I found it difficult to process this paragraph.}
-
-Using JastAdd simultaneously as a library and a build component
-\CR{(incomplete)}
+%\CR{Overall I found it difficult to process this paragraph.}
 
 %Having to use a hacky method to access private JastAdd methods to parse a second AST at runtime. (Beyond jastadd's scope?)
-Some of the features we intended to support, such as verifying that typing rules correspond to valid nodes within the object language AST, were complicated by having to parse the object language AST at runtime, using JastAdd as a library rather than a build tool.
+Using JastAdd simultaneously as a library and a build component proved difficult on several occasions.
+The largest issue was that some of the features we intended to support, such as verifying that typing rules correspond to valid nodes within the object language AST, were complicated by having to parse the object language AST at runtime, using JastAdd as a library rather than a build tool.
 JastAdd does not expose any methods for parsing an AST tree at runtime, and we had to access undocumented, private methods through Java's reflection API to get the relevant information.
@@
 
-\CR{This is all about technical challenges.  Do you also see conceptual challenges?  (You can be less specific here, but
-  some words about support for parametric polymorphism and/or subtyping could be nice; at the very least, something
-  about the relevance of syntax-directedness woudl be valuable.)}
+We also ran into a few tooling issues.
+The first was Gradle not willing to recognise JastAdd as both a library and a build component, which we eventually solved by manually providing a jar file separately from the resolved build dependency.
+The second one was our LSP server, jdtls, not recognising JastAdd imports, which we never managed to fix.
+%Gradle troubles
+%Library not recognised by LSP server
+%Difficult runtime parsing of AST files
+
+\subsubsection{Conceptual challenges}
+On the more conceptual side, a fundamental challenge of our approach is how to implement non syntax-directed rules.
+The structure of our program is based around turning typing rules into evaluable attributes on the AST.
+This means we inherently tie typing rules to a specific AST node, creating a direct link between typing rules and the syntax.
+
+However, more complex type systems may require rules that are not syntax-directed, where a rule does not apply to any specific AST node.
+A common example use case for this is subtyping, where any expression of a certain type can also be considered to be any of that type's supertypes.
+Figure \ref{typingruleexamplesub} shows a subtyping rule taken from Pierce\cite{Pierce}.
+
+With our approach, this rule would have to change the behaviour of every generated \lstinline{type()} attribute.
+Additionally, we would require an implementation supporting type polymorphism, where a single term can be evaluated to have multiple different types.
+
+\begin{figure}[]
+  \begin{equation}
+    \tag{T-Sub}
+    \frac{
+      \Gamma \vdash t : S\quad
+      S <: T
+    }{
+      \Gamma \vdash t : T
+    }
+  \end{equation}
+  \caption{A typing rule for subtyping, an example of a non syntax-directed typing rule.}
+  \label{typingruleexamplesub}
+\end{figure}
+
+%\CR{This is all about technical challenges.
+%  Do you also see conceptual challenges?
+%  (You can be less specific here, but some words about support for parametric polymorphism and/or subtyping could be nice; at the very least, something about the relevance of syntax-directedness would be valuable.)}
   % Focus on syntax directedness. Exceptions: if t has type t, it also has type t; nested rules if(xcondition): X, if(ycondition): Y
 
-\chapter{Discussion} % allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
+%\chapter{Discussion} % allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
 Focus here on a broader view of the results, talking about the relation between the different finds.\footnote{Bad practice is to display graphs in Results and then describe them textually one by one in here.
@@
 %    This is something you deal with later in JastAdd.}
-% TODO: Related work section after chapter 5, one or two pages. Various papers from CR comments and goal document.
-% TODO: Discuss Pacak in introduction
-
-\reviewquestion{Circular attributes?}
-\CR{Circular attributes: Recursion in typing rules: Datalog is built with support for ``infinite'' recursion (since it operates on finite domains,
-  it can use a special bottom-up evaluation scheme to ensure termination.)
-  If you need this in RAGs, it gets more complicated, but you'd need an example to explain this in the introduction-- since you don't directly address that in this work,
-  I wouldn't talk about it too much here (you could also skip it entirely here).  Minimum: talk about it in the discussion later.}
+
+Other related work has been done in the domain of compiling natural semantics, without the specific focus on generating type checkers.
+
+An early compiler for natural semantics was made by Mikael Pettersson in 1996\cite{Pettersson}.
+They wrote a compiler in Standard ML which generated C code, linked to a custom runtime system.
+
+Saioc and Hüttel developed a compiler for natural semantics\cite{Saioc_2022}.
+They did so by defining a parsable meta language adaptation of natural semantics, similar to the one we defined for writing typing rules.
+Their project was however focused on verifying the correctness of these definitions and finding errors in their definitions.
+This was something we had hoped to explore more in our own work, but ended up deprioritising after running into issues with JastAdd.
+% : Related work section after chapter 5, one or two pages. Various papers from CR comments and goal document.
 
@@
 \section{Future work}
-\subsection{Environments}
+\subsection{Environments}\label{futureworkenvironments}
 Implementing environments, or type contexts, would be the clear next step for this project.

### 2024-01-16 MSc.tex
Further improvements

@@
 
-  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from a subset of possible typing rules.
+  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from a subset of possible typing rules, known as 'syntax-directed' typing rules.
   This aims to provide integration with existing compilers implemented with JastAdd, as well as a playground for type systems usable as an educational tool.
@@
   % 'Syntax-directed' needs explaining if used, "a subset known as syntax-directed"
-  We find RAGs to be a flexible model for this subset of typing rules,
-  % This is too specific, works with small languages from research literature
-  and our software was able to support simple values, nodes with premises placing demands on their children, and those with rules requiring type variables.
+  We find RAGs to be a flexible model for syntax-directed typing rules, and our software was able to support small languages from research literature.
   It proved difficult to provide compile-time verification of the typing rules, and further work is required to extend the compiler to support type environments, a necessity for most real-world type systems.
@@
 \section{Abstract syntax trees}\label{abstractsyntaxtrees}
-\CR{What's a syntax?}
-An abstract syntax tree (AST) is a tree-based representation of the syntax of a program.
+The syntax of a language is the definition of the symbols and structures which make up a valid program.
+%\CR{What's a syntax?}
+An abstract syntax tree (AST) is a simplified tree-based representation of the syntax.
 Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
@@
 
-\section{Values and expressions}
+\subsection{Values and expressions}
 The main purpose of most computer software is to handle and manipulate data.
@@
   \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
-  \label{boolstrgen}
-\end{figure}
-The RAG modules written in \lstinline{.jrag} files, declare the attributes to be included in the classes generated by the AST specification.
+  \label{jragexample}
+\end{figure}
+The RAG modules are written in \lstinline{.jrag} files, declare the attributes to be included in the classes generated by the AST specification.
+Figure \ref{jragexample} shows a minimal example.
 The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.
@@
   \caption{A definition of the concrete syntax for the typing rule language in a notation similar to BNF.}
-  \label{trastspec}
+  \label{trconcretesyntaxspec}
 \end{figure}
@@
 
-\begin{figure}[h]
+\begin{figure}[]
 \centering
@@
 \end{figure}
-\begin{figure}[h]
+\begin{figure}[]
 \centering
@@
 From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.
-%TODO: Rewrite this section to be less bad
-
-This section details the implementation of our type checker generator, on a feature by feature basis.
+
+%This section details the implementation of our type checker generator, on a feature by feature basis.
 
@@
 Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
-We have chosen to evaluate our program using two example languages, inspired by the languages Booleans and Numbers introduced in Types and Programming Languages\ref{Pierce}.
+We have chosen to evaluate our program using two example languages, inspired by the languages Booleans and Numbers introduced in Types and Programming Languages\cite{Pierce}.
 We have chosen to simply the languages somewhat, to reduce the amount of equivalent cases for our program to parse, making it easier to evaluate each feature individually.
@@
 %\end{figure}
-As an example of rules utilising type variables, we have extended the Bools language with an if term, as used in Pierce's Booleans language\ref{Pierce}.
+As an example of rules utilising type variables, we have extended the Bools language with an if term, as used in Pierce's Booleans language\cite{Pierce}.
 This if term captures each of the utilities of type variables in typing rules, using it both to ensure that multiple terms are of the same type and then using that same type as the result of the rule evaluation.
@@
 
-A cleaner solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
+A more informative solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
 Instead of reporting the reason any individual rule failed, the error might instead display either all or only the best matching of the rules it tried to apply, perhaps with a list of non-holding premises for each one.
 %\CR{This is a good idea; you can even sort the rules by closest matches and only show the top ones if there are many.}
+%\subsubsection{Error handling}
+%We chose a very simple representation for type errors, based around runtime exceptions, which are thrown either when type variable unification fails or the premises of a typing rule fails to match.
+%The exceptions include a short message stating which of the two causes has lead to the error.
+%
+%This approach gave us the minimum information to distinguish between successful and failing type checking, with a very simple implementation.
+%It does however complicate the case where a single node has multiple applicable typing rules, depending on their premises.
+%In Section \ref{improvederrorhandling} we discuss alternative approaches.
+
+%\CR{Something about error handling would be good to discuss here}
 
@@
 Could be subsubsections, but not necessarily needed; maybe a bullet list of challenges and for each challenge a bullet list of potential approaches could work?}
-
-\subsubsection{Error handling}
-We chose a very simple representation for type errors, based around runtime exceptions, which are thrown either when type variable unification fails or the premises of a typing rule fails to match.
-The exceptions include a short message stating which of the two causes has lead to the error.
-
-This approach gave us the minimum information to distinguish between successful and failing type checking, with a very simple implementation.
-It does however complicate the case where a single node has multiple applicable typing rules, depending on their premises.
-In Section \ref{improvederrorhandling} we discuss alternative approaches.
-
-%\CR{Something about error handling would be good to discuss here}
 
@@
 \chapter{Source code}
-\url{https://github.com/nichobi/thesis-project}
+The source code for our project is available at \\\url{https://github.com/nichobi/thesis-project}, \\with the README.md file providing instructions for use.
+
+\vspace*{\fill}
+\hspace*{\fill}
+\includegraphics[scale=0.2]{stagbeetle.pdf}
 
@@
 
-%make sure we're on even page with the pop-sci
-\checkoddpage
-\ifoddpage
-\else
-  \newpage
-  \thispagestyle{empty}
-  \mbox{ }
-\fi
-\includepdf[pages={1}]{popsci/popsci.pdf}
+%TODO: Re-add the popsci summary
+%%make sure we're on even page with the pop-sci
+%\checkoddpage
+%\ifoddpage
+%\else
+%  \newpage
+%  \thispagestyle{empty}
+%  \mbox{ }
+%\fi
+%\includepdf[pages={1}]{popsci/popsci.pdf}
 \end{appendices}

### 2024-01-17 MSc.tex
Fix typoes and mistakes

@@
 While writing software, developers often make mistakes.
-Catching these mistakes as early as possible can prevent these bugs from ever reaching production code.
+Catching these mistakes, as early as possible, can prevent these bugs from ever reaching production code.
 %\CR{Possible papers to cite as evidence:
@@
 It focused on defining a way of expressing typing rules as a series of relations solvable by the Datalog compiler.
-Our use of RAGs instead allows us to build our analysis directly onto a representation of the program's structure.
+Our use of RAGs allows us instead to build our analysis directly onto a representation of the program's structure.
 JastAdd is also an interesting target because it has already been used to implement compilers of several languages, to which our typing rule implementations could potentially be integrated.
@@
 %\CR{what is ``this information'', concretely?}
-A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if node, with named branches pointing towards the condition and the two values.
+A statement like \verb|if (true) then {1} else {0}| may instead be represented by an if node, with named branches pointing towards the condition and the two values.
 %\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
@@
 \subsection{Type checking}
-Verifying that these constraints are upheld can assist us in catching programming mistakes, preventing the wrong data types from being used in the wrong places, which may have lead to erroneous or unpredictable behaviour.
+Verifying that these constraints are upheld can assist us in catching programming mistakes, preventing the wrong data types from being used in the wrong places, which may have led to erroneous or unpredictable behaviour.
 There are two main approaches to ensuring the constraints are upheld, dynamic and static type-checking.
@@
 
-Dynamic checking is done performed runtime, with each type checked only when needed.
-While this means no checks are performed unnecessarily, it makes it harder to verify the constraints over the whole program, as it won't be tested until each specific part is executed.
+Dynamic checking is performed at runtime, with each type checked only when needed.
+While this means no checks are performed unnecessarily, it makes it harder to verify the constraints over the whole program, as it will not be tested until each specific part is executed.
 
 Static checking is performed on the code before it is run, usually as part of compilation, preventing many faulty programs from being run at all.
-However, since it lacks knowledge of what happens at runtime, it's required to take a more conservative approach, flagging potential errors even in code which will never executed.
+However, since it lacks knowledge of what happens at runtime, it is required to take a more conservative approach, flagging potential errors even in code which will never executed.
 Static checkers analyse the structure of the source code, often in the form of an abstract syntax tree.
@@
 There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
-This enables differentiating lists that store integers from those that store strings, and ensuring that any item added to a list must be of the same type as the type parameter.
+This enables differentiating lists that store integers from those that store strings and ensuring that any item added to a list must be of the same type as the type parameter.
 %\CR{A bit vague; consider giving an example.}
@@
 \subsubsection{AST specification}\label{astdef}
-One of the key components of a JastAdd projects are the \lstinline{.ast} files, which declare the structure of the abstract syntax tree.
+One of the key components of JastAdd projects are the \lstinline{.ast} files, which declare the structure of the abstract syntax tree.
 This specification is then used by JastAdd to generate Java classes representing the nodes of the AST.
-These AST classes can have a variety of forms, supporting both abstract classes and subclasses, and can contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.
+These AST classes can have a variety of forms, supporting both abstract classes and subclasses.
+They can also contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.
 For a detailed overview of the syntax, see the reference manual\cite{jastaddref}.
@@
 \end{figure}
-The RAG modules are written in \lstinline{.jrag} files, declare the attributes to be included in the classes generated by the AST specification.
+The RAG modules are written in \lstinline{.jrag} files and declare the attributes to be included in the classes generated by the AST specification.
 Figure \ref{jragexample} shows a minimal example.
@@
 
-The largest difference lies in the notation used for the expression to be typed, where Pierce uses the object language's native syntactic form, whereas our representation is based on the corresponding AST nodes.
+The largest difference lies in the notation used for the expression to be typed.
+Pierce uses the object language's native syntactic form, whereas our representation is based on the corresponding AST nodes.
 Node names are consistent with those used by JastAdd, followed by parentheses containing variable names to denote their child nodes in the order of their definition.
@@
 Another difference is the use of the type variables, which have been made lowercase.
-This disambiguates them from the concrete types, which as AST nodes themselves are written with uppercase letters, and also aligns the syntax for type terms with that of regular terms, where concrete nodes are written uppercase and variables in lowercase.
-\lstinline{T}, which has been switched to the lowercase \lstinline{t} in order to distinguish concrete types and type variables.
-This disambiguates them from the concrete types, which as AST nodes themselves are written with uppercase letters, and also aligns the syntax for type terms with that of regular terms, where concrete nodes (such as \lstinline{If}) are written uppercase and variables (\lstinline{t1}, \lstinline{t2}, \lstinline{t3}) in lowercase.
+This disambiguates them from the concrete types, which as AST nodes themselves are written with uppercase letters, and also aligns the syntax for type terms with that of regular terms, where concrete nodes are written in uppercase and variables in lowercase.
+%\lstinline{T}, which has been switched to the lowercase \lstinline{t} in order to distinguish concrete types and type variables.
+%This disambiguates them from the concrete types, which as AST nodes themselves are written with uppercase letters, and also aligns the syntax for type terms with that of regular terms, where concrete nodes (such as \lstinline{If}) are written uppercase and variables (\lstinline{t1}, \lstinline{t2}, \lstinline{t3}) in lowercase.
 %\CR{I generally recommend avoiding the passive voice, but the use below is an example that would be OK}
@@
 It represents the flow of the program, with components and artefacts (represented as rectangular boxes) with arrows linking them to actions (elliptical boxes) which in turn have arrows to the new components and artefacts they produce.
-The dashed arrows from represent links from the support library components that exist in the current experimental state of the project, but in a finished product would instead link to component from the object language's own compilation environment.
+The dashed arrows represent links from the support library components that exist in the current experimental state of the project, but in a finished product would instead link from components out of the object language's own compilation environment.
   %\CR{Can you mark, e.g. via colours, the parts that the user provides, and the parts that will be in the generated output project?}
@@
 To express typing rules, we created a simple ASCII representation of the natural deduction notation used by Pierce\cite{Pierce}.
-Custom DSL
-Include typing rules AST specification (or BNF?)
-ASCII representation of natural deduction %BNF (is it BNF?)
+%Custom DSL
+%Include typing rules AST specification (or BNF?)
+%ASCII representation of natural deduction %BNF (is it BNF?)
 
@@
 We have chosen to evaluate our program using two example languages, inspired by the languages Booleans and Numbers introduced in Types and Programming Languages\cite{Pierce}.
-We have chosen to simply the languages somewhat, to reduce the amount of equivalent cases for our program to parse, making it easier to evaluate each feature individually.
+We have chosen to simplify the languages somewhat, to reduce the amount of equivalent cases for our program to parse, making it easier to evaluate each feature individually.
 Thus, we have defined two languages Bools and BoolsIf, where the former is a simple language consisting of boolean values, an \lstinline{OR} operator, and a zero-value.
@@
 
-A more informative solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
+A more informative solution to this might be to view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
 Instead of reporting the reason any individual rule failed, the error might instead display either all or only the best matching of the rules it tried to apply, perhaps with a list of non-holding premises for each one.
@@
 %It does however complicate the case where a single node has multiple applicable typing rules, depending on their premises.
-%In Section \ref{improvederrorhandling} we discuss alternative approaches.
+%In Section \ref{,improvederrorhandling} we discuss alternative approaches.
 
@@
 Our approach to generating type attributes was based around recursively traversing the typing rule AST to fill in each bit of information, rather than having one parent node trying to form an understanding of the entire tree.
-The plus side of this recursive approach is that each component doesn't need an intricate understanding of the nodes underneath it.
-A \lstinline{Rule} node doesn't need to know whether the right-hand side of the conclusion is a concrete type or a type variable.
+The plus side of this recursive approach is that each component does not need an intricate understanding of the nodes underneath it.
+For example, a \lstinline{Rule} node does not need to know whether the right-hand side of the conclusion is a concrete type or a type variable.
 The \lstinline{HasType} node representing the conclusion generates the structure of the return statement, calling on the right-hand side node to fill in the exact type.
@@
 
-Their work is ends up largely focused on expressing typing rules within the limits of the Datalog language, which can only compute finite relations.
+Their work ends up largely focused on expressing typing rules within the limits of the Datalog language, which can only compute finite relations.
 We have a significantly different compilation target, JastAdd, which allows us to use reference attribute grammars to evaluate the typing relations, utilising a built-in AST structure.
@@
 The current exception-based error handling is rather primitive and leaves much to be desired.
-Some improvements, such as provide more useful error messages, could be possible by making minor changes to how these exceptions are generated.
+Some improvements, such as providing more useful error messages, could be possible by making minor changes to how these exceptions are generated.
 Information such as what part of the input program the type error occurred at, what the failing typing rule was and what prerequisites it has, could be added to the exception message, for a quick usability improvement.
@@
 
-\subsection{Improved type checking algorithm}
+%\subsection{Improved type checking algorithm}
 

### 2024-01-17 popsci/popsci.tex
Add popsci summary

@@
+% USE PDFLatex!
+% to correctly render Swedish characters
+
 \documentclass{popsci}
 
-\usepackage[swedish, english]{babel}
-%\usepackage[utf8]{inputenc}
+\usepackage[utf8]{inputenc}
+\usepackage[main=swedish, british]{babel}
+
 \usepackage{fancyhdr}
@@
 \usepackage{colortbl}
+\usepackage{graphicx}
+\usepackage{flushend}
+\usepackage{lmodern}
+\usepackage{caption}
 
-\usepackage{ifxetex}
+% Please specify the presentation date
+\presentationsdag{2024-01-25}
 
-\ifxetex
-  \usepackage{fontspec}
-\else
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{lmodern}
-\fi
+% use either of these commands to specify the title of your thesis
+\examensarbete{Type Checker Generation using Reference Attribute Grammars}
+% To create a title in two rows, leave examensarbete blank and fill in examensarbeteTwoRows.
+%\examensarbeteTwoRows{Application Specific Instruction-set Processor Using a Parametrizable multi-SIMD}{Synthesizeable Model Supporting Design Space Exploration}
+\student{Nicholas Boyd Isacsson}
+%\students{Magnus Hultin}{Mr X}
+\supervisor{Christoph Reichenback (LTH)}
+\examiner{Niklas Fors (LTH)}
 
-\presentationsdag{2016-04-28}
-
-\examensarbete{}
-%To create a title in two rows, leave examensarbete blank and fill in examensarbeteTwoRows.
-\examensarbeteTwoRows{Application Specific Instruction-set Processor Using a Parametrizable multi-SIMD}{Synthesizeable Model Supporting Design Space Exploration}
-
-\author{Magnus Hultin}
-\supervisor {Flavius Gruian (LTH)}
-\examiner{Krzysztof Kuchcinski (LTH)}
-
-\title{Parametrisk processor modell för design utforskning}
+% Your pop-sci title should be different (more catchy) than your thesis title
+\title{Autogenererad typsäkerhet för programmeringsspråk}
 
@@
 
-\theabstract{Applikations-specifika processorer är allt mer vanligt för få ut rätt prestanda med så lite resurser som möjligt.
+% not more than 4 rows!
+%\theabstract{Applikations-specifika processorer är allt mer vanligt för få ut rätt prestanda med så lite resurser som möjligt.
 Detta arbete har en parametrisk modell för att kunna testa hur mycket resurser som behövs för en specifik applikation.}
+\theabstract{Typkollande är ett vanligt verktyg för att hitta buggar i datorprogram, men är relativt komplicerade att skapa.
+Jag har byggt ett program som automatiskt skapar typkollare för program utifrån en kort definition.}
 
-{\noindent För att öka prestandan i dagens processorer finns det vektorenheter och flera kärnor i processorer.
+{\noindent
+Datorprogram hanterar ofta många olika datatyper, och fel användning av dem är en av de vanligaste buggarna.
+För att hitta dessa buggar används typkollare, som systematisk går igenom hela programmet och verifierar varenda användning av variabler eller värden för att se till att rätt datatyp använts på rätt plats.
+
+\begin{figure}[!bth] % Use pictures in your pop-vet!
+\includegraphics[width=\columnwidth]{typeerror.png}
+\caption*{Ett typfel i ett vanligt programmeringsspråk, där en siffra används istället för ett booleskt (sant/falskt) värde.}
+\end{figure}
+
+Dessa typkollare är dock komplicerade att skapa, och är ofta starkt kopplade till ett specifikt programmeringsspråk.
+Det innebär att varenda programmeringsspråk behöver skapa sin egen typkollare, och leder till en otrolig mängd upprepat arbete.
+I mitt arbete har jag skapat ett verktyg för att generera typkollare per automatik, utifrån en kort och koncis definition av programmets regler.
+
+Detta görs genom att baka in kod i ett av mellanleden när programmet omvandlas från människoskriven text till maskinexekverbar kod.
+%Detta mellanled representerar programmets beteende, utan att ha med alla detaljer om hur exakt programmet är skrivet.
+Mellanledet har en struktur lik ett släktträd där allt härstammar ifrån roten, med grenar som representerar olika kodstrukturer, och som i sig kan delas av i flera grenar.
+
+Därmed kan vi enkelt skapa typkollare för nya språk på ett kortfattat vis, genom att bara definiera vilka regler som ska finnas och inte behöva fundera över hur man implementerar dem.
+Det underlättar också om man vill göra förändringar i systemet, eftersom det är mycket lättare att ändra de koncisa reglerna än att ändra de många raderna kod som kan krävas för att implementera den.
+
+I nuläget är programmet ganska begränsat i hur komplicerade typkollare det kan generera, och skulle inte kunna appliceras på något av de stora, kända programmeringsspråken.
+Däremot klarar det flera exempel från forskningslitteraturen, och tillvägagångssättet har dock visats vara flexibelt och utbyggbart.
+Man borde alltså kunna vidareutveckla programmet för ett bredare stöd i framtiden.
+
+%För att öka prestandan i dagens processorer finns det vektorenheter och flera kärnor i processorer.
 Vektorenheten finns till för att kunna utföra en operation på en mängd data samtidigt och flera kärnor gör att man kan utföra fler instruktioner samtidigt.
@@
 Man kan jämföra det med att frakta ett paket med en stor lastbil istället för att använda en mindre bil där samma paketet också skulle få plats.
-
-I mitt examensarbete har jag skrivit en modell som kan användas för att snabbt designa en processor enligt vissa parametrar.
+%
+%I mitt examensarbete har jag skrivit en modell som kan användas för att snabbt designa en processor enligt vissa parametrar.
 Dessa parametrar väljs utifrån vilket eller vilka program man tänkta köra på den.
@@
 lättare använda flera kärnor och vissa program kan använda korta eller längre vektorenheter för dess data.
-
-För att kunna välja vilken typ av processor som är rätt för den specifika applikationen krävs det ofta att man snabbt kan testa olika prototyper.
+%
+%%För att kunna välja vilken typ av processor som är rätt för den specifika applikationen krävs det ofta att man snabbt kan testa olika prototyper.
 Att implementera dessa till hårdvara kan ofta vara tidskrävande och ifall det visar sig att implementationen inte klarar dem kraven man ställt för prestanda och energieffektivitet, måste man designa för nya parametrar och mer tid har blivit slösat.
-Om den här processen istället kan göras automatiskt utifrån dessa design-parametrar kan man teoretiskt spara en massa tid.
-Modellen testades med olika multimedia program.
+%Om den här processen istället kan göras automatiskt utifrån dessa design-parametrar kan man teoretiskt spara en massa tid.
+%Modellen testades med olika multimedia program.
 Den mest beräkningsintensiva och mest upprepande delen av programmen användes.
@@
 Kärnorna som användes var ifrån MPEG och JPEG, som används för bildkomprimering och videokomprimering.
-
-Resultatet visar att det finns en prestanda vinst jämfört med generella processorer men att detta också ökar resurserna som behövs.
+%
+%\begin{figure}[!bth] % Use pictures in your pop-vet!
+%\includegraphics[width=\columnwidth]{samplePic.pdf} 
+%%\caption{En fin bild}
+%\end{figure}
+%Resultatet visar att det finns en prestanda vinst jämfört med generella processorer men att detta också ökar resurserna som behövs.
 Detta trots att den generella processorn har nästan dubbelt så hög klockfrekvens än dem applikations-specifika processorerna.

### 2024-01-26 MSc.tex
Add popsci summary
Address review comments

@@
 
-\thesisnumber{LU-CS-EX: 2023-79} % Magic Number! Do not change unless Birger Swahn asks you to do so!
+\thesisnumber{LU-CS-EX: 2024-06} % Magic Number! Do not change unless Birger Swahn asks you to do so!
 % default is Master. Uncomment the following for "kandidatarbete"/Bachelor's thesis
@@
 \title{Type Checker Generation using Reference Attribute Grammars}
-\begin{otherlanguage}{swedish}
-\svensktitel{Infoga den Svenska titeln här!}
-\end{otherlanguage}
+\svensktitel{Generering av Typkontrollerare med Referens-Attribut-Grammatik}
 
@@
 \acknowledgements{
-Sincere thanks to Christoph Reichenbach,
-whose enthusiasm for the subject inspired me to pursue this subject
-and whose guidance in matters of both implementation and writing were key to this thesis.
+Sincere thanks to Christoph Reichenbach, whose enthusiasm for the field inspired me to pursue this project and whose guidance in matters of both implementation and writing was key to this thesis.
 Additional thanks to Niklas Fors for providing several valuable insights, greatly simplifying certain components and Alexandru Dura for providing some guidance to JastAdd internals.
 
-I would also like to thank my partner Lo-Isabel Krantz Andrée, my family, and Moritz for their support throughout the duration of this project.
+I would also like to thank my partner Lo-Isabel Krantz Andrée as well as my family, friends, and Moritz for their support throughout the duration of this project.
 
@@
 %Thanks also go to the following (former) students for helping with feedback and suggestions on this template: Mikael Persson, Christoffer Lundgren, Mahmoud Nasser.
+\vspace*{\fill}
+\hspace*{\fill}
+\includegraphics[scale=0.5]{assets/cat.pdf}
 }
@@
 We define a custom parsable language for typing rules adapted from natural semantics, and introduce a strategy to compile it into RAG attributes.
-Our case studies show that simple type systems can be systematically compiled into attributes on the AST, and the tree syntax provides a natural approach for recursive evaluation.
+Our case studies show that simple type systems can be systematically compiled into attributes on the Abstract Syntax Tree (AST), and the tree syntax provides a natural approach for recursive evaluation.
 Overall, this initial exploration shows that RAGs are a flexible approach for implementing type checkers for syntax-directed typing rules, though there are difficulties when implementing rule verification at compiler runtime.
@@
 \section{Abstract syntax trees}\label{abstractsyntaxtrees}
+\begin{figure}[]
+  \centering
+  \begin{tikzpicture}[
+      every node/.style = {align=center, line width=0.8pt},
+      every path/.style = {draw, -latex},
+      astnode/.style={draw=black},
+      %greennode/.style={shape=circle, draw=green, line width=2},
+      %rednode/.style={shape=circle, draw=red, line width=2}
+    ]
+    \node (if)   [astnode]                             {If};
+    \node (then) [astnode, below=of if]                {True};
+    \node (else) [astnode, right=of then]              {False};
+    \node (cond) [astnode, left=of then]               {GreaterThan};
+    \node (one)  [astnode, below = of cond.south west] {Int\\ \textcolor{gray}{value: 1}};
+    \node (zero) [astnode, below = of cond.south east] {Int\\ \textcolor{gray}{value: 0}};
+
+    %\node () [above right = 0.3cm and -1cm of cond] {cond};
+    %\node () [above left = 0.1cm and -0.7cm of then] {then};
+    %\node () [above left= 0.3cm and -0.5cm of else] {else};
+    %\node () [above= 0.3cm of one] {left};
+    %\node () [above= 0.3cm of zero] {right};
+
+    \draw (if.south)   to (cond.north);
+    \draw (if.south)   to (then.north);
+    \draw (if.south)   to (else.north);
+    \draw (cond.south) to (one.north);
+    \draw (cond.south) to (zero.north);
+    %\draw (if.south)   to [edge label=Cond]  (cond.north);
+    %\draw (if.south)   to [edge label=Then]  (then.north);
+    %\draw (if.south)   to [edge label=Else]  (else.north);
+    %\draw (cond.south) to [edge label=Left]  (one.north);
+    %\draw (cond.south) to [edge label=Right] (zero.north);
+
+  \end{tikzpicture}
+  \caption{Example of an abstract syntax tree}
+  \label{astexample}
+\end{figure}
 The syntax of a language is the definition of the symbols and structures which make up a valid program.
@@
 %\CR{what is ``this information'', concretely?}
-A statement like \verb|if (true) then {1} else {0}| may instead be represented by an if node, with named branches pointing towards the condition and the two values.
+Figure \ref{astexample} shows an example of an abstract syntax tree representation of the expression \verb|if 1 > 0 then true else false|.
+%A statement like \verb|if (true) then {1} else {0}| may instead be represented by an if node, with named branches pointing towards the condition and the two values.
 %\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
@@
 Verifying that these constraints are upheld can assist us in catching programming mistakes, preventing the wrong data types from being used in the wrong places, which may have led to erroneous or unpredictable behaviour.
-There are two main approaches to ensuring the constraints are upheld, dynamic and static type-checking.
+There are two main approaches to ensuring the constraints are upheld, dynamic and static type checking.
 As our project generates a static type checker, we will focus our explanations on that topic, but provide a quick overview of their differences here.
@@
 This enables differentiating lists that store integers from those that store strings and ensuring that any item added to a list must be of the same type as the type parameter.
+
+Figure \ref{typesyntaxspec} shows an example specification of a type syntax, consisting of only two types, integers and booleans.
+This specification is used for all the languages used in our project.
 %\CR{A bit vague; consider giving an example.}
-\begin{figure}[h]
+\begin{figure}[]
 \begin{lstlisting}[]
@@
 \end{lstlisting}
-  \caption{Type syntax consisting of two types, for boolean and integer values, written as a JastAdd AST specification (see Section \ref{astdef})}
+  \caption{Type syntax consisting of two types, for boolean and integer values, written as a JastAdd AST specification (see Section \ref{astdef}).}
   \label{typesyntaxspec}
@@
 
-A slightly more complex case can be expressed as \lstinline{if true then true else true}.
+A somewhat more complex case can be expressed as \lstinline{if true then true else }\\\lstinline{true}.
 The only rule that matches the structure, is T-If, with \lstinline{true} values taking the place of each of the $t_i$ variables.
@@
 \subsection{JastAdd}
-JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion.\cite{JastAdd}
+JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion\cite{JastAdd}.
 Our project uses it both for the implementation of our compiler and as a compilation target, as it outputs JastAdd code.
@@
 \subsubsection{AST specification}\label{astdef}
+\begin{figure}[]
+\begin{lstlisting}[]
+Expression ::= Term;
+
+abstract Term;
+True  : Term;
+False : Term;
+Or    : Term ::= Left:Term Right:Term;
+Zero  : Term;
+
+\end{lstlisting}
+  \caption{AST specification for the Bools language.}
+  \label{astspecexample}
+\end{figure}
 One of the key components of JastAdd projects are the \lstinline{.ast} files, which declare the structure of the abstract syntax tree.
@@
 They can also contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.
+
+Figure \ref{astspecexample} shows a small example of the specificaton syntax.
+Each of the node definitions begins with a class name.
+\lstinline{True : Term;} indicates that the \lstinline{True} node is a subclass of the abstract class \lstinline{Term}.
+If a node has child nodes, these are provided after the \lstinline{::=}, and may either have a name and a type separated by a colon, or only a type.
 For a detailed overview of the syntax, see the reference manual\cite{jastaddref}.
@@
 \subsubsection{RAG Modules}
-\begin{figure}[h]
+\begin{figure}[]
 \begin{lstlisting}[language=jrag]
@@
 
-\chapter{Approach/Design} % should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:
+\chapter{Approach} % should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:
 
@@
 \end{lstlisting}
-  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to BNF.}
+  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to Backus-Naur form (BNF).}
   \label{trconcretesyntaxspec}
@@
 
-To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{Pierce}.
+To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{Pierce}, with its concrete syntax given in Figure \ref{trconcretesyntaxspec}.
 Our language can express a subset of the notation used by Pierce, with some differences.
@@
 \subsubsection{Typing rules parser}\label{typingrulesparser}
-The parser for typing rules is written using JFlex\cite{jflex}, a scanner generator, and Beaver\cite{beaver}, an LALR parser generator.
+The parser for typing rules is written using JFlex\cite{jflex}, a scanner generator, and Beaver\cite{beaver}, a parser generator.
 JFlex takes a series of regular expressions to generate a tokenizer for the parsed language.
@@
 %\CR{I would give this AST/BNF a more specific name here, to distinguish it from the other two (as done nicely already in Figure 3.1)}
-\begin{figure}[h]
+\begin{figure}[]
 \begin{lstlisting}[]
@@
 \subsection{Support library}
-In addition to the generated type-checking files, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the support library.
+In addition to the generated type checking files, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the support library.
 Some of these are key components of the type checker, but others are dummy components included solely for convenience of testing.
@@
 These children will be null, causing issues further down the line, though as this parser is only intended for testing it is of little concern.
+Our tests have been carefully written to avoid this error, and any faulty tests would result in obvious null pointer exceptions during further evaluation.
 %\CR{Does this also apply to the analysis of typing rules?  If so, you can mention it as ``future work''
@@
 It can parse simplified tree-representations of programs, and utilise these to perform the type checking.
-In a future real world use case, it is intended for our generated type-checking aspects to integrate into a pre-existing project, which already has all the necessary project files to parse an AST tree, to which our type aspects could integrate.
+In a future real world use case, it is intended for our generated type checking aspects to integrate into a pre-existing project, which already has all the necessary project files to parse an AST tree, to which our type aspects could integrate.
 
@@
 
-\subsection{Type-checking algorithm}\label{typecheckingalgorithm}
+\subsection{Type checking algorithm}\label{typecheckingalgorithm}
 
@@
 
-\begin{figure}[]
+\begin{figure}[bhpt]
 %foreach parameter in rule.conclusion.parameters:
@@
 The \lstinline{RuleSet} node is responsible for generating the overall structure of the output file, a JastAdd aspect called \lstinline{TypingRules}.
-It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation (though this feature is not yet fully working).
+It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple typing rules type the same node, they can be inserted into the same method implementation (though this feature is not yet fully working).
 
@@
 
-\subsubsection{Parameters}
-When the typing rules concern a node with parameters, these can be named arbitrarily in the typing rule.
+\subsubsection{Child nodes}
+When the typing rules concern a node with children, these are bound to free variables which can be named arbitrarily.
 %\CR{Who does the naming?  Why is the ``arbitrarily'' bit important?}
@@
 This naming format means any errors reported will use the same variable names as the user has defined in the typing rules, in order to minimise the disconnect when analysing errors from the generated code.
-However, it also means it is not possible to name a parameter any of the reserved keywords in Java.
+However, it also means it is not possible to name a variable any of the reserved keywords in Java.
 This could be alleviated by adding a prefix to the variable names, at the cost of a slight indirection for the user.
@@
 
-\subsubsection{Type variables}
+\subsubsection{Type variables}\label{typevariables}
 In the case of type variables, much of the logic is handled at runtime, to simplify the code generation.
@@
 z: a
-------------------
+--------------
 If(x, y, z): a
@@
 %\CR{``Shown in Figure~\ref{reference}'' is the current preferred form in our field}
-The code for this rule is shown in Figure \ref{iftrgen}, and begins with three parameter declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
+The code for this rule is shown in Figure \ref{iftrgen}, and begins with three local variable declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
 For each invocation of the type variable, a code section is generated, which checks if the type variable is still null, and if so assigns it to the type of the left-hand side term, or if it has been assigned checks whether the types match each other.
@@
 
-The rest of the method is similar, with premises being checked -- including a redundant check of the type variables -- before returning the type variable, or a type error being thrown.
+The rest of the method is similar, with premises being checked -- including a redundant check of the type variables, as mentioned in Section \ref{typevariables} -- before returning the type variable, or a type error being thrown.
+This type checker accurately captures the intended behaviour of the typing rule.
 
@@
 Perhaps the default implementation could be extended to compare all child elements, at the risk of potentially tricky evaluation, or these more advanced types would require manually overriding the method.
-Yet another approach may be to design a way to define type equality within the typing rules, allowing users to customise the behaviour without having to restort to manually writing JastAdd code.
+Yet another approach may be to design a way to define type equality within the typing rules, allowing users to customise the behaviour without having to resort to manually writing JastAdd code.
 
@@
 The first was Gradle not willing to recognise JastAdd as both a library and a build component, which we eventually solved by manually providing a jar file separately from the resolved build dependency.
-The second one was our LSP server, jdtls, not recognising JastAdd imports, which we never managed to fix.
+The second one was our Language Server Protocol (LSP) server, jdtls, not recognising JastAdd imports and flagging every use thereof as an error, which we never managed to fix.
 %Gradle troubles
@@
 In type systems it is generally considered valid to have multiple typing rules for the same construct, with different, non-intersecting prerequisites.
-It could for example be useful to type an addition operator for a variety of different numerical types, which may not have a common ancestor.
+It could, for example, be useful to type an addition operator for a variety of different numerical types, which may not have a common ancestor.
 
@@
 \hspace*{\fill}
-\includegraphics[scale=0.2]{stagbeetle.pdf}
+\includegraphics[scale=0.2]{assets/stagbeetle.pdf}
 
@@
 
-%TODO: Re-add the popsci summary
-%%make sure we're on even page with the pop-sci
-%\checkoddpage
-%\ifoddpage
-%\else
-%  \newpage
-%  \thispagestyle{empty}
-%  \mbox{ }
-%\fi
-%\includepdf[pages={1}]{popsci/popsci.pdf}
+%make sure we're on even page with the pop-sci
+\checkoddpage
+\ifoddpage
+\else
+  \newpage
+  \thispagestyle{empty}
+  \mbox{ }
+\fi
+\includepdf[pages={1}]{popsci/popsci.pdf}
 \end{appendices}

### 2024-01-26 popsci/popsci.tex
Address review comments

@@
 Detta arbete har en parametrisk modell för att kunna testa hur mycket resurser som behövs för en specifik applikation.}
-\theabstract{Typkollande är ett vanligt verktyg för att hitta buggar i datorprogram, men är relativt komplicerade att skapa.
-Jag har byggt ett program som automatiskt skapar typkollare för program utifrån en kort definition.}
+\theabstract{Typkontrollerare är ett vanligt verktyg för att hitta buggar i datorprogram, men är relativt komplicerade att skapa.
+Jag har byggt ett program som automatiskt skapar typkontrollerare för program utifrån en kort definition.}
 
@@
 Datorprogram hanterar ofta många olika datatyper, och fel användning av dem är en av de vanligaste buggarna.
-För att hitta dessa buggar används typkollare, som systematisk går igenom hela programmet och verifierar varenda användning av variabler eller värden för att se till att rätt datatyp använts på rätt plats.
+För att hitta dessa buggar används typkontrollerare, som systematisk går igenom hela programmet och verifierar varenda användning av variabler eller värden för att se till att rätt datatyp använts på rätt plats.
 
@@
 
-Dessa typkollare är dock komplicerade att skapa, och är ofta starkt kopplade till ett specifikt programmeringsspråk.
-Det innebär att varenda programmeringsspråk behöver skapa sin egen typkollare, och leder till en otrolig mängd upprepat arbete.
-I mitt arbete har jag skapat ett verktyg för att generera typkollare per automatik, utifrån en kort och koncis definition av programmets regler.
+Dessa typkontrollerare är dock komplicerade att skapa, och är ofta starkt kopplade till ett specifikt programmeringsspråk.
+Det innebär att varenda programmeringsspråk behöver skapa sin egen typkontrollerare, och leder till en otrolig mängd upprepat arbete.
+I mitt arbete har jag skapat ett verktyg för att generera typkontrollerare per automatik, utifrån en kort och koncis definition av programmets regler.
 
@@
 
-Därmed kan vi enkelt skapa typkollare för nya språk på ett kortfattat vis, genom att bara definiera vilka regler som ska finnas och inte behöva fundera över hur man implementerar dem.
+Därmed kan vi enkelt skapa typkontrollerare för nya språk på ett kortfattat vis, genom att bara definiera vilka regler som ska finnas och inte behöva fundera över hur man implementerar dem.
 Det underlättar också om man vill göra förändringar i systemet, eftersom det är mycket lättare att ändra de koncisa reglerna än att ändra de många raderna kod som kan krävas för att implementera den.
 
-I nuläget är programmet ganska begränsat i hur komplicerade typkollare det kan generera, och skulle inte kunna appliceras på något av de stora, kända programmeringsspråken.
+I nuläget är programmet ganska begränsat i hur komplicerade typkontrollerare det kan generera, och skulle inte kunna appliceras på något av de stora, kända programmeringsspråken.
 Däremot klarar det flera exempel från forskningslitteraturen, och tillvägagångssättet har dock visats vara flexibelt och utbyggbart.

### 2024-12-09 presentationpics.tex
Commit final changes from way back

--- 2024-01-16 MSc.tex
@@
 
-% disable spellcheck for this section
-% LTeX: SETTINGS enabled=false
 \lstdefinestyle{mystyle}{
@@
 }
-% LTeX: SETTINGS enabled=true
-
-\setcounter{secnumdepth}{4} % Number subsubsections
-%\setcounter{tocdepth}{4} % Include subsubsections in table of contents
-
-% used to display the used files at the end. Select nofilelist as a package option to disable this
-\listfiles % initialize
-
-%\geometry{showframe}
-%better like this?
-\student{Nicholas Boyd Isacsson}{nicholas@isacsson.se}
-%\students{Flavius Gruian}{Flavius.Gruian@cs.lth.se}{Camilla Lekebjer}{Camilla.Lekebjer@cs.lth.se}
-
-\thesisnumber{LU-CS-EX: 2023-79} % Magic Number! Do not change unless Birger Swahn asks you to do so!
-% default is Master. Uncomment the following for "kandidatarbete"/Bachelor's thesis
-%\thesistype{Bachelor}{Kandidatarbete}
-
-\newcommand{\CR}[1]{\textcolor{green!60!black}{[\textbf{CR}:#1]}}
-%\newcommand{\CR}[1]{}
-
-% For questions to people reviewing, remove before finalizing
-\newcommand{\reviewquestion}[1]{\textcolor{red!80!black}{[\textbf{Review question}:#1]}}
-%\newcommand{\reviewquestion}[1]{\textcolor{red}{\begin{itemize}\item Question: #1\end{itemize}}}
-%\newcommand{\reviewquestion}[1]{}
-
-%\title{Formatting a Master's Thesis}
-\title{Type Checker Generation using Reference Attribute Grammars}
-\begin{otherlanguage}{swedish}
-\svensktitel{Infoga den Svenska titeln här!}
-\end{otherlanguage}
-
-%\onelinetitle
-%\twolinestitle
-\threelinestitle
-%\fourlinestitle
-
-%\subtitle{A {\LaTeX} class}
-%\company{The Corporation AB LTD Inc}
-%\supervisors{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}{Don Jeer, \href{mailto:djeer@xy.lth.se}{\texttt{djeer@xy.lth.se}}}
-\supervisor{Christoph Reichenbach, \href{mailto:christoph.reichenbach@cs.lth.se}{\texttt{christoph.reichenbach@cs.lth.se}}}
-%\supervisor{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}
-\examiner{Niklas Fors, \href{mailto:niklas.fors@cs.lth.se}{\texttt{niklas.fors@cs.lth.se}}}
-
-\date{\today}
-%\date{January 16, 2015}
-
-\acknowledgements{
-Sincere thanks to Christoph Reichenbach,
-whose enthusiasm for the subject inspired me to pursue this subject
-and whose guidance in matters of both implementation and writing were key to this thesis.
-Additional thanks to Niklas Fors for providing several valuable insights, greatly simplifying certain components and Alexandru Dura for providing some guidance to JastAdd internals.
-
-I would also like to thank my partner Lo-Isabel Krantz Andrée, my family, and Moritz for their support throughout the duration of this project.
-
-%If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.
-%
-%We would like to thank Lennart Andersson for his feedback on this template.
-%
-%We would also like thank Camilla Lekebjer for her contribution on this template, as well as Magnus Hultin for his popular science summary class and example document.
-%
-%Thanks also go to the following (former) students for helping with feedback and suggestions on this template: Mikael Persson, Christoffer Lundgren, Mahmoud Nasser.
-}
-
-\theabstract{
-%Your abstract should capture, in English, the whole thesis with focus on the problem and solution in 150 words.
-It should be placed on a separate right-hand page, with an additional \textit{1cm} margin on both left and right.
-Avoid acronyms, footnotes, and references in the abstract if possible.
-%Leave a \textit{2cm} vertical space after the abstract and provide a few keywords relevant for your report. Use five to six words, of which at most two should be from the title.
-
-  Static type checkers are a popular method for early detection of bugs in computer programs, by ensuring the correct data types are used throughout the program.
-  % RAGs may not be the most convenient format for typing rules
-  One approach for implementing type checkers are Reference Attribute Grammars (RAGs), a formalism for specifying the static semantics of programming languages by declaring attributes on an abstract tree-representation of the program, and is used by the JastAdd meta-compilation system.
-
-  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from a subset of possible typing rules, known as 'syntax-directed' typing rules.
-  This aims to provide integration with existing compilers implemented with JastAdd, as well as a playground for type systems usable as an educational tool.
-
-  % 'Syntax-directed' needs explaining if used, "a subset known as syntax-directed"
-  We find RAGs to be a flexible model for syntax-directed typing rules, and our software was able to support small languages from research literature.
-  It proved difficult to provide compile-time verification of the typing rules, and further work is required to extend the compiler to support type environments, a necessity for most real-world type systems.
-}
-
-%\keywords{MSc, BSc, template, report, style, structure}
-\keywords{type checking, type theory, code generation, reference attribute grammars}
-
-%% Only used to display font sizes
-\makeatletter
-\newcommand\thefontsize[1]{{#1 \f@size pt\par}}
-\makeatother
-%%%%%%%%%%
-
 \begin{document}
-\renewcommand{\bibname}{References}
-
-\makefrontmatter
-\chapter{Introduction} % should give the background of your work. Important parts to cover:
-% Give the context of your work, have a short introduction to the area.
-While writing software, developers often make mistakes.
-Catching these mistakes as early as possible can prevent these bugs from ever reaching production code.
-%\CR{Possible papers to cite as evidence:
-%  \url{https://dl.acm.org/doi/abs/10.1145/1985793.1985889}
-%  \url{https://www.usenix.org/legacy/event/sec04/tech/full_papers/johnson/johnson_html/}
-%  \url{https://ieeexplore.ieee.org/abstract/document/10190394}}
-%\CR{These papers add static checkers on top of existing languages to find additional bugs; you can also
-%  mention statically typed languages, but those generally don't require frequent changes to the type checker implementation, so they're less
-%  compelling to motivate your work.}
-Type checkers are one of the most common static checking methods, ensuring that the correct data types are used.
-While some languages include type checkers in their compilers, work has also been done to add static type checkers to existing languages.
-This requires writing large amounts of specialised code on a per language basis, but has shown to be effective at detecting bugs\cite{10.1145/1985793.1985889}\cite{10.5555/1251375.1251384}\cite{10190394}.
-Furthermore, making changes to the typing rules of a language can require extensive rewrites in many different parts of the type checker's codebase.
-
-% Define the problem you are solving (or trying to solve).
-%\CR{Prior work has already looked into pluggable type checkers.
-%  Why is translating them into JastAdd interesting?
-%  (aside from theoretical questions, you can also use
-%  pragmatic reasons, such as integration into our courses.)}
-In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of Reference Attribute Grammars (RAGs).
-
-Pacak et al. have previously implemented a compiler for generating incremental type checkers using declarative logic programming language Datalog\cite{Pacak}.
-It focused on defining a way of expressing typing rules as a series of relations solvable by the Datalog compiler.
-Our use of RAGs instead allows us to build our analysis directly onto a representation of the program's structure.
-JastAdd is also an interesting target because it has already been used to implement compilers of several languages, to which our typing rule implementations could potentially be integrated.
-It is also used extensively in courses within the Department of Computer Science at LTH, where our project could serve as an educational tool, providing students an interactive way to define and test simple type systems.
-
-Typing rules form a complex, often ambiguous, and infinite relation.
-To be able to translate them into JastAdd code, we will need to define a limited subset of syntax-directed typing relations.
-We identify a number of different kinds of typing rules, used in previous work in the field\cite{Pierce}, which share a common set of features for us to implement.
-We then devise a strategy to translate these rules into snippets of iterative Java code, which can be evaluated by JastAdd.
-
-%\CR{Make clear to the reader what the challenges are:
-%      How are these two formalisms similar, how do they differ (at a high level)?
-%      You can use a simple typing rule as an example if it helps (very optional).
-%      Hint: look at determinism in evaluation (syntax-directedness), unification, but stick to intuition in this chapter; you can elaborate in Chapter 2 or later if you think it helps.}
-
-We define a custom parsable language for typing rules adapted from natural semantics, and introduce a strategy to compile it into RAG attributes.
-Our case studies show that simple type systems can be systematically compiled into attributes on the AST, and the tree syntax provides a natural approach for recursive evaluation.
-Overall, this initial exploration shows that RAGs are a flexible approach for implementing type checkers for syntax-directed typing rules, though there are difficulties when implementing rule verification at compiler runtime.
-%\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
-%  We evaluate the approach against Foo, and demonstrate Bar.
-%  Our case studies show Quux.
-%  Overall, this initial exploration shows \ldots''}
-
-% Specify your contributions. What does this particular work/report bring to the research are or to the body of knowledge?
-\section{Research questions}
-To formulate our goals more concretely, we have set out to answer the following research questions:
-%\CR{Give the RQs numbers so you can refer to them later.  Common form: RQ1, RQ2 etc.; some authors use RQ2.3 if they split RQ2 into multiple subquestions (this can happen later).}
-\newcommand{\rqone}{RQ1: What kinds of typing rules can we translate using RAGs?}
-\newcommand{\rqtwo}{RQ2: How can we implement the translations and what algorithms should we use for the output code?}
-\newcommand{\rqthree}{RQ3: What are the challenges in translating typing rules into JastAdd-based RAGs?}
-\begin{itemize}
-  \item \rqone
-    %\CR{``kinds'' of typing rules: right now the reader isn't aware that there even are different kinds (see my earlier comment on ``what the challenges are''-- sketching this out should suffice.)}
-  \item \rqtwo
-    %\CR{What does ``form'' mean, here?  How broad is ``implement''-- are you talking algorithms (which I think you are) or including technical challenges (which I think you have in the next RQ)?}
-  \item \rqthree
-    %\CR{I'd go with `challenges in translating' instead of `potential issues of translating', language-wise}
-\end{itemize}
-
-\section{Chapter outline}
-%\CR{%paper $\to$ thesis; also,
- % this always comes at the very end of the introduction}
-The rest of this thesis is outlined as follows.
-% This chapter introduces the
-Chapter 2 presents the necessary background information for the report.
-Chapter 3 describes the design and implementation of our program, followed by demonstrations of its results in chapter 4.
-Chapter 5 discusses the results in a more general sense, before the conclusion in chapter 6.
-
-\chapter{Background}
-%\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
-%\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinking ``I know why I want to read this''}
-This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
-Section \ref{abstractsyntaxtrees} explains abstract syntax trees, a simplified tree-based representation of a program, suitable for performing static analysis on.
-These are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
-%\CR{Consider saying just a little more about them here, for readers who have no idea what ASTs are and why they matter for
-%    analysing programs.}
-Section \ref{typetheory} details type systems, by explaining each of the component pieces, followed by a description of their use within static type checking.
-
-\section{Abstract syntax trees}\label{abstractsyntaxtrees}
-The syntax of a language is the definition of the symbols and structures which make up a valid program.
-%\CR{What's a syntax?}
-An abstract syntax tree (AST) is a simplified tree-based representation of the syntax.
-Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
-It is abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
-Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this semantic information.
-
-This makes it easier to write analysis based on the behaviour of a program, without concerning oneself with its exact formulation.
-Several different formulations of a program with identical behaviour may be abstracted into identical ASTs.
-The hierarchical structure of the tree lends itself especially well for recursive analyses, and eases the passing of information between connected structures by connecting them as parent and children nodes.
-%\CR{what is ``this information'', concretely?}
-A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if node, with named branches pointing towards the condition and the two values.
-%\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
-%\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}
-
-%\CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}
-%\begin{figure}[h]
-%\begin{lstlisting}[]
-%Expr   -> Expr "+" Term | Term
-%Term   -> Term "*" Factor | Factor
-%Factor -> Int | "(" Expr ")"
-%\end{lstlisting}
-%  \caption{\reviewquestion{One like this? I'm unsure how to tie it in effectively.}}
-%  \label{typesyntaxspec}
-%\end{figure}
-
-\section{Type theory}\label{typetheory}
-In this section we will discuss the elements of type theory required to understand this thesis.
-Firstly, we will discuss what types are and give some examples of their use.
-Next, we will describe type systems, which form a relation between types and expressions.
-
-\subsection{Values and expressions}
-The main purpose of most computer software is to handle and manipulate data.
-Each of these individual pieces of data is a value, which could be anything from the number 4 or the text `Hello!'.
-
-Expressions are a broader term, indicating any syntactic structure which evaluates to a value.
-This includes but is not limited to values, which evaluate to themselves, but also constructs of values, operators or function calls, etc.
-
-\subsection{Data types}
-%\CR{think about the connection between types and values here.  You may even want to talk about values in a separate subsection.}
-Data types can be considered a categorisation of values which share common properties, such as supporting numerical operations or accessing a certain element in a list of characters.
-While expressions may be arbitrarily complex, nested structures, by assigning it a type, we can perform useful analysis on it, without having to evaluate its value.
-%\CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
-%  For each concrete (and possibly very complex) operator on the left-hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
-%  This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
-%  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractions over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
-%}
-
-%several different types of data in one program, such as numbers or text.
-Most programming languages come with a number of built-in data types, from different kinds of numbers (integers, floating point) to strings or lists and arrays.
-
-%\CR{if you discuss values, you can use the intuition of ``grouping values'' by certain properties (such as supported operations) to make your point clearer.}
-Different parts of programs may have different constraints on the types of expressions, such as a print function requiring a string of text, or a plus function requiring some kind of number.
-%\CR{Very good point about types of expressions, but again, this is likely easier to understand by linking expressions and values.
-%  I strongly recommend using a definition of the form ``all values are expressions, but not all expressions are values''.}
-
-\subsection{Type checking}
-Verifying that these constraints are upheld can assist us in catching programming mistakes, preventing the wrong data types from being used in the wrong places, which may have lead to erroneous or unpredictable behaviour.
-There are two main approaches to ensuring the constraints are upheld, dynamic and static type-checking.
-As our project generates a static type checker, we will focus our explanations on that topic, but provide a quick overview of their differences here.
-%\CR{Why would we even want to ensure that they are upheld?}
-
-Dynamic checking is done performed runtime, with each type checked only when needed.
-While this means no checks are performed unnecessarily, it makes it harder to verify the constraints over the whole program, as it won't be tested until each specific part is executed.
-
-Static checking is performed on the code before it is run, usually as part of compilation, preventing many faulty programs from being run at all.
-However, since it lacks knowledge of what happens at runtime, it's required to take a more conservative approach, flagging potential errors even in code which will never executed.
-Static checkers analyse the structure of the source code, often in the form of an abstract syntax tree.
-
-%\CR{Why would I want static checking?  Dynamic only checks when needed, isn't that more efficient?}
-
-\subsection{Type systems}
-A type system typically consists of two parts, a type syntax, describing the available types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
-%\CR{Why do the types need a syntax?
-%    (There are other approaches to types; for your own sanity, I recommend deferring to Pierce here and calling this ``typical approaches'')}
-
-\subsubsection{Type syntax}\label{typesyntax}
-We're using the term `type syntax' to mean a definition of the valid types within a language, often referred to simply as the set of types.
-%\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
-%\reviewquestion{I believe Pierce simply refers to it as the "set of types".
-I seem to have got the term type syntax from EDAP15 \url{https://fileadmin.cs.lth.se/cs/Education/EDAP15/2022/web/skills/type-system.html}.
-In my opinion it works better to have a clear name for it, so I'm keeping it for now, unless you have further thoughts.}
-It can consist of fundamental types, such as integers or booleans.
-There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
-This enables differentiating lists that store integers from those that store strings, and ensuring that any item added to a list must be of the same type as the type parameter.
-%\CR{A bit vague; consider giving an example.}
-\begin{figure}[h]
-\begin{lstlisting}[]
-abstract Type;
-
-Bool : Type;
-Int  : Type;
-\end{lstlisting}
-  \caption{Type syntax consisting of two types, for boolean and integer values, written as a JastAdd AST specification (see Section \ref{astdef})}
-  \label{typesyntaxspec}
-\end{figure}
-
-\subsubsection{Typing rules}
-The typing relation, notated \lstinline{t : T}, is the link between the expressions of a language and the types of the type syntax.
-%\CR{It is \emph{the} link, even.}
-The left-hand side is an expression in the object language, typically written in its native syntax, the right-hand side a type from the aforementioned type syntax.
-
-Typing rules are a manner of inference rules, which in their simplest form, consist of no more than a typing relation.
-The typing rule \lstinline{T-True} seen in Figure \ref{typingruleexampletrue}, for example, simply assigns the type of \lstinline{Bool} to the expression \lstinline{true}.
-Additionally, the typing rules in these examples have a name, given in parentheses next to the rule, though this is merely for reference.
-
-\begin{figure}[h]
-  \begin{equation}
-    \tag{T-True}
-    \text{true} : \text{Bool}
-  \end{equation}
-  \caption{Typing rule for a true value}
-  \label{typingruleexampletrue}
-\end{figure}
-
-Typing rules may also have premises, a number of conditions that must hold for the typing rule to apply.
-%\CR{\emph{that} must hold (``which'' relates a set to a superset: ``carrots, which are vegetables'', while ``that'' restricts a set to a subset: ``vegetables that are carrots'')}
-The notational convention here is to put the premises above a bar, with the conclusion underneath it.
-
-\begin{figure}[h]
-  \begin{equation}
-    \tag{T-If}
-    \frac{
-      t_1 : \text{Bool}\quad
-      t_2 : \textsf{T}\quad
-      t_3 : \textsf{T}
-    }{
-      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \textsf{T}
-    }
-  \end{equation}
-  \caption{Typing rule for an if-expression}
-  \label{typingruleexampleif}
-\end{figure}
-
-\newcommand{\ty}[1]{\textcolor{blue!20!black}{\textsf{#1}}}
-\newcommand{\tyvar}[1]{\textsf{#1}}
-Figure \ref{typingruleexampleif} shows an example rule, typing an if-expression.
-The bottom line, the conclusion, contains the keywords of the if-expression and three free variables $t_i$ which represent any term within the language.
-Above the bar are three premises, placing constraints on the three variables, which must hold for the typing rule to apply.
-The first constrains $t_1$ to the type bool, the second two constrain $t_2$ and $t_3$ to a type variable \tyvar{T}, which represents any single arbitrary type within the type syntax.
-While \tyvar{T} may be any type, it must consistently be the same type within any single evaluation of the rule, i.e. the two terms must have the same type.
-This \tyvar{T} also appears in the conclusion, as the resulting type of the if-expression, meaning the result of the entire expression will be that of the $t_2$ and $t_3$ terms.
-%\CR{Consider typesetting T differently, e.g. as \tyvar{T}}
-%\CR{\tyvar{T}, $t_1$, $t_2$, $t_3$ are \emph{free} variables here, logically speaking}.
-
-\subsubsection{Evaluating typing rules}
-While performing static type checking, the typing rules form the basis of our analysis.
-By analysing the program, we attempt to match the syntactic form with the typing rules, to find the type of the expression.
-If no matching typing rules can be found, a type error is produced.
-
-The simplest program possible with the typing rules T-True and T-If, consists of the single value \lstinline{true}.
-It can immediately be matched to the rule T-True, giving the type \lstinline{Bool}.
-
-A slightly more complex case can be expressed as \lstinline{if true then true else true}.
-The only rule that matches the structure, is T-If, with \lstinline{true} values taking the place of each of the $t_i$ variables.
-To evaluate this rule, we are thus required to recursively evaluate the types each of our $t_i$ variables, to ensure they match the premises.
-In this case, the type of $t_1$ evaluates to \lstinline{Bool}, matching the first premise of T-If.
-The types of both $t_2$ and $t_3$, represented as the type variable \tyvar{T}, also evaluate to \lstinline{Bool}.
-Since both the uses of this type variable within the premises are consistent, all three of the rule's premises have been fulfilled, and the entire expression can be typed as \tyvar{T}, which has been found to correspond to \lstinline{Bool}.
-
-\begin{figure}[h]
-  \begin{equation}
-    \tag{T-Zero}
-    \text{zero} : \text{Int}
-  \end{equation}
-  \caption{Typing rule for a zero value}
-  \label{typingruleexamplezero}
-\end{figure}
-
-Adding a second value of a different type allows us to express incorrectly typed programs, for example the program \lstinline{if zero then zero else true}.
-Recursively evaluating this according to T-If, we find that the premise concerning $t_1$ no longer upholds, as the type of it now evaluates to \lstinline{Int}.
-Additionally, the type variable \tyvar{T} typed by both $t_2$ and $t_3$ now has contradictory use, being assigned both as the type \lstinline{Int} and \lstinline{Bool}.
-As this program cannot be found to satisfy any set of typing rules in our type system, it is found to be invalid.
-
-%\CR{Please explain the purpose of typing rules in static type checking.
-Using just your two examples you would be able to show a well-typed program, and if you add one more rule (e.g.
-for integers), you can show a type error.
-No need to build the full derivation tree, but it's important to mention recursion here.
-This entire bit may be worth a separate subsubsection.}
-
-\section{Reference Attribute Grammars}
-This section introduces the subject of Reference Attribute Grammars (RAGs), first explaining the theory behind them, then introducing JastAdd, an implementation of RAGs within a Java meta-compilation system\cite{JastAdd}.
-
-\subsection{Reference Attribute Grammars}
-Reference Attribute Grammars are an extension to Attribute Grammars (AGs), a formalism used for specifying the static semantics of programming languages.
-%\CR{Generally only for the static semantics (RAGs can also do dynamic semantics, if you allow circular attributes, but that's not worth mentioning here.)
-%Please also connect ``semantics'' to ``type checking'' (via ``static semantics'')}
-AGs provide a systematic way to define computations on the nodes of an abstract syntax tree, associating attributes with language constructs and are commonly used in the context of compiler construction, where they help define the translation of source code into executable code\cite{AGs}.
-
-RAGs extend these properties by introducing a new kind of attribute, allowing references to other nodes in the AST, and through the references accessing the attributes of other nodes.
-This provides a simplified and more flexible method for interconnectedness between nodes\cite{RAG}.
-
-%\subsection{Non-Terminal Attributes}
-%Our project also uses not\cite{higherorderattributegrammars}
-%\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are not a necessary feature for RAGs.
-%Please introduce them in a separate section: you can very directly connect them to the idea of ``type syntax''.}
-
-\subsection{JastAdd}
-JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion.\cite{JastAdd}
-Our project uses it both for the implementation of our compiler and as a compilation target, as it outputs JastAdd code.
-%\CR{It's worth telling the reader here why you are explaining some details about JastAdd.}
-
-\subsubsection{AST specification}\label{astdef}
-One of the key components of a JastAdd projects are the \lstinline{.ast} files, which declare the structure of the abstract syntax tree.
-This specification is then used by JastAdd to generate Java classes representing the nodes of the AST.
-These AST classes can have a variety of forms, supporting both abstract classes and subclasses, and can contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.
-For a detailed overview of the syntax, see the reference manual\cite{jastaddref}.
-
-\subsubsection{RAG Modules}
-\begin{figure}[h]
-\begin{lstlisting}[language=jrag]
-aspect TypingRules {
-  syn Type Zero.type() {
-    return new Int();
-  }
-}
-\end{lstlisting}
-  \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
-  \label{jragexample}
-\end{figure}
-The RAG modules are written in \lstinline{.jrag} files, declare the attributes to be included in the classes generated by the AST specification.
-Figure \ref{jragexample} shows a minimal example.
-The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.
-
-\chapter{Approach/Design} % should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:
-
-%\CR{Missing: introduction to this chapter'}
-We begin this chapter with a section presenting our language for writing typing rules, an adaptation of the natural semantics used to notate typing rules in the academic context.
-In Section \ref{sectionarchitecture} we proceed with presenting the overall architecture of our software, including an overview graph in Figure \ref{overviewgraph}, which may be useful to bear in mind while reading the rest of the report.
-Section \ref{sectionimplementation} goes into the details of our implementation, how we generate code for different parts of the typing rules.
-
-\section{Language design}\label{sectionlanguage}
-\begin{figure}[]
-\begin{lstlisting}[]
-ruleset = rule_list;
-
-rule_list =
-    rule
-  | rule_list rule;
-
-rule =
-    '[' uppercase_id ']' premises_list '---+' formula
-  | '[' uppercase_id ']' formula;
-
-premises_list =
-    formula
-  | premises_list ',' formula;
-
-formula = term ':' tyterm;
-
-term =
-    uppercase_id '(' term_list ')'
-  | uppercase_id '(' ')'
-  | uppercase_id
-  | lowercase_id;
-
-tyterm =
-    uppercase_id
-  | lowercase_id;
-
-term_list =
-    term
-  | term_list ',' term;
-
-uppercase_id = "[A-Z][A-Za-z0-9-]";
-uppercase_id = "[a-z][A-Za-z0-9-]";
-\end{lstlisting}
-  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to BNF.}
-  \label{trconcretesyntaxspec}
-\end{figure}
-%\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
-%  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
-%  This would be before the current 3.1, then.
-%  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
-%  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
-%  of the object language as well as of the types.}
-
-\begin{figure}[]
-\centering
-\begin{minipage}{.45\textwidth}
-  \centering
-  \begin{equation}
-    \tag{T-True}
-    \text{true} : \text{Bool}
-  \end{equation}
-  %\caption*{Pierce syntax}
-  %\label{fig:sub1}
-\end{minipage}%
-\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
-  \centering
-\begin{lstlisting}[]
-[T-True]
-True : Bool
-\end{lstlisting}
-  %\caption*{ASCII representation}
-  %\label{fig:sub2}
-\end{minipage}
-\caption{A typing rule for the value true in natural semantic syntax on the left and its equivalent ASCII representation on the right}
-\label{trexampletrue}
-\end{figure}
-\begin{figure}[]
-\centering
-\begin{minipage}{.45\textwidth}
-  \centering
 \begin{equation}
@@
 \end{equation}
-  %\caption*{Pierce syntax}
-  %\label{fig:sub1}
-\end{minipage}%
-\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
-  \centering
-\begin{lstlisting}[]
-[T-If]
-t1 : Bool,
-t2 : t,
-t3 : t
------------------
-If(t1, t2, t3): t
-\end{lstlisting}
-  %\caption*{ASCII representation}
-  %\label{fig:sub2}
-\end{minipage}
-\caption{A typing rule describing an if-expression in natural semantic syntax and its equivalent ASCII representation}
-\label{trexampleif}
-\end{figure}
 
-%- Figure of example natural deduction rules (from Pierce?) and ASCII representation
-%No infix notation
-%No environments
-
-To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{Pierce}.
-Our language can express a subset of the notation used by Pierce, with some differences.
-
-The largest difference lies in the notation used for the expression to be typed, where Pierce uses the object language's native syntactic form, whereas our representation is based on the corresponding AST nodes.
-Node names are consistent with those used by JastAdd, followed by parentheses containing variable names to denote their child nodes in the order of their definition.
-This change is necessary to allow us to type arbitrary languages without needing to understand their syntax and grammar, though it does make translating rules from one syntax to the other less straightforward.
-
-Another difference is the use of the type variables, which have been made lowercase.
-This disambiguates them from the concrete types, which as AST nodes themselves are written with uppercase letters, and also aligns the syntax for type terms with that of regular terms, where concrete nodes are written uppercase and variables in lowercase.
-\lstinline{T}, which has been switched to the lowercase \lstinline{t} in order to distinguish concrete types and type variables.
-This disambiguates them from the concrete types, which as AST nodes themselves are written with uppercase letters, and also aligns the syntax for type terms with that of regular terms, where concrete nodes (such as \lstinline{If}) are written uppercase and variables (\lstinline{t1}, \lstinline{t2}, \lstinline{t3}) in lowercase.
-%\CR{I generally recommend avoiding the passive voice, but the use below is an example that would be OK}
-Commas are also used to separate the premises, and the horizontal line separating premises from the conclusion is represented as a series of dashes.
-
-%\CR{Optional: you can mention Alexandru's paper about how one could add metavariables into ``native'' syntax: \url{https://dl.acm.org/doi/pdf/10.1145/3485542}
-%  (since this part has already been done, it's not something you needed to explore)}
-%\CR{Capitalising the AST node types is consistent with their naming in JastAdd-- worth pointing out.}
-
-%\CR{either forward reference to the discussion of the limitations, or remove the ``limitations'' bit here (you already say ``subset'', after all) and discuss those limitations after you define the language.}
-
-Figures \ref{trexampletrue} and \ref{trexampleif} show two rules from Pierce in both their original syntax and our ASCII representation side by side.
-The former figure describes a simple rule named T-True, which assigns the \lstinline{true} expression the type \lstinline{Bool}.
-
-%\CR{Given that it's nontrivial to separate out the different changes, I recommend first enumerating the differences abstractly and then discussing how the changes manifest in the figures, rather than starting with the figures to ``derive'' the differences.}
-Looking at the notation of the typed expressions in Figure \ref{trexampletrue}, the difference is minor, the syntactic term \lstinline{true} has been replaced by the name of the AST node \lstinline{True}.
-The notational change is more apparent in Figure \ref{trexampleif}, where instead of the full native syntactic form, the expression is written as an AST node with its children represented as parameters.
-%Note the difference between the \lstinline{true} written in Pierce, representing the native syntactic form of the expression, and the \lstinline{True} in the ASCII representation, representing the name of the corresponding AST node.
-%This difference is more apparent in Figure \ref{trexampleif}, where instead of the full syntactic form, the expression is written as an AST node with its children represented as parameters.
-
-%\CR{Important: this affects numbers used at the type level only (which is an advanced feature mostly used in undecidable
-%  typesystems like the Calculus of Constructions), and it wouldn't matter here since you can't use \ty{Succ} as a parametric
-%  type anyway (cf. igure 3.4).
-%  My problem is that right now this reads as if you can't typecheck programs with Integer literals,
-%  but those would be e.g. \textsf{Integer} AST nodes, which (to my reading) you would have no trouble writing typing rules for.
-%  (you never need to touch the actual lexeme that contains the character string that encodes the integer, in traditional
-%  type analysis).  So you shouldn't actually have any real-life problem here.}
-%Numerical values have not been implemented, but natural numbers can be expressed using Peano numbers\cite{peano}, utilising a zero value and successor function which represents the number incremented by one.
-%If a language has the syntax \lstinline{0 + 1} it may instead be represented as \lstinline{Add(Zero, Succ(Zero))}.
-
-\section{Architecture}\label{sectionarchitecture}
-In this section we will be presenting our architecture.
-%Since it contains many different components, it may be helpful to refer to the overview graph in Figure \ref{overviewgraph} while reading.
-Figure \ref{overviewgraph} provides an overview of all the components and the connections between them.
-The rest of the section goes into further detail about individual components.
-
-%\CR{Stylistically, technical documents should avoid discussing the process of reading (it's not as frowned upon as speaking directly to the reader, though), so I recommend just saying that the figure ``provides an overview over...''}
-%\CR{Please put the following in the caption of th/e figure instead.}
-%It uses rectangular boxes to represent the concrete data or artefacts of the procedure, and elliptical boxes to represent the actions performed with them.
-
-%\CR{You still want an introduction to this section that explains what is following and why}
-
-\begin{figure}[]
   \resizebox{\textwidth}{!}{%
@@
     ]
-    \node (typerules) [userinput]                    {Typing rules\\definition (\ref{typingruledefinition})};
+    \node (typerules) [userinput]                    {Typing rules\\definition };
     \node (trparsing) [right=of typerules, process] {Scan,\\parse,\\build};
-    \node (trparser)  [above=of trparsing, artifact]          {Typing rule parser\\(\ref{typingrulesparser})};
-    \node (trast)     [right=of trparsing, generated]          {Typing rules AST\\(\ref{typingrulesast})};
+    \node (trparser)  [above=of trparsing, artifact]          {Typing rule parser};
+    \node (trast)     [right=of trparsing, generated]          {Typing rules AST};
 
@@
     \node (typecheckgen)       [below=of trast, process] {Generate\\type checker};
-    \node (typingrulecompiler) [right=of trast, artifact]          {Typing rule\\compiler (\ref{sectionimplementation})};
-    \node (typechecker)        [below=of typecheckgen, generated]   {Type checker \\(\ref{typecheckingalgorithm})};
+    \node (typingrulecompiler) [right=of trast, artifact]          {Typing rule\\compiler };
+    \node (typechecker)        [below=of typecheckgen, generated]   {Type checker };
 
-    \node (olast)     [below=of typerules, userinput] {AST specification\\(\ref{astdef})};
+    \node (olast)     [below=of typerules, userinput] {AST specification};
 
@@
     \node (outputparsing)      [left=of inputast, process]     {Scan,\\parse,\\build};
-    \node (typeerrors)         [right=of outputtypechecking, generated]   {Type errors\\(\ref{errorhandling})};
+    \node (typeerrors)         [right=of outputtypechecking, generated]   {Type errors};
     \node (inputprog)          [left=of outputparsing, artifact]         {Input program};
@@
     %    \item Project files
-    %    \item Type checking algorithm (\ref{typecheckingalgorithm})
+    %    \item Type checking algorithm 
     %  \end{itemize}
     %};
-    \node (templateparser)        [below=of outputparsing, artifact]                               {Tree-based\\parser\\(\ref{treebasedparser})};
-    \node (templateprojectfiles)  at (inputast |- templateparser)           [artifact]           {Project files\\(\ref{projectfiles})};
-    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser) [artifact]           {Type checking\\aspect (\ref{typecheckingaspect})};
-    \node (templatetypesyntax)    [right=of typecheckingalgorithm, artifact]                       {Type syntax\\AST (\ref{typesyntax})};
+    \node (templateparser)        [below=of outputparsing, artifact]                               {Tree-based\\parser};
+    \node (templateprojectfiles)  at (inputast |- templateparser)           [artifact]           {Project files};
+    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser) [artifact]           {Type checking\\aspect };
+    \node (templatetypesyntax)    [right=of typecheckingalgorithm, artifact]                       {Type syntax\\AST };
     \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm) (templatetypesyntax), artifact] {};
@@
   }
-  \caption{An overview of the project components and flow.
-It represents the flow of the program, with components and artefacts (represented as rectangular boxes) with arrows linking them to actions (elliptical boxes) which in turn have arrows to the new components and artefacts they produce.
-The dashed arrows from represent links from the support library components that exist in the current experimental state of the project, but in a finished product would instead link to component from the object language's own compilation environment.
-  %\CR{Can you mark, e.g. via colours, the parts that the user provides, and the parts that will be in the generated output project?}
-  }
-  \label{overviewgraph}
-\end{figure}
 
-\subsection{Typing rule definition}\label{typingruledefinition}
-To express typing rules, we created a simple ASCII representation of the natural deduction notation used by Pierce\cite{Pierce}.
-Custom DSL
-Include typing rules AST specification (or BNF?)
-ASCII representation of natural deduction %BNF (is it BNF?)
-
-\subsubsection{Typing rules parser}\label{typingrulesparser}
-The parser for typing rules is written using JFlex\cite{jflex}, a scanner generator, and Beaver\cite{beaver}, an LALR parser generator.
-JFlex takes a series of regular expressions to generate a tokenizer for the parsed language.
-The tokens are passed on to Beaver along with a parser definition in extended Backus-Naur form (EBNF), to generate a parser.
-
-\subsubsection{Typing rules AST}\label{typingrulesast}
-%\CR{I would give this AST/BNF a more specific name here, to distinguish it from the other two (as done nicely already in Figure 3.1)}
-\begin{figure}[h]
-\begin{lstlisting}[]
-RuleSet ::= Rule*;
-
-Rule ::= <Name> Conclusion:Formula Premises:Formula*;
-
-abstract Formula;
-HasType : Formula ::= Expr:Term Ty:TyTerm;
-
-abstract Term;
-Function : Term ::= <ID> Term*;
-Value    : Term ::= <ID>;
-
-abstract TyTerm;
-TyVal : TyTerm ::= <ID>;
-TyVar : TyTerm ::= <ID>;
-\end{lstlisting}
-  \caption{Typing rules abstract syntax tree specification}
-  \label{trastspec}
-\end{figure}
-%\CR{The syntax doesn't allow parametric types; please make that explicit here.}
-
-The abstract syntax tree for the typing rules language consists of a \lstinline{RuleSet} root node.
-The root node contains only a list of rules, where each rule has a name, a conclusion and a list of premises.
-The rule's name is currently unused but is intended to be utilised in error reporting, to clearly specify what rules are available for each node, or which rule is causing an error.
-
-Both conclusion and premises are represented by the abstract HasType node, consisting of a term, which may be a function or value; and a typeterm, representing either a type value or a type variable.
-For the conclusion, this node represents what type should be assigned to the term on the left, whereas for the premises, the node represents a requirement for the node on the left to have the type on the right.
-
-Currently, our compiler does not support type parameters, so they are not included in the AST.
-To support this in future, an additional \lstinline{TyTerm} could be added which includes a list of \lstinline{TyTerm}s along with its ID.
-
-\subsection{Support library}
-In addition to the generated type-checking files, there are a number of files that are added to the output file that are identical for all generated type checkers, which we refer to as the support library.
-Some of these are key components of the type checker, but others are dummy components included solely for convenience of testing.
-We expect that in a real world use case, our software would integrate into an object language's compiler pipeline, and utilise its own parser and project files to provide the abstract syntax tree for our type checking to tie into.
-For testing purposes, we have written generic versions of these components, which work for any supported object language.
-%\CR{``Templates'' are usually things that are mostly fixed but have holes to be filled in.  I would call what you describe here the ``support library''}
-
-\subsubsection{Input program parser}\label{treebasedparser}
-%\CR{Parser for what?}
-%\CR{Is there a conection between this section and Section~\ref{typingrulesparser} wrt parsing terms of the object language?}
-For testing purposes, our support library comes with a generic tree-based parser.
-Programs are written as a tree of AST node names followed by potential parenthesis within which a comma separated list of child nodes are written.
-The parser parses the tree recursively, using Java's reflection API to find a class with a matching name, and a constructor with parameters of the same amount and types.
-Figure \ref{boolinputexample} shows an example input file for the Bools language, which will be discussed further in Section \ref{boolsast}.
-
-\begin{figure}[h]
-\begin{lstlisting}[]
-Expression(
-  Or(
-    True,
-    Or(
-      True,
-      False
-    )
-  )
-)
-\end{lstlisting}
-  \caption{Example tree-representation input for the Bools language}
-  \label{boolinputexample}
-\end{figure}
-
-This approach does have the issue of occasionally parsing invalid programs, due to the reflection API finding an additional parameterless default constructor.
-This leads the parser to always accept nodes written with no children, such as \lstinline{Expression(Or())}, even if the AST specification requires them.
-These children will be null, causing issues further down the line, though as this parser is only intended for testing it is of little concern.
-%\CR{Does this also apply to the analysis of typing rules?  If so, you can mention it as ``future work''
-%  or describe how it could be fixed (it's not a fundamental limitation, so nothing that warrants a deeper discussion)}
-
-\subsubsection{Project files}\label{projectfiles}
-For ease of use the compiler will output a complete runnable Gradle project.
-It can parse simplified tree-representations of programs, and utilise these to perform the type checking.
-In a future real world use case, it is intended for our generated type-checking aspects to integrate into a pre-existing project, which already has all the necessary project files to parse an AST tree, to which our type aspects could integrate.
-
-%\CR{Can you elaborate on if/how this would integrate with an existing language implementation?
-%  I've always interpreted everything as that you start with some `objectlanguage.jar' file of an existing language (given that you use reflection), and you get out a fresh Gradle project that uses that `objectlanguage.jar' for parsing and AST construction, but since you generate JastAdd source code, you can't really (to my understanding) add the `type()' attribute etc.
-to the existing implementation, since JastAdd needs to run over the source jrag files.
-%  Going by your overview figure, I now wonder if the generated file also incorporates all of the source code of the original project?
-Or do you reconstruct the AST specification via reflection somehow and simply mirror it in the generated project (which would lead to two different AST representations, but with a straightforward mapping)?}
-
-%\subsubsection{Executable}
-
-%\subsection{Typing rule compiler}\label{sectionimplementation}
-
-%\subsection{Output type checker}
-%\subsubsection{Tree-representation}
-%\subsubsection{Type checking algorithm}\label{typecheckingalgorithm}
-%Generate code for each node and typing rule
-%Simple algorithm
-%Recursive
-
-%\section{Method (remove this?)} % describe how you go about solving the problem you defined. Also how do you show/prove that your solution actually works, and how well does it work.
-
-\section{Implementation}\label{sectionimplementation} % if your work involved building an artefact/implementation, give the details here.
-Note, that this should not, as a rule, be a chronological description of your efforts, but a view of the result.
-There is a place for insights and lamentation later on in the report, in the Discussion section.
-
-%\subsection{Type checker generation}
-%The type checker generation is based on a single-responsibility principle, where each node of the typing rule AST is responsible for generating the code it corresponds to \CR{Not sure what you mean}.
-In this section we present our approach to translating typing rules.
-We start by presenting pseudocode for our typing rule evaluation.
-%Section \ref{typecheckingalgorithm} presenting a p the algorithm we use to evaluate typing rules in pseudocode.
-This is followed by Section \ref{codegeneration}, which explains how we generate  code to implement this algorithm.
-
-\subsection{Type-checking algorithm}\label{typecheckingalgorithm}
-
-%For each rule:
-%\begin{enumerate}
-%  \item For each parameter in the conclusion:
-%    \begin{enumerate}
-%      \item Declare a variable by the same name, initialised by fetching the child of the same index.
-%    \end{enumerate}
-%  \item For each type variable used in the rule:
-%    \begin{enumerate}
-%      \item Initialise it to null
-%      \item Iterate through each premise it appears on the right-hand side of:
-%        \begin{enumerate}
-%          \item If the type variable is set to null, evaluate the type of the left-hand side and assign it to the type variable.
-%          \item If the type variable is not set to null, evaluate the type of our left-hand side and check if it is the same type or a subtype to the type variable. If it is not, throw a type error.
-%        \end{enumerate}
-%    \end{enumerate}
-%  \item Check whether each of the premises is true:
-%    \begin{enumerate}
-%      \item Evaluate the type of the left-hand side of the premise and check if it is the same type or a subtype to the right-hand side.
-%    \end{enumerate}
-%    If all true: Return the right-hand side of the premise
-%  \item Throw a type error if we reach this point
-%\end{enumerate}
-
-\begin{figure}[]
-%foreach parameter in rule.conclusion.parameters:
-%  declare var "parameer.name" = getChild(parameter.index)
-\begin{lstlisting}[]
-foreach typevar in rule.typevars:
-  typevar.type = null;
-  foreach premise in rule.premises where premise.rhs == typevar:
-    if typevar.type == null:
-      typevar.type = premise.lhs.type()
-    else if typevar.type != premise.lhs.type():
-      throw typeerror("Type variable mismatch")
-
-// Check if all evaluated left-hand side types match their right-hand side
-if allTrue(rule.premises.map(p => p.lhs.type() == p.rhs)):
-  return rule.conclusion.rhs
-else:
-  throw typeerror("Typechecking failed")
-\end{lstlisting}
-  \caption{Pseudocode for the implementation of the type() attribute}
-  \label{psuedotype}
-\end{figure}
-
-Here we aim to give an overview of the algorithm used to evaluate the typing rules.
-Figure \ref{psuedotype} contains a pseudocode algorithm for the evaluation of a typing rule.
-It is not representative of the actual generated code, but captures its behaviour in the manner of an interpreted language.
-
-\subsection{Code generation}\label{codegeneration}
-
-In this section we will present the code we generate to implement the previously discussed algorithm.
-Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
-From the root node, calls propagate down the tree to fill in increasingly specific sections of the code.
-
-%This section details the implementation of our type checker generator, on a feature by feature basis.
-
-\subsubsection{Aspect generation}
-The \lstinline{RuleSet} node is responsible for generating the overall structure of the output file, a JastAdd aspect called \lstinline{TypingRules}.
-It sorts its \lstinline{Rule} children by which nodes they apply to, so that if multiple tying rules type the same node, they can be inserted into the same method implementation (though this feature is not yet fully working).
-
-\subsubsection{Method implementation overview}
-The implementation of a typing rule can range from a single return statement to a system of declarations of variables, type variables and control flow.
-These are explained in further detail in subsequent sections, but here follows a quick overview.
-
-Firstly, the algorithm checks if the number of premises is zero, in which case it immediately generates the conclusion, consisting of a return statement.
-If there are premises, it first generates code for the declarations of variables and both declarations and unification for type variables.
-This is followed by an if-statement to test whether the premises hold true, containing a return statement.
-Should the premises not hold, a type error is produced.
-%\CR{What if there are multiple rules that apply to the same AST node?  E.g., one rule that allows \ty{Int} addition and one that allows \ty{String} addition.}
-
-\subsubsection{Parameters}
-When the typing rules concern a node with parameters, these can be named arbitrarily in the typing rule.
-%\CR{Who does the naming?  Why is the ``arbitrarily'' bit important?}
-%\CR{ Will this work if the type system specification has a type variable called ``this'' or ``null'' or ``class''?}.
-A variable with the same name will be instantiated at the beginning of the method, fetching the corresponding child node by index.
-This naming format means any errors reported will use the same variable names as the user has defined in the typing rules, in order to minimise the disconnect when analysing errors from the generated code.
-However, it also means it is not possible to name a parameter any of the reserved keywords in Java.
-This could be alleviated by adding a prefix to the variable names, at the cost of a slight indirection for the user.
-These variables are then utilised in later premises or type variable declarations.
-
-\subsubsection{Type variables}
-In the case of type variables, much of the logic is handled at runtime, to simplify the code generation.
-Each type variable that appears in the conclusion is instantiated to a null value, and the code generated by each premise first checks if the value is still null and either assigns the variable its own type or tests to make sure the existing type matches its own.
-
-The generated code could be simplified in future work by instantiating the type variable to the type of the left-hand side of the first premise it appears in, and then only checking if each of the subsequent uses match it.
-
-\subsubsection{Premises/control flow}
-For rules containing premises, the return statement is protected by an if-statement.
-At this point we've declared variable for each of the child nodes and type variables used, so the premises can very directly be translated into boolean expressions.
-A premise \lstinline{x : Bool} translates into the expression \lstinline{x.type().matches(new Bool());}.
-This includes a call to the \lstinline{type()} of each left-hand side of the premise, recursively evaluating the necessary types.
-Since our typing rules can only reference their direct children, this recursion will always progress down the program's AST, ensuring termination.
-Each of the boolean expressions then joined by the \lstinline{&&} operator into a single if-statement, protecting the return of the right-hand side of the conclusion.
-For the case where any of the boolean expressions is false, the if-statement is followed by a thrown exception representing a type error.
-
-%\CR{Overall, I am not convinced that this strategy for presenting the type checking algorithm is very effective (though it might be better for someone who is reading the source code at the same time).
-%  I would propose to merge 3.2.2, 3.2.3, 3.2.4, 3.2.5, and to split them into (a) the type checking algorithm (while ignoring the code generation bit), and (b) the code generation aspect.
-%  I.e., for (a), pretend that you are interpreting the typing rules instead of compiling them, and please try to present an algorithm that captures
-%  the check.
-%  The algorithm can be in plain English with bullet points / numbered sub-steps, or in some pseudo-formal notation (``foreach premise $p$ in rule $r$ ...''), as long as it is clear.}
-
-\subsection{Type checking aspect}\label{typecheckingaspect}
-The type checking aspect contains the declaration of the \lstinline{type()} attribute for all AST nodes, as well as a default implementation.
-This default implementation immediately produces a type error, to cause an error when the type of an expression not defined in the typing rules is demanded.
-
-%\CR{The details of this section (esp.\@ the bit below) might fit better into Section~\ref{sectionimplementation}?}
-%\CR{What is ``type matching''?  I am not familiar with this term, and you haven't introduced it before.}
-It also contains the implementation of the type matching attribute, as defined in Figure \ref{typematches}.
-%It is defined for all \lstinline{Type} nodes, and checks whether the type's class is equal to or a subclass of the parameter type.
-%\CR{So this is not an equivalence relation, but a partial order?
-What does that mean wrt how it is used by the implementation of the type checker?} \reviewquestion{I think it would be easier to change this poorly thought out line of code than to fully evaluate and address the implications in the report :) Is that okay?}
-The current implementation of \lstinline{Type.matches(Type t)} checks that the two types are the exact same, though we also considered an implementation that checks if the Type it is called on is either of the same class or a subtype of the type \lstinline{t}.
-%\CR{is ``matches'' a good name for something that is not symmetric?}
-This would allow the language limited sub-typing at the type checker generation step, but at the risk of causing hard to predict behaviour and hidden order dependencies with type variables.
-%\CR{Can you give an example of a case where this is helpful, and one example where it might be troublesome?
-%Let's use the ``if'' rule as an example.
-If $\ty{Number} :> \ty{Int}$ (\ty{Number} is a supertype of \ty{Int}), what type will I infer if $t_2 = \ty{Number}$ and $t_3 = \ty{Int}$?
-How about the other way around?}
-
-It is also possible to override the \lstinline{matches} implementation for a specific Type, by adding a definition to a JastAdd aspect included in the output type checker.
-This could potentially be used to compare type parameters in future versions, or support for ad-hoc type polymorphism.
-%\CR{This sounds like this is a way to support ``ad-hoc type polymorphism''?}
-
-\begin{figure}[h]
-\begin{lstlisting}[language=jrag]
-syn boolean Type.matches(Type t) {
-  return getClass().equals(t.getClass());
-}
-\end{lstlisting}
-  \caption{Definition of the type matching method}
-  \label{typematches}
-\end{figure}
-
-%\subsubsection{Error handling}\label{errorhandling}
-\subsection{Error handling}\label{errorhandling}
-The current error handling consists of throwing a runtime exception at the first discovered type error, ceasing the type checking immediately.
-This method was chosen solely for its simplicity of implementation, rather than its functionality.
-Possible improvements to it are discussed in Section \ref{improvederrorhandling}.
-
-It provides a stack trace with information on see where in the type checking code the error occurs (such as what typing rule fails), but doesn't currently provide any information about what section of the input program causes the error.
-%This tends to result in errors that are more useful for debugging the typing rule implementation, rather than providing useful information to the user of the type checker, where there is much room for improvements.
-
-%\CR{This section is needlessly apologetic-- error handling was not the focus of your work.  You can add a forward reference to 6.1.2, though.}
-
-%Once typing rules have been parsed, we end up with an abstract syntax tree, in the form of Figure \ref{trastspec}, where the \lstinline{RuleSet} is the root node.
-%The \lstinline{RuleSet} node consists of a list of \lstinline{Rule} nodes, and begins by sorting each rule by which object language node it describes.
-%
-%The \lstinline{RuleSet} node then begins by generating the outline of a JastAdd aspect \lstinline{TypingRules}, and the definition for the method \lstinline{.type()} for each object language node.
-%It then calls on the \lstinline{Rule} node to fill in the implementation of the method.
-%
-%A \lstinline{Rule} consists of a name, a conclusion, and a possibly empty list of premises that need to be true for the conclusion to be returned.
-%In the simplest case of an empty list of premises, the entire content of the rule is generated by the conclusion.
-%For rules containing premises, the code becomes more involved.
-%Firstly, any variables or type variables used in the rule need to be declared, followed by a check to ensure each use of a type variable represents a node with the same type.
-%Then it generates an if-statement corresponding to our premises, by letting each \lstinline{Premise} generate a boolean expression, all of which get joined by the AND operator.
-%
-%Both \lstinline{Premise}s and \lstinline{Conclusion}s are represented by the same abstract class \lstinline{Formula}, which currently contains a single implementation \lstinline{HasType}, consisting of an expression and a type.
-%Depending on whether it is used as a premise or conclusion, it can either represent that the expression must have the specified type or that the expression should be given the specified type.
-
-%\subsection{Type checker design}
-%The provided typing rules compile into \lstinline{type()} methods, which return a value from the Type interface.
-%For the most simple typing rules, this method's implementation will consist of a single return statement.
-%
-%Types are declared in an AST specification, currently provided by the template, allowing simple type hierarchies.
-
-\chapter{Evaluation} % is the part where you present the finds. Depending on the area this part contains a subset or all of the following:
-In the first section of this chapter, we present typing rule definitions for a few simple type systems, alongside the resulting code produced as an output when they are input to our compiler.
-
-In Section \ref{sectionresearchquestions}, we evaluate the overall result of our study by answering our research questions.
-%\CR{The section has two parts, so the intro paragraph here can directly discuss their purpose.}
-%\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach.
-Sometimes this is already addressed in the \textbf{Method}, sometimes this part replaces \textbf{Method}.
-\section{Results} % contains the data (as tables, graphs) obtained via experiments  (benchmarking, polls, interviews).
-Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
-We have chosen to evaluate our program using two example languages, inspired by the languages Booleans and Numbers introduced in Types and Programming Languages\cite{Pierce}.
-We have chosen to simply the languages somewhat, to reduce the amount of equivalent cases for our program to parse, making it easier to evaluate each feature individually.
-Thus, we have defined two languages Bools and BoolsIf, where the former is a simple language consisting of boolean values, an \lstinline{OR} operator, and a zero-value.
-This is a minimal case to produce a meaningful type checker, as it contains values of two different types and a rule with premises limiting the types.
-
-The second language BoolsIf is an extension to the Bools language.
-It adds an if-expression, which typing rules utilises type variables to ensure both if and else terms have the same type, and then uses that type as its own type.
-%\CR{You can't discuss results without describing your experiments, but to describe the experiments, you should first motivate them.
-%  Since you are doing case studies, you can use Pierce to motivate your choice of languages (even if your language is subtly different somehow; I don't have the book in front of me).
-%  Make clear to the reader that you picked your languages based on the existing literature, not on a whim.}
-%\subsection{Input/output examples, with descriptions}
-\subsection{Bools}
-
-\begin{figure}[]
-\begin{lstlisting}[]
-Expression ::= Term;
-
-abstract Term;
-True  : Term;
-False : Term;
-Or    : Term ::= Left:Term Right:Term;
-Zero  : Term;
-\end{lstlisting}
-  \caption{AST specification for the Bools language}
-  \label{boolsast}
-\end{figure}
-\begin{figure}[]
-\begin{lstlisting}[]
-[T-True]
-True : Bool
-
-[T-False]
-False : Bool
-
-[T-Or]
-left : Bool,
-right : Bool
-----------------------
-Or(left, right) : Bool
-
-[T-Zero]
-Zero : Int
-\end{lstlisting}
-  \caption{Typing rules for the Bools language}
-  \label{boolstr}
-\end{figure}
-
-\begin{figure}[]
+\pagebreak
 \begin{lstlisting}[language=jrag]
@@
 
+  syn Type True.type() {
+    return new Bool();
+  }
+
+  syn Type False.type() {
+    return new Bool();
+  }
   syn Type Zero.type() {
@@
   }
-
-  syn Type True.type() {
-    return new Bool();
-  }
-
-  syn Type False.type() {
-    return new Bool();
-  }
 }
 \end{lstlisting}
-\caption{Generated typing rules for the Bools language (reformatted for readability)}
-  \label{boolstrgen}
-\end{figure}
-A simple example of a language is provided in Figure \ref{boolsast}, consisting of three values, \lstinline{True}, \lstinline{False} and \lstinline{Zero}; and the boolean operator \lstinline{Or} with two parameters, Left and Right.
-In the AST specification, there are no limitations on what terms can be provided as parameters to the \lstinline{Or} operator, allowing illogical constructs such as \lstinline{Or(True, Zero)}.
 
-To prevent cases like these, we define typing rules via the specification in Figure \ref{boolstr}.
-The typing rules for \lstinline{True}, \lstinline{False} and \lstinline{Zero} contain no premises and define the type as \lstinline{Bool} and \lstinline{Int} as appropriate.
-The typing rule for the \lstinline{Or} operator, which takes two parameters \lstinline{left} and \lstinline{right}, contains two premises, specifying that the \lstinline{Or} operator has the type \lstinline{Bool} only if both the \lstinline{left} and \lstinline{right} terms also have the type \lstinline{Bool}.
+\begin{lstlisting}[]
+[T-True]
+True : Bool
 
-The typing rules from Figure \ref{boolstr} generate the type definitions in Figure \ref{boolstrgen}.
-The simple rule definitions for the values compile into equally simple implementations, consisting of a single return statement of a newly constructed \lstinline{Type} object.
-For the \lstinline{Or} node, the code becomes slightly more complex.
-%\CR{The next bit would probably be easier if you just show the code.}
-It declares variables corresponding to the child nodes given in the parameters, followed by an if-statement to check if the premises are upheld by recursively evaluating the types of the child nodes.
-Depending on the result of this evaluation, either a \lstinline{Bool} type is returned, or a type error is thrown.
-%First two variables are defined, corresponding to the parameters in the typing rule, and assigned to the children with the index corresponding to the position among the parameters.
-%This is followed by an if-statement, recursively evaluating the children's types and comparing the result the type demanded in the premise.
-%If the condition is true, a \lstinline{Bool} type is returned.
-%If not, a runtime exception is thrown to convey a type error.
+[T-False]
+False : Bool
 
-%\CR{You may want to add a short statement that this captures the intended behaviour of the type checker.}
-This type checker accurately captures the intended behaviour of the typing rules.
+[T-Zero]
+Zero : Int
 
-\subsection{BoolsIf}
+[T-Or]
+left : Bool,
+right : Bool
+----------------------
+Or(left, right) : Bool
+\end{lstlisting}
 
-%\CR{Separate case study or next step in the same case study?  Either way, the section title should make the connection to 4.1.1 (or lack thereof) clear.}
+\begin{lstlisting}[]
+[T-True]
+True : Bool
 
-\begin{figure}[]
-\begin{lstlisting}[]
-If : Term ::= Cond:Term Then:Term Else:Term;
+[T-False]
+False : Bool
+
+[T-Zero]
+Zero : Int
+
+[T-Or]
+left : Bool,
+right : Bool
+----------------------
+Or(left, right) : Bool
 \end{lstlisting}
-  \caption{The additional term introduced in the BoolsIf language, extending the Bools AST in Figure \ref{boolsast}}
-  %\CR{Consider giving this language a different name, for clarity}
-  \label{ifast}
-\end{figure}
 
-\begin{figure}[]
-\begin{lstlisting}[]
-[T-If]
-x: Bool,
-y: a,
-z: a
-------------------
-If(x, y, z): a
-\end{lstlisting}
-  \caption{Additional typing rules for the BoolsIf language, an extension for the Bools typing rules in Figure \ref{boolstr} with an if term}
-  \label{iftr}
-\end{figure}
-\begin{figure}[]
-\begin{lstlisting}[language=jrag]
+\pagebreak
+\begin{lstlisting}[language=jrag, linewidth=350pt]
 syn Type If.type() {
@@
 \end{lstlisting}
-\caption{Generated typing rule for the if term added in the BoolsIf extension. The rest of the typing rules are identical to those in Figure \ref{boolstrgen}. (Reformatted for readability)
-  %\CR{The components of this function would be great to discuss in Section 3.
-Here in Section 4, you may want to forward reference the discussion about redundancy/possible optimisations (Section 4.2.2) and very concretely use this code and explain the changes you have in mind (for someone else to do in future work).}
-}
-  \label{iftrgen}
-\end{figure}
-
-%\begin{figure}
-%\centering
-%  \begin{minipage}[b]{.4\textwidth}
-%  \centering
-%\begin{lstlisting}[]
-%[T-If]
-%x: Bool,
-%y: a,
-%z: a
-%---------
-%If(x, y, z): a
-%\end{lstlisting}
-%  \captionof{figure}{A figure}
-%  \label{fig:test1}
-%\end{minipage}%
-%\begin{minipage}[b]{.6\textwidth}
-%  \centering
-%\begin{lstlisting}[]
-%syn Type If.type() {
-%ASTNode x = getChild(0);
-%ASTNode y = getChild(1);
-%ASTNode z = getChild(2);
-%
-%Type tyvar_a = null;
-%if(tyvar_a == null) tyvar_a = y.type();
-%else if (!tyvar_a.matches(y.type()))
-%throw new RuntimeException("Typechecking failed: Type variable mismatch");
-%if(tyvar_a == null) tyvar_a = z.type();
-%else if (!tyvar_a.matches(z.type()))
-%throw new RuntimeException("Typechecking failed: Type variable mismatch");
-%if(x.type().matches(new Bool()) && y.type().matches(tyvar_a) && z.type().matches(tyvar_a)) {
-%return tyvar_a;
-%}
-%throw new RuntimeException("Typechecking failed");
-%}
-%\end{lstlisting}
-%  \captionof{figure}{Another figure}
-%  \label{fig:test2}
-%\end{minipage}
-%\end{figure}
-As an example of rules utilising type variables, we have extended the Bools language with an if term, as used in Pierce's Booleans language\cite{Pierce}.
-This if term captures each of the utilities of type variables in typing rules, using it both to ensure that multiple terms are of the same type and then using that same type as the result of the rule evaluation.
-%\CR{``Can extend'' sounds a bit arbitrary.  Can you motivate with Pierce (even if not exactly)?}
-As shown in the AST extension in Figure \ref{ifast}, it consists of three terms: a conditional, a term to be evaluated if the condition is true and a condition to evaluate if false.
-A typing rule of this if term can be seen in Figure \ref{iftr}.
-The rule has three premises, the first of which declares the conditional must have the type \lstinline{Bool}.
-The next two declare that the two other children must be of the type \lstinline{a}, with the lowercase identifier indicating a type variable.
-This type variable \lstinline{a} also subsequently appears in the resulting type of the if term.
-
-%\CR{``Shown in Figure~\ref{reference}'' is the current preferred form in our field}
-The code for this rule is shown in Figure \ref{iftrgen}, and begins with three parameter declarations as in previous rules, followed by a new type variable declaration on line 6, which is instantiated to null.
-For each invocation of the type variable, a code section is generated, which checks if the type variable is still null, and if so assigns it to the type of the left-hand side term, or if it has been assigned checks whether the types match each other.
-Should any non-matches be found, a type error is thrown.
-
-The rest of the method is similar, with premises being checked -- including a redundant check of the type variables -- before returning the type variable, or a type error being thrown.
-
-\section{Research questions}\label{sectionresearchquestions}
-
-\subsection{\rqone}
-Our project has shown that generating type checkers using RAGs is a viable solution for a number of different language constructs.
-%To assess our implementation, we'll go through the examples used by Pierce\cite{Pierce} to see which cases we can successfully express.
-To evaluate the completeness of our implementation, we refer to Pierce\cite{Pierce} as a reference work.
-Typing rules are first introduced in chapter 8, with two simple languages in Figure 8-1 and 8-2.
-The first, Booleans (generally referred to as B) consists of only the syntactic structures \lstinline{true}, \lstinline{false} and \lstinline{if t then t else t}.
-Extracts of the typing rules for the B language were used Section \ref{sectionlanguage}.
-The second language called Numbers (NB) is an extended version of B, adding arithmetic expressions.
-This introduces the terms \lstinline{0}, \lstinline{succ t}, \lstinline{pred t} and \lstinline{iszero t}.
-
-Both the B and NB languages are fully supported by our project, and our Bools and BoolsIf languages were devised to utilise all the same language features but in a more condensed form.
-In chapter 9, Pierce introduces the Simply Typed Lambda Calculus, which typing rules our compiler does not yet support, as they require environments, a feature we'll discuss in Section \ref{futureworkenvironments}.
-
-%\CR{Please be clearer on which examples from Pierce you implemented and where support stops.}
-We have implemented support for simple syntax-directed type declarations, where a value or function always has a certain type.
-These simple rules are not of much interest in themselves but are an important backbone to the subsequent analysis.
-We also support type declarations with premises that must hold for the rule to apply, allowing us to express relationships between types and bringing with them the possibility of type errors.
-
-Type variables extend these relationships further by providing a layer of abstraction, allowing us to write rules where the types are not precisely defined.
-This enables rules where the return type of an expression depends on its child expressions, or where child expressions must have the same types.
-
-%A major feature missing from our implementation is the concept of type environments or typing contexts, providing a mapping from variables and functions to types.
-%This prevents us from generating type checkers for even simple general purpose languages.
-%
-%\CR{Please explain typing contexts and why/where they are important.  You can also reflect on whether typing contexts are necessary in the context of RAG-based analysis (cf. what we did in EDAP15).}
-%
-%\CR{I am missing the phrase ``syntax-directed'' here.}
-
-\subsection{\rqtwo}
-The basis of our approach has been to translate typing rules into \lstinline{type()} attributes added to the program's AST.
-We opted for an approach where a successfully evaluated typing rule returns a \lstinline{Type} value, defined in a separate type syntax file, while a failed rule throws a runtime exception representing a type error.
-
-In this section we will be analysing different aspects of this approach, and compare them to potential alternative approaches.
-%We have approached translating typing rules into a JastAdd attribute \lstinline{type()}, which either returns a \lstinline{Type} value, or throws a runtime exception if the expression is invalidly typed.
-
-\subsubsection{Type syntax}
-We have chosen to represent the right-hand side of type equations, the types themselves, as an abstract AST class.
-While our current implementation only supports simple types, this representation gives us a lot of flexibility for future extensions.
-Parametric types, such as lists where the type of the list contains information about the types of its elements, could be expressed as simply as \lstinline{MyList : Type ::= Type;}.
-
-The current implementation of \lstinline[language=java]{matches()} would however consider all lists to match, without comparing the child nodes.
-Perhaps the default implementation could be extended to compare all child elements, at the risk of potentially tricky evaluation, or these more advanced types would require manually overriding the method.
-Yet another approach may be to design a way to define type equality within the typing rules, allowing users to customise the behaviour without having to restort to manually writing JastAdd code.
-
-%\CR{Recursive term equivalence can become tricky when you are dealing with type variables (cf. the Hindley-Milner inference exercise in EDAP15).}
-%\CR{Type equality could also be expressed in terms of typing rules.  This approach is quite common with subtyping.}
-
-\subsubsection{Error handling}
-We have implemented premises as conditions of a single if-statement, where we check that each premise holds true before returning the type, or falling through to a type error if the premises do not hold.
-It might instead be worthwhile to check each condition individually, to be able to specify in the error message precisely which premises have failed and why.
-%This would perhaps make it easier to implement informative type errors, as you could report precisely which premise is failing.
-If paired with the current error handling of throwing an exception to convey type errors, it would make it harder to implement having multiple valid rules providing the type of an expression, as the first failure would throw an exception, stopping the evaluation of any further typing rules which may have been successful.
-%\CR{What do you mean by ``seize''?}
-An alternative error handling would suit it better, one able to collect multiple errors without seizing evaluation, perhaps merely adding all the errors to a list and checking whether it is empty before returning a type.
-
-A more informative solution to this might be view rules as a kind of pattern matching, where each rule is evaluated one by one until a matching one is found.
-Instead of reporting the reason any individual rule failed, the error might instead display either all or only the best matching of the rules it tried to apply, perhaps with a list of non-holding premises for each one.
-%\CR{This is a good idea; you can even sort the rules by closest matches and only show the top ones if there are many.}
-%\subsubsection{Error handling}
-%We chose a very simple representation for type errors, based around runtime exceptions, which are thrown either when type variable unification fails or the premises of a typing rule fails to match.
-%The exceptions include a short message stating which of the two causes has lead to the error.
-%
-%This approach gave us the minimum information to distinguish between successful and failing type checking, with a very simple implementation.
-%It does however complicate the case where a single node has multiple applicable typing rules, depending on their premises.
-%In Section \ref{improvederrorhandling} we discuss alternative approaches.
-
-%\CR{Something about error handling would be good to discuss here}
-
-\subsubsection{Variables}
-The variables in the typing rules are compiled into regular Java variables containing the AST node.
-Notably, the declarations of these variables type them merely as \lstinline{ASTNode}.
-As we currently only utilise the \lstinline{type()} attribute, which is declared for all nodes, this lost information has little impact.
-However, in the future it may become relevant to type these more specifically, to support more complex rules that require deeper analysis within the tree.
-
-\subsubsection{Type variables}
-%\CR{I am missing a discussion of type variables and unification as we used them in EDAP15 for Hindley/Milner.}
-Type variables are implemented as Java variables and as we currently only support type variables scoped within a single typing rule and lack support for type parameters, all uses of it will be known directly at the evaluation of the typing rule.
-Thus, our unification algorithm can be very simple.
-All our type variables are initiated as null, to represent the type variable's unassigned state.
-This is followed by a check for each point of use in the typing rule, to see whether the type variable is still unassigned and if so assign it to the type of the left-hand side.
-
-This approach could be improved by finding the first use of each type variable during the rule compilation, and generating a single declaration statement initialising it to the type of the corresponding expression.
-%\CR{no need to say ``much room for improvement'', just mention the specific disadvantages.}
-
-%\reviewquestion{Circular attributes?}
-%\CR{Circular attributes: Recursion in typing rules: Datalog is built with support for ``infinite'' recursion (since it operates on finite domains,
-%  it can use a special bottom-up evaluation scheme to ensure termination.)
-%  If you need this in RAGs, it gets more complicated, but you'd need an example to explain this in the introduction-- since you don't directly address that in this work,
-%  I wouldn't talk about it too much here (you could also skip it entirely here).  Minimum: talk about it in the discussion later.}
-%  %TODO: Circular attributes not necessary in our case, but may be required for other types of type analysis
-
-%\CR{Can you structure this subsection a bit more explicitly?
-Could be subsubsections, but not necessarily needed; maybe a bullet list of challenges and for each challenge a bullet list of potential approaches could work?}
-
-\subsection{\rqthree}
-
-\subsubsection{Technical challenges}
-%Splitting code between multiple JastAdd components (and java) can get messy
-Writing idiomatic JastAdd components often requires splitting a single algorithm into many different pieces of code.
-Our approach to generating type attributes was based around recursively traversing the typing rule AST to fill in each bit of information, rather than having one parent node trying to form an understanding of the entire tree.
-The plus side of this recursive approach is that each component doesn't need an intricate understanding of the nodes underneath it.
-A \lstinline{Rule} node doesn't need to know whether the right-hand side of the conclusion is a concrete type or a type variable.
-The \lstinline{HasType} node representing the conclusion generates the structure of the return statement, calling on the right-hand side node to fill in the exact type.
-
-However, this code could at times become complex, as certain lines required bits of information from several different parts of the AST, several steps apart.
-Propagating this information required adding intermediate attributes to nodes throughout the AST, leading to a complex set of interdependencies.
-If we want to change a specific token of the generated code, tracing down the exact attribute it is generated in can be tricky.
-%There is a tradeoff to be made between the single responsibility of each component making it easy to verify its correctness, but the complicated interdependence making it hard to grasp the full picture.
-
-%We attempted to keep each node in the tree responsible for generating the code it corresponds to \CR{What does that mean?
-You need to provide a bit more background}, which avoids parent nodes having to check their children's structure with \lstinline{isInstanceOf}, but also means even generating a simple rule can require a dozen calls between different nodes.
-%\CR{Overall I found it difficult to process this paragraph.}
-
-%Having to use a hacky method to access private JastAdd methods to parse a second AST at runtime. (Beyond jastadd's scope?)
-Using JastAdd simultaneously as a library and a build component proved difficult on several occasions.
-The largest issue was that some of the features we intended to support, such as verifying that typing rules correspond to valid nodes within the object language AST, were complicated by having to parse the object language AST at runtime, using JastAdd as a library rather than a build tool.
-JastAdd does not expose any methods for parsing an AST tree at runtime, and we had to access undocumented, private methods through Java's reflection API to get the relevant information.
-Unfortunately, the resulting data was hard to work with, and we could not allocate enough time to implement the planned features.
-It may be that this use case is beyond the intended scope of JastAdd, or it is perhaps only poorly documented.
-In either case, while our experimentation has shown it to be technically possible, the complexity is worth bearing in mind for future work.
-
-Not being able to perform these error checks during the generation of our type checker means our translation will emit nonfunctional typing definitions if a mistake is written in the typing rules.
-These errors will then go unreported until the output project is compiled or run, at which point the user will have to attempt to trace these errors back to the typing rule definition, to find the actual cause.
-As JastAdd itself also adds a layer of indirection, by taking the attributes defined in RAG modules and compiling them into generated Java class files, this can be particularly tricky.
-%\CR{NB: this is also somewhat true of JastAdd's own compilation approach, so there might be double indirection errors ...}
-
-We also ran into a few tooling issues.
-The first was Gradle not willing to recognise JastAdd as both a library and a build component, which we eventually solved by manually providing a jar file separately from the resolved build dependency.
-The second one was our LSP server, jdtls, not recognising JastAdd imports, which we never managed to fix.
-%Gradle troubles
-%Library not recognised by LSP server
-%Difficult runtime parsing of AST files
-
-\subsubsection{Conceptual challenges}
-On the more conceptual side, a fundamental challenge of our approach is how to implement non syntax-directed rules.
-The structure of our program is based around turning typing rules into evaluable attributes on the AST.
-This means we inherently tie typing rules to a specific AST node, creating a direct link between typing rules and the syntax.
-
-However, more complex type systems may require rules that are not syntax-directed, where a rule does not apply to any specific AST node.
-A common example use case for this is subtyping, where any expression of a certain type can also be considered to be any of that type's supertypes.
-Figure \ref{typingruleexamplesub} shows a subtyping rule taken from Pierce\cite{Pierce}.
-
-With our approach, this rule would have to change the behaviour of every generated \lstinline{type()} attribute.
-Additionally, we would require an implementation supporting type polymorphism, where a single term can be evaluated to have multiple different types.
-
-\begin{figure}[]
-  \begin{equation}
-    \tag{T-Sub}
-    \frac{
-      \Gamma \vdash t : S\quad
-      S <: T
-    }{
-      \Gamma \vdash t : T
-    }
-  \end{equation}
-  \caption{A typing rule for subtyping, an example of a non syntax-directed typing rule.}
-  \label{typingruleexamplesub}
-\end{figure}
-
-%\CR{This is all about technical challenges.
-%  Do you also see conceptual challenges?
-%  (You can be less specific here, but some words about support for parametric polymorphism and/or subtyping could be nice; at the very least, something about the relevance of syntax-directedness would be valuable.)}
-  % Focus on syntax directedness. Exceptions: if t has type t, it also has type t; nested rules if(xcondition): X, if(ycondition): Y
-
-%\chapter{Discussion} % allows for a longer discussion and interpretation of the results from the evaluation, including extrapolations and/or expected impact.
-Focus here on a broader view of the results, talking about the relation between the different finds.\footnote{Bad practice is to display graphs in Results and then describe them textually one by one in here.
-No!
-Both sections should have some discussion, but one targets individual finds and the other tries to bridge between these adopting a more overarching viewpoint.} This might also be a good place to describe your positive and negative experiences related to the work you carried out.
-% Occasionally these sections are intermingled, if this allows for a better presentation of your work.
-However, try to distinguish between measurements or hard data (results) and extrapolations, interpretations, or speculations (discussion).
-%\section{Supported features}
-%\section{Development learnings}
-
-\chapter{Related work} % Maybe move to discussion?
-% Describe related work (literature study).
-Besides listing other work in the area, mention how is it related or relevant to your work.
-The tradition in some research area is to place this part at the end of the report (check with your supervisor).
-Pacak et al. implemented a compiler for generating incremental type checkers using declarative logic programming language Datalog\cite{Pacak}.
-While the goal is similar to ours, their project defines a formalism for expressing typing rules within the Datalog language, whereas ours defines a language which aims to keep as close as possible to the traditional inference rules used for typing rules, to minimise the difficulty of translation.
-%\CR{I'd start with this difference-- the translation is targeting two rather different formalisms, which is all you need to justify the novelty.}
-
-%they take a different approach, by converting typing rules into the  \CR{second half of this sentence confuses me}.
-
-Their work is ends up largely focused on expressing typing rules within the limits of the Datalog language, which can only compute finite relations.
-We have a significantly different compilation target, JastAdd, which allows us to use reference attribute grammars to evaluate the typing relations, utilising a built-in AST structure.
-Whereas Datalog is a fully declarative language, with JastAdd we express typing relations as declared attributes on the AST, with their implementations written in imperative Java code.
-%\CR{Datalog is fully declarative, whereas JastAdd uses declarative rules that connect snippets of imperative Java code.  Is that what you meant?}
-%\CR{Some more thoughts below to help make the distinction clearer (all optional):}
-%\CR{Datalog has no direct notion of ASTs or other tree structures / algebraic terms, so trees must be encoded into relations.}
-%\CR{Datalog has a simple, built-in notion of unification that can handle cases like testing sibling relations: $\textsf{Sibling}(x, y) :- Parent(p, x), \textsf{Parent}(p, y), x \ne y.$.
-%    This is something you deal with later in JastAdd.}
-
-Other related work has been done in the domain of compiling natural semantics, without the specific focus on generating type checkers.
-
-An early compiler for natural semantics was made by Mikael Pettersson in 1996\cite{Pettersson}.
-They wrote a compiler in Standard ML which generated C code, linked to a custom runtime system.
-
-Saioc and Hüttel developed a compiler for natural semantics\cite{Saioc_2022}.
-They did so by defining a parsable meta language adaptation of natural semantics, similar to the one we defined for writing typing rules.
-Their project was however focused on verifying the correctness of these definitions and finding errors in their definitions.
-This was something we had hoped to explore more in our own work, but ended up deprioritising after running into issues with JastAdd.
-% : Related work section after chapter 5, one or two pages. Various papers from CR comments and goal document.
-
-\chapter{Conclusions} % should summarize your findings and possible improvements or recommendations.
-Our experiments have shown that reference attribute grammars are a flexible method for implementing type checker generators.
-Implementing the first end-to-end prototype was challenging and took quite a lot of work, but extending it with additional features such as type variables turned out to be a surprisingly painless experience.
-
-However, our compiler still has a long way to go before it can be used for real world use cases and general purpose languages.
-Chief among the missing features is the concept of environments, without which we are unable to support even common language features such as variables or functions.
-
-%\section{Summary of findings}
-\section{Future work}
-\subsection{Environments}\label{futureworkenvironments}
-Implementing environments, or type contexts, would be the clear next step for this project.
-However, its implementation is not entirely clear-cut, with several aspects requiring consideration.
-
-The essence of environments is the mapping of variables or functions to types, and an appropriate representation of this would need to be found.
-A method often used by traditional type checkers is a symbol table, a separate data structure mapping language constructs to their types.
-In our case, we may instead be able to leverage our reference attribute grammars to store the information within the AST.
-By connecting each use of a variable or function to the AST node representing its declaration site, we could utilise the node to store the type information.
-This approach would however also have to consider the typing of external language constructs, such as built-in functions or those imported from external libraries, for which we would not have access to the declaration site.
-
-%An additional concern is that of scoping, which different languages may have different rules for.
-
-A type checker working on these principles is certainly a possibility, though it remains to be seen how it could be generalised to the generation of typing rules for arbitrary languages.
-
-\subsection{Improved error handling}\label{improvederrorhandling}
-The current exception-based error handling is rather primitive and leaves much to be desired.
-Some improvements, such as provide more useful error messages, could be possible by making minor changes to how these exceptions are generated.
-Information such as what part of the input program the type error occurred at, what the failing typing rule was and what prerequisites it has, could be added to the exception message, for a quick usability improvement.
-
-An improvement demanding more extensive changes would be collecting multiple errors, instead of seizing the type checker at the first discovered error.
-This could be implemented by adding another type class to be returned when an error occurs, representing the `Any' type.
-Any \lstinline{matches()} call on this type would return true, to avoid one error cascading into many, but still allowing other unrelated errors to be discovered.
-This implementation would also need new functionality to collect the error messages, perhaps by splitting the type resolution and type errors into separate attributes for each node, where the type errors propagate to a program-wide collection.
-
-It may be possible to create this by wrapping around the current implementation with two additional attributes.
-One would be a total function, that returns the type evaluated by the current \lstinline{type()} method if it is successful or returns the Any type in the case of an exception.
-The second attribute would contribute an error to the error collection only in the case of an exception.
-The only other changes required would be to use the total wrapped method within the generated attributes.
-
-\subsection{Verify rules against AST}
-Major improvements to the user experience could be made by tying the typing rules to the object language AST before generating the type checker.
-Currently, there are no checks in place to verify the completeness of the typing rules.
-If a user accidentally leaves out the typing rule for a certain construct, they will receive no warning until the outputted type checker attempts to check the expression's type.
-
-Similarly, no checks are made to ensure the nodes described in the typing rules actually exist within the language's AST.
-Our project will compile invalid or misspelled typing rules without complaint, leading to errors which will not be caught until the outputted type checker is attempted to be compiled.
-
-\subsection{Handle multiple rules for the same node}
-In type systems it is generally considered valid to have multiple typing rules for the same construct, with different, non-intersecting prerequisites.
-It could for example be useful to type an addition operator for a variety of different numerical types, which may not have a common ancestor.
-
-Most of the work to support this is in place, though the first rule to fail would throw a type error exception, preventing the later rules from evaluating.
-These errors would either have to be caught, and only emitted if all rules had failed, or a new error handling system not based around exceptions could be introduced.
-Additionally, an extra layer of scoping would have to be introduced, to prevent variable naming collisions between the different rules.
-This could be solved simply by wrapping each rule within brackets, creating a separate scope for the evaluation of each rule, or relabelling the variables to prevent collisions.
-%\CR{This is about name clashes in the generated code?  You may want to consider automatic relabelling here anyway (``this'', ``new'' etc. as type variable names are likely going to cause issues) }
-
-%add(int, int): int, add(double, double): double)
-%\subsection{Better support for numbers, strings, etc.}
-
-%\CR{Not sure that there is anything needed here}
-
-\subsection{Improved type checking algorithm}
-
-% Should use consistent formatting when it comes to Names ("FirstName LastName", or "F. LastName")
-%\printbibliography
-\makebibliography{MSc} % is a must in a scientific report. {\LaTeX} and \texttt{bibtex} offer great support for  handling references and automatically generating bibliographies.
-
-\begin{appendices} %should contain lengthy details of the experimental setup, mathematical proofs, code download information, and shorter code snippets.
-Avoid longer code listings.
-Source code should rather be made available for download on a website or on-line repository of your choosing.
-\chapter{Source code}
-The source code for our project is available at \\\url{https://github.com/nichobi/thesis-project}, \\with the README.md file providing instructions for use.
-
-\vspace*{\fill}
-\hspace*{\fill}
-\includegraphics[scale=0.2]{stagbeetle.pdf}
-
-% display used packages information unless noflielist is used in the cslthse-msc package option
-\printfilelist
-
-%TODO: Re-add the popsci summary
-%%make sure we're on even page with the pop-sci
-%\checkoddpage
-%\ifoddpage
-%\else
-%  \newpage
-%  \thispagestyle{empty}
-%  \mbox{ }
-%\fi
-%\includepdf[pages={1}]{popsci/popsci.pdf}
-\end{appendices}
-
 \end{document}
